### 2022-01-21 scr/gbm_proteomics_data_analysis_proteolytic_products_prep_figures_v3_gh_format.Rmd
first commit

---
title: "GBM Proteomics and Proteolytic processing data analysis - Reproducible report"
author: "Miguel Cosenza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document
always_allow_html: true
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

```{r results='hide'}
## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)

source(here("scr/helper_functions.R"))

theme_set(theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0, size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=8),
                legend.text = element_text(size = 7),
                legend.title = element_text(size = 8),
                legend.key.height= unit(3, 'mm'),
                legend.key.width= unit(3, 'mm'),
                legend.position="bottom"))
```

The following report holds the code, results and interpretation of the proteomics data analysis of Initial-Recurrent glioblastoma samples of matched patients.

# Sample annotation info

```{r echo=TRUE}
sample_annotation <- read_csv(here("data/sample_annotation.csv"))

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",patient, sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
  filter(recurrence %in% c("prim", "rec"))

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")
```

## Sample annotation

### Brief note on sample annotation:

Twenty-two patient matched samples (corresponding to 11 patients) were processed for protein extraction, TMT-labelled and submitted via high-pH fractionation and further analyzed via LC-MS/MS.

```{r echo=FALSE, error=TRUE}
kbl(sample_annotation, booktabs = TRUE) %>%
  kable_styling(font_size = 14, bootstrap_options = c("striped", "hover"))
```

# Brief note on data analysis and processing:

Three proteomics data analysis approaches were used to explore the spectral data and will be briefly described:

1. General large scale proteomics.
2. Proteolytic processing.
3. Proteogenomics

The FragPipe bioinformatic pipeline was used for each of these approaches with varying parameters depending on the type of explorative approach.

__1. General large scale proteomics analyses__ : In order to explore the differential abundance of proteins between Recurrent and Initial tumor, a fully-tryptic search was performed against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides. A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__2. Proteolytic processing__: The large scale evaluation of differential proteolytic activity between recurrent vs primary tumor was executed via a semi-tryptic search against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides.  A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__3. Proteogenomics__: The explorative proteogenomics analysis started with the generation of a GBM-specific database from publicly available RNA-seq sequencing data from tumor specimens. Fastq files were obtain from the SRA BioProject PRJNA627121 (Cell Rep, 2021 Mar 2;34(9):108787). Eight samples containing paired reads were selected, based on: 1) the higher number of reads per file; 2) having a similar proportion of males and females and 3) not bearing isocitrate-dehyderogenase (IDH)-1 mutation (IDH WT). None of these patients undergo previous treatment, therefore potential variants in the sequencing data are not expected to be associated to treatment-related mutagenic processes. Spectral data was then searched against this custom database using a fully-tryptic approac. No minimum intensity threshold was set for reporter ion intensity for quantitation.

Exceptionally when stated differently within each sub-approach described above, all these searches were executed using the following parameters: MSFragger was used as a search engine, setting a fully or semi-tryptic specificity allowing for 1 missed cleavages. Precursor mass tolerance was set to -20/20 and fragment mass tolerance of 20 ppm with mass calibration and parameter optimization resulting in 5 ppm. Peptide N-terminal acetylation and peptide N-terminal TMT labeling were set as variable modifications. TMT labelling at K and carbamidomethylation of C were set as fixed modifications. MSBooster was used for deep-learning based predictions of retention time and spectra. Predicted features were used by Percolator for post-processing scoring and false discovery rate (FDR) control via target-decoy competition of all peptide-to-spectrum matches (PSMs) obtained from MSFragger search. ProteinProphet was used as a protein inference algorithm keeping all PSMs with probability score bigger than 0.9. Final report and FDR estimation was based on the filtered PSM and protein lists before protein scoring.

Relative quantitation of identified peptides within each sample was performed via their reported ion intensities using TMT-integrator. Only PSMs coming from unique peptides with a minimum probability of 0.9 and an isotopic purity of at least 50% were considered for quantitation. Quantitative values were normalized via median centering and summarized by protein/gene and peptide using a virtual reference channel, to finally generate protein and peptide normalized abundance matrices of abundance used for further statistical processing in combination with the sample annotation.

# General proteomics analysis

## Initial data loading and wrangling

Load id/quant data and prep coverage information

```{r message=FALSE}
## tmt-integrator output loading 
tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
  janitor::clean_names()

identified_peptides <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_peptide.tsv")) %>%
  janitor::clean_names()

fasta_ident_prots <- read.fasta(here("data/specific_search_fragpipe17/specific_no_ptms_2/mix_2/protein.fas"), seqtype = "AA", as.string = TRUE)

```

Wrangle and merge protein ID/quant with sample annotation

```{r}
# wide expression matrix with NAs
expr_matrix <- dplyr::select(tmt_protdata_frag,
                             Protein = index, starts_with("x"))

long_mat <- pivot_longer(expr_matrix,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_2 <- long_mat %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq <- long_mat_2 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",patient, sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
  filter(recurrence %in% c("prim", "rec"))

## Annotated abundance data in long format ----

quant_annotated <- left_join(long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

## Wide xpression matrices without NAs
expr_matnona <- column_to_rownames(expr_matrix,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()
```

## ID coverage

### Coverage plot

```{r}
ids_summary <- quant_annotated %>% 
  dplyr::select(Protein, mixture, Abundance) %>% 
  distinct(Protein, mixture, .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "In all three",
                   `Nr of IDs` = nrow(expr_matnona))) %>%
  bind_rows(.,
            tibble(mixture = "Overall",
                   `Nr of IDs` = nrow(expr_matrix)))
```

* Number of identified proteins: `r nrow(expr_matrix)`

* Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`

The initial analysis with MaxQuant yielded 2090 IDs in all samples (without missing values) and 5176 overall.

```{r fig.height=4, fig.width=7}
ggplot(data = ids_summary,
       aes(y = `Nr of IDs`, x = mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 5, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Proteins by Mixture") +
  ylab("Nr of IDed and Quant Proteins") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
        axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey"),
        panel.border = element_rect(colour = "black", fill=NA, size=1.5),
        axis.title=element_text(size=12,face="bold"))
```  
## Exploratory analysis and visualizations

### Visualization of missing values

```{r fig.height=7, fig.width=8}
vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values") + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))
```

### QC plots after TMT integrator pre-processing and normalization (Protein level)  {.tabset .tabset-fade}

#### By TMT channel/sample {.unnumbered}

##### General Abudance distribution by TMT channel/sample {.unlisted .unnumbered}

```{r fig.height=6, fig.width=9, warning=FALSE}
qcplot <- ggplot(quant_annotated,
                 mapping = aes(x = Channel_mix, 
                               y = Abundance, 
                               fill = recurrence)) +
          geom_boxplot() + 
          #facet_grid(.~recurrence, scales = "free") +
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot)

```

The median-centering after TMT-integrator seem to show a slight mixture-dependent centered median per sample. The paired distribution of initial and recurrent tumor allows the biological effect to be more important than any potential TMT-mixture-based batch effect.

The different abundance distribution can also be explained by a different distribution of missing values.

PCA and sPLS-DA analysis allow to confirm this.

#### By Protein (General)

```{r}
# select/sample N proteins to observe their abundance distribution  
which_prots <- sample_proteins(x = quant_annotated$Protein,
                               size = 50,
                               seed = 156)
```

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=5.5, fig.width=8, warning=FALSE}

qcplot_prots <- ggplot(quant_annotated %>% 
                         filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
                               y = Abundance)) +
          geom_boxplot() + 
          geom_jitter(color="black", size=0.4, alpha=0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
               x = "Protein ID ") + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot_prots)
```

#### By Protein and Batch {.unnumbered}

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=5, fig.width=10, warning=FALSE}
qcplot_prots2 <- ggplot(quant_annotated %>% 
                         filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
                               y = Abundance, 
                               fill = mixture)) +
          geom_boxplot() + 
          geom_jitter(color="black", size=0.4, alpha=0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
                       x = "Protein ID ") + 
          facet_grid(.~mixture, scales = "free") +
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot_prots2)
```

The abundance distribution of individual proteins can be considered as normaly distributed in general, which makes then suitable for inferential statistics via linear models.

### Exploratory PCA analisys

```{r echo=FALSE}
## Get predictors for mixtures
expr_matrix2 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotation2$paired_id)

paired_id <- sample_annotation2$paired_id
recurrence <- sample_annotation2$recurrence
patient <- sample_annotation2$patient
mixture <- sample_annotation2$mixture

## get predictor for mixtures wo patient 6
expr_matrixwo6 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotationwo6$paired_id)

paired_idwo6 <- sample_annotationwo6$paired_id
recurrencewo6 <- sample_annotationwo6$recurrence
patientwo6 <- sample_annotationwo6$patient
mixturewo6 <- sample_annotationwo6$mixture

## Prep matrix without missing values for PCA

expr_matnona <- column_to_rownames(expr_matrix2,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnona) <- paste(sample_annotation2$paired_id, 
                                sample_annotation2$mixture,
                                sep = "_")

t_expr_matnona <- t(expr_matnona)

expr_mat <- column_to_rownames(expr_matrix2,
                               "Protein") %>%
          as.matrix()

t_expr_mat <- t(expr_mat)

## Matrix without patient 6

expr_matnonawo6 <- column_to_rownames(expr_matrixwo6,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnonawo6) <- paste(sample_annotationwo6$paired_id, 
                                sample_annotationwo6$mixture,
                                sep = "_")

t_expr_matnonawo6 <- t(expr_matnonawo6)

expr_matwo6 <- column_to_rownames(expr_matrixwo6,
                               "Protein") %>%
          as.matrix()

t_expr_matwo6 <- t(expr_matwo6)
```

#### General PCA

##### Percentage of explained variance per component

The PCA was applied on the expression matrix without missing values

```{r}
pca_res = pca(t_expr_matnona, ncomp = 10, center = TRUE, scale = TRUE)
```

```{r fig.width=7, fig.height=5}
plot(pca_res)
```

~45% of variance based on protein abundance can be explained by components 1 and 2 after PCA.

##### PCA plot on samples based on protein abundance

```{r}
# preprocess pca results 
pca_variates <- pca_res$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id") %>%
  mutate(Stage = if_else(condition = recurrence == "prim", 
                         true = "Initial", 
                         false = "Recurrent"),
         Mix = case_when(mixture == "tmt_1" ~ "Mix1",
                         mixture == "tmt_2" ~ "Mix2",
                         mixture == "tmt_3" ~ "Mix3"))

pca_variateswopat6 <- pca_variates %>% 
  filter(patient != "x6")

```

```{r fig.width=7.3, fig.height=5}
ggplot(data = pca_variates,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins") +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

There is no evident batch effect after PCA analysis.

There is a good number of individual patients that show important differences in their PCA variates values in recurrent vs primary.

```{r fig.width=7.3, fig.height=5}
ggplot(data = pca_variateswopat6,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins (patient 6 excluded)") +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r}
pca_plot <- ggplot(data = pca_variateswopat6,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = Mix, color = Stage)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5), size = 2) +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) +
theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
      axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(colour = "black", fill=NA, size=0.5),
      axis.title = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 8),
      legend.key.height= unit(3, 'mm'),
      legend.key.width= unit(3, 'mm'),
      legend.position="bottom")
```

```{r}
ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

### Exploratory sPLS-DA

```{r}
plsda_rest1 <- plsda(X = t_expr_matnona,
                     Y = recurrence,
                     ncomp = 3)
```

```{r fig.width=6, fig.height=5}
plotIndiv(plsda_rest1,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

```

Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.

Therefore we decided to exclude patient 6 from further analyses.

```{r}
plsda_rest2 <- plsda(X = t_expr_matnona,
                     Y = mixture,
                     ncomp = 3)
```

```{r fig.width=6, fig.height=5}
plotIndiv(plsda_rest2,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

```

We can say that the biological differences between initial and recurrent status are stronger than the differences that could come from potential batch effects.

### Exploratory sPLS-DA without patient 6

```{r}
plsda_restwo6 <- plsda(X = t_expr_matnonawo6,
                     Y = recurrencewo6,
                     ncomp = 3)
```

```{r}
plotIndiv(plsda_restwo6,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim wo Pat6", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

```

#### GGplot-based visualization of Exploratory sPLS-DA

```{r}
# preprocess pca results 
plsda_variates <- plsda_restwo6$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id")

plsda_variateswpat6 <- plsda_rest1$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id")

```

```{r fig.width=7.3, fig.height=5}
ggplot(data = plsda_variateswpat6,
       aes(x = comp1, y = comp2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
  labs(title = "sPLS-DA plot of samples\nBased on the abundance values of proteins") +
  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.

Therefore we decided to exclude patient 6 from further analyses.

```{r fig.width=7.3, fig.height=5}
ggplot(data = plsda_variates,
       aes(x = comp1, y = comp2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
  labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

The difference between initial and recurrent tumor is better resolved in the sPLS-DA plot when excluding patient 6.

## Sparcity reduction

Only proteins that are identified and quantified in 2 out of 3 TMT mixtures will be kept for the inferential analysis. The proteins missing in 1 mixture will be imputed with `missForest`.

__Function to select proteins to exclude:__

```{r}
sel_proteins_missing <- function(long_matrix,
                                 threshold) {

na_count <- group_by(long_matrix,
                     Protein, mixture) %>%
    summarise(na_count = sum(is.na(Abundance)),
              total = n()) %>% 
    ungroup() %>% 
    mutate(NA_fraction = na_count/total)

na_count_perbatch <- na_count %>%
  group_by(Protein) %>%
  summarise(na_per_batch = sum(NA_fraction)) %>%
  ungroup()

proteins2exclude <- na_count_perbatch %>%
  filter(na_per_batch > threshold) %>%
  pull(Protein)

return(proteins2exclude)
}

```

__Excluded proteins:__

```{r}
proteins2exclude <- sel_proteins_missing(quant_annotated,
                                         threshold = 1)
```

Exclude proteins from initial abundance matrix.

```{r}
expr_matrix_filt <- expr_matrix2 %>%
  filter(!Protein %in% proteins2exclude)
```

### Visualize missing values (before and after filtering)

```{r}
nofiltnas_plot <- vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("All quant proteins. Total =", nrow(expr_matrix))) + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

filtnas_plot <- vis_miss(expr_matrix_filt) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("Only present in 2/3 TMT mixtures. Total =", nrow(expr_matrix_filt))) + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))
```

```{r fig.height=6, fig.width=12}
cowplot::plot_grid(nofiltnas_plot, filtnas_plot, ncol = 2)
```

## `missForest` imputation:

```{r}
mat_filt <- expr_matrix_filt %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt <- t(mat_filt)

if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){

 t_imp_mat_filt <- missForest::missForest(t_mat_filt)
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

 saveRDS(t_imp_mat_filt,
         file = here("rds/missforest_imp_mat_filt.Rds"))

} else {

 t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

}

```

## Inferential analysis

__Protein to Gene annotation__

```{r}
prot2gene <- tmt_protdata_frag %>%
  dplyr::select(Protein = index, Gene = gene)
```

__Prep protein abundance matrix__

```{r}
mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")
```

### Set up design matrix

Initial vs recurrent samples are patient-matched, therefore the design matrix is setup so the linear model can account for the patient random effect.

```{r}
design_wo6 <- model.matrix(~patientwo6+recurrencewo6)

rownames(design_wo6) <- sample_annotationwo6$paired_id
```

Check design matrix:

```{r}
print(design_wo6)
```

```{r}
mat_filt_impwo6[1:3,1:12]
```

### Fit limma

```{r}
source(here("scr/fit_limmawo6.R"))
```

```{r}
# fit limma
limma_tab_wo6 <- fit_limmawo6(mat_filt_impwo6, 
                         design_wo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect wo Pat6",
                         prot2gene = prot2gene)

#get proteins increased in recurrence
increased_in_rec <- limma_tab_wo6 %>%
  filter(logFC > 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)

#get proteins decreased in recurrence
decreased_in_rec <- limma_tab_wo6 %>%
  filter(logFC < 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)
```

### Volcano plots

```{r fig.width=8, fig.height=6}
our_volcano(limma_tab_wo6, 
            FC_cutoff = 0, 
            pval_cutoff = 0.05, 
            color_diffex = "red", 
            color_nondifex = "#2a9d8f", 
            interesting_proteins = NULL, 
            vert_line_col = "red",
            hline_col = "red", 
            hline_pos = 0.05, 
            linetype = "dashed",
            increased_in = "Recurrent", 
            comparison_title = "Recurrent vs Initial") +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

```{r fig.width=8, fig.height=6, echo=FALSE}
protlabels <- limma_tab_wo6 %>%
      filter(logFC > 0 | logFC < 0,
             adj.P.Val < 0.05)

signi_hits <- nrow(protlabels)

signi_up <- limma_tab_wo6 %>%
      filter(logFC > 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

signi_down <- limma_tab_wo6 %>%
      filter(logFC < 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

size <- 0.4

volcano <- ggplot(data = limma_tab_wo6,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      ggrepel::geom_text_repel(data = protlabels,
                               aes(label = Gene), size = 1.5) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Primary")+
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

```

```{r echo=FALSE}
ggsave(plot = volcano, 
       filename = here::here("figures/volcano_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano, 
       filename = here::here("figures/volcano_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

```{r fig.width=8, fig.height=6, echo=FALSE}
protlabels <- limma_tab_wo6 %>%
      filter(logFC > 0 | logFC < 0,
             adj.P.Val < 0.05,
             Gene %in% c("ASAH1", "GPNMB", "SYNM"))

signi_hits <- nrow(protlabels)

size <- 0.4

volcano2 <- ggplot(data = limma_tab_wo6,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      ggrepel::geom_text_repel(data = protlabels,
                               aes(label = Gene), size = 3.5) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Primary")+
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

```

```{r echo=FALSE}
ggsave(plot = volcano2, 
       filename = here::here("figures/volcano2_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano2, 
       filename = here::here("figures/volcano2_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

### Boxplots specific proteins

```{r}
quant_anwgene <- left_join(quant_annotated, prot2gene)

tab_2_paired_boxplot <-  quant_anwgene %>%
  filter(Gene %in% c("ASAH1", "SYNM", "GPNMB"))
# prep data
  slim_data <- pivot_longer(expr_matrix_filt, 
                          cols = 2:length(names(expr_matrix_filt)), 
                          names_to = c("patient","condition"),
                          values_to = "Intensity",
                          names_sep = "\\_") %>% 
  group_by(Protein)

  slim_data2 <- pivot_wider(slim_data, 
                          values_from = Intensity, 
                          names_from = condition) %>%
    filter(patient != "x6")

  slim_data3 <- slim_data2 %>% 
    mutate(Abs_diff = rec-prim) %>%
    mutate(rank = row_number(Abs_diff)) %>% 
    dplyr::rename(Initial = prim, Recurrent = rec) %>%
    ungroup() %>%
    filter(Protein %in% c("Q13510", "Q14956", "O15061")) %>%
    mutate(Gene = case_when(Protein == "Q13510" ~ "ASAH1",
                              Protein == "Q14956" ~ "GPNMB",
                              Protein == "O15061" ~ "SYNM"))

```

```{r fig.width=7, fig.height=3.5}
ggpaired(slim_data3,
         cond1 = "Initial",
         cond2 = "Recurrent", fill = "condition",
         ylab = "Normalized Abundance",
         label = NULL,
         repel = TRUE,
         facet.by = "Gene", point.size = 0.2, line.size = 0.1) +
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom",
        strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) 
```

```{r echo=FALSE}
paired_boxplots <- ggpaired(slim_data3,
         cond1 = "Initial",
         cond2 = "Recurrent", fill = "condition",
         ylab = "Normalized Abundance",
         label = NULL,
         repel = TRUE,
         facet.by = "Gene", point.size = 0.2, line.size = 0.1) +
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.height= unit(2, 'mm'),
        legend.key.width= unit(2, 'mm'),
        legend.position="bottom",
        strip.text = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360)) 
```

```{r echo=FALSE}
ggsave(plot = paired_boxplots, 
       filename = here::here("figures/paired_box_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 61,
       height = 45)

ggsave(plot = paired_boxplots, 
       filename = here::here("figures/paired_box_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 61,
       height = 45)
```

## ID coverage (final)

### Coverage plot

```{r}
ids_summary2 <- quant_annotated %>% 
  dplyr::select(Protein, mixture, Abundance) %>% 
  distinct(Protein, mixture, .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "Quant_in_all",
                   `Nr of IDs` = nrow(expr_matnona))) %>%
  bind_rows(.,
            tibble(mixture = "Quant_2/3_mixtures",
                   `Nr of IDs` = nrow(mat_filt_impwo6))) %>%
  bind_rows(.,
            tibble(mixture = "Overall",
                   `Nr of IDs` = nrow(expr_matrix)))

ids_summary2 <- mutate(ids_summary2,
                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
                                           mixture == "tmt_2" ~ "Mix_2",
                                           mixture == "tmt_3" ~ "Mix_3",
                                           mixture == "Quant_in_all" ~ "Quant. in all",
                                           mixture == "Overall" ~ "Overall",
                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
  mutate(Mixture = factor(Mixture, levels = c("Overall","Quant. in 2/3", "Quant. in all",
                                               "Mix_3", "Mix_2", "Mix_1")))
```

Number of identified proteins: `r nrow(expr_matrix)`

Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`

Number of identified and quantified proteins 2 out of 3 batches: `r nrow(mat_filt_impwo6)`

```{r fig.width=7, fig.height=4}
ggplot(data = ids_summary2,
       aes(y = `Nr of IDs`, x = mixture)) +
  geom_col(fill = "#008683") +

  geom_text(size = 5, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Proteins by Mixture") +
  ylab("Nr of IDed and Quant Proteins") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 14, angle = 0),
        axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 12, angle = 0),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey"),
        panel.border = element_rect(colour = "black", fill=NA, size=1.5),
        axis.title=element_text(size=12, face="bold"))
```
```{r echo=FALSE}
barplot_cover <- ggplot(data = ids_summary2,
       aes(y = `Nr of IDs`, x = Mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 2, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r echo=FALSE}
ggsave(plot = barplot_cover, 
       filename = here::here("figures/barplot_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = barplot_cover, 
       filename = here::here("figures/barplot_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

## Median abundance distribution of filtered and inputed samples

```{r}
expr_matrix_filtwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6"))

long_mat_f <- pivot_longer(expr_matrix_filtwo6,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_3 <- long_mat_f %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq2 <- long_mat_3 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

paired_annot_uniq2$paired_id
## Annotated abundance data in long format ----

quant_annotated_f <- left_join(long_mat_3,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"),
         paired_id = factor(paired_id, levels = paired_annot_uniq2$paired_id)) %>%
  mutate(Sample = str_replace_all(paired_id, pattern = "prim", replacement = "init"),
         Stage = if_else(recurrence == "prim",
                         true = "Initial",
                         false = "Recurrent"))

```

### By TMT channel/sample {.unnumbered}

#### General Abudance distribution by TMT channel/sample {.unlisted .unnumbered}

```{r}
qcplot_f <- ggplot(quant_annotated_f,
                 mapping = aes(x = Sample, 
                               y = Abundance, 
                               fill = Stage)) +
          geom_boxplot(outlier.size=0.1, size = 0.25) + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0, size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_blank(),
                panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                axis.title=element_text(size=8),
                axis.title.x = element_blank(),
                legend.text = element_text(size = 7),
                legend.title = element_text(size = 8),
                legend.key.height= unit(3, 'mm'),
                legend.key.width= unit(3, 'mm'),
                legend.position="bottom")

```

```{r fig.height=4, fig.width=5, warning=FALSE, echo=FALSE}
print(qcplot_f)
```
```{r echo=FALSE}
ggsave(plot = qcplot_f, 
       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = qcplot_f, 
       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

## Over-representation analyses/tests (ORAs)

### Prep tabular info for ORAs

```{r}
tab_sig_prots_tryptic <- bind_rows(tibble(protein = increased_in_rec,
                                  characteristic = "up-regulated"),
                           tibble(protein = decreased_in_rec,
                                  characteristic = "down-regulated"))

limma_tab_wo62_tryptic <- limma_tab_wo6

unip2symb_tryptic <- bitr(limma_tab_wo62_tryptic$Protein, 
                  fromType = "UNIPROT", 
                  toType = c("SYMBOL", "ENTREZID"),
                  OrgDb = org.Hs.eg.db) %>%
  dplyr::rename(protein = UNIPROT) %>%
  mutate(Protein = protein)

tab_sig_prots_tryptic <- left_join(tab_sig_prots_tryptic, unip2symb_tryptic)

unip2symbIDed_tryptic <- bitr(identified_proteins$protein_id, 
                  fromType = "UNIPROT", 
                  toType = c("SYMBOL", "ENTREZID"),
                  OrgDb = org.Hs.eg.db)

limma_tab_wo62_tryptic <- left_join(limma_tab_wo62_tryptic, unip2symb_tryptic)

geneList <- limma_tab_wo62_tryptic$logFC
names(geneList) <- limma_tab_wo62_tryptic$Protein
geneList <-  sort(geneList, decreasing = TRUE)

geneListr <- limma_tab_wo62_tryptic$logFC
names(geneListr) <- limma_tab_wo62_tryptic$ENTREZID
geneListr <-  sort(geneListr, decreasing = TRUE)
```

### Reactome

```{r}
group_comparison_react <- compareCluster(ENTREZID~characteristic, 
                                              data=tab_sig_prots_tryptic, 
                                              fun="enrichPathway",
                                              organism = "human",
                                              pvalueCutoff = 0.05,
                                              pAdjustMethod = "BH",
                                              qvalueCutoff = 0.2,
                                              universe = unip2symbIDed_tryptic$ENTREZID,
                                              minGSSize = 10,
                                              maxGSSize = 1000,
                                              readable = TRUE)
```

#### Dotplot Reactome

```{r}
enrichplot::dotplot(group_comparison_react, x = "characteristic") + 
          xlab("Quant in Recurrent status") +
  scale_color_continuous(low="red", high="blue",
            guide=guide_colorbar(reverse=TRUE),breaks = c(0.01, 0.04)) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r}
dotplot_react <- enrichplot::dotplot(group_comparison_react, x = "characteristic") + 
          xlab("Quant in Recurrent status") +
  scale_color_continuous(low="red", high="blue",
            guide=guide_colorbar(reverse=TRUE),breaks = c(0.01, 0.04)) +
  scale_size_continuous(breaks = c(0.10, 0.25)) +
  theme(axis.text.x = element_text(hjust = 0.4, vjust = 0.1, size = 6, angle = 340),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 6),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.height= unit(2, 'mm'),
        legend.key.width= unit(2, 'mm'),
        legend.position="bottom")
```
```{r}
ggsave(plot = dotplot_react, 
       filename = here::here("figures/dotplot_reactome_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 91,
       height = 90)

ggsave(plot = dotplot_react, 
       filename = here::here("figures/dotplot_reactome_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 91,
       height = 90)
```

#### Cnet plot Reactome

```{r}
enrichplot::cnetplot(group_comparison_react) + 
  theme(legend.text = element_text(size = 7),
                legend.title = element_text(size = 8),
                legend.key.height= unit(3, 'mm'),
                legend.key.width= unit(3, 'mm'),
                legend.position="bottom")
```

```{r}
cnet_plot <- enrichplot::cnetplot(group_comparison_react) + 
  theme(legend.text = element_text(size = 7),
        text = element_text(size = 4),
                legend.title = element_text(size = 8),
                legend.key.height= unit(3, 'mm'),
                legend.key.width= unit(3, 'mm'),
                legend.position="bottom")
```

```{r}
ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 110,
       height = 110)

ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 110,
       height = 110)

ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.svg"), 
       device = "svg",
       units = "mm",
       width = 110,
       height = 110)
```

## Protein coverage plots

```{r}
source(here("scr/protein_peptide_coverage_plot.R"))
```

### Annotate peptides from interesting proteins

```{r}
source(here("scr/annotate_peptides.R"))
```

```{r}
# get protein to peptide pairs
pept2prot_inter <- identified_peptides %>%
  filter(gene %in% c("ASAH1", "SYNM", "GPNMB")) %>%
  dplyr::select(Peptide = sequence, Genes = protein_id)

annotated_peptides <- annotate_peptides(expr_mat = pept2prot_inter,
                                        fasta = fasta_ident_prots)

annotated_peptides2 <- annotated_peptides %>%
  mutate(peptide_length = str_length(Peptide))
```

### ASAH1

```{r}
feat_asah1 <- drawProteins::get_features("Q13510")

feat_asah1df <- drawProteins::feature_to_dataframe(feat_asah1)
#Get peptides as coverage features for plotting

cov_feat_asah1 <- get_coverage(annotated_peptides = annotated_peptides2, 
                               id = "Q13510")
#Merge coverage feature info with protein sequence features info:

fullfeat_asah1 <- add_peptides(feat_asah1df,
                               peptide_coverage_data = cov_feat_asah1)
# drawCanvas
canvas <- draw_canvas(fullfeat_asah1)

# draw the protein chain
wchain <- draw_chains(canvas, fullfeat_asah1)

# draw protein domains
wdomainsasah1 <- draw_domains(wchain, data = fullfeat_asah1)
#pasa <- drawProteins::draw_folding(wdomainsasah1, fullfeat_asah1)
cov_feat_asah1$order <- 1.4

pepcov_asah1 <- draw_peptides(wdomainsasah1, cov_feat_asah1)
pepcov_asah1
```

```{r}
ggsave(plot = pepcov_asah1, 
       filename = here::here("figures/asah1_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_asah1, 
       filename = here::here("figures/asah1_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

### SYNM coverage plot
Get features from Uniprot

```{r}
feat_SYNM <- drawProteins::get_features("O15061")
feat_SYNMdf <- drawProteins::feature_to_dataframe(feat_SYNM)
#Get peptides as coverage features for plotting

cov_feat_SYNM <- get_coverage(annotated_peptides = annotated_peptides2, 
                               id = "O15061")
#Merge coverage feature info with protein sequence features info:

fullfeat_SYNM <- add_peptides(feat_SYNMdf,
                               peptide_coverage_data = cov_feat_SYNM)
# drawCanvas
canvasSYNM <- draw_canvas(fullfeat_SYNM)

# draw the protein chain
wchainSYNM <- draw_chains(canvasSYNM, fullfeat_SYNM)
cov_feat_SYNM$order <- 1.4

pepcov_SYNM <- draw_peptides(wchainSYNM, cov_feat_SYNM)
pepcov_SYNM
```

```{r}
ggsave(plot = pepcov_SYNM, 
       filename = here::here("figures/SYNM_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_SYNM, 
       filename = here::here("figures/SYNM_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

# Analysis of proteolytic processing

## Load semi-specific search results

```{r}
tmt_peptdata_frag <- read_tsv(here("data/semi_specific_search_fragpipe17/tmt-report/abundance_peptide_MD.tsv")) %>% 
  janitor::clean_names()

pept_ids_mix1 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_1/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("empty")))

pept_ids_mix2 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_2/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("x"), starts_with("empty"), "master"))

pept_ids_mix3 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_3/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("x"), starts_with("empty"), "master"))

identified_proteins_semi <- read_tsv(here("data/semi_specific_search_fragpipe17/combined_protein.tsv")) %>% 
  janitor::clean_names()

```

## Wrangle and merge peptide ID/quant with sample annotation

```{r}
# wide expression matrix with NAs
pept_matrix <- dplyr::select(tmt_peptdata_frag,
                             index, starts_with("x"))

pept_long_mat <- pivot_longer(pept_matrix,
                         cols = starts_with("x"),
                         names_to = c("patient"),
                         values_to = "Abundance") %>%
  mutate(recurrence = str_extract(patient, "[^0-9]+$")) %>%
  mutate(patient = str_remove_all(patient, "[^0-9]+$"))

pept_long_mat_2 <- pept_long_mat %>%
  mutate(paired_id_semi = paste(patient, recurrence, sep = "_"))

sample_annotation2 <- sample_annotation2 %>%
  mutate(paired_id_semi = paste(patient, recurrence, sep = "_"))

## Annotated abundance data in long format ----

pept_quant_annotated <- left_join(pept_long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id_semi", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

## Wide xpression matrices without NAs
pept_expr_matnona <- column_to_rownames(pept_matrix,
                            "index") %>%
          as.matrix() %>%
          na.omit()
```

Wrangle and merge peptide ids for further annotation based on they proteolytic specificity

```{r}
peptide_joined1 <- full_join(pept_ids_mix1, pept_ids_mix2) %>%
                    mutate(duplicated = duplicated(peptide))

peptide_joined <- full_join(peptide_joined1, pept_ids_mix3) %>%
                    mutate(duplicated = duplicated(peptide))

peptide_ids_filtered <- filter(peptide_joined,
                               peptide %in% peptide_joined$peptide) %>%
                    distinct(peptide, assigned_modifications,observed_modifications, .keep_all = TRUE) %>%
                    dplyr::rename(Peptide = peptide) %>%
                    dplyr::mutate(Genes = protein_id)

```

### Annotate peptides with cleavage information

```{r}
fasta1 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_1/protein.fas"),
                            seqtype = "AA", as.string = TRUE)

fasta2 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_2/protein.fas"),
                             seqtype = "AA", as.string = TRUE)

fasta3 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_3/protein.fas"),
                             seqtype = "AA", as.string = TRUE)

fasta22 <- append(fasta1, fasta2)

fasta <- append(fasta22, fasta3)

fasta <- discard(fasta,
                 duplicated(names(fasta)))

attr(fasta[[1]], "Annot")

extr_annt <- function(x){
  ext <- attr(x, "Annot")
}

names <- purrr::map_chr(fasta, .f = extr_annt)

seqinr::write.fasta(fasta,
                    file.out = here("data/semi_specific_search_fragpipe17/protein_combined.fas"),
                    names = names)

if(!file.exists(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))){
  cleavage_annoated_peptides <- annotate_peptides(expr_mat = peptide_ids_filtered, 
                                                  fasta = fasta,
                                                  decoy_tag = "rev_")

  write_tsv(cleavage_annoated_peptides, here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))
} else {
  cleavage_annoated_peptides <- read_tsv(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))
}

cleavage_annoated_peptides <- dplyr::mutate(cleavage_annoated_peptides,
                                            Genes = protein_id)

peptides_annotated <- left_join(peptide_ids_filtered, cleavage_annoated_peptides,
                                by = c("Peptide", "protein_id"))

```

## Sparcity reduction

Only peptides that were identified and quantified in all 3 mixtures would be used for further inferential analyses for proteolytic processing.

```{r}
pept_matrix_nona <- pept_matrix %>% 
  na.omit()

pept_nona <- pept_matrix_nona %>%
  dplyr::select(index) %>%
  separate(col = index, 
           sep = "_", into = c("protein", "peptide"))

pept_matrix_nonawop6 <- pept_matrix %>% 
  dplyr::select(-starts_with("x6")) %>%
  na.omit()

```

## Coverage information (peptide-level)

```{r}
ids_summarypep <- pept_quant_annotated %>% 
  dplyr::select(index, mixture, Abundance) %>% 
  distinct(index, mixture, .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "In all three",
                   `Nr of IDs` = nrow(pept_matrix_nona)))# %>%
  #bind_rows(.,
  #          tibble(mixture = "Overall",
  #                `Nr of IDs` = nrow(pept_matrix)))

ids_summarypep <- mutate(ids_summarypep,
                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
                                           mixture == "tmt_2" ~ "Mix_2",
                                           mixture == "tmt_3" ~ "Mix_3",
                                           mixture == "In all three" ~ "Quant. in all",
                                           mixture == "Overall" ~ "Overall",
                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
  mutate(Mixture = factor(Mixture, levels = c("Overall", "Quant. in all",
                                               "Mix_3", "Mix_2", "Mix_1")))
```

```{r fig.height=4, fig.width=7}
ggplot(data = ids_summarypep,
       aes(y = `Nr of IDs`, x = Mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 5, position = position_stack(vjust = 0.5), 
            aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Peptides by Mixture") +
  ylab("Nr of IDed and Quant Peptides") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```
```{r}
barplot_pept_cover <- ggplot(data = ids_summarypep,
       aes(y = `Nr of IDs`, x = Mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 2, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
  coord_flip() +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```  
```{r}
ggsave(plot = barplot_pept_cover, 
       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 38,
       height = 50)

ggsave(plot = barplot_pept_cover, 
       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 38,
       height = 50)
```

## Label N-terminal peptides

```{r echo=TRUE}
source(here("scr/annotate_nterm.R"))
```

```{r echo=TRUE}
nterannot <- annotate_nterm(peptides_annotated,
                            tmtmass = 229.1629) %>%
  clean_names()
```

### Merge peptide annotation with TMT integrator quant report

```{r}
# all quantified peptides
tmt_reprt_annot <- left_join(tmt_peptdata_frag, nterannot)

check_na <- filter(tmt_reprt_annot, is.na(semi_type))

excluded_tmt_reprt_annot <- anti_join(nterannot, tmt_reprt_annot) %>% 
                    clean_names()

#only those present in all mixtures
tmt_peptdata_fragnona <- tmt_peptdata_frag %>%
  filter(peptide %in% pept_nona$peptide)

tmt_reprt_annotnona <- left_join(tmt_peptdata_fragnona, nterannot) 
```

## Summarize feature counts per feature

```{r echo=TRUE}
source(here("scr/summarize_peptide_counts.R"))
```

```{r echo=TRUE}
summary_count <- summarize_peptide_counts(tmt_reprt_annot) 
```

### Barplot

These are the ones that would be used for inferential statistics.

```{r echo=TRUE}
to_count_info <- tmt_reprt_annot %>% 
   filter(index %in% pept_matrix_nona$index) %>%
                      dplyr::select(index, specificity, nterm, semi_type, tmt_tag,
                                    aa_after, aa_before, following_10_resid, previous_10_resid, start_position, end_position) %>%
  mutate(specificity = if_else(str_detect(semi_type, "unspecific"), 
                               true = "semi_specific",
                               false = specificity),
         duplicated = duplicated(index)) %>%
  distinct()

to_count_info_semi <- to_count_info %>%
                           dplyr::select(index,specificity) %>%
                           distinct() %>%
  mutate(duplicated = duplicated(index))

false_duplication <- to_count_info_semi %>%
  filter(duplicated) %>%
  pull(index)

to_count_info_semi <- to_count_info_semi %>%
  mutate(specificity = if_else(index %in% false_duplication, 
                               true = "semi_specific",
                               false = specificity)) %>%
  dplyr::select(-duplicated) %>%
  distinct()

  n_semi <- dplyr::count(to_count_info_semi, specificity) %>% 
                      dplyr::rename(feature_type = specificity) %>%
                      dplyr::mutate(category = "specificity")

  n_total <- tibble(feature_type = "Total",
                    n = length(unique(to_count_info$index)),
                    category = "Total")

  n_term <- dplyr::count(to_count_info, nterm) %>% 
                      dplyr::rename(feature_type = nterm) %>%
                      dplyr::mutate(category = "N-term")

  summary_count_pept_nona <- bind_rows(n_semi,
                                       n_term,
                                       n_total)

total_sumcount <- summary_count_pept_nona %>%
  filter(feature_type == "Total") %>%
  pull(n)

summary_countnona_no_tot <- summary_count_pept_nona %>%
  dplyr::filter(category != "Total") %>%
  dplyr::rename(`Feature category` = category)
```

```{r fig.width=5, fig.height=2, echo=FALSE}
ggplot(summary_countnona_no_tot,
       aes(x = feature_type, y = n, fill = `Feature category`)) + 
  coord_flip()+ 
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 2.5) + 
  scale_y_continuous(limits = c(0, 17000), breaks = c(0, 7000, 15000)) + 
  labs(y = "Number of IDs", x = "Feature type") +
  facet_wrap(~`Feature category`, nrow = 1, scales="free") +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom", 
        strip.background = element_blank(),
        strip.text = element_blank())
```

```{r echo=FALSE}
summary_pept_annot <- ggplot(summary_countnona_no_tot,
       aes(x = feature_type, y = n, fill = `Feature category`)) + 
  coord_flip()+ 
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 2) + 
  scale_y_continuous(limits = c(0, 20000), breaks = c(0, 7000, 15000)) + 
  labs(y = "Number of IDs", x = "Feature type", 
       caption = paste0("Total = ",total_sumcount)) +
  facet_wrap(~`Feature category`, nrow = 1, scales="free") +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom", 
        strip.background = element_blank(),
        strip.text = element_blank())

```
```{r}
ggsave(plot = summary_pept_annot, 
       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 76,
       height = 50)

ggsave(plot = summary_pept_annot, 
       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 76,
       height = 50)
```

## Inferential analysis (peptide-level)

### Prep design matrix

```{r}
design_limmawo6 <- model.matrix(~patientwo6+recurrencewo6)

sample_annotationwo6 <- sample_annotationwo6 %>%
  mutate(paired_id_semi = paste(patient, recurrence, sep = ""))

rownames(design_limmawo6) <- sample_annotationwo6$paired_id_semi
```

### Prep abundance matrices

```{r}
pep_matwo6 <- dplyr::select(pept_matrix_nonawop6,
                             index, rownames(design_limmawo6))  %>%
  column_to_rownames("index") %>%
  as.matrix()

```

```{r}
colnames(pep_matwo6)
```
```{r}
rownames(design_limmawo6)
```

The column names in our peptide abundance matrix have the same order as the row names in our design matrix.

### Peptide-level model fit

```{r}
fit_limmapeptwo6 <- function(mat, design, method,Limma){
  limmafit <- lmFit(mat, design, method = method)
  limmafit <- eBayes(limmafit)
  limma_tab <- topTable(limmafit, coef = "recurrencewo6rec", number = Inf, adjust.method = "BH") %>%
  mutate(Protein = rownames(.),
         Limma = Limma)

  return(limma_tab)
}
```

```{r echo=TRUE, warning=FALSE, message=FALSE}
fit_mod_peptwo6 <- fit_limmapeptwo6(pep_matwo6, 
                         design_limmawo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect")

compar_tab_peptwo6 <-  fit_mod_peptwo6 %>%
                    separate(Protein, 
                             into = c("protein", "peptide"), 
                             remove = FALSE) %>%
  dplyr::rename(index = Protein)

```

### Extract coefficients and feature specific FDR correction

#### Interesting features

```{r}
features1 <- tmt_reprt_annotnona %>%
                    dplyr::select(peptide, index, specificity, nterm, 
                                  semi_type, is_terminal)

interesting_features1 <- features1 %>%
                    filter(specificity == "semi_specific",
                           is_terminal == "not_terminal") 
```

```{r}
# load function for feature specific fdr correction
source(here("scr/feature_specific_fdr_correction.R"))
```

```{r}
# load function for volcano plot
source(here("scr/plot_volcano.R"))
```

```{r}
# excluding patient 6
compar_tab_correctedwo6_pept <- feature_fdr_correction(compar_tab_peptwo6,
                                               features1,
                                               interesting_features1)

compar_tab_interestingwo6_pept <- compar_tab_correctedwo6_pept %>% 
  dplyr::rename(Protein = protein) %>%
  filter(fdr_correction == 'feature-specific') %>%
  left_join(.,prot2gene)

increased_recwo6_pept <- compar_tab_interestingwo6_pept %>%
                    filter(logFC>0,
                           adj.P.Val < 0.05) %>% 
  pull(peptide)

decreased_recwo6_pept <- compar_tab_interestingwo6_pept %>%
                    filter(logFC<0,
                           adj.P.Val < 0.05) %>% 
  pull(peptide)

check_increased_pept <- cleavage_annoated_peptides %>%
  filter(Peptide %in% increased_recwo6_pept)

```

```{r}
length(increased_recwo6_pept)
```

```{r}
length(decreased_recwo6_pept)
```

`r length(increased_recwo6_pept)` semi-specific peptides are increased in Recurrent tumor and `r length(decreased_recwo6_pept)` are decreased in Recurrent tumor.

### Volcano peptides

```{r fig.width=8, fig.height=6}
recwo6_pept22 <- compar_tab_interestingwo6_pept %>%
  filter(adj.P.Val < 0.05)

size <- 0.4

volcano_pept <- ggplot(data = compar_tab_interestingwo6_pept,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC > 0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC < -0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
                 size = size) +
      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Primary") + 
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

```

```{r}
ggsave(plot = volcano_pept, 
       filename = here::here("figures/volcano_proteolysis_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano_pept, 
       filename = here::here("figures/volcano_proteolysis_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

## Proportional abundance of proteolytic peptides

### Prepare data

__Wrangle data__

```{r}
pept_semi_annot <- peptides_annotated %>%
  dplyr::select(Peptide, specificity, is_terminal)

pept_quant_to_summary <- pept_quant_annotated %>%
  separate(col = index, 
           into = c("Protein", "Peptide"), 
           sep = "\\_", 
           remove = FALSE) %>%
  left_join(.,pept_semi_annot)

pep_quant_presummary <- pept_quant_to_summary %>%
  dplyr::select(Protein, Peptide, recurrence, paired_id, 
                Abundance, specificity, is_terminal)

```

__Prepare data-frame with abudance percentage of semi-specific peptides per sample__

```{r}
pept_summary_int_all <- pep_quant_presummary %>%
  group_by(paired_id, recurrence) %>%
  summarise(Sum_All = sum(Abundance, na.rm = TRUE))

pept_summary_int_semi <- pep_quant_presummary %>%
  filter(specificity == "semi_specific",
         is_terminal == "not_terminal") %>%
  group_by(paired_id, recurrence) %>%
  summarise(Sum_Semi = sum(Abundance, na.rm = TRUE))

pept_sum_summary <- left_join(pept_summary_int_all, pept_summary_int_semi) %>%
  mutate(Percentage = Sum_Semi/Sum_All * 100,
         Stage = if_else(recurrence == "prim",
                         true = "Initial",
                         false = "Recurrence"))

```

### Generate plot % of abundance of semi-specific peptides

```{r}
plotSumEM <- ggplot(pept_sum_summary, 
                    aes(x = Stage, 
                        y = Percentage, fill = Stage, 
                        cex.axis = 1.5)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5) +
  # Box plot with jittered points
  # 0.2 : degree of jitter in x direction
  # geom_jitter(shape=16, position=position_jitter(0.2))
  ylab("Proportional intensity of semi-tryptic peptides [%]") +
  geom_signif(
    comparisons = list(c("Initial", "Recurrence")),
    map_signif_level = TRUE
  ) + 
  stat_compare_means(method="t.test") +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

```

```{r}
plotSumEM
```

```{r}
ggsave(plot = plotSumEM, 
       filename = here::here("figures/box_perc_semispec_pepts.tiff"), 
       device = "tiff",
       units = "mm",
       width = 40,
       height = 50)

ggsave(plot = plotSumEM, 
       filename = here::here("figures/box_perc_semispec_pepts.eps"), 
       device = "eps",
       units = "mm",
       width = 40,
       height = 50)
```

## Comparative analysis of semi-specific peptides vs protein abundance

### Prepare data

#### Get log2FC Summary of abundances of semi-specific peptides

```{r}
#taking output from limma semi-specific
log2_semi_tryptic <- dplyr::select(compar_tab_interestingwo6_pept,
                                   logFC, Protein, peptide, specificity, 
                                   is_terminal,
                                   Gene) %>%
  filter(specificity == "semi_specific",
         is_terminal == "not_terminal") %>%
  dplyr::rename(logFC_semi_peptide = logFC) %>%
  dplyr::select(-c(specificity, is_terminal))

```

#### Get log2 Summary of protein abundances based only on fully-tryptic peptides

```{r}
med_prot_quant_specific <- pep_quant_presummary %>%
  group_by(Protein, paired_id) %>%
  summarise(`median(Abundance)` = median(Abundance, na.rm = TRUE)) %>%
  ungroup()
```

##### Prep wide matrix of median protein abundances from only fully-tryptic peptides

```{r}
# prep wide matrix for median proteins abundances (arousing only from fully-tryptic peptides)
wide_premat_prot_only_tryp <- med_prot_quant_specific %>%
  pivot_wider(id_cols = c("Protein"),
              values_from = `median(Abundance)`, 
              names_from = paired_id) %>%
  # keep only proteins present in 2 / 3 of the samples
  filter(!Protein %in% proteins2exclude)

mat_prot_only_tryp <- wide_premat_prot_only_tryp %>%
  column_to_rownames("Protein") %>%
  as.matrix()
```

__missForest-based imputation__

```{r}
t_mat_prot_only_tryp <- t(mat_prot_only_tryp)

if(!file.exists(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))){

 imp_prot_only_tryp <- missForest::missForest(t_mat_prot_only_tryp)
 t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp

 mat_prot_only_tryp <- t(t_mat_prot_only_tryp)

 saveRDS(imp_prot_only_tryp,
         file = here("rds/missforest_imp_mat_prot_only_tryp.Rds"))

} else {

 imp_prot_only_tryp <- readRDS(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))
 t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp

 mat_imp_prot_only_tryp <- t(t_mat_imp_prot_only_tryp)

}

```

__Run limma on this new data to generate log2 FCs of proteins from only fully-tryptic peptides__

```{r}
split <- str_split_fixed(colnames(mat_imp_prot_only_tryp), pattern = "\\_", n = 2)

patient_tryp <- split[,1]
recurrence <- split[,2]

design_tryp <- model.matrix(~patient_tryp+recurrence)

rownames(design_tryp) <- colnames(mat_imp_prot_only_tryp)
```

```{r}
rownames(design_tryp) 
```

```{r}
colnames(mat_prot_only_tryp)
```

```{r}
fit_limmawo6_tryp <- function(mat, design, method,Limma, prot2gene){
  limmafit <- lmFit(mat, design, method = method)
  limmafit <- eBayes(limmafit)
  limma_tab <- topTable(limmafit, coef = "recurrencerec", number = Inf, adjust.method = "BH") %>%
  mutate(Protein = rownames(.),
         Limma = Limma)

  limma_tab <- left_join(limma_tab, prot2gene)

  return(limma_tab)
}
```

```{r}
limma_tab_wo6tryp <- fit_limmawo6_tryp(mat_imp_prot_only_tryp, 
                         design_tryp, 
                         method = 'robust', 
                         Limma = "Robust - Proteins w only tryptic peptides",
                         prot2gene = prot2gene)

```

```{r}
log2_fully_tryptic <- dplyr::select(limma_tab_wo6tryp,
                                   logFC, Protein, Gene) %>%
  dplyr::rename(logFC_fully_tryp_protein = logFC) 
```

#### Merge log2FCs of semi-tryptic peptides vs proteins with fully tryptic peptides

```{r}
log2semipept2_log2protein_spec <- left_join(log2_fully_tryptic, log2_semi_tryptic) %>%
  na.omit() %>%
  mutate(DA_Protein = if_else(Protein %in% tab_sig_prots_tryptic$protein,
                              true = TRUE,
                              false = FALSE),
         DA_peptide = if_else(peptide %in% recwo6_pept22$peptide,
                              true = TRUE,
                              false = FALSE))

```

#### log2FC Proteins vs Semi-tryp peptides (scatter plot)

```{r}
inter_scater <- log2semipept2_log2protein_spec %>%
  filter(Protein %in% check_increased_pept$protein_id)

inter_scater$Gene[duplicated(inter_scater$Gene)] <- NA

#geom_point(data = dataarg %>% filter(logFC < -FC_cutoff,
#                                           adj.P.Val < pval_cutoff),
#                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red") +
#      geom_point(data = dataarg %>% filter(logFC > FC_cutoff,
#                                           adj.P.Val > pval_cutoff),
#                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f") +

scatter_proteolysis_plot <- ggplot(log2semipept2_log2protein_spec, 
                                   aes(x = logFC_fully_tryp_protein, 
                                       y = logFC_semi_peptide)) + 
  geom_smooth(method=lm, se = FALSE, linetype="dashed", size = 0.5, 
              color = "black") + 
  geom_point(aes(color = DA_Protein, shape = DA_peptide), size = 0.6) +
  scale_color_manual(values = c("#2a9d8f", "red")) +
  xlab("log2(FC) - Protein abundances") + 
  ylab("log2(FC) - Semi-specific peptides") + 
  ggrepel::geom_text_repel(data = inter_scater,
                               aes(label = Gene), 
                           size = 1,
                           box.padding = 0.5,
                           max.overlaps = 25) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

cor.test(log2semipept2_log2protein_spec$logFC_fully_tryp_protein,
    log2semipept2_log2protein_spec$logFC_semi_peptide, method = "pearson")
```

```{r}
ggsave(plot = scatter_proteolysis_plot, 
       filename = here::here("figures/scatter_proteolysis_plot_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 63)

ggsave(plot = scatter_proteolysis_plot, 
       filename = here::here("figures/scatter_proteolysis_plot_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 63)
```

## Analysis of differential amino acid usage (iceLogo)

```{r}
source(here("scr/get_cleavage_area.R"))
```

```{r}
cleave_areas <- get_cleave_area(cleavage_annoated_peptides)
```

```{r}
library(dagLogo)
```

**Load background proteome**

Used the identified protein sequences as background proteome.

```{r}
proteome_ided <- prepareProteome(fasta = here("data/semi_specific_search_fragpipe17/protein_combined.fas"), 
                                 species = "Homo sapiens")
```

```{r}
increased_Rec <- filter(cleave_areas$cleave_area20,
                            Peptide %in% increased_recwo6_pept)

decreased_Rec <- filter(cleave_areas$cleave_area20,
                            Peptide %in% decreased_recwo6_pept)

write_tsv(x = increased_Rec,
          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))

write_tsv(x = decreased_Rec,
          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))

increased_Rec_4ice <- filter(cleave_areas$cleave_area20,
                            Peptide %in% increased_recwo6_pept) %>%
                    pull(cleave_area20)

decreased_Rec_4ice <- filter(cleave_areas$cleave_area20,
                            Peptide %in% decreased_recwo6_pept) %>%
                    pull(cleave_area20)
```

**Format peptide sequences**

```{r}
if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))){

  form_peptidesincreased_4ice <- formatSequence(increased_Rec_4ice, 
                                  proteome = proteome_ided)

  write_rds(form_peptidesincreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))
} else {

  form_peptidesincreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))

}
```

```{r}
if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))){

  form_peptidesdecreased_4ice <- formatSequence(decreased_Rec_4ice, 
                                  proteome = proteome_ided)

  write_rds(form_peptidesdecreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))
} else {

  form_peptidesdecreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))

}
```

#### DAU increased

```{r}
bg_mod_ztest_increased <- buildBackgroundModel(form_peptidesincreased_4ice,
                                           proteome = proteome_ided,
                           background = "wholeProteome",
                           testType = "ztest")
```

```{r}
dau_nogroup_increased <- testDAU(form_peptidesincreased_4ice, 
                           dagBackground = bg_mod_ztest_increased)
```

```{r}
dagHeatmap(dau_nogroup_increased) 
```

#### DAU decreased

```{r}
bg_mod_ztest_decreased <- buildBackgroundModel(form_peptidesdecreased_4ice,
                                           proteome = proteome_ided,
                           background = "wholeProteome",
                           testType = "ztest")
```

```{r}
dau_nogroup_decreased <- testDAU(form_peptidesdecreased_4ice, 
                           dagBackground = bg_mod_ztest_decreased)
```

```{r}
dagHeatmap(dau_nogroup_decreased) 
```

# Proteogenomics

## Initial data loading and wrangling

Load id/quant data and prep coverage information

```{r message=FALSE}
## tmt-integrator output loading 
tmt_protdata_frag_progen <- read_tsv(here("data/proteogenomics/sras_cellreports_2021_paper_razor/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

tmt_peptdata_frag_progen <- read_tsv(here("data/proteogenomics/sras_cellreports_2021_paper_razor/tmt-report/abundance_peptide_MD.tsv")) %>% 
  janitor::clean_names()

identified_proteins_progen <- read_tsv(here("data/proteogenomics/sras_cellreports_2021_paper_razor/combined_protein.tsv")) %>% 
  janitor::clean_names()

identified_peptides_progen <- read_tsv(here("data/proteogenomics/sras_cellreports_2021_paper_razor/combined_peptide.tsv")) %>% 
  janitor::clean_names()

```

```{r}
# wide expression matrix with NAs
expr_matrix_progen <- dplyr::select(tmt_protdata_frag_progen,
                             Protein = index, 
                             starts_with("x"))

long_mat_progen <- pivot_longer(expr_matrix_progen,
                         cols = starts_with("x"),
                         names_to = c("paired_id_semi"),
                         values_to = "Abundance") %>%
  mutate(recurrence = str_extract(paired_id_semi, "[^0-9]+$")) %>%
  mutate(patient = str_remove_all(paired_id_semi, "[^0-9]+$"))

long_mat_progen_2 <- long_mat_progen %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq_progen <- long_mat_progen_2 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

# wide peptide abundace matrix with NAs
expr_matrix_pept_progen <- dplyr::select(tmt_peptdata_frag_progen,
                             index, peptide, starts_with("x")) %>%
  dplyr::select(!starts_with("x6"))

expr_matrixnona_pept_progen <- expr_matrix_pept_progen %>%
  na.omit()

long_mat_pept_progen <- pivot_longer(expr_matrix_pept_progen,
                         cols = starts_with("x"),
                         names_to = c("paired_id_semi"),
                         values_to = "Abundance") %>%
  mutate(recurrence = str_extract(paired_id_semi, "[^0-9]+$")) %>%
  mutate(patient = str_remove_all(paired_id_semi, "[^0-9]+$"))

long_mat_pept_progen_2 <- long_mat_pept_progen %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq_pept_progen <- long_mat_pept_progen_2 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

## Annotated abundance data in long format ----

quant_annotated_progen <- left_join(long_mat_progen_2,
                             sample_annotation2) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

quant_annotated_pept_progen <- left_join(long_mat_pept_progen_2,
                             sample_annotation2) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

## Wide xpression matrices without NAs
expr_matnona_progen <- column_to_rownames(expr_matrix_progen,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()

# peptide to protein mapping from combined_peptides.tsv 
protein2peptideall <- dplyr::select(identified_peptides_progen,
                                    protein, 
                                    Peptide = sequence) %>% 
          separate(col = protein,
                   into = c("sp", "Genes", "ensembl"),
                   sep = "\\|") %>%
          dplyr::select(Genes, Peptide) %>%
          distinct() %>%
          mutate(Peptide = str_remove_all(Peptide, "\\[.+?\\]")) %>%
          mutate(Peptide = str_remove_all(Peptide, "^n"))

# peptide to protein mappong from tmt abundance report (quantified and normalized)
protein2peptide_tmtrept <- tmt_peptdata_frag_progen %>%
  dplyr::select(index, protein_id, peptide)

```

## Localization of identified peptides and mapping to variant regions

```{r}
# list of identified peptides associated with 'variant' proteins
var_prots <- protein2peptideall %>%
                    mutate(variant = if_else(str_detect(Genes, "\\_"),
                           true = "variant",
                           false = "canonical")) %>%
                    filter(variant == "variant") %>% 
                    pull(Genes) %>%
                    unique()
```

```{r echo=TRUE}
variant_peptides_1 <- filter(protein2peptideall,
                             Genes %in% var_prots)

# prep fasta containing only identified protein sequences
fastavar <- seqinr::read.fasta(here("data/proteogenomics/sras_cellreports_2021_paper_razor/mix_1/protein.fas"), 
                            seqtype = "AA", 
                            as.string = TRUE)

fastavar2 <- seqinr::read.fasta(here("data/proteogenomics/sras_cellreports_2021_paper_razor/mix_2/protein.fas"), 
                            seqtype = "AA", 
                            as.string = TRUE)

fastavar3 <- seqinr::read.fasta(here("data/proteogenomics/sras_cellreports_2021_paper_razor/mix_3/protein.fas"), 
                            seqtype = "AA", 
                            as.string = TRUE)

fasta1 <- append(fastavar,
                fastavar2)

fasta <- append(fasta1,
                fastavar3)

fasta <- discard(fasta,
                 duplicated(names(fasta)))

```

```{r echo=TRUE}
if(!file.exists(here("results/proteogenomics/cellreps_2021/cleavage_annoated_peptides_razor.tsv"))){
  cleavage_annoated_peptides <- annotate_peptides(expr_mat = variant_peptides_1, 
                                                  fasta = fasta,
                                                  decoy_tag = "rev_")

  write_tsv(cleavage_annoated_peptides, "results/proteogenomics/cellreps_2021/cleavage_annoated_peptides_razor.tsv")
} else {
  cleavage_annoated_peptides <- read_tsv(here("results/proteogenomics/cellreps_2021/cleavage_annoated_peptides_razor.tsv"))
}

```

### Localized peptides within protein sequence and confirm identification of SAAVs

```{r echo=TRUE}
# tabulate location of the variant
annotated_w_variant_location <- cleavage_annoated_peptides %>%
          as_tibble() %>%
          mutate(protein_id_var = protein_id,
                 variant_type = if_else(str_detect(protein_id, "\\:"), 
                                        true = "other",
                                        false = "saav")) %>%
          mutate(protein_id = str_replace_all(protein_id, "\\.", "\\_")) %>%
          mutate(protein_id = str_replace_all(protein_id, "\\:", "\\_")) %>%
          separate(protein_id, 
                   into = c("ense_id", "variant_location"), 
                   sep = "\\_",
                   extra = "merge") %>%
                    na.omit() %>%
          mutate(variant_type = if_else(condition = str_detect(variant_location, "[0-9]$"),
                                        true = "other",
                                        false = variant_type))

# generate a long list of variant locations

locat2prot <- str_split(annotated_w_variant_location$variant_location,
                        pattern = "\\_")

names(locat2prot) <- annotated_w_variant_location$ense_id %>%
                    na.omit()

locat2protdf <- list2DF(locat2prot)

long2prot <- locat2protdf %>%
          pivot_longer(everything(),
                       values_to = "variant_location",
                       names_to = "ense_id") %>% 
          distinct() %>%
          mutate(variant_num_location = readr::parse_number(variant_location))

long2prot_worivar <- left_join(long2prot, 
                               dplyr::select(annotated_w_variant_location,
                                             protein_id_var, ense_id)) %>%
  distinct()

annotated_wo_variant_location <- annotated_w_variant_location %>%
          dplyr::select(-variant_location)

# merge long table of variant locations with the annotation of peptide position within 
# their associated protein sequences

protein2varlocat <- left_join(annotated_wo_variant_location, long2prot_worivar) %>%
          distinct() %>%
          mutate(start_num = as.numeric(start_position),
                 end_num = as.numeric(end_position)) %>%
          mutate(variant_peptide = variant_num_location >= as.numeric(start_position) & variant_num_location <= as.numeric(end_position)) %>% # evaluate if the falls on the position of a variant sequence
          dplyr::select(protein_id_var, ense_id, protein_description, Peptide, start_position, 
                        start_num,end_position,end_num, variant_location, 
                        variant_num_location, variant_peptide, variant_type)

pept2variant <- protein2varlocat %>%
  dplyr::select(Peptide, variant_peptide, variant_type) %>%
  distinct() %>%
  na.omit() %>%
  filter(variant_type == "saav")
```

```{r echo=TRUE}
table(pept2variant$variant_peptide)
```  
232 peptides were identified as mapping to single amino acid variant regions of their corresponding annotated proteins sequence.

## Visualize quantitative features of identified SAAVs

### Prep abundance matrices

```{r}
pep_prematwo6 <- dplyr::select(expr_matrix_pept_progen,
                               index, peptide, rownames(design_limmawo6))

pep_matwo6 <- dplyr::select(expr_matrix_pept_progen,
                             index, rownames(design_limmawo6))  %>%
  column_to_rownames("index") %>%
  as.matrix()

pep_matnona_wo6 <- pep_matwo6 %>%
  na.omit()

```

### Filter, merge and summarize SAAV abundances

```{r}
# extract indexing, peptide and sequence information from normalized data
index2genepeptide <- tmt_peptdata_frag_progen %>% 
  dplyr::select(index, peptide, gene, protein_id)

# table of interesting features (saavs)
features1 <- pept2variant %>%
  dplyr::rename(peptide = Peptide) %>%
  left_join(., index2genepeptide) %>%
  na.omit()

interesting_features1 <- features1 %>%
                    filter(variant_peptide == TRUE)

# data frame of identified SAAVs and their annotation
protein2varlocat_saav <- protein2varlocat %>%
  dplyr::rename(peptide = Peptide) %>%
  filter(peptide %in% interesting_features1$peptide,
         variant_peptide == TRUE) %>%
  left_join(., index2genepeptide) %>%
  mutate(gene_variant = paste(gene, variant_location, sep = "_"))

# mapping variant to index id
genevariant2index <- protein2varlocat_saav %>%
  dplyr::select(index, gene_variant)

# abundance matrix of peptides mapping to variant locations
pep_premat_variantwo6 <- filter(pep_prematwo6,
                                peptide %in% protein2varlocat_saav$peptide) %>%
  left_join(., genevariant2index) %>%
  relocate(gene_variant)

# summarize SAAVs (gene+variant location)
# i.e. take the median abundance of all the peptides mapping to the same SAAV

pep_premat_sumgenarwo6 <- pep_premat_variantwo6 %>%
  group_by(gene_variant) %>%
  summarise_at(vars(starts_with("x")), .funs = function(x){median(x, na.rm = TRUE)}) %>%
  ungroup()

full_info_sumgenvar <- pep_premat_sumgenarwo6 %>%
  left_join(., protein2varlocat_saav) %>%
  filter(variant_type == "saav")

write_tsv(protein2varlocat_saav,
          here("results/proteogenomics/cellreps_2021/all_peptides_mapping_to_saavs_gbm.tsv"))

write_tsv(pep_premat_sumgenarwo6,
          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs.tsv"))

write_tsv(full_info_sumgenvar,
          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs_full.tsv"))

pep_matvariantswo6or <- pep_premat_sumgenarwo6 %>%
  column_to_rownames("gene_variant") %>%
  as.matrix()

pep_matvariantswo6 <- pep_premat_sumgenarwo6 %>%
  filter(str_detect(gene_variant, "CAP1_", 
                    negate = TRUE)) %>%
  column_to_rownames("gene_variant") %>%
  as.matrix()

pep_matvariantswo6nona <- pep_matvariantswo6 %>%
  na.omit()
```

The variant-containing peptides identified are able to identify `r nrow(pep_premat_sumgenarwo6)` single amino acid mutations (SAAVs).

Among these identified SAAVs, `r nrow(pep_matvariantswo6nona)` were identified in all mixtures and samples.

### Heatmap of all SAAVs observed

```{r}
library(circlize)
```

```{r}
library(ComplexHeatmap)
```

```{r}
split_patstage <- str_split_fixed(string = colnames(pep_matvariantswo6nona), 
                                  pattern = "[:digit:]+", 
                                  n = 2)

patient <- str_extract(string = colnames(pep_matvariantswo6nona),
                       pattern = "[:digit:]+")

patientstage <- colnames(pep_matvariantswo6nona) %>%
  str_replace(., pattern = "prim", 
              replacement = "Init") %>%
  str_replace(., pattern = "rec", 
              replacement = "Rec")

stage <- split_patstage[,2] %>%
  if_else(condition = . == "prim",
          true = "Initial",
          false = "Recurrent")

pep_matvariantswo6or_hm <- pep_matvariantswo6or
pep_matvariantswo6nona_hm <- pep_matvariantswo6nona

colnames(pep_matvariantswo6or_hm) <- patientstage
colnames(pep_matvariantswo6nona_hm) <- patientstage
```

```{r fig.height=20, fig.width=7}
col_fun_or = colorRamp2(c(min(pep_matvariantswo6or_hm, na.rm = TRUE), 
                       median(pep_matvariantswo6or_hm, na.rm = TRUE), 
                       max(pep_matvariantswo6or_hm, na.rm = TRUE)), 
                     c("#2a9d8f", "white", "red"))

Heatmap(pep_matvariantswo6or_hm, 
        col = col_fun_or,
        column_split = stage, 
        row_names_gp = gpar(fontsize = 5.5),
        column_names_gp = gpar(fontsize = 7),
        cluster_rows = FALSE, cluster_columns = FALSE,
        heatmap_legend_param = list(
        title = "log2-\nNormalized\nAbundance"
    ))
```

### Heatmap of SAAVs observed on every sample

```{r fig.height=7.5, fig.width=7}
col_fun_sel = colorRamp2(c(min(pep_matvariantswo6or_hm, na.rm = TRUE), 
                       median(pep_matvariantswo6or_hm, na.rm = TRUE), 
                       max(pep_matvariantswo6or_hm, na.rm = TRUE)), 
                     c("#2a9d8f", "white", "red"))

Heatmap(pep_matvariantswo6nona_hm, 
        col = col_fun_sel,
        column_split = stage, 
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 7),
        heatmap_legend_param = list(
        title = "log2-\nNormalized\nAbundance"
    ))
```

### Limma on SAAVs

### Prep design matrix

```{r}
design_limmawo6_saav <- model.matrix(~patient+stage)

sample_annotationwo6 <- sample_annotationwo6 %>%
  mutate(paired_id_semi = paste(patient, recurrence, sep = ""))

rownames(design_limmawo6_saav) <- sample_annotationwo6$paired_id_semi
```

### Prep abundance matrices

```{r}
colnames(pep_matvariantswo6nona)
```

```{r}
rownames(design_limmawo6_saav)
```

#### SAAV-level model fit

```{r}
fit_limmapeptwo6 <- function(mat, design, method,Limma){
  limmafit <- lmFit(mat, design, method = method)
  limmafit <- eBayes(limmafit)
  limma_tab <- topTable(limmafit, coef = "stageRecurrent", number = Inf, adjust.method = "BH") %>%
  mutate(Protein = rownames(.),
         Limma = Limma)

  return(limma_tab)
}
```

```{r echo=TRUE, warning=FALSE, message=FALSE}
fit_mod_saavs <- fit_limmapeptwo6(pep_matvariantswo6nona, 
                                  design_limmawo6_saav, 
                                  method = 'robust', 
                                  Limma = "Robust - w Patient effect")

compar_tab_saavs <-  fit_mod_saavs %>% 
  rownames_to_column("gene_variant") 
```

```{r}
increased_recwo6_saavs <- compar_tab_saavs %>%
                    filter(logFC>0,
                           adj.P.Val < 0.05) %>% 
  pull(gene_variant)

decreased_recwo6_saavs <- compar_tab_saavs %>%
                    filter(logFC<0,
                           adj.P.Val < 0.05) %>% 
  pull(gene_variant)

```

```{r}
length(increased_recwo6_saavs)
```
```{r}
length(decreased_recwo6_saavs)
```

#### Volcano SAAVs

```{r fig.width=8, fig.height=6}
diff_saavs <- compar_tab_saavs %>%
                    filter(adj.P.Val < 0.05)

size <- 1.5

volcano_saavs <- ggplot(data = compar_tab_saavs,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = compar_tab_saavs %>% filter(logFC > 0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = compar_tab_saavs %>% filter(logFC < -0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
                 size = size) +
      geom_point(data = compar_tab_saavs %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = compar_tab_saavs %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      ggrepel::geom_text_repel(data = diff_saavs,
                               aes(label = gene_variant), size = 4) +
      xlab("logFC - Recurrent / Primary") + 
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

```{r echo=FALSE}
ggsave(plot = volcano_saavs, 
       filename = here::here("figures/volcanosaavs_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano_saavs, 
       filename = here::here("figures/volcanosaavs_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

### Boxplot of differentially abundant SAAV

```{r}
prelongsaav <- pep_matvariantswo6nona %>%
  as.data.frame() %>%
  rownames_to_column("gene_variant") %>%
   pivot_longer(cols = starts_with("x"),
                         names_to = c("patient"),
                         values_to = "Abundance") %>%
  mutate(Stage = str_extract(patient, "[^0-9]+$")) %>%
  mutate(Patient = str_remove_all(patient, "[^0-9]+$")) %>%
  filter(gene_variant == increased_recwo6_saavs)
```

```{r}
diff_abund_saav <- ggplot(prelongsaav, 
                    aes(x = Stage, 
                        y = Abundance, fill = Stage, 
                        cex.axis = 1.5)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.2) +
  # Box plot with jittered points
  # 0.2 : degree of jitter in x direction
  # geom_jitter(shape=16, position=position_jitter(0.2))
  ylab("log2-Normalized Abundance") +
  geom_signif(
    comparisons = list(c("Initial", "Recurrence")),
    map_signif_level = TRUE
  ) + 
  stat_compare_means(method="t.test") +
  ggtitle(increased_recwo6_saavs) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")

```

### Boxplot DA SAAV

```{r}
  slim_data2saav <- pivot_wider(prelongsaav %>% dplyr::select(-patient), 
                          values_from = Abundance, 
                          names_from = Stage) #%>%
  #dplyr::select(-patient)

  slim_data3saav <- slim_data2saav %>% 
    mutate(Abs_diff = rec-prim) %>%
    mutate(rank = row_number(Abs_diff)) %>% 
    dplyr::rename(Initial = prim, Recurrent = rec) %>%
    ungroup()

```

```{r fig.width=7, fig.height=3.5}
ggpaired(slim_data3saav,
         cond1 = "Initial",
         cond2 = "Recurrent", 
         fill = "condition",
         ylab = "log2-Normalized Abundance",
         label = NULL,
         repel = TRUE, point.size = 0.2, line.size = 0.1) +
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
            axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
            panel.background = element_blank(),
            panel.grid.major = element_blank(),
            panel.border = element_rect(colour = "black", fill=NA, size=0.5),
            axis.title = element_text(size = 8),
            legend.text = element_text(size = 6),
            legend.title = element_text(size = 8),
            legend.key.height= unit(3, 'mm'),
            legend.key.width= unit(3, 'mm'),
            legend.position="bottom",
            strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) + 
  ggtitle(increased_recwo6_saavs)
```

```{r}
paired_box_saavs <- ggpaired(slim_data3saav,
         cond1 = "Initial",
         cond2 = "Recurrent", 
         fill = "condition",
         ylab = "log2-Normalized Abundance",
         label = NULL,
         repel = TRUE, point.size = 0.2, line.size = 0.1) +
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
            axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
            panel.background = element_blank(),
            panel.grid.major = element_blank(),
            panel.border = element_rect(colour = "black", fill=NA, size=0.5),
            axis.title = element_text(size = 8),
            legend.text = element_text(size = 6),
            legend.title = element_text(size = 8),
            legend.key.height= unit(3, 'mm'),
            legend.key.width= unit(3, 'mm'),
            legend.position="bottom",
            strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) 
```

```{r echo=FALSE}
ggsave(plot = paired_box_saavs, 
       filename = here::here("figures/paired_box_saav_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 61,
       height = 50)

ggsave(plot = paired_box_saavs, 
       filename = here::here("figures/paired_box_saav_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 61,
       height = 50)
```

```{r}
sessionInfo()
```

### 2022-01-21 scr/annotate_nterm.R

@@ -1,0 +1,43 @@
+# function for annotation of N-terminal modifications
+
+annotate_nterm <- function(peptidestsv, # peptide.tsv table
+                           tmtmass = 304.2072, # either 304.2072 for 16plex or 229.1629 for 10/11plex
+                           protease_specificity = "R|K") # or R for argc 
+{
+  
+  require(dplyr)
+  require(stringr)
+  
+  if (tmtmass == 304.2072){
+    
+    nterm_tmt <- "N-term\\(304.2072\\)"
+    ktmt <- "K\\(304.2072\\)"
+    
+  } else if (tmtmass == 229.1629){
+    
+    nterm_tmt <- "N-term\\(229.1629\\)"
+    ktmt <- "K\\(229.1629\\)"                    
+    
+  } else {
+    
+    error("Please check your tmtmass argument input.")
+    
+  }
+  
+  nterm_annotated <- peptidestsv %>% 
+    mutate(nterm = case_when(str_detect(assigned_modifications, nterm_tmt) ~ "TMT-labelled",
+                             str_detect(assigned_modifications, "N-term\\(42.0106\\)") ~ "acetylated",
+                             TRUE ~ "free")) %>%
+    mutate(tmt_tag = case_when(str_detect(assigned_modifications, nterm_tmt) ~ "nterm",
+                               str_detect(assigned_modifications, ktmt) ~ "lysine",
+                               str_detect(assigned_modifications, ktmt,
+                                          negate = TRUE) & str_detect(assigned_modifications, "N-term\\(42.0106\\)") ~ "untagged_acetylated",
+                               str_detect(assigned_modifications, ktmt,
+                                          negate = TRUE) & nterm == "acetylated" ~ "untagged_acetylated",
+                               str_detect(assigned_modifications, ktmt,
+                                          negate = TRUE) & nterm == "free" ~ "untagged_free",
+                               TRUE ~ "untagged"))
+  
+  return(nterm_annotated)
+  
+}

### 2022-01-21 scr/annotate_peptides.R

@@ -1,0 +1,171 @@
+## Annotate peptides function
+
+annotate_peptides <- function(expr_mat, fasta,
+                              decoy_tag = "^rev",# input should be a vector of peptide sequences and the fasta file
+                              specificity = "R|K"){ 
+  
+  pept2prot <- expr_mat %>% 
+    dplyr::select(Peptide, Genes) %>% 
+    mutate(Peptide = as.character(Peptide),
+           Genes = as.character(Genes))
+  
+  peptide_sequence <- pept2prot$Peptide
+  
+  
+  if (!is.null(decoy_tag)){
+    fasta <- fasta %>% 
+      purrr::discard(str_detect(names(.), decoy_tag))
+  }
+  
+  annotation <- data.frame()
+  
+  for (i in 1:length(peptide_sequence)){
+    
+    # extract the list element that matches the peptide sequence
+    list_elem <- fasta %>%
+      purrr::keep(str_detect(names(.), fixed(pept2prot$Genes[i])))
+    
+    
+    if(is_empty(list_elem)){
+      protein_id <- NA
+      protein_description <- NA
+      protein_length <- NA
+      Peptide <- peptide_sequence[i]
+      peptide_position <- NA
+      start_position <- NA
+      end_position <- NA
+      following_10_resid <- NA
+      previous_10_resid <- NA
+      previous_all_resid <- NA
+      aa_after <- NA
+      aa_before <- NA
+      last_aa <- NA
+      
+      temprow <- cbind(protein_id,
+                       protein_description,
+                       protein_length,
+                       Peptide, 
+                       start_position,
+                       end_position, 
+                       last_aa, aa_after,
+                       aa_before, 
+                       following_10_resid, 
+                       previous_10_resid,
+                       previous_all_resid)
+      
+      annotation <- rbind(annotation,
+                          temprow)
+      
+      row.names(annotation) <- NULL
+      
+    } else {
+      
+      # extract the annotation info stored in the fasta header, check for an Uniprot fasta pattern and extract protein ID and description
+      if(str_detect(attr(list_elem[[1]], "Annot"), 
+                    pattern = "^\\>sp\\|") | str_detect(attr(list_elem[[1]], "Annot"), 
+                                                        pattern = "^\\>tr\\|")){
+        split1 <- str_split_fixed(attr(list_elem[[1]], "Annot"), 
+                                  pattern = "\\|", 
+                                  n = 3)
[…]

### 2022-01-21 scr/feature_specific_fdr_correction.R

@@ -1,0 +1,33 @@
+## function for feature-specific FDR correction
+## used specifically in this case to apply FDR correction only on semi-specific peptides
+
+feature_fdr_correction <- function(toptable,
+                                   features_table,
+                                   interesting_features_table){
+  
+  # merge limma output with feature annotation  ----
+  
+  tab_limma_feature_annot <- left_join(toptable, 
+                                       features_table,
+                                       by = c("peptide","index"))
+  
+  # filter limma table based on interesting feature ----
+  
+  tab_limma_subsetin <- filter(tab_limma_feature_annot,
+                               peptide %in% interesting_features_table$peptide) %>%
+    # apply FDR correction only on the subset
+    mutate(adj.P.Val = p.adjust(p = P.Value, 
+                                method = "BH")) %>%
+    mutate(fdr_correction = 'feature-specific')
+  
+  # create a table of non-interesting features containing the globally adjusted p-values
+  tablimma_subsetout <- filter(tab_limma_feature_annot,
+                               !peptide %in% interesting_features_table$peptide) %>%
+    mutate(fdr_correction = 'global')
+  
+  output_limma3 <- bind_rows(tab_limma_subsetin,
+                             tablimma_subsetout) 
+  
+  return(output_limma3)
+  
+}

### 2022-01-21 scr/fit_limmawo6.R

@@ -1,0 +1,15 @@
+# Function to fit the linear model with limma, generate the summary table and merge 
+# with gene annotation per protein
+
+fit_limmawo6 <- function(mat, design, method,Limma, prot2gene){
+  limmafit <- lmFit(mat, design, method = method)
+  limmafit <- eBayes(limmafit)
+  limma_tab <- topTable(limmafit, coef = "recurrencewo6rec", number = Inf, adjust.method = "BH") %>%
+    mutate(Protein = rownames(.),
+           Limma = Limma)
+  
+  limma_tab <- left_join(limma_tab, prot2gene)
+  
+  return(limma_tab)
+}
+

### 2022-01-21 scr/get_cleavage_area.R

@@ -1,0 +1,47 @@
+### Prepare cleavage motifs for iceLogo analysis
+
+get_cleave_area <- function(peptide_annotation){
+  
+  nterm_semi_pepts <- peptide_annotation %>% 
+    filter(semi_type == "semi_Nterm") %>%
+    dplyr::select(-previous_all_resid) %>%
+    mutate(previous_10_resid = if_else(is.na(previous_10_resid), true = "", false = previous_10_resid),
+           following_10_resid = if_else(is.na(following_10_resid), true = "", false = following_10_resid)) %>%
+    mutate(prev_10_pad = str_pad(previous_10_resid, side = "left", pad = "X", width = 10),
+           following_10_pad = str_pad(following_10_resid, side = "right", pad = "X", width = 10)) %>%
+    mutate(cleave_area = paste0(prev_10_pad, Peptide, following_10_pad)) %>%
+    mutate(cleave_area20 = str_sub(cleave_area,
+                                   start = 1,
+                                   end = 20)) %>%
+    mutate(len_20 = str_count(cleave_area20),
+           len_clearea = str_count(cleave_area),
+           len_follow = str_count(following_10_resid),
+           len_prev = str_count(previous_10_resid),
+           len_followpad = str_count(prev_10_pad),
+           len_prevpad = str_count(following_10_pad))
+  
+  cterm_semi_pepts <- peptide_annotation %>% 
+    filter(semi_type == "semi_Cterm") %>%
+    dplyr::select(-previous_all_resid) %>%
+    mutate(previous_10_resid = if_else(is.na(previous_10_resid), true = "", false = previous_10_resid),
+           following_10_resid = if_else(is.na(following_10_resid), true = "", false = following_10_resid)) %>%
+    mutate(prev_10_pad = str_pad(previous_10_resid, side = "left", pad = "X", width = 10),
+           following_10_pad = str_pad(following_10_resid, side = "right", pad = "X", width = 10)) %>%
+    mutate(cleave_area = paste0(prev_10_pad, Peptide, following_10_pad)) %>%
+    mutate(cleave_area20 = str_sub(cleave_area,
+                                   -20)) %>%
+    #mutate(cleave_area20) 
+    mutate(len_20 = str_count(cleave_area20),
+           len_clearea = str_count(cleave_area),
+           len_follow = str_count(following_10_resid),
+           len_prev = str_count(previous_10_resid),
+           len_followpad = str_count(prev_10_pad),
[…]

### 2022-01-21 scr/helper_functions.R

@@ -1,0 +1,155 @@
+### helper functions
+
+## sample proteins for profile plots and QC plots ----
+sample_proteins <- function(x, # char vector of protein IDs
+                            size, # number of proteins to sample
+                            seed = 363 # random seed number 
+) {
+  
+  # select/sample N proteins to observe their abundance distribution  
+  set.seed(seed)
+  
+  proteins <- x %>% 
+    unique() %>%
+    str_remove_all(pattern = "Biognosys") # get unique protein IDs
+  
+  which_prots <- sample(proteins, 
+                        size = size, 
+                        replace = FALSE)
+  
+}
+
+
+## eliminate proteins based on missing values ----
+
+na_filter <- function(data, max_na_per_group, logtrans, minInt_inpute, fraction_Inpute = 5){
+  
+  data_long <- pivot_longer(data = data,
+                            cols = matches("^S"),
+                            values_to = "Intensity",
+                            names_to = c("Group-Sample")) %>%
+    separate(col = "Group-Sample", 
+             into = c("Group", "Sample"),
+             sep = "\\-")
+  
+  if (logtrans){
+    data_long <- mutate(data_long,
+                        Intensity = log2(Intensity))
+  }
+  
+  na_count <- group_by(data_long,
+                       ID, Group) %>%
+    summarise(na_count = sum(is.na(Intensity)),
+              total = n()) %>% 
+    ungroup() %>% 
+    mutate(NA_fraction = na_count/total)
+  
+  nafraction <- group_by(na_count,
+                         ID) %>%
+    summarise(max_nafraction = max(NA_fraction),
+              min_nafraction = min(NA_fraction)) %>%
+    ungroup()
+  
+  included_prots <- dplyr::filter(nafraction,
+                                  max_nafraction <= max_na_per_group)
+  
+  
+  
+  if (minInt_inpute){
+    
+    min_intensityprot <- data_long %>% 
+      group_by(ID) %>% 
+      mutate(min_int = min(Intensity, na.rm = TRUE)) %>%
+      mutate(th_of_min_int = min_int/fraction_Inpute)
+    
+    data_long_nonas1 <- min_intensityprot %>% 
+      mutate(Intensity = ifelse(is.na(Intensity),
+                                yes = th_of_min_int,
+                                no = Intensity),
+             `Group-Sample` = paste(Group,"-",Sample, sep = "")) %>%
+      dplyr::select(ID, `Group-Sample`, Intensity) %>%
[…]

### 2022-01-21 scr/our_volcano.R

@@ -1,0 +1,68 @@
+# function to create volcano plots for visualization in the HTML report
+
+### volcano plots ----  
+
+our_volcano <- function(dataarg, 
+                        FC_cutoff = 1, 
+                        pval_cutoff = 0.05, 
+                        color_diffex = "red", 
+                        color_nondifex = "#2a9d8f", 
+                        interesting_proteins = NULL, 
+                        vert_line_col = "red",
+                        hline_col = "red", 
+                        hline_pos = 0.05, 
+                        linetype = "dashed",
+                        #vprot_ylim = c(0,2),
+                        #vprot_xlim = c(-2,2),
+                        increased_in, 
+                        comparison_title) {
+  
+  signi_hits <- filter(dataarg, 
+                       adj.P.Val <= pval_cutoff) %>% 
+    pull(Protein)
+  
+  if(is_null(interesting_proteins)){
+    
+    protlabels <- dataarg %>%
+      filter(adj.P.Val <= pval_cutoff)
+    
+  } else {
+    
+    protlabels <- dataarg %>%
+      filter(Gene %in% interesting_proteins)
+    
+  }
+  
+  
+  volcano <- ggplot(data = dataarg,
+                    mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
+    geom_point()+
+    geom_point(data = dataarg %>% filter(logFC > FC_cutoff,
+                                         adj.P.Val < pval_cutoff),
+               mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red")+
+    geom_point(data = dataarg %>% filter(logFC < -FC_cutoff,
+                                         adj.P.Val < pval_cutoff),
+               mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red") +
+    geom_point(data = dataarg %>% filter(logFC > FC_cutoff,
+                                         adj.P.Val > pval_cutoff),
+               mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f") +
+    geom_point(data = dataarg %>% filter(logFC < -FC_cutoff,
+                                         adj.P.Val > pval_cutoff),
+               mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f") +
+    ggrepel::geom_text_repel(data = protlabels,
+                             aes(label = Gene)) +
+    geom_hline(yintercept = -log10(hline_pos),
+               color = "red", linetype = "dashed") +
[…]

### 2022-01-21 scr/plot_volcano.R

@@ -1,0 +1,33 @@
+## function to generate a volcano plot
+
+plot_volcano <- function(output_limma3){
+  tovolc_all <- output_limma3 %>% 
+    mutate(Differentially_expressed = case_when(adj.P.Val <= 0.05 ~ TRUE,
+                                                TRUE ~ FALSE))
+  
+  tovolc_featu <- output_limma3 %>% 
+    filter(fdr_correction == "feature-specific") %>%
+    mutate(Differentially_expressed = case_when(adj.P.Val <= 0.05 ~ TRUE,
+                                                TRUE ~ FALSE))
+  volcanoes <- ggplot(data = tovolc_all, 
+                      mapping = aes(x = logFC,
+                                    y = -log10(adj.P.Val),
+                                    color = Differentially_expressed,
+                                    label = ID)) + 
+    geom_point(alpha = 0.5) + 
+    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
+    #ggtitle("Volcano plot of all limma-tested features") + 
+    theme(legend.position = "none")
+  
+  volcano_features <- ggplot(data = tovolc_featu, 
+                             mapping = aes(x = logFC,
+                                           y = -log10(adj.P.Val),
+                                           color = Differentially_expressed,
+                                           label = ID)) + 
+    geom_point(alpha = 0.5) + 
+    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
+    #ggtitle("Volcano plot of selected/interesting limma-tested features") + 
+    theme(legend.position = "none")
+  
+  print(volcano_features)
+}

### 2022-01-21 scr/protein_peptide_coverage_plot.R

@@ -1,0 +1,51 @@
+## peptide coverage plots
+
+get_coverage <- function(annotated_peptides, id){
+  
+  mappedid <- dplyr::filter(annotated_peptides,
+                            protein_id == id) %>% 
+    dplyr::mutate(type = "PEPTIDE",
+                  begin = as.numeric(start_position),
+                  end = as.numeric(end_position), 
+                  length = as.numeric(peptide_length)) %>%
+    dplyr::rename(description = Peptide,
+                  accession = protein_id) %>% 
+    dplyr::select(type, description, begin, end, length, accession) %>%
+    dplyr::arrange(end) %>%
+    dplyr::mutate(order = 1)
+  
+}
+add_peptides <- function(feature_df,
+                         peptide_coverage_data){
+  
+  taxid <- feature_df$taxid[1]
+  entryName <- feature_df$entryName[1]
+  
+  pept_feat <- dplyr::mutate(peptide_coverage_data,
+                             taxid = taxid,
+                             entryName = entryName) %>%
+    dplyr::select(names(feature_df)) # added cleave window information
+  
+  feature_df <- mutate(feature_df)
+  
+  wpept_features <- bind_rows(feature_df, pept_feat)
+  
+  return(wpept_features)
+  
+}
+draw_peptides <- function(p, data){
+  data$order <- 1.4
+  p + ggplot2::geom_rect(data = data,
+                         mapping=ggplot2::aes(xmin=begin,
+                                              xmax=end,
+                                              ymin=order-0.05,
+                                              ymax=order+0.05),
+                         colour = "black") +
+    annotate("text", x = -50, y = 1.4, label = "Peptides") +
+    theme_bw(base_size = 12) + # white background
+    theme(panel.grid.minor=element_blank(),
+          panel.grid.major=element_blank()) +
+    theme(axis.ticks = element_blank(),
+          axis.text.y = element_blank()) +
+    theme(panel.border = element_blank())
+}

### 2022-01-21 scr/summarize_peptide_counts.R

@@ -1,0 +1,43 @@
+# function to summarize peptide counts per annotated feature in terms of
+# N-terminal modification and proteolytic specificity
+
+summarize_peptide_counts <- function(tmt_report_annotated){
+  
+  require(dplyr)
+  require(tibble)
+  
+  to_count_info <- tmt_report_annotated %>% 
+    dplyr::select(peptide, specificity, nterm, semi_type, tmt_tag)
+  
+  to_count_info_acetyl <- to_count_info %>% 
+    dplyr::filter(nterm == "acetylated")
+  
+  n_semi <- dplyr::count(to_count_info, specificity) %>% 
+    dplyr::rename(feature_type = specificity) %>%
+    dplyr::mutate(category = "specificity")
+  
+  n_term <- dplyr::count(to_count_info, nterm) %>% 
+    dplyr::rename(feature_type = nterm) %>%
+    dplyr::mutate(category = "N-term")
+  
+  n_semi_type <- dplyr::count(to_count_info, semi_type) %>% 
+    dplyr::rename(feature_type = semi_type) %>%
+    dplyr::mutate(category = "Semi type")
+  
+  n_tmt_tag <- dplyr::count(to_count_info, tmt_tag) %>% 
+    dplyr::rename(feature_type = tmt_tag)  %>%
+    dplyr::mutate(category = "TMT location")
+  
+  n_total <- tibble(feature_type = "Total",
+                    n = nrow(to_count_info),
+                    category = "Total")
+  
+  summary_count <- bind_rows(n_semi,
+                             n_term,
+                             n_semi_type,
+                             n_tmt_tag,
+                             n_total)
+  
+  return(summary_count)
+  
+}

### 2022-01-25 scr/gbm_proteomics_data_analysis_proteolytic_products_prep_figures_v3_gh_format.Rmd
modify Rmd and md; add figures; add results

@@
 
+pep_quant_presummary_semi <- pep_quant_presummary %>%
+  filter(specificity == "semi_specific",
+         is_terminal == "not_terminal")
 ```
@@
                          false = "Recurrence"))
+
+# summary of sum of log2 semi-specific peptides per sample
+
+pept_summary_semi_1 <- pep_quant_presummary_semi %>% 
+  group_by(paired_id, recurrence) %>%
+  summarise(Sum_All = sum(Abundance, na.rm = TRUE)) %>%
+  mutate(Stage = if_else(recurrence == "prim",
+                         true = "Initial",
+                         false = "Recurrence"),
+         `Summed Abundances` = Sum_All)
 
@@
 ```{r}
-plotSumEM
-```
-
-```{r}
 ggsave(plot = plotSumEM, 
@@
        filename = here::here("figures/box_perc_semispec_pepts.eps"), 
+       device = "eps",
+       units = "mm",
+       width = 40,
+       height = 50)
+```
+
+### Generate plot of sum of abundances semi-specific peptides
+
+```{r}
+sum_semi_abunds <- ggplot(pept_summary_semi_1, 
+                    aes(x = Stage, 
+                        y = `Summed Abundances`, fill = Stage, 
+                        cex.axis = 1.5)) +
+  geom_boxplot() +
+  #geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5,
+  #             ) +
+  geom_jitter(position=position_jitter(0.2)) + 
+  # Box plot with jittered points
+  # 0.2 : degree of jitter in x direction
+  # geom_jitter(shape=16, position=position_jitter(0.2))
+  ylab("Sum of log2-intensities of semi-tryptic peptides") +
+  geom_signif(
+    comparisons = list(c("Initial", "Recurrence")),
+    map_signif_level = TRUE
+  ) + 
+  stat_compare_means(method="t.test") +
+  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
+        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
+        panel.background = element_blank(),
+        panel.grid.major = element_blank(),
+        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
+        axis.title = element_text(size = 8),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 8),
+        legend.key.height= unit(3, 'mm'),
+        legend.key.width= unit(3, 'mm'),
+        legend.position="bottom")
+```
+
+```{r}
+ggsave(plot = sum_semi_abunds, 
+       filename = here::here("figures/box_sum_semispec_pepts.tiff"), 
+       device = "tiff",
+       units = "mm",
+       width = 40,
+       height = 50)
+
+ggsave(plot = sum_semi_abunds, 
+       filename = here::here("figures/box_sum_semispec_pepts.eps"), 
        device = "eps",
@@
 write_tsv(x = decreased_Rec,
-          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))
+          file = here("results/semi_tryptic/proteolytic_products_decreased_in_Recurrent.tsv"))
 

### 2022-11-28 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
[ADD] rmarkdown notebook and pdf of elisa proteomics analysis

---
title: "ASAH1 rGBM Manuscript ELISA Figure"
author: "Justin Sing, University of Toronto"
date: "`r format(Sys.time(), '%a %B %d %X %Z %Y')`"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{fancyhdr}
- \usepackage[fit, breakall]{truncate}
  \usepackage{tcolorbox}
  \newtcolorbox{myquote}{colback=red!5!white, colframe=red!75!black}
  \renewenvironment{quote}{\begin{myquote}}{\end{myquote}}
output:
  pdf_document: 
    keep_tex: no
    latex_engine: xelatex
    toc: yes
    toc_depth: 4
  html_notebook: default
editor_options:
  chunk_output_type: console
---

\pagestyle{fancy}
<!-- \fancyhead[LO,CE]{ASAH1 rGBM Manuscript} -->
\fancyfoot[CO,CE]{Justin Sing, University of Toronto}
\fancyfoot[LE,RO]{\thepage}

<!-- # Required Libraries -->
```{r message=FALSE, warning=FALSE, include=FALSE}
#' @title Try load package, install from correct package manager if not installed
#' @param package (character) string name of package to load
#' @param s (character) source to install package from it not installed
installIfNeeded <- function(package, s = "CRAN") {
  # s: "CRAN" or "BIO"
  if (s == "CRAN") {
    if (! requireNamespace(package, quietly=TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE, quietly = TRUE)
    } else {
      library(package, character.only = TRUE, quietly = TRUE)
    }
  } else if (s == "BIO") {
    if (! requireNamespace("BiocManager", quietly=TRUE)) {
      install.packages("BiocManager")
    }
    if (! requireNamespace(package, quietly=TRUE)) {
      BiocManager::install(package)
      library(package, character.only = TRUE, quietly = TRUE)
    } else {
      library(package, character.only = TRUE, quietly = TRUE)
    }
  } else {
    stop(sprintf("Unknown source \"%s\".", s))
  }
}

# Data I/O and Wrangling
installIfNeeded("data.table")
installIfNeeded("dplyr")
# Visualization
installIfNeeded("ggplot2")
installIfNeeded("grid")
installIfNeeded("gridExtra")
installIfNeeded("ggpubr")
# Table Visualization
installIfNeeded("gt")
# Statistics
installIfNeeded("rstatix")
installIfNeeded("broman")
# Machine Learning 
installIfNeeded("caret")
installIfNeeded("plotROC")
installIfNeeded("mixOmics", "BIO")
```

<!-- # Helper Functions -->
```{r message=FALSE, warning=FALSE, include=FALSE}
#' @title Master Trainer and Prediction
#' @description Top Function to perform predictions for testing logistic regression, logistic regression + elastic regularization, random forest and k-Nearest Neightbours
#' 
#' @param train_df (data.frame) A dataframe that contains labels and features to run through prediction models. Has to have the following structure:
#' \item{class}{(character) sample/observation label. i.e. ctrl, iGBM, or rGBM sample.}
#' \item{feature_1}{(numeric} a specific feature measurement such as protein ASAH1)}
#' \item{feature_n}{(numeric) another specific feature.}
#' @param pred_col_prob (character) A character corresponding to the class of probabilities you want to extract for. i.e. "GBM"
#' @param multiclass (logical) Run predictions using multiclass modelling and prediction if more than 2 classes.
#' 
#' @return
#' 
#' returns a list object that contains trained model objects, lift data and roc data.
#' 
#' \item(model){(glmnet) Logistic Regression with Elastic Net Regularization Model }
#' \item(model_rf){(rf) Random Forest Model}
#' \item(model_logit){(glm) Logistic Regression Model}
#' \item(model_boostedlogit){(knn) k-Nearest Neighbours Models}
#' \item(master_lift){(data.table) lift data for all models}
#' \item(roc_data){(data.table) roc data for all models}
run_predictions <- function(train_df, pred_col_prob="GBM", multiclass=FALSE){

  ##*****************************************************
  ##  Model Training  ----

  if (!multiclass){
    # specify the cross-validation method
    ctrl <- trainControl(method = "LOOCV", search = "grid", classProbs = T, savePredictions = "final", summaryFunction = twoClassSummary, verboseIter = T)

    # fit models and use LOOCV to evaluate performance
    model <- train( class ~ ., data = train_df, na.action = na.omit, method = "glmnet", family="binomial", trControl = ctrl, metric = "ROC")
    model_rf <- train( class ~ ., data = train_df, na.action = na.omit, method = "rf", verbose=1, trControl = ctrl, metric = "ROC")
    model_logit <- train( class ~ . , data = train_df, na.action = na.omit, method = "glm", family="binomial", trControl = ctrl, metric = "ROC")
    model_boostedlogit <- train( class ~ . , data = train_df, na.action = na.omit, method = "knn", trControl = ctrl, metric = "ROC")

  } else{
    # specify the cross-validation method
    ctrl <- trainControl(method = "LOOCV", number = 10, search = "grid", classProbs = T, savePredictions = "final", verboseIter = T)

    # fit models and use LOOCV to evaluate performance
    model <- train( class ~ . , data = train_df, na.action = na.omit, method = "glmnet", family="multinomial", trControl = ctrl, metric = "ROC")
    model_rf <- train( class ~ ., data = train_df, na.action = na.omit, method = "rf", verbose=1, trControl = ctrl, metric = "ROC")
    model_logit <- train( class ~ . , data = train_df, na.action = na.omit, method = "multinom", family="binomial", trControl = ctrl, metric = "ROC")
    model_boostedlogit <- train( class ~ . , data = train_df, na.action = na.omit, method = "knn", trControl = ctrl, metric = "ROC")
  }

  ##*****************************************************
  ##  Generate Lift Plot Data ----

  if(!multiclass)  {
    for_lift = data.frame(Class = model$pred[model$pred$alpha==model$bestTune$alpha & model$pred$lambda==model$bestTune$lambda ,]$obs,  
                          glmnet = model$pred[model$pred$alpha==model$bestTune$alpha & model$pred$lambda==model$bestTune$lambda , pred_col_prob])
    for_lift_rf = data.frame(Class = model_rf$pred[model_rf$pred$mtry==model_rf$bestTune$mtry,]$obs,  
                             rf = model_rf$pred[model_rf$pred$mtry==model_rf$bestTune$mtry ,pred_col_prob])
    if (!multiclass){
      for_lift_logit = data.frame(Class = model_logit$pred$obs,  
                                  logit = model_logit$pred[,pred_col_prob])
    } else {
      for_lift_logit = data.frame(Class = model_logit$pred[model_logit$pred$decay==model_logit$bestTune$decay,]$obs,  
                                  logit = model_logit$pred[model_logit$pred$decay==model_logit$bestTune$decay,pred_col_prob])
    }
    for_lift_boostedlogit = data.frame(Class = model_boostedlogit$pred[model_boostedlogit$pred$k==model_boostedlogit$bestTune$k,]$obs,  
                                       knn = model_boostedlogit$pred[model_boostedlogit$pred$k==model_boostedlogit$bestTune$k ,pred_col_prob])

    lift_obj = lift(Class ~ glmnet, data = for_lift, class = pred_col_prob)
    lift_obj_rf = lift(Class ~ rf, data = for_lift_rf, class = pred_col_prob)
    lift_obj_logit = lift(Class ~ logit, data = for_lift_logit, class = pred_col_prob)
    lift_obj_boostedlogit = lift(Class ~ knn, data = for_lift_boostedlogit, class = pred_col_prob)

    master_lift <- rbindlist(list(lift_obj$data, lift_obj_rf$data, lift_obj_logit$data, lift_obj_boostedlogit$data))
    master_lift$liftModelVar <- factor(master_lift$liftModelVar, levels = c("logit", "glmnet", "knn", "rf"))

  } else {
    master_lift <- data.frame()
  }

  ##*****************************************************
  ## Generate ROC Prediction Data ----

  if (!multiclass){
    roc_data <- rbindlist(list(Logit=model_logit$pred, 
                               `Elastic Logit`=model$pred[model$pred$alpha==model$bestTune$alpha & model$pred$lambda==model$bestTune$lambda, ],
                               KNN=model_boostedlogit$pred[model_boostedlogit$pred$k==model_boostedlogit$bestTune$k, ], 
                               RF=model_rf$pred[model_rf$pred$mtry==model_rf$bestTune$mtry, ]), 
                          idcol = "Model", use.names = TRUE, fill = TRUE)
    roc_data$Model <- factor(roc_data$Model, levels=c("Logit", "Elastic Logit", "KNN", "RF"))
  } else {
    roc_data <- rbindlist(list(Logit=model_logit$pred[model_logit$pred$decay==model_logit$bestTune$decay, ], 
                               `Elastic Logit`=model$pred[model$pred$alpha==model$bestTune$alpha & model$pred$lambda==model$bestTune$lambda, ], 
                               KNN=model_boostedlogit$pred[model_boostedlogit$pred$k==model_boostedlogit$bestTune$k, ], 
                               RF=model_rf$pred[model_rf$pred$mtry==model_rf$bestTune$mtry, ]), 
                          idcol = "Model", use.names = TRUE, fill = TRUE)
    roc_data$Model <- factor(roc_data$Model, levels=c("Logit", "Elastic Logit", "KNN", "RF"))
  }

  ##*****************************************************
  ##  Return a list object with each model object, ----
  ##  lift data and roc data

  return(list(model=model, model_rf=model_rf, model_logit=model_logit, model_boostedlogit=model_boostedlogit, master_lift=master_lift, roc_data=roc_data))
}

#' @title Permutation Test Statistics
#' @description Perform permutation test to compare two groups
#' 
#' @param data (data.frame) if paired, dataframe must have comparison groups in two different columns with the associated log2 conc, else comparison groups must be in a single column with the grouping name. 
#' @param gene (char) character of gene being tested
#' @param response_var (char) column containing response data
#' @param group1 (char) character of group 1. i.e. control
#' @param group2 (char) character of group 2. i.e. treatment
#' @param method (char) character of either 'paired.perm.test' or 'perm.test'
#' @param grouping_column (Char) character of column containing group labels
#' 
#' @return returns a data.frame of data and test p-value
do_perm_test <- function(data, gene, response_var, group1, group2, method, grouping_column){
  data <- as.data.table(data)
  results <- data.frame(gene=gene, .y.=response_var, group1=group1, group2=group2, method=method, p=NA )
  if (method=="paired.perm.test"){
    message(sprintf("INFO: Performing %s", method))
    results$p <- broman::paired.perm.test(data[[group2]] - data[[group1]])
  } else if (method=="perm.test"){
    message(sprintf("INFO: Performing %s", method))
    results$p <- broman::perm.test(data[data[[grouping_column]]==group1][[response_var]], data[data[[grouping_column]]==group2][[response_var]], n.perm = 20000, var.equal = FALSE)
  }
  message(sprintf("INFO: Returning results"))
  return( results )
}

#' @title Subset a data frame based on groups
#' @description Subset a data frame based on groups
#' 
#' @param data (data.frame) if paired, dataframe must have comparison groups in two different columns with the associated log2 conc, else comparison groups must be in a single column with the grouping name. 
#' @param filter_groups (char) vector of character of groups to filter for
#' @param method (char) character of either 'paired.perm.test' or 'perm.test'
#' @param grouping_column (Char) character of column containing groups labels
#' 
#' @return returns a subset of data
filter_data <- function(data, filter_groups, method, grouping_column){
  data %>%
    dplyr::filter(!!rlang::sym(grouping_column) %in% filter_groups) %>%
    dplyr::select(Number.x, !!rlang::sym(grouping_column), gene, log2_conc) -> data_sub
  if (method=="paired.perm.test"){
    data_sub %>% 
      dplyr::group_by(Number.x) %>% 
      dplyr::add_count() %>%
      dplyr::ungroup() %>% 
      dplyr::filter(n==8) %>%
      dplyr::select(-n) -> data_sub
  }
  data_sub
}
```

<!-- # Data File Paths -->
```{r message=FALSE, warning=FALSE, include=FALSE}
patient_clinical_data <- "../../data/elisa_plasma_proteomics/Patients_information_FINAL.csv"
patient_longitudinal_mapping_file <- "../../data/elisa_plasma_proteomics/Patients_for_ELISA_longitudinal_mapping.csv"
elisa_data_file <- "../../data/elisa_plasma_proteomics/ELISA_data.csv"
```

<!-- # Load Data -->
```{r echo=FALSE, warning=FALSE}
clinical_data <- fread(patient_clinical_data)
patient_longitudinal_mapping_data <- fread(patient_longitudinal_mapping_file)

patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "Overall_Lab_ID", all = TRUE)
patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y

elisa_data <- fread(elisa_data_file)

data_long <- merge(elisa_data, patient_meta_data, by = "LAB_ID")
# Remove patient 11, for some reason they are kept
data_long %>%
  dplyr::filter(Number.x!=11) -> data_long

patient_meta_data %>%
  dplyr::filter(!is.na(manifestation_type) & ELISA=="X") %>%
  dplyr::group_by(Number.x) %>%
  dplyr::add_count() %>%
  dplyr::select(`Unique Patient Number`=Number.x, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
  unique() %>% dplyr::ungroup() -> recurrent_patients_meta

# message(sprintf("There are %s patients with intial and recurrent ELISA measurements", nrow(recurrent_patients_meta)))

patient_meta_data %>%
  dplyr::filter(!is.na(manifestation_type)) %>%
  dplyr::select(Number.x, Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, `Mass Spectrometry`, Lipidomics, `RT-qPCR`, IHC, ELISA) %>%
  dplyr::group_by(Number.x) %>%
  dplyr::add_count() %>%
  unique() %>%
  dplyr::ungroup()-> meta_dt
```

<!-- # Distribution of ELISA Measurements -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
# iGBM vs rGBM ----
## Build gg dataframe for plotting ----
data_long %>%
  dplyr::filter(!is.na(manifestation_type) | patient_state==0) %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(Manifestation = plyr::mapvalues(manifestation_type, from=c("Initial manifestation", "Second manifestation"), to=c("iGBM", "rGBM"))) %>%
  dplyr::mutate(Manifestation = ifelse(is.na(Manifestation), "Ctrl", Manifestation)) %>%
  dplyr::mutate(Manifestation = factor(Manifestation, levels = c("Ctrl", "iGBM", "rGBM"))) -> pdata
pdata <- as.data.table(pdata)

## Perform Statistical-Test ----

comparisons <- matrix(c("Ctrl", "Ctrl", "iGBM", "iGBM", "rGBM", "rGBM"), ncol=2)
methods <- c("perm.test", "perm.test", "paired.perm.test")
# methods <- c("perm.test", "perm.test", "perm.test")

stats_frame <- parallel::mclapply(seq_len(nrow(comparisons)), function(iterator){
  comp <- comparisons[iterator, ]
  method <- methods[iterator]
  data_sub <- filter_data(pdata, comp, method, "Manifestation" )

  uni_genes <- data_sub %>% dplyr::pull(gene) %>% unique()

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "Manifestation", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "Manifestation")
  }, mc.cores = 4) %>%
    rbindlist() %>%
    unique()
}, mc.cores=3) %>% rbindlist()

stat_test <- stats_frame %>%
  dplyr::group_by(group1, group2) %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance(p.col="p", output.col = "p.signif",
                   cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                   symbols = c("****", "***", "**", "*", "ns")) %>%
  dplyr::arrange(gene) %>%
  add_xy_position(data=pdata, formula= log2_conc ~ Manifestation)

stat_test %>%
  mutate(y.position = c(6.64, 6.64, 6.64, 8.70004, 8.70004, 8.70004, 11.49704, 11.49704, 11.49704, 13.62804, 13.62804, 13.62804)) -> stat_test

stat_test_igm_v_rgm <- stat_test

## Build the plot ----
ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = "jco", line.color = "grey", line.size = 0.4) +
  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "Initial manifestation", "Second manifestation")) +
  # geom_jitter(aes(label=Number.x)) +
  # ggtitle("Initial vs Recurrence") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
         col=guide_legend(title="Healthy Control Baseline")) -> p1

p1 + 
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(8, 18, 9, 12)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1

## Build the table ----
stat_test %>%
  dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
  dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
  dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
  tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p") %>%
  gt() %>%
  tab_header(title = md("**Statistical Comparisons Between Population Groups per Genes**")) %>%
  cols_label(
    Comparison = md("**Comparison**"),
    `Statistical Test` = md("**Statistical Test**"),
    ASAH1 = md("**ASAH1**"),
    GPNMB = md("**GPNMB**"),
    `MMP-9` = md("**MMP-9**"),
    SYNEMIN = md("**SYNEMIN**"),
  ) %>% 
  tab_footnote(
    footnote = md("Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples."),
    locations = cells_column_labels("Statistical Test")
  ) %>% 
  tab_footnote(
    footnote = md("Values in cells represent the non-adjusted p-value from the corresponding statiscal test."),
    locations = cells_column_labels(c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9"))
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(ASAH1),
      rows = ASAH1 < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(GPNMB),
      rows = GPNMB < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(`MMP-9`),
      rows = `MMP-9` < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(SYNEMIN),
      rows = SYNEMIN < 0.05
    )
  ) -> stat_comp_table

# Pre-surgery vs Post-surgery ----
## Build gg dataframe for plotting ----
data_long %>%
  dplyr::filter(is.na(manifestation_type) | patient_state==0) %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(`Collection Type` = plyr::mapvalues(collection_type, from=c("ctrl", "E1", "E2"), to=c("Ctrl", "pre-\nsurgery", "post-\nsurgery"))) %>%
  # dplyr::mutate(`Collection Type` = ifelse(is.na(`Collection Type`), "Ctrl", `Collection Type`)) %>%
  dplyr::mutate(collection_type2 = factor(`Collection Type`, levels = c("Ctrl", "pre-\nsurgery", "post-\nsurgery"))) -> pdata2

## Perform Statistical-Test ----
comparisons <- matrix(c("Ctrl", "Ctrl", "pre-\nsurgery", "pre-\nsurgery", "post-\nsurgery", "post-\nsurgery"), ncol=2)
methods <- c("perm.test", "perm.test", "paired.perm.test")
# methods <- c("perm.test", "perm.test", "perm.test")

stats_frame <- parallel::mclapply(seq_len(nrow(comparisons)), function(iterator){
  comp <- comparisons[iterator, ]
  method <- methods[iterator]
  data_sub <- filter_data(pdata2, comp, method, "collection_type2")

  uni_genes <- data_sub %>% dplyr::pull(gene) %>% unique()

  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
    if (method=="paired.perm.test"){
      .d %>%
        dplyr::select(-gene) %>%
        tidyr::pivot_wider(id_cols = "Number.x", names_from = "collection_type2", values_from = "log2_conc") -> .d_wide
    } else {
      .d -> .d_wide
    }
    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "collection_type2")
  }, mc.cores = 4) %>%
    rbindlist() %>%
    unique()
}, mc.cores=3) %>% rbindlist()

stat_test_coord <- pdata2 %>%
  group_by(gene) %>%
  t_test(log2_conc ~ collection_type2) %>%
  add_xy_position() %>%
  dplyr::select(gene, .y., group1, group2, n1, n2, df, y.position, groups, xmin, xmax)

stat_test <- stats_frame %>%
  dplyr::group_by(group1, group2) %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance(p.col="p", output.col = "p.signif",
                   cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
                   symbols = c("****", "***", "**", "*", "ns")) %>%
  merge(stat_test_coord, by=c("gene", ".y.", "group1", "group2")) %>%
  dplyr::select(-p.adj) %>%
  dplyr::mutate(p=round(p, 6))

stat_test %>%
  mutate(y.position = c(7.42688, 7.42688, 7.42688, 8.60988, 8.60988, 8.60988, 10.82388, 10.82388, 10.82388, 13.57988, 13.57988, 13.57988)) -> stat_test

stat_test_pre_v_post_surgery <- stat_test

## Build the plot ----
ggpaired(data=pdata2, id = "Number.x", x = "collection_type2", y = "log2_conc", fill = "collection_type2", palette = "jco", line.color = "grey", line.size = 0.4) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata2$`collection_type2`)), "Dark2"), labels = c("Ctrl", "pre-surgery", "post-surgery")) +
  # stat_compare_means(paired = TRUE) +
  # ggnewscale::new_scale_color() +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
  # scale_color_manual(label="Mean Conc.", values="red") +
  # ggtitle("Pre-Surgery vs Post-Surgery") +
  labs(x="", y="log2( Concentration )") +
  facet_wrap(~gene, scales = "free_y") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical",) +
  guides(fill=guide_legend(title="EDTA Blood\nCollection Type", ncol=1),
         col=guide_legend(title="Healthy Control Baseline")) -> p2

p2 +
  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.15, hide.ns = T, size=4, bracket.size=1) +
  geom_point(data=data.frame(Manifestation=c("pre-\nsurgery", "pre-\nsurgery", "pre-\nsurgery", "pre-\nsurgery"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(10, 15, 9, 12)),
             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p2

supp_figure_pre_v_post <- p2

## Build the table ----
stat_test %>%
  dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
  dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
  dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
  tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p") %>%
  gt() %>%
  tab_header(title = md("**Statistical Comparisons Between Population Groups per Genes**")) %>%
  cols_label(
    Comparison = md("**Comparison**"),
    `Statistical Test` = md("**Statistical Test**"),
    ASAH1 = md("**ASAH1**"),
    GPNMB = md("**GPNMB**"),
    `MMP-9` = md("**MMP-9**"),
    SYNEMIN = md("**SYNEMIN**"),
  ) %>% 
  tab_footnote(
    footnote = md("Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples."),
    locations = cells_column_labels("Statistical Test")
  ) %>% 
  tab_footnote(
    footnote = md("Values in cells represent the non-adjusted p-value from the corresponding statiscal test."),
    locations = cells_column_labels(c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9"))
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(ASAH1),
      rows = ASAH1 < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(GPNMB),
      rows = GPNMB < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(`MMP-9`),
      rows = `MMP-9` < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(SYNEMIN),
      rows = SYNEMIN < 0.05
    )
  ) -> stat_pre_post_surgery_comp_table

# Build a master table ----
rbind(stat_test_igm_v_rgm %>%
        dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
        dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
        dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
        tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p"),
      stat_test_pre_v_post_surgery %>%
        dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
        dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
        dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
        tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p")) %>%
  gt() %>%
  tab_row_group(
    label = md("**Control vs Initial GBM vs Recurrent GBM Plasma Samples**"),
    rows = 1:3
  ) %>%
  tab_row_group(
    label = md("**Control vs Pre-Surgery vs Post-Surgery Plasma Samples**"),
    rows = 4:6
  ) %>%
  tab_header(title = md("**Statistical Comparisons Between Population Groups per Genes**")) %>%
  cols_label(
    Comparison = md("**Comparison**"),
    `Statistical Test` = md("**Statistical Test**"),
    ASAH1 = md("**ASAH1**"),
    GPNMB = md("**GPNMB**"),
    `MMP-9` = md("**MMP-9**"),
    SYNEMIN = md("**SYNEMIN**"),
  ) %>% 
  tab_footnote(
    footnote = md("Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples."),
    locations = cells_column_labels("Statistical Test")
  ) %>% 
  tab_footnote(
    footnote = md("Values in cells represent the non-adjusted p-value from the corresponding statistical test."),
    locations = cells_column_labels(c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9"))
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(ASAH1),
      rows = ASAH1 < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(GPNMB),
      rows = GPNMB < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(`MMP-9`),
      rows = `MMP-9` < 0.05
    )
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#F9E3D6"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = c(SYNEMIN),
      rows = SYNEMIN < 0.05
    )
  ) -> stat_comp_table

# Generate plot of rGBM to iGBM Ratios ----

data_long %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::add_count() %>%
  dplyr::filter(n==2) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>%
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) -> data_long_igm_rgm_ratios

data_long_igm_rgm_ratios %>%
  dplyr::filter(sample_comparison!="ctrl") %>%
  dplyr::filter(sample_comparison!="pre-surgery/post-surgery") %>%
  dplyr::group_by(gene, sample_comparison) %>%
  dplyr::mutate(myaxis = paste0(sample_comparison, "\n", "n=", n())) %>%
  ggplot() +
  # geom_violin(aes(x=as.factor(sample_comparison), y=log2(ratio), fill=as.factor(sample_comparison)), width=0.9) +
  geom_boxplot(aes(x=as.factor(gene), y=log2(ratio), fill=as.factor(gene)), width=0.3, alpha=0.9) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Dark2")) +
  # ggnewscale::new_scale_color() +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc+sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc-sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
  # scale_color_manual(label="Mean Conc.", values="red") +
  geom_jitter(aes(x=as.factor(gene), y=log2(ratio), col=as.factor(Number.x)), alpha=0.7) +
  # ggtitle("iGBM / rGBM Ratios") +
  labs(x="", y=expression(paste("log2 ", bgroup("(", frac('initial GBM', 'recurrent GBM'), ")"), " Ratio"))) +
  # facet_wrap(~gene) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical") +
  guides(fill="none", col="none") -> p3

supp_figure_ratios <- p3

data_long %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::add_count() %>%
  dplyr::filter(n==2) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>%
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Second manifestation"]
      .d_second <- .d[manifestation_type=="Initial manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBM/iGBM", "pre-surgery/post-surgery"))) -> data_long_rgm_igm_ratios

data_long_rgm_igm_ratios %>%
  dplyr::filter(sample_comparison!="ctrl") %>%
  dplyr::filter(sample_comparison!="pre-surgery/post-surgery") %>%
  dplyr::group_by(gene, sample_comparison) %>%
  dplyr::mutate(myaxis = paste0(sample_comparison, "\n", "n=", n())) %>%
  ggplot() +
  # geom_violin(aes(x=as.factor(sample_comparison), y=log2(ratio), fill=as.factor(sample_comparison)), width=0.9) +
  geom_boxplot(aes(x=as.factor(gene), y=log2(ratio), fill=as.factor(gene)), width=0.3, alpha=0.9) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Dark2")) +
  # ggnewscale::new_scale_color() +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc+sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc-sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
  # scale_color_manual(label="Mean Conc.", values="red") +
  geom_jitter(aes(x=as.factor(gene), y=log2(ratio), col=as.factor(Number.x)), alpha=0.7) +
  # ggtitle("rGBM / iGBM Ratios") +
  labs(x="", y=expression(paste("log2 ", bgroup("(", frac('recurrent GBM', 'initial GBM'), ")"), " Ratio"))) +
  # facet_wrap(~gene) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="vertical") +
  guides(fill="none", col="none") -> p3

supp_figure_ratios2 <- p3

# Master figure ----
# subfigure_6 <- ggarrange(p1, p2, p3, ncol = 3)
subfigure_6 <- p1
subfigure_6
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.png", plot = subfigure_6, width = 21, height = 11, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.svg", plot = subfigure_6, width = 21, height = 11, dpi = 200)
```

<!-- # Dimensionality Reduction -->

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
# Dimensionality Reduction ----
## Data wrnagling ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type %in% c("Initial manifestation", "Second manifestation") | sample_comparison=="ctrl") %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) %>%
  dplyr::mutate(class = ifelse(manifestation_type=="Initial manifestation", "iGBM", class)) %>%
  dplyr::mutate(class = ifelse(manifestation_type=="Second manifestation", "rGBM", class)) %>%
  dplyr::mutate(class = ifelse(is.na(manifestation_type), "Ctrl", class)) -> pdata

pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction ----
data_wide %>%
  dplyr::mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

## sPLS-DA ----
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)

MyResult.splsda <- mixOmics::splsda(train_df[, -1], as.factor(train_df[, 1]), keepX = c(4,4))

## Predict boundary planes of separation ----
background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")

## Prepare data frame for plotting ----
plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=as.factor(train_df[, 1]), as.data.frame(MyResult.splsda$variates$X))
plsda_data$patient_id[plsda_data$class=="Ctrl"] <- "Ctrl"
plsda_data$patient_id <- as.factor(plsda_data$patient_id)
plsda_data %>%
  dplyr::arrange(class) -> plsda_data
plsda_data <- as.data.table(plsda_data)

## Get average trajectories of iGBM to rGBM in low dimension ----
plsda_data[patient_id!="Ctrl" ] %>%
  dplyr::group_by(class) %>%
  dplyr::add_count() %>%
  dplyr::group_by(class, n) %>%
  dplyr::summarise(avg_comp1=sum(comp1)/n,
                   avg_comp2=sum(comp2)/n) %>%
  dplyr::ungroup() %>%
  unique() %>%
  dplyr::mutate(patient_id="avg_all") %>%
  as.data.table() -> average_comp

slope <- (average_comp$avg_comp2[2] - average_comp$avg_comp2[1])/(average_comp$avg_comp1[2] - average_comp$avg_comp1[1])
b = -0.25

average_comp[class=='rGBM']$avg_comp2 = slope*-1.5 + b
average_comp[class=='rGBM']$avg_comp1 = -1.5
average_comp[class=='iGBM']$avg_comp2 = slope*3 + b
average_comp[class=='iGBM']$avg_comp1 = 3

## Prepare decision boundaries dataframe fo plotting ----
for(i in 1:length(background))
{
  if(!is.null(background[[i]]))
    background[[i]]=data.frame(id=i,col=names(background)[i], background[[i]])
}

background = do.call(rbind,background)

## Build plot ----
### Average Trajectories
p <- ggplot() +
  geom_polygon(data = background,aes(x=Var1, y=Var2,
                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
  scale_fill_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  scale_shape_manual(values=c(15, 16, 17)) +
  scale_color_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
  ggnewscale::new_scale_color() +
  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
            arrow = arrow(type = "closed",
                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "right", legend.box="vertical",) +
  guides(shape="none",
         col_org=guide_legend(title="g", ncol=1),
         col=guide_legend(title="Trajectory", ncol=1),
         col_new=guide_legend(title="g", ncol=1))

splsda_avg_traj_fig <- p

# ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)

### Individual trajectories
p <- ggplot() +
  geom_polygon(data = background,aes(x=Var1, y=Var2,
                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
  scale_fill_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  scale_shape_manual(values=c(15, 16, 17)) +
  scale_color_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
  ggnewscale::new_scale_color() +
  geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
            arrow = arrow(type = "closed",
                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "right", legend.box="vertical",) +
  guides(shape="none",
         col_org=guide_legend(title="g", ncol=1),
         col=guide_legend(title="Patient ID\nTrajectory", ncol=2),
         col_new=guide_legend(title="g", ncol=1))

splsda_ind_traj_fig <- p

# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig2_subfigure_a.png", plot = splsda_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig2_subfigure_a.svg", plot = splsda_ind_traj_fig, width = 11, height = 8.5, dpi = 200)

```

<!-- # Ratios -->

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
## Data wrangling ----
### iGBM over rGBM ratio
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

### rGBM over iGBM
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_second$conc / .d_initial$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
  dplyr::filter(grepl(";", agg_lab_id)) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

### Prepare final table with ratios
pdata %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="iGBM") -> pdata

pdata2 %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="rGBM") -> pdata2

pdata <- rbindlist(list(pdata, pdata2))

pdata <- merge(pdata, meta_dt, by="Number.x")

### Pivot Table to wide
pdata %>%
  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide

# Prepare final table for analysis
train_df <- data_wide
train_df <- na.omit(train_df)  
train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
# Make Valid Column Names 
colnames(train_df) <- make.names(colnames(train_df))
train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.table(train_df)
train_df$group=0

## Plot Protein Ratios Against Each Other ----
p_pa_pb_ratio_scatter1 <- ggplot(train_df, aes(x=ASAH1, y=GPNMB, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="ASAH1 Ratio", y="GPNMB Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter2 <- ggplot(train_df, aes(x=ASAH1, y=MMP.9, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="ASAH1 Ratio", y="MMP-9 Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter3 <- ggplot(train_df, aes(x=ASAH1, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="ASAH1 Ratio", y="SYNEMIN Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter4 <- ggplot(train_df, aes(x=MMP.9, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="MMP-9 Ratio", y="SYNEMIN Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter5 <- ggplot(train_df, aes(x=GPNMB, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="GPNMB Ratio", y="SYNEMIN Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter6 <- ggplot(train_df, aes(x=GPNMB, y=MMP.9, col=class, shape=as.factor(Number.x))) +
  geom_point(size=3) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  labs(x="GPNMB Ratio", y="MMP-9 Ratio") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "bottom", legend.box="horizontal",) +
  guides(shape=guide_legend(title="Patient ID", ncol=5),
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Class", ncol=2, byrow = T))

p_pa_pb_ratio_scatter <- ggpubr::ggarrange(p_pa_pb_ratio_scatter1, p_pa_pb_ratio_scatter2, p_pa_pb_ratio_scatter3, 
                                           p_pa_pb_ratio_scatter4, p_pa_pb_ratio_scatter5, p_pa_pb_ratio_scatter6,
                                           ncol=3, nrow=2, common.legend = TRUE, legend="bottom")

# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_a.png", plot = p_pa_pb_ratio_scatter, width = 15, height = 9, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_a.svg", plot = p_pa_pb_ratio_scatter, width = 15, height = 9, dpi = 200)

## sPLS-DA of Ratios ----

MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")

plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
plsda_data$patient_id <- as.factor(plsda_data$patient_id)
plsda_data %>%
  dplyr::arrange(class) -> plsda_data
plsda_data <- as.data.table(plsda_data)

plsda_data %>%
  dplyr::group_by(class) %>%
  dplyr::add_count() %>%
  dplyr::group_by(class, n) %>%
  dplyr::summarise(avg_comp1=sum(comp1)/n,
                   avg_comp2=sum(comp2)/n) %>%
  dplyr::ungroup() %>%
  unique() %>%
  dplyr::mutate(patient_id="avg_all") %>%
  as.data.table() -> average_comp

slope <- (average_comp$avg_comp2[2] - average_comp$avg_comp2[1])/(average_comp$avg_comp1[2] - average_comp$avg_comp1[1])
b = -0.25

for(i in 1:length(background))
{
  if(!is.null(background[[i]]))
    background[[i]]=data.frame(id=i,col=names(background)[i],
                               background[[i]])
}

background = do.call(rbind,background)

p <- ggplot() +
  geom_polygon(data = background,aes(x=Var1, y=Var2,
                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
  scale_fill_manual(values=c("#388ECC", "#F68B33")) +
  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  scale_shape_manual(values=c(15, 16, 17)) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  ggnewscale::new_scale_color() +
  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
            arrow = arrow(type = "closed",
                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "right", legend.box="vertical",) +
  guides(shape="none",
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
         col=guide_legend(title="Trajectory", ncol=1))

splsda_ratio_avg_traj_fig <- p

# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)

p <- ggplot() +
  geom_polygon(data = background,aes(x=Var1, y=Var2,
                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
  scale_fill_manual(values=c( "#388ECC", "#F68B33")) +
  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
  scale_shape_manual(values=c(15, 16, 17)) +
  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
  ) +
  ggnewscale::new_scale_color() +
  geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
            arrow = arrow(type = "closed",
                          length=unit(0.1, "inches")), size=1.5, alpha=0.8) +
  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "right", legend.box="vertical",) +
  guides(shape="none",
         colour_new=guide_legend(title="Class", ncol=1, byrow = T),         
         col=guide_legend(title="Patient ID\nTrajectory", ncol=2))

splsda_ratio_ind_traj_fig <- p

# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.png", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.svg", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)

## Time to Recurrence using Ratios ----
train_df_Sub <- train_df[class=="rGBM"] %>%
  dplyr::arrange(Time.to.reccurence..days.)

train_df_Sub %>%
  tidyr::pivot_longer(cols = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9"), names_to = "gene", values_to = "ratio") %>%
  dplyr::group_by(gene, Time.to.reccurence..days.) %>%
  # dplyr::summarise(ratio=mean(ratio)) %>%
  dplyr::mutate(Time.to.reccurence..days. = as.numeric(Time.to.reccurence..days.)) %>%
  dplyr::ungroup() -> train_df_Sub_long

p_ratio_ttr <- ggplot(train_df_Sub_long) +
  geom_point(aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene, col=as.factor(Number.x)), size=3) +
  ggnewscale::new_scale("color") +
  # geom_line(aes(x=as.factor(Time.to.reccurence..days.), y=ratio, group=gene, col=gene)) +
  geom_smooth(aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene, col=gene), method = "lm", alpha=0.2) +
  stat_cor(method = "pearson", aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene)) +
  labs(x="Time to Recurrence (Days)", y=expression(paste("log2 ", bgroup("(", frac('recurrent GBM', 'initial GBM'), ")"), " Ratio"))) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=17),
        axis.title = element_text(size = 17),
        axis.text = element_text(size=16),
        strip.text = element_text(size=16),
        legend.title = element_text(size=16),
        legend.text = element_text(size=15),
        legend.position = "right", legend.box="vertical") +
  guides(col=guide_legend(title="Gene"), 
         colour_new=guide_legend(title="Patient ID", ncol = 2)) +
  facet_wrap(~gene)

train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
train_df_Sub$Time.to.reccurence..days. <- as.numeric(train_df_Sub$Time.to.reccurence..days.)
lm_ttr <- lmerTest::lmer(Time.to.reccurence..days. ~ SYNEMIN + GPNMB + ASAH1 + MMP.9 + (1 | Type.of.resection), data=train_df_Sub )
summary(lm_ttr)
anova(lm_ttr)

lm_ttr <- lm(Time.to.reccurence..days. ~ SYNEMIN + GPNMB + ASAH1 + MMP.9, data=train_df_Sub )
summary(lm_ttr)
# car::Anova(lm_ttr, type="III")

ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.png", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.svg", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)

```

<!-- # ROC Prediction Analysis -->

```{r fig.height=8, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}
# iGBM vs Ctrl protein conc ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Initial manifestation" | sample_comparison=="ctrl") %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_1.1 <- run_predictions(train_df)

## ROC Figure 
g <- ggplot(pred_obj_1.1$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.1$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_1.1 <- as.data.table(calc_auc(g))

g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.1[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.1$master_lift$liftModelVar)), "Dark2")[1])
g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.1[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.1$master_lift$liftModelVar)), "Dark2")[2])
g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.1[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.1$master_lift$liftModelVar)), "Dark2")[3])
g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.1[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.1$master_lift$liftModelVar)), "Dark2")[4])
subfigure_1.1 <- g

# rGBM vs Ctrl protein conc ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Second manifestation" | sample_comparison=="ctrl") %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_1.2 <- run_predictions(train_df)

## ROC Figure 
g <- ggplot(pred_obj_1.2$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.2$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_1.2 <- as.data.table(calc_auc(g))

g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.2[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.2$master_lift$liftModelVar)), "Dark2")[1])
g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.2[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.2$master_lift$liftModelVar)), "Dark2")[2])
g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.2[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.2$master_lift$liftModelVar)), "Dark2")[3])
g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.2[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.2$master_lift$liftModelVar)), "Dark2")[4])
subfigure_1.2 <- g

# iGBM vs rGBM protein conc----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Second manifestation" | manifestation_type=="Initial manifestation") %>%
  dplyr::mutate(class = plyr::mapvalues(manifestation_type, from=c("Initial manifestation","Second manifestation"), to=c("iGBM", "rGBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_1.5 <- run_predictions(train_df, pred_col_prob = "rGBM")

## ROC Figure 
g <- ggplot(pred_obj_1.5$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.5$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_1.5 <- as.data.table(calc_auc(g))

g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.5[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.5$master_lift$liftModelVar)), "Dark2")[1])
g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.5[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.5$master_lift$liftModelVar)), "Dark2")[2])
g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.5[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.5$master_lift$liftModelVar)), "Dark2")[3])
g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.5[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.5$master_lift$liftModelVar)), "Dark2")[4])
subfigure_1.5 <- g

pred_obj_1.5$roc_data %>%
  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.5_roc_data

# iGBM vs rGBM ratio ----
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_second$conc / .d_initial$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
  dplyr::filter(grepl(";", agg_lab_id)) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

pdata %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="iGBM") -> pdata

pdata2 %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="rGBM") -> pdata2

pdata <- rbindlist(list(pdata, pdata2))

## Pivot Table
pdata %>%
  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
## Make Valid Column Names 
colnames(train_df) <- make.names(colnames(train_df))
train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_1.3 <- run_predictions(train_df, pred_col_prob = "rGBM")

## ROC Figure 
g <- ggplot(pred_obj_1.3$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_1.3 <- as.data.table(calc_auc(g))

g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.3[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2")[1])
g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.3[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2")[2])
g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.3[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2")[3])
g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.3[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2")[4])
subfigure_1.3 <- g

pred_obj_1.3$roc_data %>%
  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.3_roc_data

## Merged Plot
pred_obj_1.3$roc_data %>%
  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.3_roc_data

# iGBM vs rGBM with clinical data ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Second manifestation" | manifestation_type=="Initial manifestation") %>%
  dplyr::mutate(class = plyr::mapvalues(manifestation_type, from=c("Initial manifestation","Second manifestation"), to=c("iGBM", "rGBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_1.4 <- run_predictions(train_df, pred_col_prob = "rGBM")

## ROC Figure 
g <- ggplot(pred_obj_1.4$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_1.4 <- as.data.table(calc_auc(g))

g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.4[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2")[1])
g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.4[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2")[2])
g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.4[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2")[3])
g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.4[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2")[4])
subfigure_1.4 <- g

pred_obj_1.4$roc_data %>%
  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.4_roc_data

## Merged Plot ----

master_roc_data <- rbindlist(list(`Ctrl vs iGBM`=pred_obj_1.1$roc_data, `Ctrl vs rGBM`=pred_obj_1.2$roc_data, `iGBM vs rGBM protein`=tmp_pred_obj_1.5_roc_data, `iGBM vs rGBM protein + clinical`=tmp_pred_obj_1.4_roc_data, `iGBM vs rGBM protein ratios`=tmp_pred_obj_1.3_roc_data), idcol = "Comparison")

master_auc_data <- rbindlist(list(`Ctrl vs iGBM`=auc_tbl_1.1, `Ctrl vs rGBM`=auc_tbl_1.2, `iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")

master_auc_data[Model %in% c("Elastic Logit")] %>%
  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
  tidyr::pivot_wider(id_cols = "Comparison", names_from = Model, values_from = AUC) -> auc_annotation_tbl

table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))))
auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)

g <- ggplot(master_roc_data[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Comparison, linetype=Comparison)) +
  geom_roc(n.cuts=0, size=1.5) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "Dark2")) +
  # scale_linetype_manual(values=c("solid", "dotted"), labels=c("Elastic Logit")) + 
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity") +
  annotation_custom(auc_grob, xmin=0.23, xmax=0.90, ymin=0.10, ymax=0.38) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=13),
        legend.title = element_text(size=14),
        legend.text = element_text(size=13))

ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.png", plot = g, width = 11, height = 8.5, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.svg", plot = g, width = 11, height = 8.5, dpi = 200)
subfigure_1 <- g
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Generate table of prediction metrics ----
pred_col_prob="GBM"

pred_mods_dt <- rbindlist(list(`Ctrl vs iGBM`=pred_obj_1.1$model$pred, 
                               `Ctrl vs rGBM`=pred_obj_1.2$model$pred, 
                               `iGBM vs rGBM protein`=pred_obj_1.5$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM"))), 
                               `iGBM vs rGBM protein + clinical`=pred_obj_1.4$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM"))), 
                               `iGBM vs rGBM protein ratios`=pred_obj_1.3$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM")))), idcol = "Comparison")

conf_mats <-lapply(split(pred_mods_dt[, .(obs, pred)], f=pred_mods_dt$Comparison), function(.d){
  caret::confusionMatrix(.d$obs, .d$pred, positive=pred_col_prob)
})
metric_table <- rbindlist(lapply(conf_mats, function(x){data.frame(metric=names(x$byClass), values=x$byClass)}), idcol = "Comparison")

metric_table %>%
  dplyr::group_by(Comparison, metric) %>% 
  dplyr::filter(metric %in% c("Balanced Accuracy", "Sensitivity", "Specificity", "Pos Pred Value", "Neg Pred Value")) %>%
  dplyr::mutate(values = sprintf("%s%%", round(values*100, 2))) %>%
  dplyr::select(Comparison, metric, values) -> metric_table

metric_table %>%
  tidyr::pivot_wider(id_cols="Comparison", names_from = metric, values_from = values) -> performance_dt

performance_dt %>% 
  dplyr::arrange(Comparison) %>%
  ungroup() %>%
  gt() %>%
  gt::tab_header(md("**Logistic Regression with Elastic Net Regularization Performance Metrics**")) %>%
  cols_label(.list = list(Comparison=md("**Comparison**"), `Balanced Accuracy`=md("**Balanced Accuracy**"), Sensitivity=md("**Sensitivity**"), Specificity=md("**Specificity**"), `Pos Pred Value`=md("**Pos Pred Value**"), `Neg Pred Value`=md("**Neg Pred Value**"))) -> logit_net_comp_perf_met_table
```

<!-- # Prediction for Ctrl vs iGBM -->
```{r fig.height=8, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}
# iGBM vs Ctrl protein conc ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Initial manifestation" | sample_comparison=="ctrl") %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_2.1 <- run_predictions(train_df)

## ROC Figure 
g <- ggplot(pred_obj_2.1$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_2.1$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_2 <- as.data.table(calc_auc(g))

auc_tbl_2 %>%
  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
  dplyr::select(AUC) -> auc_annotation_tbl

table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)

g +
  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=13),
        legend.title = element_text(size=14),
        legend.text = element_text(size=13)) -> g
g
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_a.png", plot = g, width = 9, height = 8, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_a.svg", plot = g, width = 9, height = 8, dpi = 200)

subfigure_2 <- g

```

<!-- # Prediction for iGBM vs rGBM  -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
# iGBM vs rGBM protein conc ----
data_long %>% 
  dplyr::group_by(gene, Number.x) %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::filter(manifestation_type=="Second manifestation" | sample_comparison=="ctrl") %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata

## Pivot Table
pdata %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide

## Only use proteins for prediction
data_wide %>%
  mutate(n = row_number()) %>%
  dplyr::select(Number.x, class, n, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_3 <- run_predictions(train_df)

## ROC Figure 
g <- ggplot(pred_obj_3$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_3 <- as.data.table(calc_auc(g))

auc_tbl_3 %>%
  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
  dplyr::select(AUC) -> auc_annotation_tbl

table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)

g +
  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=13),
        legend.title = element_text(size=14),
        legend.text = element_text(size=13)) -> g
g
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_b.png", plot = g, width = 9, height = 8, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_b.svg", plot = g, width = 9, height = 8, dpi = 200)

subfigure_3 <- g
```

<!-- # Prediction for iGBM vs rGBM ratio -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
# iGBM vs rGBM ratio ----
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_second$conc / .d_initial$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
  dplyr::filter(grepl(";", agg_lab_id)) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

pdata %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="iGBM") -> pdata

pdata2 %>%
  dplyr::select(Number.x, gene, log2_conc) %>%
  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
  dplyr::mutate(class="rGBM") -> pdata2

pdata <- rbindlist(list(pdata, pdata2))

## Pivot Table
pdata %>%
  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

## Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
## Make Valid Column Names 
colnames(train_df) <- make.names(colnames(train_df))
train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_4 <- run_predictions(train_df, pred_col_prob = "rGBM")

## ROC Figure 
g <- ggplot(pred_obj_4$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_4$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_4 <- as.data.table(calc_auc(g))

auc_tbl_4 %>%
  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
  dplyr::select(AUC) -> auc_annotation_tbl

table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)

g +
  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=13),
        legend.title = element_text(size=14),
        legend.text = element_text(size=13)) -> g
g
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_d.png", plot = g, width = 9, height = 8, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_d.svg", plot = g, width = 9, height = 8, dpi = 200)

subfigure_4 <- g
```

<!-- # Protein Measurements and Clinical Data -->
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
data_long %>%
  dplyr::filter(!is.na(manifestation_type) | patient_state==0) %>%
  dplyr::mutate(log2_conc = log2(conc+1)) %>%
  dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(Manifestation = plyr::mapvalues(manifestation_type, from=c("Initial manifestation", "Second manifestation"), to=c("iGBM", "rGBM"))) %>%
  dplyr::mutate(Manifestation = ifelse(is.na(Manifestation), "Ctrl", Manifestation)) %>%
  dplyr::mutate(Manifestation = factor(Manifestation, levels = c("Ctrl", "iGBM", "rGBM"))) %>%
  dplyr::filter(Manifestation!="Ctrl") %>%
  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, log2_conc, Manifestation) %>%
  tidyr::pivot_wider(id_cols = c("Number.x", "Manifestation",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Manifestation'), names_from = "gene", values_from = "log2_conc" ) %>%
  dplyr::ungroup() -> data_wide #'Latency (days)'

# data_wide$`Age at surgery (years)` <- factor(data_wide$`Age at surgery (years)`, levels = unique(data_wide$`Age at surgery (years)`))
data_wide$`Tumor localization` <- factor(data_wide$`Tumor localization`, levels = unique(data_wide$`Tumor localization`))
data_wide$`Tumor localization` <- as.numeric(data_wide$`Tumor localization`)
data_wide$`Type of resection` <- factor(data_wide$`Type of resection`, levels = unique(data_wide$`Type of resection`))
data_wide$`Type of resection`  <- as.numeric(data_wide$`Type of resection` )
data_wide$`MGMT promotor methylation` <- factor(data_wide$`MGMT promotor methylation`, levels = unique(data_wide$`MGMT promotor methylation`))
data_wide$`MGMT promotor methylation` <- as.numeric(data_wide$`MGMT promotor methylation`)
data_wide$`p53 accumulation` <- factor(data_wide$`p53 accumulation`, levels = unique(data_wide$`p53 accumulation`))
data_wide$`p53 accumulation` <- as.numeric(data_wide$`p53 accumulation` )
data_wide$`Ki67-Li` <- factor(data_wide$`Ki67-Li`, levels = unique(data_wide$`Ki67-Li`))
data_wide$`Ki67-Li` <- as.numeric(data_wide$`Ki67-Li` )
data_wide$`EGFR vIII` <- factor(data_wide$`EGFR vIII`, levels = unique(data_wide$`EGFR vIII`))
data_wide$`EGFR vIII` <- as.numeric(data_wide$`EGFR vIII` )
data_wide$Treatment <- factor(data_wide$Treatment, levels = unique(data_wide$Treatment))
data_wide$Treatment <- as.numeric(data_wide$Treatment)
# data_wide$`Time-to-reccurence (days)` <- factor(data_wide$`Time-to-reccurence (days)`, levels = unique(data_wide$`Time-to-reccurence (days)`))
# data_wide$`Latency (days)` <- factor(data_wide$`Latency (days)`, levels = unique(data_wide$`Latency (days)`))

# Encode Category Sex
data_wide$Sex <- plyr::mapvalues(data_wide$Sex, from = c("w", "m"), to = c(0, 1))
data_wide$Sex <- as.factor(data_wide$Sex)
data_wide$Sex <- as.numeric(data_wide$Sex)

data_wide  %>%
  dplyr::select(Number.x, -`Age at surgery (years)`, -Sex, `Tumor localization`, -`Type of resection`, -`MGMT promotor methylation`, -`p53 accumulation`, -`Ki67-Li`, -`EGFR vIII`, -Treatment, -`Time-to-reccurence (days)`, Manifestation, ASAH1, SYNEMIN,GPNMB,`MMP-9`) -> data_wide

# Prediction
train_df <- data_wide
train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x)
# Make Valid Column Names 
colnames(train_df) <- make.names(colnames(train_df))
train_df %>% dplyr::rename(class=Manifestation) -> train_df
train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
colnames(train_df) <- make.names(colnames(train_df))
train_df <- as.data.frame(train_df)
pred_obj_5 <- run_predictions(train_df, pred_col_prob = "rGBM")

# ROC
g <- ggplot(pred_obj_5$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
  geom_roc(n.cuts=0) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_5$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
  coord_equal() +
  style_roc(x="1 - Specificity", y="Sensitivity")

auc_tbl_5 <- as.data.table(calc_auc(g))

auc_tbl_5 %>%
  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
  dplyr::select(AUC) -> auc_annotation_tbl

table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)

g +
  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=13),
        legend.title = element_text(size=14),
        legend.text = element_text(size=13)) -> g
g
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_c.png", plot = g, width = 9, height = 8, dpi = 200)
# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_c.svg", plot = g, width = 9, height = 8, dpi = 200)

subfigure_5 <- g
```

\blandscape

# Master Figure
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
mataster_figure <- ggarrange(ggarrange(subfigure_6, 
                                       splsda_ratio_avg_traj_fig, ncol=2, labels = c("A", "B")),
                             ggarrange(ggplot() + 
                                         annotate("text", x = 4, y = 25, size=8, label = "Kaplan-Meyer plots") + 
                                         theme_void(), ncol=1, labels = c("C")), 
                             ggarrange(subfigure_1, 
                                       p_ratio_ttr, ncol=2, labels = c("D", "E")),
                             ncol = 1, nrow=3)
mataster_figure
ggsave("../../figures/elisa_plasma_proteomics/elisa_master_figure.png", plot = mataster_figure, width = 25, height = 21, dpi = 200)
ggsave("../../figures/elisa_plasma_proteomics/elisa_master_figure.svg", plot = mataster_figure, width = 25, height = 21, dpi = 200)

```

**Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA). (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM populations. (**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$. (**c**)  Kaplan-Meyer plots showing... (**d**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins. (**e**) The potential of recurrent GBM / initial GBM protein ratios for predicting time to recurrence. Note, each point represents a single patient, so we may not be able to directly state a specific ratio can be used to infer to time to recurrence. However, we can still state given a specific patients ratio, we can predict that patients time to recurrence.

> **Note:** I'm not entirely sure about plot e, and if my thinking/interpretation is right, but let me know what you think.

> **Note:** I excluded the pre-surgery vs post-surgery distribution boxplots, because I didn't do any prediction analyses on that set. We can still include it in the supplement if we thing there is anything interesting there to show?

\elandscape

\pagebreak

# Supplemental Figures and Tables

```{r message=FALSE, warning=FALSE, include=FALSE}
fig_counter <- 1
tab_counter <- 1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8.5, fig.width=11}

supp_figure_0 <- ggarrange(supp_figure_pre_v_post, ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), ncol=2, nrow=1, labels=c("A", "B"))
supp_figure_0
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_0, width = 18, height = 16, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_0, width = 18, height = 16, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
```
**Supplemental Figure** `r knitr::asis_output(md(sprintf("**%s**", fig_counter)))` `r fig_counter <- fig_counter + 1`**: Distribution of Pre- vs Post-Surgery Plasma Proteomics and Ratios of Initial and Recurrent GBM Plasma Proteomics.**  (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, pre-surgery and post-surgery populations.  (**b**) Demonstrates the log transformed $\frac{initial \quad GBM}{recurrent \quad GBM}$ ratio (top) as well as the inverse ratio (bottom) for the potential use of identifying a patients state to recurrence using these ratios.

\pagebreak

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
supp_figure_1 <- ggarrange(subfigure_2, subfigure_3, subfigure_5, subfigure_4, ncol=2, nrow=2, labels=c("A", "B", "C", "D"))

supp_figure_1
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_1, width = 18, height = 18, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_1, width = 18, height = 18, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.png", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.svg", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
```
**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Variation in Predictive Performance using Different Models.** Testing parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms on different prediction tasks. (**a**) Most models have AUCs greater than 60% when predicting control vs initial GBM. (**b**) Most models have comparable AUCs, except for k-nearest neighbours when predicting for initial GBM vs recurrent GBM. (**c**) Including clinical data for *Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection* reduces predictive performance for all models. (**d**) All models, except random forest, have AUCs above 80% when predicting initial GBM vs recurrent GBM using protein ratios.

\pagebreak

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=8.5, dpi = 200}
supp_figure_2 <- ggarrange(splsda_avg_traj_fig, splsda_ind_traj_fig, ncol=1, nrow=2, labels=c("A", "B"))

supp_figure_2
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_2, width = 11, height = 8.5, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_2, width = 11, height = 8.5, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = splsda_avg_traj_fig, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = splsda_avg_traj_fig, width = 9, height = 8, dpi = 200)

# ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
# ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)

```
**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Individual Patient Trajectory in Low Dimensional Space.** (**a**) A sparse partial least squares discriminant analysis shows separation between the different population groups, as well as an overall trajectory of a initial GBM patient going towards a recurrent GBM patient in a low dimensional space. (**b**) Individual trajectories demonstrate the individual variance of patients path to recurrence in a low-dimensional space.

\pagebreak

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21, dpi=200}
supp_figure_3 <- ggarrange( p_pa_pb_ratio_scatter, splsda_ratio_ind_traj_fig, ncol=1, nrow=2, labels=c("A", "B"), heights = c(1.5,1))

supp_figure_3
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_3, width = 18, height = 18, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_3, width = 18, height = 18, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = p_pa_pb_ratio_scatter, width = 11, height = 8.5, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = p_pa_pb_ratio_scatter, width = 11, height = 8.5, dpi = 200)

ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = splsda_ratio_ind_traj_fig, width = 9, height = 8, dpi = 200)
ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = splsda_ratio_ind_traj_fig, width = 9, height = 8, dpi = 200)
```
**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Information Contained in Plasma Proteomic Ratios for Initial GBM vs Recurrent GBM Stratification.** (**a**) Pairwise protein ratios shows promise of individual protein-pair ratios being able to stratify initial vs recurrent plasma GBM samples. (**b**) Individual trajectories demonstrate the individual variance of patients path to recurrence in a low-dimensional space using ratios.

\pagebreak

\blandscape

```{r echo=FALSE, message=FALSE, warning=FALSE}
logit_net_comp_perf_met_table %>%
  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))
logit_net_comp_perf_met_table %>%
  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))

logit_net_comp_perf_met_table
```

**Supplemental Table `r as.roman(tab_counter)` `r tab_counter <- tab_counter + 1`: Logistic Regression with Elastic Net Regularization Performance Metrics**

```{r echo=FALSE, message=FALSE, warning=FALSE}
stat_comp_table %>%
  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))
stat_comp_table %>%
  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.tex", tab_counter))

stat_comp_table
```

**Supplemental Table `r as.roman(tab_counter)` `r tab_counter <- tab_counter + 1`: Statistical Comparisons Between Population Groups per Genes**

\elandscape

\pagebreak

# Methods

## Statistical Test for Comparison of Log2 Distributions

Log2 transformed distributions of protein concentrations were compared pairwise for control, initial GBM, and recurrent GBM samples for each protein, ASAH1, GNMPB, MMP-9 and SYNEMIN. Statistical p-values were computed using a permutation test, that utilizes t-test statistics but performs random permutations of the data for computing a p-value. Note, for paired sample data, initial GBM vs recurrent GBM, a paired permutation test was performed instead. Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382 ), permutation tests were performed using broman (v0.80).

## Dimensionaly Reduction

To visualize the samples plasma proteomics on a different dimension, we applied a sparse Partial Least Squares Discriminant Analysis (sPLS-DA), which maximizes the covariance between the explanatory and response variable(s). This was performed on the log2 transformed protein concentrations, as well as the log2 transformed ratios ($\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$) of each protein. Individual trajectories are displayed as vector arrows connecting initial GBM patient samples to their corresponding recurrent GBM patient samples, in the sPLS-DA space using the first two components that contain the most amount of explained variance. An average trajectory direction can be computed by taking the average of the individual trajectory vectors, to visualize an overall trajectory of initial GBM state to recurrent GBM state. To visualize decision boundaries for class separation (control, initial GBM and recurrent GBM samples), we applied a maximum distance approach. [1] Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382), sPLS-DA analysis was performed using mixOmics (v6.18.1).

## Prediction Analysis

Prediction tasks were performed using parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms. General predictive tasks were performed on: controls vs initial GBM samples, controls vs recurrent GBM samples, initial GBM samples vs recurrent GBM samples. Data used in prediction tasks are log2 transformed protein concentrations, log2 transformed protein concentrations supplemented with clinical data (*Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection*), and log2 transformed ratios of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$ protein measurements. Model training and prediction performances are performed on the entire data set using leave-one-out cross-validation due to the small sample size. Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382), prediction analyses were performed using caret (v6.0-93).

# References

1. Rohart, F, B Gautier, A Singh, and K-A Lê Cao. 2017. “MixOmics: An R Package for ‘Omics Feature Selection and Multiple Data Integration.” PLoS Computational Biology 13 (11). Cold Spring Harbor Labs Journals.

### 2022-11-28 tables/elisa_plasma_proteomics/supp_table_2.tex

\setlength{\LTpost}{0mm}
\begin{longtable}{llrrrr}
\caption*{
{\large \textbf{Statistical Comparisons Between Population Groups per Genes}}
} \\ 
\toprule
\textbf{Comparison} & \textbf{Statistical Test}\textsuperscript{1} & \textbf{ASAH1}\textsuperscript{2} & \textbf{GPNMB}\textsuperscript{2} & \textbf{MMP-9}\textsuperscript{2} & \textbf{SYNEMIN}\textsuperscript{2} \\ 
\midrule
\multicolumn{6}{l}{\textbf{Control vs Pre-Surgery vs Post-Surgery Plasma Samples}} \\ 
\midrule
Ctrl vs post-
surgery & Permutation Test & 0.7522500 & 0.1694500 & 0.1216000 & 0.2528000 \\ 
Ctrl vs pre-
surgery & Permutation Test & 0.0223500 & 0.9930500 & 0.0421500 & 0.0599500 \\ 
pre-
surgery vs post-
surgery & Paired Permutation Test & 0.0089720 & 0.2421880 & 0.4755860 & 0.5427860 \\ 
\midrule
\multicolumn{6}{l}{\textbf{Control vs Initial GBM vs Recurrent GBM Plasma Samples}} \\ 
\midrule
Ctrl vs iGBM & Permutation Test & 0.1804000 & 0.4020500 & 0.0805000 & 0.0021000 \\ 
Ctrl vs rGBM & Permutation Test & 0.2538500 & 0.2506500 & 0.4599500 & 0.1405500 \\ 
iGBM vs rGBM & Paired Permutation Test & 0.1992188 & 0.6123047 & 0.1171875 & 0.1054688 \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{1}Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples.\\
\textsuperscript{2}Values in cells represent the non-adjusted p-value from the corresponding statistical test.\\
\end{minipage}

### 2022-12-15 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
[ADD] Kaplan-Meir plots and lm of time to recurrence

@@
 installIfNeeded("broman")
+installIfNeeded("pwr")
+# Survival Analysis
+installIfNeeded("survival")
+installIfNeeded("survminer")
 # Machine Learning 
@@
 
+\pagebreak
+
 <!-- # Distribution of ELISA Measurements -->
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
+```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
 # iGBM vs rGBM ----
@@
 ## Build the plot ----
-ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = "jco", line.color = "grey", line.size = 0.4) +
+colfunc <- colorRampPalette(c("red","yellow","springgreen","royalblue"))
+ttr_categorizer <- function(data) {
+  data <- as.numeric(data)
+  if (is.na(data)){
+    return(NA)
+  }
+  if (data >= 100 & data < 200){
+    return("100-200")
+  } else if (data >= 200 & data < 300){
+    return("200-300")
+  } else if (data >= 300 & data < 400){
+    return("300-400")
+  } else if (data >= 400 & data < 500){
+    return("400-500")
+  } else if (data >= 500 & data < 600){
+    return("500-600")
+  }
+}
+# ttr_categorizer <- function(data) {
+#   data <- as.numeric(data)
+#   if (is.na(data)){
+#     return(NA)
+#   }
+#   if (data >= 100 & data < 150){
+#     return("100-150")
+#   } else if (data >= 150 & data < 200){
+#     return("150-200")
+#   } else if (data >= 200 & data < 250){
+#     return("200-250")
+#   } else if (data >= 250 & data < 300){
+#     return("250-300")
+#   } else if (data >= 300 & data < 350){
+#     return("300-350")
+#   } else if (data >= 350 & data < 400){
+#     return("350-400")
+#   } else if (data >= 400 & data < 450){
+#     return("400-450")
+#   } else if (data >= 450 & data < 500){
+#     return("450-500")
+#   } else if (data >= 500 & data < 550){
+#     return("500-550")
+#   } else if (data >= 550 & data < 600){
+#     return("550-600")
+#   }
+# }
+# ttr_categorizer <- function(data, split_value) {
+#   if (data <= split_value) {
+#     return(0)
+#   } else {
+#     return(1)
+#   }
+# }
+# ttr_categorizer <- function(data) {
+#   if (data >= 103 & data < 190){
+#     return("early")
+#   } else if (data >= 190 & data < 243){
+#     return("middle")
+#   } else if (data >= 243 & data <= 521){
+#     return("late")
+#   }
+# }
+pdata$`ttr_group` <- sapply(as.numeric(pdata$`Time-to-reccurence (days)`), ttr_categorizer)
+ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = colfunc(length(unique(pdata$`ttr_group`))), line.color = "ttr_group", line.size = 0.4) +
   # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
@@
   guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
-         col=guide_legend(title="Healthy Control Baseline")) -> p1
+         col=guide_legend(title="Time to Recurrence")) -> p1
 
@@
              aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1
+# p1
+
+## Average Conc per ttr
+pdata %>%
+  dplyr::filter(Manifestation!="Ctrl") %>%
+  dplyr::select(gene, Number.x, Manifestation, log2_conc, ttr_group) %>%
+  dplyr::mutate(Manifestation = plyr::mapvalues(Manifestation, from=c("iGBM", "rGBM"), to=c("initial\nGBM", "recurrent\nGBM"))) %>%
+  dplyr::group_by(gene, Manifestation, ttr_group) %>%
+  dplyr::mutate(avg_log2_conc_per_mani_ttr = median(log2_conc, na.rm = T)) -> pdata_ttr
+
+p1_avg_ttr <- ggplot(data = pdata_ttr, aes(x=Manifestation, y=log2_conc)) +
+  geom_boxplot() +
+  geom_line(aes(x=Manifestation, y=avg_log2_conc_per_mani_ttr, col=ttr_group, group=ttr_group), inherit.aes = FALSE) +
+  labs(x="", y="log2( Concentration )") +
+  facet_wrap(~gene, scales = "free_y") +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "bottom", legend.box="vertical",) +
+  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
+         col=guide_legend(title="Avg. log2 ( concentration )\nTime to Recurrence"))
 
@@
   ) -> stat_comp_table
+
+## Time to Recurrence ----
+train_df_Sub <- pdata %>%
+  dplyr::filter( !(Number.x %in% c("32", "33"))) %>%
+  dplyr::filter(Manifestation!="Ctrl") %>%
+  dplyr::select(Number.x, Time.to.reccurence..days.=`Time-to-reccurence (days)`, class=Manifestation, gene, log2_conc) %>%
+  # dplyr::mutate(Time.to.reccurence..days. = ifelse(class=="iGBM", 0, Time.to.reccurence..days.)) %>%
+  # dplyr::mutate(class = ifelse(class=="iGBM", 0, 1) ) %>%
+  dplyr::rowwise() %>%
+  # dplyr::mutate(ttr_group = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
+  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 253.5)) %>%
+  # dplyr::select(-Time.to.reccurence..days.) %>%
+  dplyr::mutate(gene=make.names(gene)) %>%
+  tidyr::pivot_wider(id_cols =c("Number.x", "Time.to.reccurence..days.", "class"), names_from = "gene", values_from = "log2_conc") %>%
+  # tidyr::pivot_wider(id_cols = c("Number.x", "Time.to.reccurence..days."), names_from = c("class"), values_from = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9")) %>%
+  # dplyr::mutate(ASAH1 = (ASAH1_rGBM+1) /( ASAH1_iGBM+1),
+  #               SYNEMIN = (SYNEMIN_rGBM+1)/(SYNEMIN_iGBM+1),
+  #               GPNMB = (GPNMB_rGBM+1)/(GPNMB_iGBM+1),
+  #               MMP.9 = (MMP.9_rGBM+1)/(MMP.9_iGBM+1)) %>%
+  # dplyr::select(Number.x, Time.to.reccurence..days., ASAH1, SYNEMIN, GPNMB, MMP.9 ) %>% 
+  # unique() %>%
+  dplyr::arrange(Time.to.reccurence..days.)
+train_df_Sub <- na.omit(train_df_Sub)
+train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
+
+train_df_Sub %>%
+  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
+  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
+                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
+                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
+                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
+  dplyr::rowwise() %>%
+  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
+  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
+  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
+  unique() -> ttr_delta
+
+# ttr_delta$Time.to.reccurence..days. <- as.numeric(as.factor(ttr_delta$Time.to.reccurence..days.))
+lm_ttr <- lm(Time.to.reccurence..days. ~ ASAH1_delta + SYNEMIN_delta + GPNMB_delta + MMP.9_delta, data=ttr_delta )
+summary(lm_ttr)
+anova(lm_ttr)
+
+power_calc <- pwr.f2.test(u = 4, f2 = sqrt(summary(lm_ttr)$adj.r.squared), sig.level = 0.05, power = 0.8)
+print(sprintf("The estimated sample size needed for 80%% power is: %s", round(power_calc$v) + 5))
+
+ttr_categorizer <- function(data) {
+  if (data >= 103 & data < 190){
+    return("early")
+  } else if (data >= 190 & data < 243){
+    return("middle")
+  } else if (data >= 243 & data <= 521){
+    return("late")
+  }
+}
+
+train_df_Sub %>%
+  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
+  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
+                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
+                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
+                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
+  dplyr::rowwise() %>%
+  dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
+  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
+  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
+  unique() -> ttr_delta
+
+tmp_pdata <- ttr_delta %>% tidyr::pivot_longer(cols = c(ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta), names_to="gene", values_to="log2_conc")
+tmp_pdata$Time.to.reccurence..days. <- factor(tmp_pdata$Time.to.reccurence..days., levels=c("early", "middle", "late"))
+ggplot(data = tmp_pdata, aes(x=Time.to.reccurence..days., y=log2_conc, col=Time.to.reccurence..days.)) +
+  geom_point() +
+  labs(x="", y="log2( Concentration )") +
+  facet_wrap(~gene, scales = "free_y") +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "bottom", legend.box="vertical",) +
+  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
+         col=guide_legend(title="Time to Recurrence"))
 
@@
 ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.svg", plot = subfigure_6, width = 21, height = 11, dpi = 200)
+
+p1_avg_ttr
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a_avg.png", plot = p1_avg_ttr, width = 21, height = 11, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a_avg.svg", plot = p1_avg_ttr, width = 21, height = 11, dpi = 200)
 ```
@@
 
-train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
-train_df_Sub$Time.to.reccurence..days. <- as.numeric(train_df_Sub$Time.to.reccurence..days.)
-lm_ttr <- lmerTest::lmer(Time.to.reccurence..days. ~ SYNEMIN + GPNMB + ASAH1 + MMP.9 + (1 | Type.of.resection), data=train_df_Sub )
-summary(lm_ttr)
-anova(lm_ttr)
-
-lm_ttr <- lm(Time.to.reccurence..days. ~ SYNEMIN + GPNMB + ASAH1 + MMP.9, data=train_df_Sub )
-summary(lm_ttr)
+# train_df_Sub <- train_df %>%
+#   dplyr::select(Number.x, Time.to.reccurence..days., class, ASAH1, SYNEMIN, GPNMB, MMP.9) %>%
+#   tidyr::pivot_wider(id_cols = c("Number.x", "Time.to.reccurence..days."), names_from = c("class"), values_from = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9")) %>%
+#   dplyr::arrange(Time.to.reccurence..days.)
+#   train_df_Sub <- na.omit(train_df_Sub)
+# 
+# train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
+# train_df_Sub$Time.to.reccurence..days. <- as.numeric(train_df_Sub$Time.to.reccurence..days.)
+# lm_ttr <- lmerTest::lmer(Time.to.reccurence..days.
+~ ASAH1_iGBM + ASAH1_rGBM + SYNEMIN_iGBM + SYNEMIN_rGBM + GPNMB_iGBM + GPNMB_rGBM + MMP.9_iGBM + MMP.9_rGBM + (1 | Type.of.resection), data=train_df_Sub )
+# summary(lm_ttr)
+# anova(lm_ttr)
+# 
+# lm_ttr <- lm(Time.to.reccurence..days. ~ ASAH1_iGBM + ASAH1_rGBM + SYNEMIN_iGBM + SYNEMIN_rGBM + GPNMB_iGBM + GPNMB_rGBM + MMP.9_iGBM + MMP.9_rGBM, data=train_df_Sub )
+# summary(lm_ttr)
 # car::Anova(lm_ttr, type="III")
 
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.png", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.svg", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_e.png", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_e.svg", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
 
@@
   mutate(n = row_number()) %>%
-  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`) -> data_wide
+  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`, `Time-to-reccurence (days)`) -> data_wide
+data_wide %>%
+  dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
 
@@
   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.4_roc_data
+
+# iGBM vs rGBM ratio  with clinical data ----
+# data_long %>% 
+#   dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+#   dplyr::group_by(gene, Number.x) %>%
+#   dplyr::group_map(function(.d, ...){
+#     .d <- as.data.table(.d)
+#     .d %>% 
+#       dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+#       dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+#                        agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+#     # print(.d)
+#     if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+#       .d_initial <- .d[manifestation_type=="Initial manifestation"]
+#       .d_second <- .d[manifestation_type=="Second manifestation"]
+#       tmp$ratio <- .d_initial$conc / .d_second$conc
+#     } else if (nrow(.d)==2) {
+#       .d_initial <- .d[collection_type=="E1"]
+#       .d_second <- .d[collection_type=="E2"]
+#       tmp$ratio <- .d_initial$conc / .d_second$conc
+#     } else {
+#       tmp$ratio <- .d$conc
+#     }
+#     return(tmp)
+#   }, .keep = TRUE) %>%
+#   rbindlist() %>%
+#   dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
+#   dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
+#   dplyr::mutate(log2_conc = log2(ratio)) %>%
+#   # dplyr::group_by(gene, Number.x) %>%
+#   # dplyr::add_count() %>%
+#   # dplyr::filter(n==2) %>%
+#   dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
+# 
+# data_long %>% 
+#   dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+#   dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
+#   dplyr::group_by(gene, Number.x) %>%
+#   dplyr::group_map(function(.d, ...){
+#     .d <- as.data.table(.d)
+#     .d %>% 
+#       dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+#       dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+#                        agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+#     # print(.d)
+#     if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+#       .d_initial <- .d[manifestation_type=="Initial manifestation"]
+#       .d_second <- .d[manifestation_type=="Second manifestation"]
+#       tmp$ratio <- .d_second$conc / .d_initial$conc
+#     } else if (nrow(.d)==2) {
+#       .d_initial <- .d[collection_type=="E1"]
+#       .d_second <- .d[collection_type=="E2"]
+#       tmp$ratio <- .d_initial$conc / .d_second$conc
+#     } else {
+#       tmp$ratio <- .d$conc
+#     }
+#     return(tmp)
+#   }, .keep = TRUE) %>%
+#   rbindlist() %>%
+#   dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+#   dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
+#   dplyr::filter(grepl(";", agg_lab_id)) %>%
+#   dplyr::mutate(log2_conc = log2(ratio)) %>%
+#   # dplyr::group_by(gene, Number.x) %>%
+#   # dplyr::add_count() %>%
+#   # dplyr::filter(n==2) %>%
+#   dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
+# 
+# pdata %>%
+#   dplyr::select(Number.x, gene, log2_conc) %>%
+#   tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+#   dplyr::mutate(class="iGBM") -> pdata
+# 
+# pdata2 %>%
+#   dplyr::select(Number.x, gene, log2_conc) %>%
+#   tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+#   dplyr::mutate(class="rGBM") -> pdata2
+# 
+# 
+# pdata <- rbindlist(list(pdata, pdata2))
+# 
+# pdata <- merge(pdata, patient_meta_data[, .(Number.x, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `Time-to-reccurence (days)`)], by="Number.x", all.x=T)
+# 
+# ## Pivot Table
+# pdata %>%
+#   dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, Sex, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`, `Time-to-reccurence (days)`) -> data_wide
+# # data_wide %>%
+# #   dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
+# 
+# ## Prediction
+# train_df <- data_wide
+# train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
+# train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
+# ## Make Valid Column Names 
+# colnames(train_df) <- make.names(colnames(train_df))
+# train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
+# colnames(train_df) <- make.names(colnames(train_df))
+# train_df <- as.data.frame(train_df)
+# pred_obj_1.6 <- run_predictions(train_df, pred_col_prob = "rGBM")
+# 
+# ## ROC Figure 
+# g <- ggplot(pred_obj_1.6$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
+#   geom_roc(n.cuts=0) + 
+#   scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+#   coord_equal() +
+#   style_roc(x="1 - Specificity", y="Sensitivity")
+# 
+# auc_tbl_1.6 <- as.data.table(calc_auc(g))
+# 
+# g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.6[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[1])
+# g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.6[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[2])
+# g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.6[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[3])
+# g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.6[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[4])
+# subfigure_1.6 <- g
+# 
+# pred_obj_1.6$roc_data %>%
+#   dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+#   dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+#   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.6_roc_data
+# 
+# ## Merged Plot
+# pred_obj_1.6$roc_data %>%
+#   dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+#   dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+#   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.6_roc_data
 
@@
 
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.png", plot = g, width = 11, height = 8.5, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.svg", plot = g, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.png", plot = g, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.svg", plot = g, width = 11, height = 8.5, dpi = 200)
 subfigure_1 <- g
@@
 
+<!-- # Latency Data -->
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
+latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
+latency_data <- fread(latency_data_file)
+# Clean names
+latency_data <- rstatix::make_clean_names(latency_data)
+latency_data %>%
+  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
+                SYNM.ng.ml = log2(SYNM.ng.ml+1),
+                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
+                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
+  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
+                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
+                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
+                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data
+
+fit <- survfit(Surv(Latency..days., status) ~ ASAH1, data = latency_data)
+print(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+surv_p1 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          )
+
+fit <- survfit(Surv(Latency..days., status) ~ SYNM, data = latency_data)
+print(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+surv_p2 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          )
+
+fit <- survfit(Surv(Latency..days., status) ~ GPNMB, data = latency_data)
+print(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+surv_p3 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          )
+
+fit <- survfit(Surv(Latency..days., status) ~ MMP9, data = latency_data)
+print(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+surv_p4 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          )
+
+survival_plot_grid <- ggarrange(surv_p1$plot, surv_p2$plot, surv_p3$plot, surv_p4$plot, ncol = 4)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.png", plot = survival_plot_grid, width = 18, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.svg", plot = survival_plot_grid, width = 18, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.png", plot = surv_p1$plot, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.svg", plot = surv_p1$plot, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.png", plot = surv_p2$plot, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.svg", plot = surv_p2$plot, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.png", plot = surv_p3$plot, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.svg", plot = surv_p3$plot, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.png", plot = surv_p4$plot, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.svg", plot = surv_p4$plot, width = 11, height = 7, dpi = 200)
+```
+
 \blandscape
@@
                                        splsda_ratio_avg_traj_fig, ncol=2, labels = c("A", "B")),
-                             ggarrange(ggplot() + 
-                                         annotate("text", x = 4, y = 25, size=8, label = "Kaplan-Meyer plots") + 
-                                         theme_void(), ncol=1, labels = c("C")), 
+                             ggarrange(survival_plot_grid, ncol=1, labels = c("C")), 
                              ggarrange(subfigure_1, 

### 2022-12-15 tables/elisa_plasma_proteomics/supp_table_2.tex

@@
 Ctrl vs post-
-surgery & Permutation Test & 0.7522500 & 0.1694500 & 0.1216000 & 0.2528000 \\ 
+surgery & Permutation Test & 0.7485000 & 0.1685000 & 0.1231000 & 0.2574000 \\ 
 Ctrl vs pre-
-surgery & Permutation Test & 0.0223500 & 0.9930500 & 0.0421500 & 0.0599500 \\ 
+surgery & Permutation Test & 0.0208000 & 0.9939500 & 0.0411000 & 0.0617000 \\ 
 pre-
@@
 \midrule
-Ctrl vs iGBM & Permutation Test & 0.1804000 & 0.4020500 & 0.0805000 & 0.0021000 \\ 
-Ctrl vs rGBM & Permutation Test & 0.2538500 & 0.2506500 & 0.4599500 & 0.1405500 \\ 
+Ctrl vs iGBM & Permutation Test & 0.1836500 & 0.3994500 & 0.0830000 & 0.0022500 \\ 
+Ctrl vs rGBM & Permutation Test & 0.2484000 & 0.2455000 & 0.4659500 & 0.1344500 \\ 
 iGBM vs rGBM & Paired Permutation Test & 0.1992188 & 0.6123047 & 0.1171875 & 0.1054688 \\ 

### 2022-12-15 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
[ADD] KM analysis for time to recurrence and combination of ASAH1 and SYNMIN

@@
           palette = "Dark2"
-          )
+          ) +
+  labs(x="Latency Time (Days)")
 
@@
           palette = "Dark2"
-          )
+          ) +
+  labs(x="Latency Time (Days)")
+
+fit <- survfit(Surv(Latency..days., status) ~ ASAH1 + SYNM, data = latency_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+surv_p5 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = F,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          ) +
+  labs(x="Latency Time (Days)")  +
+  guides(col=guide_legend(nrow=2,byrow=TRUE))
+
+surv_p5_with_ci <- ggsurvplot(fit,
+          pval = TRUE, conf.int = T,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          ) +
+  labs(x="Latency Time (Days)")  +
+  guides(col=guide_legend(nrow=2,byrow=TRUE))
 
@@
 ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.svg", plot = surv_p4$plot, width = 11, height = 7, dpi = 200)
+
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_.png", plot = surv_p5$plot, width = 11, height = 7, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_.svg", plot = surv_p5$plot, width = 11, height = 7, dpi = 200)
+# 
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_with_ci.png", plot = surv_p5_with_ci$plot, width = 11, height = 7, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_with_ci.svg", plot = surv_p5_with_ci$plot, width = 11, height = 7, dpi = 200)
+
+# survival model with time to recurrence; combined group of all 4 proteins
+ttr_surv_data <- merge(latency_data, patient_meta_data[, .(LAB_ID, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `Time-to-reccurence (days)`)], by.x="lab_id", by.y="LAB_ID", all.x=T)
+ttr_surv_data$`Time-to-reccurence (days)`[ ttr_surv_data$`Time-to-reccurence (days)`=="/" ] = ""
+ttr_surv_data$`Time-to-reccurence (days)` <- as.numeric(ttr_surv_data$`Time-to-reccurence (days)`)
+ttr_surv_data <- rstatix::make_clean_names(ttr_surv_data)
+
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p1 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p2 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ GPNMB, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p3 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ MMP9, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p4 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2"
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1 + SYNM, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p5 <- ggsurvplot(fit,
+          pval = TRUE, conf.int = F,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
+  guides(col=guide_legend(nrow=2,byrow=TRUE))
+
+ttr_surv_p5_with_ci <- ggsurvplot(fit,
+          pval = TRUE, conf.int = T,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
+  guides(col=guide_legend(nrow=2,byrow=TRUE))
 ```
@@
 (**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
-(**c**)  Kaplan-Meyer plots showing...
+(**c**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of longer survival.
 (**d**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
@@
 
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8.5, fig.width=11, dpi = 200}
+
+ttr_survival_plot_grid <- ggarrange(ttr_surv_p1$plot, ttr_surv_p2$plot, ttr_surv_p3$plot, ttr_surv_p4$plot, ncol = 4)
+
+supp_figure_4 <- ggarrange(ttr_survival_plot_grid, ggarrange(surv_p5$plot, ttr_surv_p5$plot, labels = c("B", "C")), labels = "A", nrow=2, ncol=1)
+supp_figure_4
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_4, width = 18, height = 18, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_4, width = 18, height = 18, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = ttr_survival_plot_grid, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = ttr_survival_plot_grid, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.png", fig_counter), plot = ttr_surv_p1$plot, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.svg", fig_counter), plot = ttr_surv_p1$plot, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.png", fig_counter), plot = ttr_surv_p2$plot, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.svg", fig_counter), plot = ttr_surv_p2$plot, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.png", fig_counter), plot = ttr_surv_p3$plot, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.svg", fig_counter), plot = ttr_surv_p3$plot, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.png", fig_counter), plot = ttr_surv_p4$plot, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.svg", fig_counter), plot = ttr_surv_p4$plot, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = surv_p5$plot, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = surv_p5$plot, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.png", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.svg", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
+```
+**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Survival and Time to Recurrence Analysis.** (**a**) Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of tumor recurrence occurring at a later date.
+(**b**) Kaplan-Meyer plots demonstrating the combination of both ASAH1 and SYNEMIN and their contribution of survival probability and time to recurrence.
+
+\elandscape
+
+\pagebreak
+
+\blandscape
+
 ```{r echo=FALSE, message=FALSE, warning=FALSE}

### 2022-12-15 tables/elisa_plasma_proteomics/supp_table_2.tex
[UPDATE} figures and tables

@@
 Ctrl vs post-
-surgery & Permutation Test & 0.7485000 & 0.1685000 & 0.1231000 & 0.2574000 \\ 
+surgery & Permutation Test & 0.7487500 & 0.1669500 & 0.1272000 & 0.2579500 \\ 
 Ctrl vs pre-
-surgery & Permutation Test & 0.0208000 & 0.9939500 & 0.0411000 & 0.0617000 \\ 
+surgery & Permutation Test & 0.0203000 & 0.9942000 & 0.0420000 & 0.0594500 \\ 
 pre-
@@
 \midrule
-Ctrl vs iGBM & Permutation Test & 0.1836500 & 0.3994500 & 0.0830000 & 0.0022500 \\ 
-Ctrl vs rGBM & Permutation Test & 0.2484000 & 0.2455000 & 0.4659500 & 0.1344500 \\ 
+Ctrl vs iGBM & Permutation Test & 0.1836000 & 0.3916500 & 0.0839000 & 0.0024000 \\ 
+Ctrl vs rGBM & Permutation Test & 0.2547000 & 0.2447000 & 0.4589000 & 0.1386000 \\ 
 iGBM vs rGBM & Paired Permutation Test & 0.1992188 & 0.6123047 & 0.1171875 & 0.1054688 \\ 

### 2023-01-03 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
[ADD] Update figures and redo ratio sPLS-DA analysis

@@
 pdata$`ttr_group` <- sapply(as.numeric(pdata$`Time-to-reccurence (days)`), ttr_categorizer)
-ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = colfunc(length(unique(pdata$`ttr_group`))), line.color = "ttr_group", line.size = 0.4) +
+ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
   # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
   # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "Initial manifestation", "Second manifestation")) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
   # geom_jitter(aes(label=Number.x)) +
@@
         legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
-         col=guide_legend(title="Time to Recurrence")) -> p1
+  guides(fill=guide_legend(title="", ncol=3),
+         col='none') -> p1_org
+
+p1_org + 
+  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
+  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(8, 18, 9, 12)),
+             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1_org
+
+# Exclude outlier patient number.x 54 
+ggpaired(data = pdata[Number.x!=54], id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
+  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
+  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
+  # geom_jitter(aes(label=Number.x)) +
+  # ggtitle("Initial vs Recurrence") +
+  labs(x="", y="log2( Concentration )") +
+  facet_wrap(~gene, scales = "free_y") +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "bottom", legend.box="vertical",) +
+  guides(fill=guide_legend(title="", ncol=3),
+         col='none') -> p1
 
@@
   stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
-  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(8, 18, 9, 12)),
+  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(7, 15, 8, 11)),
              aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1
@@
         legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="EDTA Blood\nCollection Type", ncol=1),
+  guides(fill=guide_legend(title="EDTA Blood\nCollection Type", ncol=3),
          col=guide_legend(title="Healthy Control Baseline")) -> p2
@@
 
+subfigure_6_org <- p1_org
+
 p1_avg_ttr
@@
 ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_e.svg", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
+
+```
+
+# Subsetting and hold out patient for Validating Ratios
+
+I still think there is value in usiing the ratios, so I performed the sPLS-DA on the ratio data again, but instead of using all the ratios and inverse ratios from all patients, I do the following:
+
+1. Randomly sample a single patient to use as a test patient, to make predictions on
+2. Randomly sample n random patients from remaining patients and compute iGBM/rGBM ratios
+3. For the rest of the patients I compute rGBM/iGBM ratios
+
+This results in the following data used in the sPLS-DA model
+
+**Training Data**
+
+| Patient ID | Class | ASAH1      | SYNMEIN    | GPNMB      | MMP-9      |
+|------------|-------|------------|------------|------------|------------|
+| 6          | iGBM  | -4.3219281 | 2.6629410  | 0.3162515  | 3.8886097  |
+| 10         | rGBM  | 2.2103599  | -3.8779037 | 2.5628539  | -3.1538943 |
+| 16         | rGBM  | 1.6956066  | 0.8506559  | 0.1659252  | -1.9814540 |
+| 23         | rGBM  | -0.0257631 | -3.6121574 | 1.6151451  | -1.7166787 |
+| 26         | iGBM  | 1.7744403  | 0.4268049  | 4.2786146  | -1.5475206 |
+| 29         | rGBM  | 3.3645724  | 0.4959143  | -0.6652013 | 0.5074684  |
+| 31         | iGBM  | -1.6599246 | -2.8570919 | -1.9497226 | -1.3115642 |
+| 34         | iGBM  | -0.1768778 | 0.7865737  | 2.1875247  | 1.1604031  |
+
+**Test Data**
+| Patient ID | Class | ASAH1      | SYNMEIN    | GPNMB      | MMP-9      |
+|------------|-------|------------|------------|------------|------------|
+| 35          | rGBM  | -0.02599521 | -4.115453  | -2.892362  | -1.679008  |
+
+This way we do not duplicate patient IDs (i.e.
+having iGBM/rGBM and rGBM/iGBM for the same patient) and we have a single patient we can make a prediction on to see where in the PLS-DA dimension they would land.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
+data_long2 <- data_long
+
+## Data wrangling ----
+### iGBM over rGBM ratio
+data_long2 %>% 
+  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::group_map(function(.d, ...){
+    .d <- as.data.table(.d)
+    .d %>% 
+      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+    # print(.d)
+    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+      .d_initial <- .d[manifestation_type=="Initial manifestation"]
+      .d_second <- .d[manifestation_type=="Second manifestation"]
+      tmp$ratio <- .d_initial$conc / .d_second$conc
+    } else if (nrow(.d)==2) {
+      .d_initial <- .d[collection_type=="E1"]
+      .d_second <- .d[collection_type=="E2"]
+      tmp$ratio <- .d_initial$conc / .d_second$conc
+    } else {
+      tmp$ratio <- NA #.d$conc
+    }
+    return(tmp)
+  }, .keep = TRUE) %>%
+  rbindlist() %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
+  dplyr::mutate(log2_conc = log2(ratio)) %>%
+  # dplyr::group_by(gene, Number.x) %>%
+  # dplyr::add_count() %>%
+  # dplyr::filter(n==2) %>%
+  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
+
+### rGBM over iGBM
+data_long2 %>% 
+  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::group_map(function(.d, ...){
+    .d <- as.data.table(.d)
+    .d %>% 
+      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+    # print(.d)
+    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+      .d_initial <- .d[manifestation_type=="Initial manifestation"]
+      .d_second <- .d[manifestation_type=="Second manifestation"]
+      tmp$ratio <- .d_second$conc / .d_initial$conc
+    } else if (nrow(.d)==2) {
+      .d_initial <- .d[collection_type=="E1"]
+      .d_second <- .d[collection_type=="E2"]
+      tmp$ratio <- .d_initial$conc / .d_second$conc
+    } else {
+      tmp$ratio <- NA #.d$conc
+    }
+    return(tmp)
+  }, .keep = TRUE) %>%
+  rbindlist() %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
+  dplyr::filter(grepl(";", agg_lab_id)) %>%
+  dplyr::mutate(log2_conc = log2(ratio)) %>%
+  # dplyr::group_by(gene, Number.x) %>%
+  # dplyr::add_count() %>%
+  # dplyr::filter(n==2) %>%
+  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
+
+### Prepare final table with ratios
+pdata %>%
+  dplyr::select(Number.x, gene, log2_conc) %>%
+  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+  dplyr::mutate(class="iGBM") -> pdata
+
+pdata2 %>%
+  dplyr::select(Number.x, gene, log2_conc) %>%
+  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+  dplyr::mutate(class="rGBM") -> pdata2
+
+pdata <- rbindlist(list(pdata, pdata2))
+
+pdata <- merge(pdata, meta_dt, by="Number.x")
+
+### Pivot Table to wide
+pdata %>%
+  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide
+
+## Remove non-paired patients
+data_wide %>%
+  dplyr::group_by(Number.x) %>%
+  dplyr::add_count() %>%
+  dplyr::filter(n==2) %>%
+  dplyr::ungroup() %>%
+  dplyr::select(-n) -> data_wide
+
+# Prepare final table for analysis
+train_df <- data_wide
+train_df <- na.omit(train_df)  
+train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
+# Make Valid Column Names 
+colnames(train_df) <- make.names(colnames(train_df))
+train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.table(train_df)
+train_df$group=0
+
+# Sample for Test data
+test_patients <- sample(unique(train_df$Number.x), 1)
+test_df <- train_df %>%
+  dplyr::filter(Number.x %in% test_patients) %>%
+  dplyr::filter(class=="rGBM")
+
+# Refine Training data
+train_df %>%
+  dplyr::filter(!(Number.x %in% test_patients)) -> train_df
+# Sample patients to select half patients for iGMB/rGBM and half patients for rGBM/iGBM
+igbm_train_patients <- sample(unique(train_df$Number.x), length(unique(train_df$Number.x))/2)
+rgbm_train_patients <- unique(train_df$Number.x)[ !unique(train_df$Number.x) %in% igbm_train_patients]
+
+train_df %>%
+  dplyr::filter((Number.x %in% igbm_train_patients & class=='iGBM') | (Number.x %in% rgbm_train_patients & class=='rGBM')) -> train_df
+
+## sPLS-DA of Ratios ----
+
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4))
+
+pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
+
+background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist", xlim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))), ylim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))))
+
+plsda_data <- cbind( patient_id=na.omit(data_wide %>% dplyr::filter(!(Number.x %in% test_patients)))  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+plsda_data$patient_id <- as.factor(plsda_data$patient_id)
+plsda_data %>%
+  dplyr::arrange(class) -> plsda_data
+plsda_data <- as.data.table(plsda_data)
+
+plsda_data %>%
+  dplyr::group_by(class) %>%
+  dplyr::add_count() %>%
+  dplyr::group_by(class, n) %>%
+  dplyr::summarise(avg_comp1=sum(comp1)/n,
+                   avg_comp2=sum(comp2)/n) %>%
+  dplyr::ungroup() %>%
+  unique() %>%
+  dplyr::mutate(patient_id="avg_all") %>%
+  as.data.table() -> average_comp
+
+slope <- (average_comp$avg_comp2[2] - average_comp$avg_comp2[1])/(average_comp$avg_comp1[2] - average_comp$avg_comp1[1])
+b = -0.25
+
+for(i in 1:length(background))
+{
+  if(!is.null(background[[i]]))
+    background[[i]]=data.frame(id=i,col=names(background)[i],
+                               background[[i]])
+}
+
+background = do.call(rbind,background)
+
+p <- ggplot() +
+  geom_polygon(data = background,aes(x=Var1, y=Var2,
+                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
+  scale_fill_manual(values=c("#388ECC", "#F68B33")) +
+  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
+  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
+  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
+  scale_shape_manual(values=c(15, 16, 17)) +
+  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
+                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
+  ) +
+  ggnewscale::new_scale_color() +
+  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+            arrow = arrow(type = "closed", ends='both',
+                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  ggnewscale::new_scale_color() +
+  geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
+  scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
+  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "right", legend.box="vertical",) +
+  guides(shape="none",
+         colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+         colour_new=guide_legend(title="Trajectory", ncol=1, byrow = T),
+         col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
+
+splsda_better_ratio_avg_traj_fig <- p
+splsda_better_ratio_avg_traj_fig
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+
+# p <- ggplot() +
+#   geom_polygon(data = background,aes(x=Var1, y=Var2,
+#                                      fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
+#   scale_fill_manual(values=c( "#388ECC", "#F68B33")) +
+#   # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
+#   # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
+#   geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
+#   scale_shape_manual(values=c(15, 16, 17)) +
+#   scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
+#                                                                expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
+#   ) +
+#   ggnewscale::new_scale_color() +
+#   geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
+#             arrow = arrow(type = "closed", ends='both',
+#                           length=unit(0.1, "inches")), size=1.5, alpha=0.8) +
+#   ggnewscale::new_scale_color() +
+#   geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
+#   scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
+#   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
+#   theme_classic() +
+#   theme(plot.title = element_text(hjust = 0.5, size=17),
+#         axis.title = element_text(size = 17),
+#         axis.text = element_text(size=16),
+#         strip.text = element_text(size=16),
+#         legend.title = element_text(size=16),
+#         legend.text = element_text(size=15),
+#         legend.position = "right", legend.box="vertical",) +
+#   guides(shape="none",
+#          colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+#          colour_new=guide_legend(title="Patient ID\nTrajectory", ncol=2),         
+#          col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
+# 
+# splsda_randomized_ratio_ind_traj_fig <- p
+# splsda_randomized_ratio_ind_traj_fig
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.png", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.svg", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
 
@@
 
+# Full
 master_roc_data <- rbindlist(list(`Ctrl vs iGBM`=pred_obj_1.1$roc_data, `Ctrl vs rGBM`=pred_obj_1.2$roc_data, `iGBM vs rGBM protein`=tmp_pred_obj_1.5_roc_data, `iGBM vs rGBM protein + clinical`=tmp_pred_obj_1.4_roc_data, `iGBM vs rGBM protein ratios`=tmp_pred_obj_1.3_roc_data), idcol = "Comparison")
@@
 master_auc_data <- rbindlist(list(`Ctrl vs iGBM`=auc_tbl_1.1, `Ctrl vs rGBM`=auc_tbl_1.2, `iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
+
+# Only iGMB and rGBM
+master_roc_data <- rbindlist(list(`iGBM vs rGBM protein`=tmp_pred_obj_1.5_roc_data, `iGBM vs rGBM protein + clinical`=tmp_pred_obj_1.4_roc_data, `iGBM vs rGBM protein ratios`=tmp_pred_obj_1.3_roc_data), idcol = "Comparison")
+
+master_auc_data <- rbindlist(list(`iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
 
@@
 
-g <- ggplot(master_roc_data[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Comparison, linetype=Comparison)) +
+g <- ggplot(master_roc_data[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Comparison), linetype='solid') +
+  geom_roc(n.cuts=0, size=1.5) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "Dark2")) +
+  # scale_linetype_manual(values=c("solid", "dotted"), labels=c("Elastic Logit")) + 
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity") +
+  annotation_custom(auc_grob, xmin=0.25, xmax=0.95, ymin=0.10, ymax=0.30) +
+  theme(axis.title = element_text(size=14),
+        axis.text = element_text(size=13),
+        legend.title = element_text(size=14),
+        legend.text = element_text(size=13))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.png", plot = g, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.svg", plot = g, width = 11, height = 8.5, dpi = 200)
+subfigure_1 <- g
+
+## Separate ROCs for controls
+
+master_roc_data <- rbindlist(list(`Ctrl vs iGBM`=pred_obj_1.1$roc_data, `Ctrl vs rGBM`=pred_obj_1.2$roc_data), idcol = "Comparison")
+
+master_auc_data <- rbindlist(list(`Ctrl vs iGBM`=auc_tbl_1.1, `Ctrl vs rGBM`=auc_tbl_1.2), idcol = "Comparison")
+
+master_auc_data[Model %in% c("Elastic Logit")] %>%
+  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
+  tidyr::pivot_wider(id_cols = "Comparison", names_from = Model, values_from = AUC) -> auc_annotation_tbl
+
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))))
+auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
+
+g <- ggplot(master_roc_data[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Comparison), linetype='solid') +
   geom_roc(n.cuts=0, size=1.5) + 
@@
 
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.png", plot = g, width = 11, height = 8.5, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_d.svg", plot = g, width = 11, height = 8.5, dpi = 200)
-subfigure_1 <- g
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_roc_controls.png", plot = g, width = 11, height = 8.5, dpi = 200)
+# ggsave("../../figures/elisa_plasma_proteomics/elisa_roc_controls.svg", plot = g, width = 11, height = 8.5, dpi = 200)
+roc_controls <- g
 ```
@@
 print(fit)
-
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$ASAH1)
 # Change color, linetype by strata, risk.table color by strata
 surv_p1 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high (n=%s)", n_data[1]), sprintf("ASAH1=low (n=%s)", n_data[2]))
           ) +
@@
 
+surv_p1 <- surv_p1$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13))
+
 fit <- survfit(Surv(Latency..days., status) ~ SYNM, data = latency_data)
 print(fit)
-
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$SYNM)
 # Change color, linetype by strata, risk.table color by strata
 surv_p2 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("SYNM=high (n=%s)", n_data[1]), sprintf("SYNM=low (n=%s)", n_data[2]))
           ) +
@@
 
+surv_p2 <- surv_p2$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13))
+
 fit <- survfit(Surv(Latency..days., status) ~ GPNMB, data = latency_data)
 print(fit)
-
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$GPNMB)
 # Change color, linetype by strata, risk.table color by strata
 surv_p3 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("GPNMB=high (n=%s)", n_data[1]), sprintf("GPNMB=low (n=%s)", n_data[2]))
           ) +
@@
 
+surv_p3 <- surv_p3$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13))
+
 fit <- survfit(Surv(Latency..days., status) ~ MMP9, data = latency_data)
 print(fit)
-
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$MMP9)
 # Change color, linetype by strata, risk.table color by strata
 surv_p4 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("MMP9=high (n=%s)", n_data[1]), sprintf("MMP9=low (n=%s)", n_data[2]))
           ) +
   labs(x="Latency Time (Days)")
+
+surv_p4 <- surv_p4$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13))
 
@@
 surv_p5 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = F,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = F,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
           ) +
   labs(x="Latency Time (Days)")  +
-  guides(col=guide_legend(nrow=2,byrow=TRUE))
+  guides(col=guide_legend(nrow=4,byrow=TRUE))
+
+surv_p5 <- surv_p5$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
 
 surv_p5_with_ci <- ggsurvplot(fit,
-          pval = TRUE, conf.int = T,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = T,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
           ) +
@@
   guides(col=guide_legend(nrow=2,byrow=TRUE))
+
+surv_p5_with_ci <- surv_p5_with_ci$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
 
@@
 print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$ASAH1)
 # Change color, linetype by strata, risk.table color by strata
 ttr_surv_p1 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
           ) +
@@
 
+ttr_surv_p1 <- ttr_surv_p1$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
+
 fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data)
 print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$SYNM)
 # Change color, linetype by strata, risk.table color by strata
 ttr_surv_p2 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
           ) +
@@
 
+ttr_surv_p2 <- ttr_surv_p2$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
+
 fit <- survfit(Surv(Time.to.reccurence..days., status) ~ GPNMB, data = ttr_surv_data)
 print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$GPNMB)
 # Change color, linetype by strata, risk.table color by strata
 ttr_surv_p3 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
           ) +
@@
 
+ttr_surv_p3 <- ttr_surv_p3$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
+
 fit <- survfit(Surv(Time.to.reccurence..days., status) ~ MMP9, data = ttr_surv_data)
 print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$MMP9)
 # Change color, linetype by strata, risk.table color by strata
 ttr_surv_p4 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2"
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
           ) +
   labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+ttr_surv_p4 <- ttr_surv_p4$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
 
@@
 ttr_surv_p5 <- ggsurvplot(fit,
-          pval = TRUE, conf.int = F,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = F,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
           ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high\n(n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
           ) +
@@
 
+ttr_surv_p5 <- ttr_surv_p5$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
+
 ttr_surv_p5_with_ci <- ggsurvplot(fit,
-          pval = TRUE, conf.int = T,
-          risk.table = TRUE, # Add risk table
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = T,
+          risk.table = FALSE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
@@
   guides(col=guide_legend(nrow=2,byrow=TRUE))
+
+ttr_surv_p5_with_ci <- ttr_surv_p5_with_ci$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=13),
+        legend.position = "bottom")
 ```
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
+# mataster_figure <- ggarrange(ggarrange(subfigure_6, 
+#                                        splsda_ratio_avg_traj_fig, ncol=2, labels = c("A", "B")),
+#                              ggarrange(survival_plot_grid, ncol=1, labels = c("C")), 
+#                              ggarrange(subfigure_1, 
+#                                        p_ratio_ttr, ncol=2, labels = c("D", "E")),
+#                              ncol = 1, nrow=3)
+
 mataster_figure <- ggarrange(ggarrange(subfigure_6, 
-                                       splsda_ratio_avg_traj_fig, ncol=2, labels = c("A", "B")),
-                             ggarrange(survival_plot_grid, ncol=1, labels = c("C")), 
+                                       ggarrange(ttr_surv_p1, ttr_surv_p2, ncol=2), ncol=2, labels = c("A", "B")),
                              ggarrange(subfigure_1, 
-                                       p_ratio_ttr, ncol=2, labels = c("D", "E")),
-                             ncol = 1, nrow=3)
+                                       splsda_better_ratio_avg_traj_fig, ncol=2, labels = c("C", "D")),
+                             ncol = 1, nrow=2)
 mataster_figure
@@
 
-**Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA).
+<!-- **Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA).
 (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM populations.
+Note, a patient was omitted due to being an outlier, see Supp Figure X for full set of patients.
 (**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
@@
 However, we can still state given a specific patients ratio, we can predict that patients time to recurrence.
+-->
+**Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA).
+(**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM populations.
+Note, a patient was omitted due to being an outlier, see Supp Figure X for full set of patients.
+(**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
+(**c**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of longer survival.
+(**d**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
+(**e**) The potential of recurrent GBM / initial GBM protein ratios for predicting time to recurrence.
+Note, each point represents a single patient, so we may not be able to directly state a specific ratio can be used to infer to time to recurrence.
+However, we can still state given a specific patients ratio, we can predict that patients time to recurrence.
 
@@
 
-supp_figure_0 <- ggarrange(supp_figure_pre_v_post, ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), ncol=2, nrow=1, labels=c("A", "B"))
+supp_figure_0 <- ggarrange(ggarrange(subfigure_6_org, supp_figure_pre_v_post, ncol=1, nrow=2, labels=c("A", "B")), ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), ncol=2, nrow=1, labels=c("", "C"))
 supp_figure_0
@@
 
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_6_org, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_6_org, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
 ```
-**Supplemental Figure** `r knitr::asis_output(md(sprintf("**%s**", fig_counter)))` `r fig_counter <- fig_counter + 1`**: Distribution of Pre- vs Post-Surgery Plasma Proteomics and Ratios of Initial and Recurrent GBM Plasma Proteomics.**  (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, pre-surgery and post-surgery populations.
-(**b**) Demonstrates the log transformed $\frac{initial \quad GBM}{recurrent \quad GBM}$ ratio (top) as well as the inverse ratio (bottom) for the potential use of identifying a patients state to recurrence using these ratios.
+**Supplemental Figure** `r knitr::asis_output(md(sprintf("**%s**", fig_counter)))` `r fig_counter <- fig_counter + 1`**: Distribution of Pre- vs Post-Surgery Plasma Proteomics and Ratios of Initial and Recurrent GBM Plasma Proteomics.**  (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM population for all patients.
+(**b**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, pre-surgery and post-surgery populations.
+(**c**) Demonstrates the log transformed $\frac{initial \quad GBM}{recurrent \quad GBM}$ ratio (top) as well as the inverse ratio (bottom) for the potential use of identifying a patients state to recurrence using these ratios.
 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
-supp_figure_1 <- ggarrange(subfigure_2, subfigure_3, subfigure_5, subfigure_4, ncol=2, nrow=2, labels=c("A", "B", "C", "D"))
+supp_figure_1 <- ggarrange(ggarrange(roc_controls, subfigure_2, subfigure_3, ncol=3, nrow=1, labels=c("A", "B", "C")), ggarrange(subfigure_5, subfigure_4, labels=c("D", "E")), ncol=1, nrow=2)
 
@@
 
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.png", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.svg", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = roc_controls, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = roc_controls, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.png", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.svg", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_e.png", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_e.svg", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
 ```
-**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Variation in Predictive Performance using Different Models.** Testing parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms on different prediction tasks.
-(**a**) Most models have AUCs greater than 60% when predicting control vs initial GBM.
-(**b**) Most models have comparable AUCs, except for k-nearest neighbours when predicting for initial GBM vs recurrent GBM.
-(**c**) Including clinical data for *Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection* reduces predictive performance for all models.
-(**d**) All models, except random forest, have AUCs above 80% when predicting initial GBM vs recurrent GBM using protein ratios.
+**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Variation in Predictive Performance using Different Models.** Separation on Controls and testing parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms on different prediction tasks.
+(**a**)  The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating controls (health patients) from patients with GBM using an elastic logistic regression.
+(**b**) Most models have AUCs greater than 60% when predicting control vs initial GBM.
+(**c**) Most models have comparable AUCs, except for k-nearest neighbours when predicting for initial GBM vs recurrent GBM.
+(**d**) Including clinical data for *Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection* reduces predictive performance for all models.
+(**e**) All models, except random forest, have AUCs above 80% when predicting initial GBM vs recurrent GBM using protein ratios.
 

### 2023-01-04 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
[ADD] Evaluating sPLD-DA and ratios

@@
 installIfNeeded("mixOmics", "BIO")
+# parallilization
+installIfNeeded("future")
 ```
@@
 }
+
+#' Calculate the ratio of iGBM to rGBM and rGBM to iGBM for a given data frame
+#'
+#' @param data_long2 A data frame with columns "gene", "Number.x", "patient_state", "Age at surgery (years)", "Sex", "Tumor localization", "Type of resection", "MGMT promotor methylation", "p53 accumulation", "Ki67-Li", "EGFR vIII", "Treatment", "Time-to-reccurence (days)", "Latency (days)", and "conc".
+#' @return A data frame with the calculated ratios.
+get_ratio_data <- function(data_long2){
+  ## Data wrangling ----
+  ### iGBM over rGBM ratio
+  data_long2 %>% 
+    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+    dplyr::group_by(gene, Number.x) %>%
+    dplyr::group_map(function(.d, ...){
+      .d <- as.data.table(.d)
+      .d %>% 
+        dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+      # print(.d)
+      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+        .d_initial <- .d[manifestation_type=="Initial manifestation"]
+        .d_second <- .d[manifestation_type=="Second manifestation"]
+        tmp$ratio <- .d_initial$conc / .d_second$conc
+      } else if (nrow(.d)==2) {
+        .d_initial <- .d[collection_type=="E1"]
+        .d_second <- .d[collection_type=="E2"]
+        tmp$ratio <- .d_initial$conc / .d_second$conc
+      } else {
+        tmp$ratio <- NA #.d$conc
+      }
+      return(tmp)
+    }, .keep = TRUE) %>%
+    rbindlist() %>%
+    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
+    dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
+    dplyr::mutate(log2_conc = log2(ratio)) %>%
+    # dplyr::group_by(gene, Number.x) %>%
+    # dplyr::add_count() %>%
+    # dplyr::filter(n==2) %>%
+    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
+
+  ### rGBM over iGBM
+  data_long2 %>% 
+    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+    dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
+    dplyr::group_by(gene, Number.x) %>%
+    dplyr::group_map(function(.d, ...){
+      .d <- as.data.table(.d)
+      .d %>% 
+        dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+      # print(.d)
+      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+        .d_initial <- .d[manifestation_type=="Initial manifestation"]
+        .d_second <- .d[manifestation_type=="Second manifestation"]
+        tmp$ratio <- .d_second$conc / .d_initial$conc
+      } else if (nrow(.d)==2) {
+        .d_initial <- .d[collection_type=="E1"]
+        .d_second <- .d[collection_type=="E2"]
+        tmp$ratio <- .d_initial$conc / .d_second$conc
+      } else {
+        tmp$ratio <- NA #.d$conc
+      }
+      return(tmp)
+    }, .keep = TRUE) %>%
+    rbindlist() %>%
+    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+    dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
+    dplyr::filter(grepl(";", agg_lab_id)) %>%
+    dplyr::mutate(log2_conc = log2(ratio)) %>%
+    # dplyr::group_by(gene, Number.x) %>%
+    # dplyr::add_count() %>%
+    # dplyr::filter(n==2) %>%
+    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
+
+  ### Prepare final table with ratios
+  pdata %>%
+    dplyr::select(Number.x, gene, log2_conc) %>%
+    tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+    dplyr::mutate(class="iGBM") -> pdata
+
+  pdata2 %>%
+    dplyr::select(Number.x, gene, log2_conc) %>%
+    tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+    dplyr::mutate(class="rGBM") -> pdata2
+
+  pdata <- rbindlist(list(pdata, pdata2))
+
+  pdata <- merge(pdata, meta_dt, by="Number.x")
+
+  ### Pivot Table to wide
+  pdata %>%
+    dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide
+
+  ## Remove non-paired patients
+  data_wide %>%
+    dplyr::group_by(Number.x) %>%
+    dplyr::add_count() %>%
+    dplyr::filter(n==2) %>%
+    dplyr::ungroup() %>%
+    dplyr::select(-n) -> data_wide
+
+  # Prepare final table for analysis
+  train_df <- data_wide
+  train_df <- na.omit(train_df)  
+  train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
+  # Make Valid Column Names 
+  colnames(train_df) <- make.names(colnames(train_df))
+  train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
+  colnames(train_df) <- make.names(colnames(train_df))
+  train_df <- as.data.table(train_df)
+  train_df$group=0
+
+  return(train_df)
+}
+
+#' Generate train and test datasets for leave-one-patient-out cross-validation
+#'
+#' This function takes in a dataset and returns a list containing train and test datasets
+#' for leave-one-patient-out cross-validation. The test dataset consists of a single patient
+#' with the "rGBM" class. If the `test_patient_id` parameter is not `NULL`, the test dataset
+#' is subsetted using the provided patient ID. If the `test_patient_id` parameter is `NULL`,
+#' a random patient with the "rGBM" class is chosen for the test dataset. The train dataset
+#' consists of all other patients. If the `unique_random_train_patient_per_class` parameter is 
+#' `TRUE`, the train dataset is further subsetted to include a unique patient for each "iGBM"
+#' and "rGBM" class. The `test` parameter can be used to set the seed for the random number 
+#' generator when sampling patients for the test and train datasets. This is useful for testing 
+#' purposes.
+#'
+#' @param data A data frame containing the "Number.x" and "class" columns.
+#' @param test_patient_id An optional patient ID to use for the test dataset. If not provided,
+#' a random patient with the "rGBM" class is chosen.
+#' @param unique_random_train_patient_per_class A logical value indicating whether to include
+#' a unique patient for each "iGBM" and "rGBM" class in the train dataset.
+#' @param test A logical value indicating whether to set the seed for the random number
+#' generator for testing purposes.
+#' @return A list containing the train and test datasets.
+#' @export
+#' @examples
+#' get_train_test_for_loocv(data)
+get_train_test_for_loocv <- function(data, test_patient_id=NULL, unique_random_train_patient_per_class = TRUE, test = FALSE){
+  # Set seed for testing
+  if (test){
+    set.seed(2023)
+  }
+
+  if (is.null(test_patient_id)){
+    # Sample single sample rGBM for Test data
+    test_patients <- sample(unique(data$Number.x), 1)
+    test_df <- subset(data, Number.x %in% test_patients & class == "rGBM")
+  } else {
+    # Subset for test data using provided test_patient_id
+    test_df <- subset(data, Number.x %in% test_patient_id & class == "rGBM")
+    test_patients <- test_patient_id
+  }
+
+  # Subset data to remove sample in test data
+  data <- subset(data, !(Number.x %in% test_patients))
+
+  if (unique_random_train_patient_per_class) {
+    # Sample a unique patient for each iGBM and rGBM class
+    igbm_train_patients <- sample(unique(data$Number.x[data$class == "iGBM"]), length(unique(data$Number.x))/2)
+    rgbm_train_patients <- setdiff(data$Number.x, igbm_train_patients)
+
+    # Filter data to obtain training data with a unique patient for each iGBM and rGBM class
+    train_df <- subset(data, (Number.x %in% igbm_train_patients & class == "iGBM") | 
+                         (Number.x %in% rgbm_train_patients & class == "rGBM"))
+  } else {
+    train_df <- data
+  }
+
+  return(list(train_df = train_df, test_df = test_df))
+}
+
+#' Calculate leave-one-patient-out cross-validation performance for splsda model
+#'
+#' This function takes in a dataset and a vector of patient IDs and returns a proportion
+#' of correct predictions using a leave-one-patient-out cross-validation approach. A splsda
+#' model is fit using the train data and used to make predictions on the test data. The
+#' `unique_random_train_patient_per_class` parameter can be used to specify whether to 
+#' include a unique patient for each "iGBM" and "rGBM" class in the train dataset. The
+#' function returns a proportion of correct predictions.
+#'
+#' @param ratio_data A data frame containing the ratio data and class labels.
+#' @param patient_ids A vector of patient IDs to use for the leave-one-patient-out
+#' cross-validation.
+#' @param unique_random_train_patient_per_class A logical value indicating whether to include
+#' a unique patient for each "iGBM" and "rGBM" class in the train dataset.
+#' @return A proportion of correct predictions.
+#' @examples
+#' splsda_loocv_perf(ratio_data, patient_ids)
+splsda_loocv_perf <- function(ratio_data, patient_ids, unique_random_train_patient_per_class=TRUE){
+  # Vector to store accuracy of LOOCV
+  acc_vec <- numeric(length(patient_ids))
+  i <- 1
+  for (test_patient_id in patient_ids){
+    # Split data into train and testing
+    train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=test_patient_id, unique_random_train_patient_per_class=unique_random_train_patient_per_class)
+    train_df <- train_test$train_df
+    test_df <- train_test$test_df
+    # Perform splsda
+    MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+    # Make prediction
+    pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
+    # evaluate the prediction accuracy for the first two components
+    predict.comp2 <- pred_test$class$mahalanobis.dist[,2]
+    # compute confusion matrix
+    conf_mat <- table(factor(predict.comp2, levels = levels(train_df$class)), test_df$class)
+    # print(conf_mat)
+    # compute accuracy
+    acc <- (conf_mat[1,1] + conf_mat[2,2])/(conf_mat[1,1] + conf_mat[2,2] + conf_mat[1,2] + conf_mat[2,1])
+    # message(sprintf("INFO: Test Patient %s - predicted class = %s - model accuracy = %s", test_patient_id, predict.comp2, acc))
+    # Store accuracy
+    acc_vec[i] <- acc
+    # update index
+    i <- i + 1
+  }
+  # Compute proportion of correct predictions
+  mean(acc_vec==1)
+}
+
 ```
@@
 tmp_pdata$Time.to.reccurence..days. <- factor(tmp_pdata$Time.to.reccurence..days., levels=c("early", "middle", "late"))
-ggplot(data = tmp_pdata, aes(x=Time.to.reccurence..days., y=log2_conc, col=Time.to.reccurence..days.)) +
-  geom_point() +
-  labs(x="", y="log2( Concentration )") +
-  facet_wrap(~gene, scales = "free_y") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
-         col=guide_legend(title="Time to Recurrence"))
+# ggplot(data = tmp_pdata, aes(x=Time.to.reccurence..days., y=log2_conc, col=Time.to.reccurence..days.)) +
+#   geom_point() +
+#   labs(x="", y="log2( Concentration )") +
+#   facet_wrap(~gene, scales = "free_y") +
+#   theme_classic() +
+#   theme(plot.title = element_text(hjust = 0.5, size=17),
+#         axis.title = element_text(size = 17),
+#         axis.text = element_text(size=16),
+#         strip.text = element_text(size=16),
+#         legend.title = element_text(size=16),
+#         legend.text = element_text(size=15),
+#         legend.position = "bottom", legend.box="vertical",) +
+#   guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
+#          col=guide_legend(title="Time to Recurrence"))
 
@@
 subfigure_6 <- p1
-subfigure_6
+# subfigure_6
 ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.png", plot = subfigure_6, width = 21, height = 11, dpi = 200)
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
+```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
 # Dimensionality Reduction ----
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
+```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
 ## Data wrangling ----
@@
   geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
-            arrow = arrow(type = "closed",
+            arrow = arrow(type = "closed", ends='both',
                           length=unit(0.1, "inches")), size=2, alpha=0.8) +
@@
   geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
-            arrow = arrow(type = "closed",
+            arrow = arrow(type = "closed", ends='both',
                           length=unit(0.1, "inches")), size=1.5, alpha=0.8) +
@@
 **Test Data**
+
 | Patient ID | Class | ASAH1      | SYNMEIN    | GPNMB      | MMP-9      |
@@
 ## Data wrangling ----
-### iGBM over rGBM ratio
-data_long2 %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- NA #.d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
-
-### rGBM over iGBM
-data_long2 %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_second$conc / .d_initial$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- NA #.d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
-  dplyr::filter(grepl(";", agg_lab_id)) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
-
-### Prepare final table with ratios
-pdata %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="iGBM") -> pdata
-
-pdata2 %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="rGBM") -> pdata2
-
-pdata <- rbindlist(list(pdata, pdata2))
-
-pdata <- merge(pdata, meta_dt, by="Number.x")
-
-### Pivot Table to wide
-pdata %>%
-  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide
-
-## Remove non-paired patients
-data_wide %>%
-  dplyr::group_by(Number.x) %>%
-  dplyr::add_count() %>%
-  dplyr::filter(n==2) %>%
-  dplyr::ungroup() %>%
-  dplyr::select(-n) -> data_wide
-
-# Prepare final table for analysis
-train_df <- data_wide
-train_df <- na.omit(train_df)  
-train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
-# Make Valid Column Names 
-colnames(train_df) <- make.names(colnames(train_df))
-train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.table(train_df)
-train_df$group=0
-
-# Sample for Test data
-test_patients <- sample(unique(train_df$Number.x), 1)
-test_df <- train_df %>%
-  dplyr::filter(Number.x %in% test_patients) %>%
-  dplyr::filter(class=="rGBM")
-
-# Refine Training data
-train_df %>%
-  dplyr::filter(!(Number.x %in% test_patients)) -> train_df
-# Sample patients to select half patients for iGMB/rGBM and half patients for rGBM/iGBM
-igbm_train_patients <- sample(unique(train_df$Number.x), length(unique(train_df$Number.x))/2)
-rgbm_train_patients <- unique(train_df$Number.x)[ !unique(train_df$Number.x) %in% igbm_train_patients]
-
-train_df %>%
-  dplyr::filter((Number.x %in% igbm_train_patients & class=='iGBM') | (Number.x %in% rgbm_train_patients & class=='rGBM')) -> train_df
+# Get ratio data
+message("Get Ratio Data")
+ratio_data <- get_ratio_data(data_long2)
+message("Get Test Train Data")
+train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=NULL, unique_random_train_patient_per_class=T)
+train_df <- train_test$train_df
+test_df <- train_test$test_df
+test_patients <- test_df$Number.x
 
 ## sPLS-DA of Ratios ----
-
+message("Perform sPLS-DA")
 MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4))
@@
 pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
-
+message("Perform background predictions")
 background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist", xlim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))), ylim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))))
-
-plsda_data <- cbind( patient_id=na.omit(data_wide %>% dplyr::filter(!(Number.x %in% test_patients)))  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+message("Prepare data for plotting")
+plsda_data <- cbind( patient_id=train_df  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
 plsda_data$patient_id <- as.factor(plsda_data$patient_id)
@@
 
+message("Gnerate sPLS-DA Plot")
 p <- ggplot() +
@@
 splsda_better_ratio_avg_traj_fig
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-
-# p <- ggplot() +
-#   geom_polygon(data = background,aes(x=Var1, y=Var2,
-#                                      fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
-#   scale_fill_manual(values=c( "#388ECC", "#F68B33")) +
-#   # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
-#   # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
-#   geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
-#   scale_shape_manual(values=c(15, 16, 17)) +
-#   scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-#                                                                expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-#   ) +
-#   ggnewscale::new_scale_color() +
-#   geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
-#             arrow = arrow(type = "closed", ends='both',
-#                           length=unit(0.1, "inches")), size=1.5, alpha=0.8) +
-#   ggnewscale::new_scale_color() +
-#   geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
-#   scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
-#   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
-#   theme_classic() +
-#   theme(plot.title = element_text(hjust = 0.5, size=17),
-#         axis.title = element_text(size = 17),
-#         axis.text = element_text(size=16),
-#         strip.text = element_text(size=16),
-#         legend.title = element_text(size=16),
-#         legend.text = element_text(size=15),
-#         legend.position = "right", legend.box="vertical",) +
-#   guides(shape="none",
-#          colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
-#          colour_new=guide_legend(title="Patient ID\nTrajectory", ncol=2),         
-#          col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
-# 
-# splsda_randomized_ratio_ind_traj_fig <- p
-# splsda_randomized_ratio_ind_traj_fig
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.png", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.svg", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-
+ggsave("../../figures/elisa_plasma_proteomics/elisa_splada_ratio_random.png", plot = splsda_better_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_splada_ratio_random.svg", plot = splsda_better_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+
+```
+
+# Comparison of Using All patients or Subsetting for Unique Patients per Class
+
+Lets compare to see if it makes a difference whether we use all patients for both iGBM/rGBM and rGBM/iGBM classes vs only using a unique set of patients for iGBM/rGBM and rGBM/iGBM (i.e.
+no duplicate patient in each class)
+
+1. I perform a leave one out cross validation, leaving one patient out and using the rest of the data for training. 
+2. The left out patient gets using for assessing model performance accuracy. 
+3. The mean accuracy of each left out sample is computed
+4. Steps 1-3 is repeated 100 times to assess the variation in mean accuracy of the LOOCV
+5. 1-4 is performed for both cases of using all patients and using unique subset patients
+
+```{r echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}
+# Get ratio data
+ratio_data <- get_ratio_data(data_long2)
+# Get unique set of patient ids
+patient_ids <- unique(ratio_data$Number.x)
+
+plan(multisession(workers = 8))
+
+# Compute mean accuracy when using unique patients per class for 100 repeats
+mean_acc_rep_unique <- future.apply::future_replicate(100, splsda_loocv_perf(ratio_data, patient_ids, unique_random_train_patient_per_class=TRUE))
+# hist(mean_acc_rep_unique)
+message("Unique Patient Class Split: 95% Confidence Interval:")
+print(c(quantile(mean_acc_rep_unique, .025), quantile(mean_acc_rep_unique, .975)))
+
+# Compute mean accuracy when using all patients per class for 100 repeats
+mean_acc_rep_all <- future.apply::future_replicate(100, splsda_loocv_perf(ratio_data, patient_ids, unique_random_train_patient_per_class=FALSE))
+# hist(mean_acc_rep_all)
+message("All Patients per Class: 95% Confidence Interval:")
+print(c(quantile(mean_acc_rep_all, .025), quantile(mean_acc_rep_all, .975)))
+
+plan(sequential)
+
+df_plot <- rbind(data.frame(method="Unique Patient Class Split", mean_acc=mean_acc_rep_unique), data.frame(method="All Patients per Class", mean_acc=mean_acc_rep_all))
+df_plot$method <- factor(df_plot$method, levels = c("All Patients per Class", "Unique Patient Class Split"))
+
+ggplot(df_plot, aes(x=mean_acc, fill=method)) +
+  geom_histogram(binwidth = 0.15, colour = "black") +
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
+  scale_fill_brewer(palette = "Dark2") +
+  stat_bin(binwidth = 0.15, aes(label=..count..), vjust=-0.5, geom = "text") +
+  facet_wrap(~method) +
+  labs(x="Mean Accuracy (LOOCV)", y="Count") +
+  ggtitle("Mean Accuracy of LOOCV for 100 Repeats") +
+  theme_classic() + 
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15)) 
 ```
@@
 
-table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))))
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 13)
 auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
@@
   annotation_custom(auc_grob, xmin=0.25, xmax=0.95, ymin=0.10, ymax=0.30) +
-  theme(axis.title = element_text(size=14),
-        axis.text = element_text(size=13),
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
         legend.title = element_text(size=14),
-        legend.text = element_text(size=13))
+        legend.text = element_text(size=13)) +
+  guides(col="none")
 
@@
 print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$ASAH1)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$ASAH1)
 # Change color, linetype by strata, risk.table color by strata
@@
           legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high (n=%s)", n_data[1]), sprintf("ASAH1=low (n=%s)", n_data[2]))
+          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
           ) +
@@
         axis.text = element_text(size=15),
-        legend.text = element_text(size=13))
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'))
 
@@
 print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$SYNM)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$SYNM)
 # Change color, linetype by strata, risk.table color by strata
@@
           legend.title = "",
-          legend.labs = c(sprintf("SYNM=high (n=%s)", n_data[1]), sprintf("SYNM=low (n=%s)", n_data[2]))
+          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
           ) +
@@
 print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$GPNMB)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$GPNMB)
 # Change color, linetype by strata, risk.table color by strata
@@
           legend.title = "",
-          legend.labs = c(sprintf("GPNMB=high (n=%s)", n_data[1]), sprintf("GPNMB=low (n=%s)", n_data[2]))
+          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
           ) +
@@
 print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Latency..days.)]$MMP9)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$MMP9)
 # Change color, linetype by strata, risk.table color by strata
@@
           legend.title = "",
-          legend.labs = c(sprintf("MMP9=high (n=%s)", n_data[1]), sprintf("MMP9=low (n=%s)", n_data[2]))
+          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
           ) +
@@
         axis.text = element_text(size=15),
-        legend.text = element_text(size=13),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
         legend.position = "bottom")
@@
         axis.text = element_text(size=15),
-        legend.text = element_text(size=13),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
         legend.position = "bottom")
 
-survival_plot_grid <- ggarrange(surv_p1$plot, surv_p2$plot, surv_p3$plot, surv_p4$plot, ncol = 4)
+survival_plot_grid <- ggarrange(surv_p1, surv_p2, surv_p3, surv_p4, ncol = 4)
 ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.png", plot = survival_plot_grid, width = 18, height = 7, dpi = 200)
@@
 
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.png", plot = surv_p1$plot, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.svg", plot = surv_p1$plot, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.png", plot = surv_p2$plot, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.svg", plot = surv_p2$plot, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.png", plot = surv_p3$plot, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.svg", plot = surv_p3$plot, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.png", plot = surv_p4$plot, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.svg", plot = surv_p4$plot, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.png", plot = surv_p1, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.svg", plot = surv_p1, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.png", plot = surv_p2, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.svg", plot = surv_p2, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.png", plot = surv_p3, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.svg", plot = surv_p3, width = 11, height = 7, dpi = 200)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.png", plot = surv_p4, width = 11, height = 7, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.svg", plot = surv_p4, width = 11, height = 7, dpi = 200)
 
@@
           legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high, SYNM=high\n(n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
           ) +
   labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
-  guides(col=guide_legend(nrow=2,byrow=TRUE))
+  guides(col=guide_legend(nrow=4,byrow=TRUE))
 
@@
 Note, a patient was omitted due to being an outlier, see Supp Figure X for full set of patients.
-(**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
-(**c**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of longer survival.
-(**d**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
-(**e**) The potential of recurrent GBM / initial GBM protein ratios for predicting time to recurrence.
-Note, each point represents a single patient, so we may not be able to directly state a specific ratio can be used to infer to time to recurrence.
-However, we can still state given a specific patients ratio, we can predict that patients time to recurrence.
-
-> **Note:** I'm not entirely sure about plot e, and if my thinking/interpretation is right, but let me know what you think.
-
-> **Note:** I excluded the pre-surgery vs post-surgery distribution boxplots, because I didn't do any prediction analyses on that set.
-We can still include it in the supplement if we thing there is anything interesting there to show?
+(**b**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of tumor recurrence occurring at a later date.
+(**c**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
+(**d**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
+One patient was randomly left out as a test patient (black + point), 4 samples were randomly sampled to represent $\frac{initial \quad GBM}{recurrent \quad GBM}$ (blue points) and remaining 4 samples were used to represent $\frac{recurrent \quad GBM}{initial \quad GBM}$.
 
@@
 
-ttr_survival_plot_grid <- ggarrange(ttr_surv_p1$plot, ttr_surv_p2$plot, ttr_surv_p3$plot, ttr_surv_p4$plot, ncol = 4)
-
-supp_figure_4 <- ggarrange(ttr_survival_plot_grid, ggarrange(surv_p5$plot, ttr_surv_p5$plot, labels = c("B", "C")), labels = "A", nrow=2, ncol=1)
+ttr_survival_plot_grid <- ggarrange(surv_p1, surv_p2, surv_p3, surv_p4, ncol = 4, labels="A")
+bottom_plots <- ggarrange(ggarrange(ttr_surv_p3, ttr_surv_p4, ncol=2), ggarrange(surv_p5, ttr_surv_p5, ncol=2), labels=c("B", "C"))
+
+supp_figure_4 <- ggarrange(ttr_survival_plot_grid, bottom_plots, nrow=2, ncol=1)
 supp_figure_4
 
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_4, width = 18, height = 18, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_4, width = 18, height = 18, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_4, width = 25, height = 15, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_4, width = 25, height = 15, dpi = 200)
 
@@
 
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.png", fig_counter), plot = ttr_surv_p1$plot, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.svg", fig_counter), plot = ttr_surv_p1$plot, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.png", fig_counter), plot = ttr_surv_p2$plot, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.svg", fig_counter), plot = ttr_surv_p2$plot, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.png", fig_counter), plot = ttr_surv_p3$plot, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.svg", fig_counter), plot = ttr_surv_p3$plot, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.png", fig_counter), plot = ttr_surv_p4$plot, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.svg", fig_counter), plot = ttr_surv_p4$plot, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = surv_p5$plot, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = surv_p5$plot, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.png", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.svg", fig_counter), plot = ttr_surv_p5$plot, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.png", fig_counter), plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.svg", fig_counter), plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.png", fig_counter), plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.svg", fig_counter), plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.png", fig_counter), plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.svg", fig_counter), plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.png", fig_counter), plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.svg", fig_counter), plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = surv_p5, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = surv_p5, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
+
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.png", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
+ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.svg", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
 ```

### 2023-01-04 tables/elisa_plasma_proteomics/supp_table_2.tex

@@
 Ctrl vs post-
-surgery & Permutation Test & 0.7487500 & 0.1669500 & 0.1272000 & 0.2579500 \\ 
+surgery & Permutation Test & 0.7476000 & 0.1693500 & 0.1244000 & 0.2567000 \\ 
 Ctrl vs pre-
-surgery & Permutation Test & 0.0203000 & 0.9942000 & 0.0420000 & 0.0594500 \\ 
+surgery & Permutation Test & 0.0195000 & 0.9946500 & 0.0436500 & 0.0625000 \\ 
 pre-
@@
 \midrule
-Ctrl vs iGBM & Permutation Test & 0.1836000 & 0.3916500 & 0.0839000 & 0.0024000 \\ 
-Ctrl vs rGBM & Permutation Test & 0.2547000 & 0.2447000 & 0.4589000 & 0.1386000 \\ 
+Ctrl vs iGBM & Permutation Test & 0.1880500 & 0.4017500 & 0.0797000 & 0.0023000 \\ 
+Ctrl vs rGBM & Permutation Test & 0.2551500 & 0.2481500 & 0.4577500 & 0.1374500 \\ 
 iGBM vs rGBM & Paired Permutation Test & 0.1992188 & 0.6123047 & 0.1171875 & 0.1054688 \\ 

### 2023-02-01 results/elisa_plasma_proteomics/manuscript_elisa_figure_analysis.Rmd
[RENAME] notebook to analysis, since more exploratory analyses were included

--- 2023-01-04 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
@@
 ```{r message=FALSE, warning=FALSE, include=FALSE}
+cs <- function(.x) {    # center/scale vector
+  print(.x)
+  out <- .x - mean(.x)  # center
+  out / sd(out)         # scale
+}
+
+median_centre_data <- function(train_df){
+  values_to = "measure_col"
+  measure_col="measure_col"
+  train_df %>%
+  tibble::rowid_to_column(var = "sample_id_col") %>%
+  tidyr::pivot_longer(cols = c(ASAH1, MMP.9, SYNEMIN, GPNMB), names_to = "feature_id_col", values_to = "measure_col") ->
+  train_long
+
+  normalized_df = train_long %>% group_by(sample_id_col) %>% 
+      mutate(median_run = median(measure_col, na.rm = TRUE)) %>% 
+      ungroup()
+  old_measure_col = paste("preNorm", measure_col, sep = "_")
+  normalized_df = normalized_df %>% mutate(median_global = median(!!(sym(measure_col)), 
+    na.rm = TRUE), `:=`(!!(old_measure_col), !!(sym(measure_col)))) %>% 
+    mutate(diff_norm = median_global - median_run) %>% mutate(`:=`(!!(sym(measure_col)), 
+    !!(sym(measure_col)) + diff_norm))
+
+  normalized_df = switch("default", all = normalized_df, 
+      default = normalized_df, minimal = normalized_df %>% 
+        dplyr::select(c(sample_id_col, feature_id_col, 
+          measure_col, old_measure_col)))
+
+  # ggplot(normalized_df, aes(x=as.factor(sample_id_col), y=measure_col, col=sample_id_col)) +
+  #   geom_boxplot()
+
+  normalized_df %>%
+    dplyr::select(sample_id_col, class, feature_id_col, measure_col) %>%
+    tidyr::pivot_wider(id_cols = c("sample_id_col", "class"), names_from = "feature_id_col", values_from = "measure_col") %>%
+    tibble::column_to_rownames("sample_id_col") -> train_df
+
+  return(train_df)
+}
+
 #' @title Master Trainer and Prediction
@@
       .d %>% 
-        dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+        dplyr::group_by(gene, Number.x, patient_state) %>% 
         dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
@@
         .d_second <- .d[collection_type=="E2"]
-        tmp$ratio <- .d_initial$conc / .d_second$conc
+        tmp$ratio <- .d_second$conc / .d_initial$conc
       } else {
@@
     rbindlist() %>%
-    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "post-surgery/pre-surgery"))) %>%
     dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
@@
 
-  pdata <- merge(pdata, meta_dt, by="Number.x")
-
   ### Pivot Table to wide
   pdata %>%
-    dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide
+    dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide
 
@@
     # Perform splsda
-    MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+    # MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+    MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2)], train_df$class, keepX = c(4,4)) 
     # Make prediction
-    pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
+    # pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
+    pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2)], test_df$class)
     # evaluate the prediction accuracy for the first two components
@@
 
+splsda <- function(train_df, test_df){
+
+    # Perform splsda
+    MyResult.splsda <- mixOmics::splsda(train_df, train_df$class, keepX = c(4,4)) 
+    # Make prediction
+    pred_test <- predict(MyResult.splsda, test_df, test_df$class)
+    # evaluate the prediction accuracy for the first two components
+    predict.comp2 <- pred_test$class$mahalanobis.dist[,2]
+    # compute confusion matrix
+    conf_mat <- table(factor(predict.comp2, levels = levels(train_df$class)), test_df$class)
+    # print(conf_mat)
+    # compute accuracy
+    acc <- (conf_mat[1,1] + conf_mat[2,2])/(conf_mat[1,1] + conf_mat[2,2] + conf_mat[1,2] + conf_mat[2,1])
+
+    return(list(model=MyResult.splsda, acc=acc))
+}
 ```
@@
 
-patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "Overall_Lab_ID", all = TRUE)
-patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y
+patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "LAB_ID", all = TRUE)
+# patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y
 
@@
   dplyr::ungroup()-> meta_dt
+
+patient_id_meta <- unique(data_long[, .(LAB_ID, Number.x, collection_type, manifestation_type, patient_state, `Time-to-reccurence (days)`)])
 ```
@@
     .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
+      dplyr::group_by(gene, Number.x, patient_state) %>% 
       dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
@@
 
+org_class <- train_df$class
+train_df$class <- sample(train_df$class)
+
 MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1)], train_df$class, keepX = c(4,4))
 background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")
 
-plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+plsda_data <- cbind( patient_id=na.omit(data_wide %>% dplyr::filter(Number.x!=32))  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
 plsda_data$patient_id <- as.factor(plsda_data$patient_id)
@@
   ) +
-  ggnewscale::new_scale_color() +
-  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
-            arrow = arrow(type = "closed", ends='both',
-                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  # ggnewscale::new_scale_color() +
+  # ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=paste(patient_id, `Time-to-reccurence (days)`, sep=": "))) +
+  # ggnewscale::new_scale_color() +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+  #           arrow = arrow(type = "closed", ends='both',
+  #                         length=unit(0.1, "inches")), size=2, alpha=0.8) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
   guides(shape="none",
+         colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
          colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Trajectory", ncol=1))
+         col=guide_legend(title="Class", ncol=1, byrow = T))
+  # guides(shape="none",
+  #        colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+  #        colour_new='none',
+  #        col=guide_legend(title="Trajectory", ncol=1))
 
 splsda_ratio_avg_traj_fig <- p
+
+background %>%
+  dplyr::filter(Var2 < 1.59058 & Var2 > -1.56044 & Var1 > -2.5462 & Var1 < 2.36565) -> background_line
+
+lm_eq <- lm(Var2 ~ Var1, data=background_line)
+coef(lm_eq)
+
+plsda_data %>%
+  dplyr::rowwise() %>%
+  dplyr::mutate(new_x = (comp1 + -1/coef(lm_eq)[2]*comp2  - (-1/coef(lm_eq)[2]*(-coef(lm_eq)[1])))/(1+(-1/coef(lm_eq)[2])^2),
+                new_y = (-1/coef(lm_eq)[2]*comp1 + ((-1/coef(lm_eq)[2])^2)*comp2  + (-coef(lm_eq)[1]))/(1+(-1/coef(lm_eq)[2])^2)) -> plsda_data_proj
+
+ggplot() +
+  # geom_point(data=background_line, aes(x=Var1, y=Var2)) +
+  geom_abline(slope = coef(lm_eq)[2], intercept = coef(lm_eq)[1]) +
+  geom_abline(slope = -1/coef(lm_eq)[2], intercept = -coef(lm_eq)[1]) +
+  geom_point(data=plsda_data_proj, aes(x=new_x, y=new_y, col=class))
+
+ggplot() +
+  geom_polygon(data = background,aes(x=Var1, y=Var2,
+                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
+  scale_fill_manual(values=c("#388ECC", "#F68B33")) +
+  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
+  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
+  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3, alpha=0.45) +
+  scale_shape_manual(values=c(15, 16, 17)) +
+  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
+                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
+  ) +
+  # ggnewscale::new_scale_color() +
+  # ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=`Time-to-reccurence (days)`)) +
+  ggnewscale::new_scale_color() +
+  ggnewscale::new_scale('shape') +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+  #           arrow = arrow(type = "closed", ends='both',
+  #                         length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  geom_abline(slope = -1/coef(lm_eq)[2], intercept = -coef(lm_eq)[1]) +
+  geom_point(data=plsda_data_proj, aes(x=new_x, y=new_y, col=class, shape=class), size=3)  +
+  scale_shape_manual(values=c(15, 16, 17)) +
+  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
+                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
+  ) +
+  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "right", legend.box="vertical",) +
+  guides(shape="none",
+         shape_new="none",
+         colour_new="none",
+         col=guide_legend(title="Class", ncol=1))
+
+ggplot() +
+  geom_point(data=merge(plsda_data_proj %>% dplyr::filter(class=='rGBM') %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=class, y=-new_x, col=`Time-to-reccurence (days)`)) +
+  ggrepel::geom_label_repel(data=merge(plsda_data_proj %>% dplyr::filter(class=='rGBM') %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=class, y=-new_x, col=`Time-to-reccurence (days)`, label=`Time-to-reccurence (days)`)) +
+  labs(y="Projected X-variate 1", x=NULL) +
+  scale_x_discrete(labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5, size=17),
+        axis.title = element_text(size = 17),
+        axis.text = element_text(size=16),
+        strip.text = element_text(size=16),
+        legend.title = element_text(size=16),
+        legend.text = element_text(size=15),
+        legend.position = "right", legend.box="vertical",) +
+  guides(col='none')
 
@@
 test_patients <- test_df$Number.x
+train_df <- rbindlist(list(train_df, test_df))
 
@@
 MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2)], train_df$class, keepX = c(4,4))
 
 pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
+pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2)], test_df$class)
 message("Perform background predictions")
 background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist", xlim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))), ylim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))))
+background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist", xlim = c(min(c(MyResult.splsda$variates$X)), max(c(MyResult.splsda$variates$X))), ylim = c(min(c(MyResult.splsda$variates$X)), max(c(MyResult.splsda$variates$X))))
 message("Prepare data for plotting")
@@
   ) +
+  # ggnewscale::new_scale_color() +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+            # arrow = arrow(type = "closed", ends='both',
+                          # length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  # ggnewscale::new_scale_color() +
+  # geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
+  # scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
   ggnewscale::new_scale_color() +
-  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
-            arrow = arrow(type = "closed", ends='both',
-                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
-  ggnewscale::new_scale_color() +
-  geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
-  scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
+  ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=`Time-to-reccurence (days)`)) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
          colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
-         colour_new=guide_legend(title="Trajectory", ncol=1, byrow = T),
-         col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
+         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
+         col='none')
+  # guides(shape="none",
+  #        colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+  #        colour_new=guide_legend(title="Trajectory", ncol=1, byrow = T),
+  #        col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
 
@@
 ratio_data <- get_ratio_data(data_long2)
+ratio_data <- data_wide
+colnames(ratio_data) <- make.names(colnames(ratio_data))
+ratio_data$class <- factor(ratio_data$class, levels=c("iGBM", "rGBM"))
+ratio_data <- ratio_data[Number.x!=32]
 # Get unique set of patient ids
@@
 ggplot(df_plot, aes(x=mean_acc, fill=method)) +
-  geom_histogram(binwidth = 0.15, colour = "black") +
+  geom_histogram(binwidth = 0.05, colour = "black") +
   scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
   scale_fill_brewer(palette = "Dark2") +
-  stat_bin(binwidth = 0.15, aes(label=..count..), vjust=-0.5, geom = "text") +
+  stat_bin(binwidth = 0.05, aes(label=..count..), vjust=-0.5, geom = "text") +
   facet_wrap(~method) +
@@
 train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
 pred_obj_1.1 <- run_predictions(train_df)
@@
 train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
 pred_obj_1.2 <- run_predictions(train_df)
@@
 train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
 pred_obj_1.5 <- run_predictions(train_df, pred_col_prob = "rGBM")
@@
 # iGBM vs rGBM ratio ----
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+
+## Median Centre data before computing ratios
+
+data_long[patient_state==1] %>%
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::add_count() %>%
+  dplyr::filter(n==2) %>% 
+  dplyr::ungroup() -> tmp_select
+
+tmp_select %>%
+  dplyr::select(gene, Number.x, conc, class=manifestation_type) %>%
+  dplyr::mutate(conc = log2(conc+1)) -> tmp
+tmp %>%
+  tidyr::pivot_wider(id_cols = c(Number.x, class), names_from = gene, values_from = conc ) -> tmp
+colnames(tmp) <- make.names(colnames(tmp))
+tmp <- na.omit(tmp)
+
+tmp_df_proteins <- median_centre_data(tmp[, c(2:6)])
+tmp_median_centred <- cbind(tmp[, 1], tmp_df_proteins)
+
+tmp_median_centred %>%
+  dplyr::rename(manifestation_type=class, `MMP-9`=MMP.9) %>%
+  tidyr::pivot_longer(cols = c(ASAH1, `MMP-9`, SYNEMIN, GPNMB), names_to = "gene", values_to = "conc") -> tmp_median_centred
+
+data_long_tmp <- merge(tmp_select[, c("Number.x", "gene", "patient_state", "LAB_ID", "collection_type", "manifestation_type")], tmp_median_centred, by=c("Number.x", "manifestation_type", "gene"), all.y=T)
+
+ggplot(data_long_tmp, aes(x=as.factor(Number.x), y=conc, col=Number.x)) +
+    geom_boxplot()
+
+data_long_tmp %>% 
   dplyr::group_by(gene, Number.x) %>%
@@
       .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
+      tmp$ratio <- .d_initial$conc - .d_second$conc
     } else if (nrow(.d)==2) {
@@
   dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
+  dplyr::mutate(log2_conc = ratio) %>%
   # dplyr::group_by(gene, Number.x) %>%
@@
 
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+data_long_tmp %>% 
+  # dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
   dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
@@
       .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_second$conc / .d_initial$conc
+      tmp$ratio <- .d_second$conc - .d_initial$conc
     } else if (nrow(.d)==2) {
@@
   dplyr::filter(grepl(";", agg_lab_id)) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
+  dplyr::mutate(log2_conc = ratio) %>%
   # dplyr::group_by(gene, Number.x) %>%
@@
 train_df <- data_wide
+train_df <- data_wide[Number.x!=32]
 train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
@@
   mutate(n = row_number()) %>%
-  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`, `Time-to-reccurence (days)`) -> data_wide
-data_wide %>%
-  dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
+  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB,  Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`) -> data_wide
+# data_wide %>%
+  # dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
 
@@
 train_df <- as.data.frame(train_df)
+train_df_proteins <- median_centre_data(train_df[, c(1:5)])
+train_df <- cbind(train_df_proteins, train_df[, c(6:9)])
 pred_obj_1.4 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+# coef(pred_obj_1.5$model$finalModel, pred_obj_1.5$model$bestTune$lambda)
+# coef(pred_obj_1.41$model$finalModel, pred_obj_1.41$model$bestTune$lambda)
+# coef(pred_obj_1.4$model$finalModel, pred_obj_1.4$model$bestTune$lambda)
 
@@
   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.4_roc_data
+
+# iGBM vs rGBM with clinical data Only ----
+data_long %>% 
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
+  dplyr::mutate(log2_conc = log2(conc+1)) %>%
+  dplyr::filter(manifestation_type=="Second manifestation" | manifestation_type=="Initial manifestation") %>%
+  dplyr::mutate(class = plyr::mapvalues(manifestation_type, from=c("Initial manifestation","Second manifestation"), to=c("iGBM", "rGBM"))) -> pdata
+
+## Pivot Table
+pdata %>%
+  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
+  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
+  dplyr::ungroup() -> data_wide
+
+## Only use proteins for prediction
+data_wide %>%
+  mutate(n = row_number()) %>%
+  dplyr::select(Number.x, class, n,  Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`) -> data_wide
+# data_wide %>%
+  # dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
+
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+# train_df_proteins <- median_centre_data(train_df[, c(1:5)])
+# train_df <- cbind(train_df_proteins, train_df[, c(6:9)])
+pred_obj_1.41 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+# coef(pred_obj_1.5$model$finalModel, pred_obj_1.5$model$bestTune$lambda)
+# coef(pred_obj_1.4$model$finalModel, pred_obj_1.4$model$bestTune$lambda)
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.41$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.41 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.41[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.41[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.41[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.41[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.41 <- g
+
+pred_obj_1.41$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.41_roc_data
 
@@
 
-master_auc_data <- rbindlist(list(`Ctrl vs iGBM`=auc_tbl_1.1, `Ctrl vs rGBM`=auc_tbl_1.2, `iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
+master_auc_data <- rbindlist(list(`iGBM vs rGBM protein + clinical`=auc_tbl_1.41, `Ctrl vs iGBM`=auc_tbl_1.1, `Ctrl vs rGBM`=auc_tbl_1.2, `iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
 
 # Only iGMB and rGBM
-master_roc_data <- rbindlist(list(`iGBM vs rGBM protein`=tmp_pred_obj_1.5_roc_data, `iGBM vs rGBM protein + clinical`=tmp_pred_obj_1.4_roc_data, `iGBM vs rGBM protein ratios`=tmp_pred_obj_1.3_roc_data), idcol = "Comparison")
-
-master_auc_data <- rbindlist(list(`iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
+master_roc_data <- rbindlist(list(`iGBM vs rGBM clinical`=tmp_pred_obj_1.41_roc_data, `iGBM vs rGBM protein`=tmp_pred_obj_1.5_roc_data, `iGBM vs rGBM protein + clinical`=tmp_pred_obj_1.4_roc_data, `iGBM vs rGBM protein ratios`=tmp_pred_obj_1.3_roc_data), idcol = "Comparison")
+
+master_auc_data <- rbindlist(list(`iGBM vs rGBM clinical`=auc_tbl_1.41, `iGBM vs rGBM protein`=auc_tbl_1.5, `iGBM vs rGBM protein + clinical`=auc_tbl_1.4, `iGBM vs rGBM protein ratios`=auc_tbl_1.3), idcol = "Comparison")
 
@@
 
+<!-- # Combination Proteins -->
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
+# iGBM vs rGBM protein conc----
+data_long %>% 
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
+  dplyr::mutate(log2_conc = log2(conc+1)) %>%
+  dplyr::filter(manifestation_type=="Second manifestation" | manifestation_type=="Initial manifestation") %>%
+  dplyr::mutate(class = plyr::mapvalues(manifestation_type, from=c("Initial manifestation","Second manifestation"), to=c("iGBM", "rGBM"))) -> pdata
+
+## Pivot Table
+pdata %>%
+  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
+  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
+  dplyr::ungroup() -> data_wide
+
+## Only use proteins for prediction
+data_wide %>%
+  mutate(n = row_number()) %>%
+  dplyr::select(Number.x, class, n, ASAH1, `MMP-9`, SYNEMIN, GPNMB) -> data_wide
+
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
+train_df <- train_df %>% dplyr::select(class, ASAH1)
+train_df$dummy_var <- 0
+pred_obj_1.51 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.51$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.51$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.51 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.51[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.51$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.51[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.51$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.51[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.51$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.51[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.51$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.51 <- g
+
+pred_obj_1.51$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.51_roc_data
+
+## 2 Proteins
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
+train_df <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN)
+pred_obj_1.52 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.52$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.52$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.52 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.52[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.52$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.52[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.52$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.52[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.52$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.52[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.52$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.52 <- g
+
+pred_obj_1.52$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.52_roc_data
+
+## 3 Protein
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
+train_df <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN, GPNMB)
+pred_obj_1.53 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.53$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.53$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.53 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.53[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.53$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.53[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.53$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.53[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.53$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.53[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.53$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.53 <- g
+
+pred_obj_1.53$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.53_roc_data
+
+## 4 Protein
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+train_df <- median_centre_data(train_df)
+train_df <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN, GPNMB, MMP.9)
+pred_obj_1.54 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.54$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.54$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.54 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.54[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.54$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.54[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.54$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.54[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.54$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.54[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.54$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.54 <- g
+
+pred_obj_1.54$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.54_roc_data
+
+master_Table <- rbindlist(list(`1 Protein`=tmp_pred_obj_1.51_roc_data, `1-2 Proteins`=tmp_pred_obj_1.52_roc_data, `1-3 Proteins`=tmp_pred_obj_1.53_roc_data, `1-4 Proteins`=tmp_pred_obj_1.54_roc_data), idcol = 'Number of Proteins')
+
+master_auc_data <- rbindlist(list(`1 Protein`=auc_tbl_1.51, `1-2 Proteins`=auc_tbl_1.52, `1-3 Proteins`=auc_tbl_1.53, `1-4 Proteins`=auc_tbl_1.54), idcol = "Number of Proteins")
+
+master_auc_data[Model %in% c("Elastic Logit")] %>%
+  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
+  tidyr::pivot_wider(id_cols = "Number of Proteins", names_from = Model, values_from = AUC) -> auc_annotation_tbl
+
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 13)
+auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
+
+g <- ggplot(master_Table[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=`Number of Proteins`), linetype='solid') +
+  geom_roc(n.cuts=0, size=1.5) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "Dark2")) +
+  # scale_linetype_manual(values=c("solid", "dotted"), labels=c("Elastic Logit")) + 
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity") +
+  annotation_custom(auc_grob, xmin=0.25, xmax=0.95, ymin=0.10, ymax=0.30) +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.title = element_text(size=14),
+        legend.text = element_text(size=13)) +
+  guides(col="none")
+g
+# iGBM vs rGBM ratio ----
+
+## Median Centre data before computing ratios
+
+data_long[patient_state==1] %>%
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::add_count() %>%
+  dplyr::filter(n==2) %>% 
+  dplyr::ungroup() -> tmp_select
+
+tmp_select %>%
+  dplyr::select(gene, Number.x, conc, class=manifestation_type) %>%
+  dplyr::mutate(conc = log2(conc+1)) -> tmp
+tmp %>%
+  tidyr::pivot_wider(id_cols = c(Number.x, class), names_from = gene, values_from = conc ) -> tmp
+colnames(tmp) <- make.names(colnames(tmp))
+tmp <- na.omit(tmp)
+
+tmp_df_proteins <- median_centre_data(tmp[, c(2:6)])
+tmp_median_centred <- cbind(tmp[, 1], tmp_df_proteins)
+
+tmp_median_centred %>%
+  dplyr::rename(manifestation_type=class, `MMP-9`=MMP.9) %>%
+  tidyr::pivot_longer(cols = c(ASAH1, `MMP-9`, SYNEMIN, GPNMB), names_to = "gene", values_to = "conc") -> tmp_median_centred
+
+data_long_tmp <- merge(tmp_select[, c("Number.x", "gene", "patient_state", "LAB_ID", "collection_type", "manifestation_type")], tmp_median_centred, by=c("Number.x", "manifestation_type", "gene"), all.y=T)
+
+ggplot(data_long_tmp, aes(x=as.factor(Number.x), y=conc, col=Number.x)) +
+    geom_boxplot()
+
+data_long_tmp %>% 
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::group_map(function(.d, ...){
+    .d <- as.data.table(.d)
+    .d %>% 
+      dplyr::group_by(gene, Number.x, patient_state) %>% 
+      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+    # print(.d)
+    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+      .d_initial <- .d[manifestation_type=="Initial manifestation"]
+      .d_second <- .d[manifestation_type=="Second manifestation"]
+      tmp$ratio <- .d_initial$conc - .d_second$conc
+    } else if (nrow(.d)==2) {
+      .d_initial <- .d[collection_type=="E1"]
+      .d_second <- .d[collection_type=="E2"]
+      tmp$ratio <- .d_initial$conc / .d_second$conc
+    } else {
+      tmp$ratio <- .d$conc
+    }
+    return(tmp)
+  }, .keep = TRUE) %>%
+  rbindlist() %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
+  dplyr::mutate(log2_conc = ratio) %>%
+  # dplyr::group_by(gene, Number.x) %>%
+  # dplyr::add_count() %>%
+  # dplyr::filter(n==2) %>%
+  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
+
+data_long_tmp %>% 
+  # dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::group_map(function(.d, ...){
+    .d <- as.data.table(.d)
+    .d %>% 
+      dplyr::group_by(gene, Number.x, patient_state) %>% 
+      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
+                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
+    # print(.d)
+    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
+      .d_initial <- .d[manifestation_type=="Initial manifestation"]
+      .d_second <- .d[manifestation_type=="Second manifestation"]
+      tmp$ratio <- .d_second$conc - .d_initial$conc
+    } else if (nrow(.d)==2) {
+      .d_initial <- .d[collection_type=="E1"]
+      .d_second <- .d[collection_type=="E2"]
+      tmp$ratio <- .d_initial$conc / .d_second$conc
+    } else {
+      tmp$ratio <- .d$conc
+    }
+    return(tmp)
+  }, .keep = TRUE) %>%
+  rbindlist() %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
+  dplyr::filter(grepl(";", agg_lab_id)) %>%
+  dplyr::mutate(log2_conc = ratio) %>%
+  # dplyr::group_by(gene, Number.x) %>%
+  # dplyr::add_count() %>%
+  # dplyr::filter(n==2) %>%
+  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
+
+pdata %>%
+  dplyr::select(Number.x, gene, log2_conc) %>%
+  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+  dplyr::mutate(class="iGBM") -> pdata
+
+pdata2 %>%
+  dplyr::select(Number.x, gene, log2_conc) %>%
+  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
+  dplyr::mutate(class="rGBM") -> pdata2
+
+pdata <- rbindlist(list(pdata, pdata2))
+
+## Pivot Table
+pdata %>%
+  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide
+
+## Prediction
+train_df <- data_wide
+train_df <- data_wide[Number.x!=32]
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
+train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
+## Make Valid Column Names 
+colnames(train_df) <- make.names(colnames(train_df))
+train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+
+## 1 Protein ----
+train_df_sub <- train_df %>% dplyr::select(class, ASAH1)
+train_df_sub$dummy_var <- 0
+pred_obj_1.31 <- run_predictions(train_df_sub, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.31$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.31$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.31 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.31[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.31$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.31[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.31$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.31[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.31$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.31[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.31$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.31 <- g
+
+pred_obj_1.31$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.31_roc_data
+
+## 2 Protein ----
+train_df_sub <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN)
+pred_obj_1.32 <- run_predictions(train_df_sub, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.32$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.32$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.32 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.32[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.32$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.32[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.32$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.32[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.32$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.32[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.32$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.32 <- g
+
+pred_obj_1.32$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.32_roc_data
+
+## 3 Protein ----
+train_df_sub <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN, GPNMB)
+pred_obj_1.33 <- run_predictions(train_df_sub, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.33$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.33$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.33 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.33[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.33$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.33[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.33$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.33[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.33$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.33[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.33$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.33 <- g
+
+pred_obj_1.33$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.33_roc_data
+
+## 4 Protein ----
+train_df_sub <- train_df %>% dplyr::select(class, ASAH1, SYNEMIN, GPNMB, MMP.9)
+pred_obj_1.34 <- run_predictions(train_df_sub, pred_col_prob = "rGBM")
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.34$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.34$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.34 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.34[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.34$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.34[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.34$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.34[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.34$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.34[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.34$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.34 <- g
+
+pred_obj_1.34$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.34_roc_data
+
+## Merged Plot
+master_Table <- rbindlist(list(`1 Protein`=tmp_pred_obj_1.31_roc_data, `1-2 Proteins`=tmp_pred_obj_1.32_roc_data, `1-3 Proteins`=tmp_pred_obj_1.33_roc_data, `1-4 Proteins`=tmp_pred_obj_1.34_roc_data), idcol = 'Number of Proteins')
+
+master_auc_data <- rbindlist(list(`1 Protein`=auc_tbl_1.31, `1-2 Proteins`=auc_tbl_1.32, `1-3 Proteins`=auc_tbl_1.33, `1-4 Proteins`=auc_tbl_1.34), idcol = "Number of Proteins")
+
+master_auc_data[Model %in% c("Elastic Logit")] %>%
+  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
+  tidyr::pivot_wider(id_cols = "Number of Proteins", names_from = Model, values_from = AUC) -> auc_annotation_tbl
+
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 13)
+auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
+
+g <- ggplot(master_Table[Model %in% c("Elastic Logit")], aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=`Number of Proteins`), linetype='solid') +
+  geom_roc(n.cuts=0, size=1.5) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "Dark2")) +
+  # scale_linetype_manual(values=c("solid", "dotted"), labels=c("Elastic Logit")) + 
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity") +
+  annotation_custom(auc_grob, xmin=0.25, xmax=0.95, ymin=0.10, ymax=0.30) +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.title = element_text(size=14),
+        legend.text = element_text(size=13)) +
+  guides(col="none")
+g
+
+```
+
 <!-- # Latency Data -->
@@
 
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data)
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data %>% dplyr::select(-MMP.9.ng.ml))
 print(fit)
@@
   labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
-  guides(col=guide_legend(nrow=4,byrow=TRUE))
+  guides(col=guide_legend(nrow=1,byrow=TRUE))
 
@@
         legend.position = "bottom")
+
+## All 4
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1 + SYNM + GPNMB + MMP9, data = ttr_surv_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p5 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = F,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(
+            sprintf("ASAH1=high, SYNM=high, GPNMB=high, MMP9=high (n=1)"), 
+            sprintf("ASAH1=high, SYNM=low, GPNMB=high, MMP9=low (n=1)"),  
+            sprintf("ASAH1=low, SYNM=high, GPNMB=low , MMP9=high (n=2)"), 
+            sprintf("ASAH1=high, SYNM=high, GPNMB=low , MMP9=low (n=1)"),
+            sprintf("ASAH1=high, SYNM=low , GPNMB=high, MMP9=low (n=1)"),
+            sprintf("ASAH1=high, SYNM=low , GPNMB=low , MMP9=high (n=1)"),
+            sprintf("ASAH1=high, SYNM=low , GPNMB=low , MMP9=low (n=2)"),
+            sprintf("ASAH1=low, SYNM=high, GPNMB=high, MMP9=high (n=1)"),
+            sprintf("ASAH1=low, SYNM=high, GPNMB=high, MMP9=low (n=1)"),
+            sprintf("ASAH1=low, SYNM=high, GPNMB=low , MMP9=high (n=2)"),
+            sprintf("ASAH1=low, SYNM=low , GPNMB=high, MMP9=low (n=3)"),
+            sprintf("ASAH1=low, SYNM=low , GPNMB=low , MMP9=high (n=1)"),
+            sprintf("ASAH1=low, SYNM=low , GPNMB=low , MMP9=low (n=2)")
+            )
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
+  guides(col=guide_legend(nrow=4,byrow=TRUE))
+
+ttr_surv_p5 <- ttr_surv_p5$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = "bottom")
+
+ttr_surv_p5_with_ci <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = T,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high\n(n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
+  guides(col=guide_legend(nrow=2,byrow=TRUE))
+
+ttr_surv_p5_with_ci <- ttr_surv_p5_with_ci$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = "bottom")
+
+## COXPH
+fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
+print(summary(fit))
+
+library(finalfit)
+
+dependent_var <- "Surv(Time.to.reccurence..days., status)"
+explanatory_vars <- c("ASAH1.ng.ml", "SYNM.ng.ml", "GPNMB.ng.ml", "MMP.9.ng.ml")
+
+ttr_surv_data %>% 
+    coxphmulti(dependent_var, explanatory_vars) %>% 
+    cox.zph() %>% 
+    {zph_result <<- .} %>% 
+    plot(var=3)
+
+zph_result
+
+ttr_surv_data %>%
+  finalfit(dependent_var, explanatory_vars) %>%
+  gt()
+
+ttr_surv_data %>%
+  hr_plot(dependent_var, explanatory_vars, dependent_label="Time to Recurrence")
+
+ttr_surv_data %>%
+  surv_plot(dependent_var, explanatory_vars, xlab="Time (days)", pval=TRUE, legend="none")
+
+ttr_surv_data %>%
+  ff_plot(dependent_var, explanatory_vars, dependent_label="Time to Recurrence")
 ```

### 2023-02-06 results/elisa_plasma_proteomics/manuscript_elisa_figure_analysis.Rmd
[ADD] finalized ELISA plots, pngs and svgs, and add cox-ph model

@@
 
-## All 4
+## All 4 ----
 fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1 + SYNM + GPNMB + MMP9, data = ttr_surv_data)
@@
 
-## COXPH
+## COXPH ----
 fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)

### 2023-02-06 results/elisa_plasma_proteomics/manuscript_figures_finalized.Rmd
[ADD] finalized ELISA plots, pngs and svgs, and add cox-ph model

--- 2023-01-04 results/elisa_plasma_proteomics/manuscript_elisa_figure_final.Rmd
@@
 ```{r message=FALSE, warning=FALSE, include=FALSE}
+#' median_centre_data
+#'
+#' @param train_df  (data.frame) A dataframe that contains labels and features to run through prediction models. Has to have the following structure:
+#' \item{class}{(character) sample/observation label. i.e. ctrl, iGBM, or rGBM sample.}
+#' \item{feature_1}{(numeric} a specific feature measurement such as protein ASAH1)}
+#' \item{feature_n}{(numeric) another specific feature.}
+#' @param pred_col_prob (character) A character corresponding to the class of probabilities you want to extract for. i.e. "GBM"
+#' @param multiclass (logical) Run predictions using multiclass modelling and prediction if more than 2 classes.
+#' 
+#' @return a data.frame with columns sample_id_col, class, and the measures normalized by median
+#' 
+#' This function takes in a train data.frame and returns the data after normalizing the measures 
+#' by subtracting the median of each sample from each value and then adding the median of the whole data.
+#' The input data.frame is also pivoted for easier analysis.
+median_centre_data <- function(train_df){
+  values_to = "measure_col"
+  measure_col="measure_col"
+  train_df %>%
+  tibble::rowid_to_column(var = "sample_id_col") %>%
+  tidyr::pivot_longer(cols = c(ASAH1, MMP.9, SYNEMIN, GPNMB), names_to = "feature_id_col", values_to = "measure_col") ->
+  train_long
+
+  normalized_df = train_long %>% group_by(sample_id_col) %>% 
+      mutate(median_run = median(measure_col, na.rm = TRUE)) %>% 
+      ungroup()
+  old_measure_col = paste("preNorm", measure_col, sep = "_")
+  normalized_df = normalized_df %>% mutate(median_global = median(!!(sym(measure_col)), 
+    na.rm = TRUE), `:=`(!!(old_measure_col), !!(sym(measure_col)))) %>% 
+    mutate(diff_norm = median_global - median_run) %>% mutate(`:=`(!!(sym(measure_col)), 
+    !!(sym(measure_col)) + diff_norm))
+
+  normalized_df = switch("default", all = normalized_df, 
+      default = normalized_df, minimal = normalized_df %>% 
+        dplyr::select(c(sample_id_col, feature_id_col, 
+          measure_col, old_measure_col)))
+
+  # ggplot(normalized_df, aes(x=as.factor(sample_id_col), y=measure_col, col=sample_id_col)) +
+  #   geom_boxplot()
+
+  normalized_df %>%
+    dplyr::select(sample_id_col, class, feature_id_col, measure_col) %>%
+    tidyr::pivot_wider(id_cols = c("sample_id_col", "class"), names_from = "feature_id_col", values_from = "measure_col") %>%
+    tibble::column_to_rownames("sample_id_col") -> train_df
+
+  return(train_df)
+}
+
+median_centre_recurrent_data <- function(data_long){
+  data_long[patient_state==1] %>%
+    dplyr::group_by(gene, Number.x) %>%
+    dplyr::add_count() %>%
+    dplyr::filter(n==2) %>% 
+    dplyr::ungroup() -> tmp_select
+
+  tmp_select %>%
+    dplyr::select(gene, Number.x, conc, class=manifestation_type) %>%
+    dplyr::mutate(conc = log2(conc+1)) -> tmp
+  tmp %>%
+    tidyr::pivot_wider(id_cols = c(Number.x, class), names_from = gene, values_from = conc ) -> tmp
+  colnames(tmp) <- make.names(colnames(tmp))
+  tmp <- na.omit(tmp)
+
+  tmp_df_proteins <- median_centre_data(tmp[, c(2:6)])
+  tmp_median_centred <- cbind(tmp[, 1], tmp_df_proteins)
+
+  tmp_median_centred %>%
+    dplyr::rename(manifestation_type=class, `MMP-9`=MMP.9) %>%
+    tidyr::pivot_longer(cols = c(ASAH1, `MMP-9`, SYNEMIN, GPNMB), names_to = "gene", values_to = "conc") -> tmp_median_centred
+
+  data_long_tmp <- merge(tmp_select[, c("Number.x", "gene", "patient_state", "LAB_ID", "collection_type", "manifestation_type")], tmp_median_centred, by=c("Number.x", "manifestation_type", "gene"), all.y=T)
+  return(data_long_tmp)
+}
+
 #' @title Master Trainer and Prediction
@@
 #' @return A data frame with the calculated ratios.
-get_ratio_data <- function(data_long2){
+get_ratio_data <- function(data_long2, already_log_transformed=F){
   ## Data wrangling ----
@@
   data_long2 %>% 
-    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+    dplyr::rowwise() %>%
+    dplyr::mutate(conc = ifelse(already_log_transformed==T, conc, conc+1)) %>% # Add one to avoid cases of 0 division
     dplyr::group_by(gene, Number.x) %>%
@@
         .d_second <- .d[manifestation_type=="Second manifestation"]
+        if (!already_log_transformed){
           tmp$ratio <- .d_initial$conc / .d_second$conc
+        } else {
+          tmp$ratio <- .d_initial$conc - .d_second$conc
+        }
       } else if (nrow(.d)==2) {
@@
         .d_second <- .d[collection_type=="E2"]
+        if (!already_log_transformed){
           tmp$ratio <- .d_initial$conc / .d_second$conc
+        } else {
+          tmp$ratio <- .d_initial$conc - .d_second$conc
+        }
       } else {
@@
     dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-    dplyr::mutate(log2_conc = log2(ratio)) %>%
+    dplyr::rowwise() %>%
+    dplyr::mutate(log2_conc = ifelse(already_log_transformed==T, ratio, log2(ratio))) %>%
     # dplyr::group_by(gene, Number.x) %>%
@@
   data_long2 %>% 
-    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
+    dplyr::rowwise() %>%
+    dplyr::mutate(conc = ifelse(already_log_transformed==T, conc, conc+1)) %>% # Add one to avoid cases of 0 division
     dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
@@
         .d_second <- .d[manifestation_type=="Second manifestation"]
+        if (!already_log_transformed){
           tmp$ratio <- .d_second$conc / .d_initial$conc
+        } else {
+          tmp$ratio <- .d_second$conc - .d_initial$conc
+        }
       } else if (nrow(.d)==2) {
@@
         .d_second <- .d[collection_type=="E2"]
-        tmp$ratio <- .d_initial$conc / .d_second$conc
+        if (!already_log_transformed){
+          tmp$ratio <- .d_second$conc / .d_initial$conc
+        } else {
+          tmp$ratio <- .d_second$conc - .d_initial$conc
+        }
       } else {
@@
     rbindlist() %>%
-    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
+    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "post-surgery/pre-surgery"))) %>%
     dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
     dplyr::filter(grepl(";", agg_lab_id)) %>%
-    dplyr::mutate(log2_conc = log2(ratio)) %>%
+    dplyr::rowwise() %>%
+    dplyr::mutate(log2_conc = ifelse(already_log_transformed==T, ratio, log2(ratio))) %>%
     # dplyr::group_by(gene, Number.x) %>%
@@
 ```{r echo=FALSE, warning=FALSE}
+# Get patient meta data ----
 clinical_data <- fread(patient_clinical_data)
 patient_longitudinal_mapping_data <- fread(patient_longitudinal_mapping_file)
-
-patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "Overall_Lab_ID", all = TRUE)
-patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y
-
-elisa_data <- fread(elisa_data_file)
-
-data_long <- merge(elisa_data, patient_meta_data, by = "LAB_ID")
-# Remove patient 11, for some reason they are kept
-data_long %>%
-  dplyr::filter(Number.x!=11) -> data_long
-
-patient_meta_data %>%
-  dplyr::filter(!is.na(manifestation_type) & ELISA=="X") %>%
-  dplyr::group_by(Number.x) %>%
-  dplyr::add_count() %>%
-  dplyr::select(`Unique Patient Number`=Number.x, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-  unique() %>% dplyr::ungroup() -> recurrent_patients_meta
-
-# message(sprintf("There are %s patients with intial and recurrent ELISA measurements", nrow(recurrent_patients_meta)))
+patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "LAB_ID", all = TRUE)
 
@@
   dplyr::ungroup()-> meta_dt
+
+# Load actual ELISA data ----
+elisa_data <- fread(elisa_data_file)
+
+# Merge ELISA data with patient meta data ----
+data_long <- merge(elisa_data, patient_meta_data, by = "LAB_ID")
 ```
@@
 methods <- c("perm.test", "perm.test", "paired.perm.test")
-# methods <- c("perm.test", "perm.test", "perm.test")
 
@@
 ## Build the plot ----
-colfunc <- colorRampPalette(c("red","yellow","springgreen","royalblue"))
-ttr_categorizer <- function(data) {
-  data <- as.numeric(data)
-  if (is.na(data)){
-    return(NA)
-  }
-  if (data >= 100 & data < 200){
-    return("100-200")
-  } else if (data >= 200 & data < 300){
-    return("200-300")
-  } else if (data >= 300 & data < 400){
-    return("300-400")
-  } else if (data >= 400 & data < 500){
-    return("400-500")
-  } else if (data >= 500 & data < 600){
-    return("500-600")
-  }
-}
-# ttr_categorizer <- function(data) {
-#   data <- as.numeric(data)
-#   if (is.na(data)){
-#     return(NA)
-#   }
-#   if (data >= 100 & data < 150){
-#     return("100-150")
-#   } else if (data >= 150 & data < 200){
-#     return("150-200")
-#   } else if (data >= 200 & data < 250){
-#     return("200-250")
-#   } else if (data >= 250 & data < 300){
-#     return("250-300")
-#   } else if (data >= 300 & data < 350){
-#     return("300-350")
-#   } else if (data >= 350 & data < 400){
-#     return("350-400")
-#   } else if (data >= 400 & data < 450){
-#     return("400-450")
-#   } else if (data >= 450 & data < 500){
-#     return("450-500")
-#   } else if (data >= 500 & data < 550){
-#     return("500-550")
-#   } else if (data >= 550 & data < 600){
-#     return("550-600")
-#   }
-# }
-# ttr_categorizer <- function(data, split_value) {
-#   if (data <= split_value) {
-#     return(0)
-#   } else {
-#     return(1)
-#   }
-# }
-# ttr_categorizer <- function(data) {
-#   if (data >= 103 & data < 190){
-#     return("early")
-#   } else if (data >= 190 & data < 243){
-#     return("middle")
-#   } else if (data >= 243 & data <= 521){
-#     return("late")
-#   }
-# }
-pdata$`ttr_group` <- sapply(as.numeric(pdata$`Time-to-reccurence (days)`), ttr_categorizer)
-ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4) +
-  # stat_compare_means(paired = TRUE, label.x = 1.5, geom = "label") +
-  # stat_compare_means(comparisons = my_comparisons, group.by="gene", label.y = c(4, 7, 7), label.x = c(1, 2, 3)) +
+
+ggpaired(data = pdata, id = "Number.x", x = "Manifestation", y = "log2_conc", fill = "Manifestation", palette = 'grey', line.color = "grey", line.size = 0.4, notch = TRUE) +
   scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata$Manifestation)), "Dark2"), labels = c("Ctrl", "iGBM", "rGBM")) +
@@
   labs(x="", y="log2( Concentration )") +
-  facet_wrap(~gene, scales = "free_y") +
+  facet_wrap(~gene, nrow = 2, scales = "free_y") +
   theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
+  theme(panel.background = element_rect(fill="white"),
+        # panel.grid.major.y = element_line(colour = "black"), 
+        plot.title = element_text(hjust = 0.5, size=17),
+        # axis.line = element_blank(),
+        axis.ticks.x = element_blank(),
         axis.title = element_text(size = 17),
@@
         legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="", ncol=3),
+  guides(fill='none',
          col='none') -> p1_org
+
+dat <- ggplot_build(p1_org)$data[[1]]
+dat <- cbind(dat, gene=sort(unique(pdata$gene))[dat$PANEL])
+
+p1_org <- p1_org + geom_segment(data=dat, 
+                 aes(x=xmin, xend=xmax, 
+                    y=middle, yend=middle), 
+                 colour="white", size=1)
 
@@
   stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
-  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(8, 18, 9, 12)),
+  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(6, 16, 8, 11)),
              aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1_org
@@
         legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="", ncol=3),
+  guides(fill="none",
          col='none') -> p1
+
+dat <- ggplot_build(p1)$data[[1]]
+dat <- cbind(dat, gene=sort(unique(pdata$gene))[dat$PANEL])
+
+p1 <- p1 + geom_segment(data=dat, 
+                 aes(x=xmin, xend=xmax, 
+                    y=middle, yend=middle), 
+                 colour="white", size=1)
 
@@
   stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.2, hide.ns = T, size=4, bracket.size=1) +
-  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(7, 15, 8, 11)),
+  geom_point(data=data.frame(Manifestation=c("iGBM", "iGBM", "iGBM", "iGBM"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(6, 15, 8, 11)),
              aes(x=Manifestation, y=y, group=gene), alpha=0) -> p1
-# p1
-
-## Average Conc per ttr
-pdata %>%
-  dplyr::filter(Manifestation!="Ctrl") %>%
-  dplyr::select(gene, Number.x, Manifestation, log2_conc, ttr_group) %>%
-  dplyr::mutate(Manifestation = plyr::mapvalues(Manifestation, from=c("iGBM", "rGBM"), to=c("initial\nGBM", "recurrent\nGBM"))) %>%
-  dplyr::group_by(gene, Manifestation, ttr_group) %>%
-  dplyr::mutate(avg_log2_conc_per_mani_ttr = median(log2_conc, na.rm = T)) -> pdata_ttr
-
-p1_avg_ttr <- ggplot(data = pdata_ttr, aes(x=Manifestation, y=log2_conc)) +
-  geom_boxplot() +
-  geom_line(aes(x=Manifestation, y=avg_log2_conc_per_mani_ttr, col=ttr_group, group=ttr_group), inherit.aes = FALSE) +
-  labs(x="", y="log2( Concentration )") +
-  facet_wrap(~gene, scales = "free_y") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
-         col=guide_legend(title="Avg. log2 ( concentration )\nTime to Recurrence"))
 
@@
 
-## Time to Recurrence ----
-train_df_Sub <- pdata %>%
-  dplyr::filter( !(Number.x %in% c("32", "33"))) %>%
-  dplyr::filter(Manifestation!="Ctrl") %>%
-  dplyr::select(Number.x, Time.to.reccurence..days.=`Time-to-reccurence (days)`, class=Manifestation, gene, log2_conc) %>%
-  # dplyr::mutate(Time.to.reccurence..days. = ifelse(class=="iGBM", 0, Time.to.reccurence..days.)) %>%
-  # dplyr::mutate(class = ifelse(class=="iGBM", 0, 1) ) %>%
-  dplyr::rowwise() %>%
-  # dplyr::mutate(ttr_group = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
-  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 253.5)) %>%
-  # dplyr::select(-Time.to.reccurence..days.) %>%
-  dplyr::mutate(gene=make.names(gene)) %>%
-  tidyr::pivot_wider(id_cols =c("Number.x", "Time.to.reccurence..days.", "class"), names_from = "gene", values_from = "log2_conc") %>%
-  # tidyr::pivot_wider(id_cols = c("Number.x", "Time.to.reccurence..days."), names_from = c("class"), values_from = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9")) %>%
-  # dplyr::mutate(ASAH1 = (ASAH1_rGBM+1) /( ASAH1_iGBM+1),
-  #               SYNEMIN = (SYNEMIN_rGBM+1)/(SYNEMIN_iGBM+1),
-  #               GPNMB = (GPNMB_rGBM+1)/(GPNMB_iGBM+1),
-  #               MMP.9 = (MMP.9_rGBM+1)/(MMP.9_iGBM+1)) %>%
-  # dplyr::select(Number.x, Time.to.reccurence..days., ASAH1, SYNEMIN, GPNMB, MMP.9 ) %>% 
-  # unique() %>%
-  dplyr::arrange(Time.to.reccurence..days.)
-train_df_Sub <- na.omit(train_df_Sub)
-train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
-
-train_df_Sub %>%
-  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
-  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
-                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
-                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
-                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
-  dplyr::rowwise() %>%
-  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
-  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
-  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
-  unique() -> ttr_delta
-
-# ttr_delta$Time.to.reccurence..days. <- as.numeric(as.factor(ttr_delta$Time.to.reccurence..days.))
-lm_ttr <- lm(Time.to.reccurence..days. ~ ASAH1_delta + SYNEMIN_delta + GPNMB_delta + MMP.9_delta, data=ttr_delta )
-summary(lm_ttr)
-anova(lm_ttr)
-
-power_calc <- pwr.f2.test(u = 4, f2 = sqrt(summary(lm_ttr)$adj.r.squared), sig.level = 0.05, power = 0.8)
-print(sprintf("The estimated sample size needed for 80%% power is: %s", round(power_calc$v) + 5))
-
-ttr_categorizer <- function(data) {
-  if (data >= 103 & data < 190){
-    return("early")
-  } else if (data >= 190 & data < 243){
-    return("middle")
-  } else if (data >= 243 & data <= 521){
-    return("late")
-  }
-}
-
-train_df_Sub %>%
-  dplyr::group_by(Number.x, Time.to.reccurence..days.) %>%
-  dplyr::mutate(ASAH1_delta = ASAH1[class=="rGBM"] - ASAH1[class=="iGBM"],
-                SYNEMIN_delta = SYNEMIN[class=="rGBM"] - SYNEMIN[class=="iGBM"],
-                GPNMB_delta = GPNMB[class=="rGBM"] - GPNMB[class=="iGBM"],
-                MMP.9_delta = MMP.9[class=="rGBM"] - MMP.9[class=="iGBM"]) %>%
-  dplyr::rowwise() %>%
-  dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.))) %>%
-  # dplyr::mutate(Time.to.reccurence..days. = ttr_categorizer(as.numeric(Time.to.reccurence..days.), 207.5)) %>%
-  dplyr::select(Number.x, Time.to.reccurence..days., ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta) %>%
-  unique() -> ttr_delta
-
-tmp_pdata <- ttr_delta %>% tidyr::pivot_longer(cols = c(ASAH1_delta, SYNEMIN_delta, GPNMB_delta, MMP.9_delta), names_to="gene", values_to="log2_conc")
-tmp_pdata$Time.to.reccurence..days. <- factor(tmp_pdata$Time.to.reccurence..days., levels=c("early", "middle", "late"))
-# ggplot(data = tmp_pdata, aes(x=Time.to.reccurence..days., y=log2_conc, col=Time.to.reccurence..days.)) +
-#   geom_point() +
-#   labs(x="", y="log2( Concentration )") +
-#   facet_wrap(~gene, scales = "free_y") +
-#   theme_classic() +
-#   theme(plot.title = element_text(hjust = 0.5, size=17),
-#         axis.title = element_text(size = 17),
-#         axis.text = element_text(size=16),
-#         strip.text = element_text(size=16),
-#         legend.title = element_text(size=16),
-#         legend.text = element_text(size=15),
-#         legend.position = "bottom", legend.box="vertical",) +
-#   guides(fill=guide_legend(title="GBM\nOccurence", ncol=1),
-#          col=guide_legend(title="Time to Recurrence"))
-
-# Pre-surgery vs Post-surgery ----
-## Build gg dataframe for plotting ----
-data_long %>%
-  dplyr::filter(is.na(manifestation_type) | patient_state==0) %>%
-  dplyr::mutate(log2_conc = log2(conc+1)) %>%
-  dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(`Collection Type` = plyr::mapvalues(collection_type, from=c("ctrl", "E1", "E2"), to=c("Ctrl", "pre-\nsurgery", "post-\nsurgery"))) %>%
-  # dplyr::mutate(`Collection Type` = ifelse(is.na(`Collection Type`), "Ctrl", `Collection Type`)) %>%
-  dplyr::mutate(collection_type2 = factor(`Collection Type`, levels = c("Ctrl", "pre-\nsurgery", "post-\nsurgery"))) -> pdata2
-
-## Perform Statistical-Test ----
-comparisons <- matrix(c("Ctrl", "Ctrl", "pre-\nsurgery", "pre-\nsurgery", "post-\nsurgery", "post-\nsurgery"), ncol=2)
-methods <- c("perm.test", "perm.test", "paired.perm.test")
-# methods <- c("perm.test", "perm.test", "perm.test")
-
-stats_frame <- parallel::mclapply(seq_len(nrow(comparisons)), function(iterator){
-  comp <- comparisons[iterator, ]
-  method <- methods[iterator]
-  data_sub <- filter_data(pdata2, comp, method, "collection_type2")
-
-  uni_genes <- data_sub %>% dplyr::pull(gene) %>% unique()
-
-  gene_stat_frame <- parallel::mclapply(split(data_sub, data_sub$gene), function(.d){
-    if (method=="paired.perm.test"){
-      .d %>%
-        dplyr::select(-gene) %>%
-        tidyr::pivot_wider(id_cols = "Number.x", names_from = "collection_type2", values_from = "log2_conc") -> .d_wide
-    } else {
-      .d -> .d_wide
-    }
-    message(sprintf("INFO: Computing %s for %s for comparison %s", method, unique(.d$gene), paste(comp, collapse=", ") ))
-    do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "collection_type2")
-  }, mc.cores = 4) %>%
-    rbindlist() %>%
-    unique()
-}, mc.cores=3) %>% rbindlist()
-
-stat_test_coord <- pdata2 %>%
-  group_by(gene) %>%
-  t_test(log2_conc ~ collection_type2) %>%
-  add_xy_position() %>%
-  dplyr::select(gene, .y., group1, group2, n1, n2, df, y.position, groups, xmin, xmax)
-
-stat_test <- stats_frame %>%
-  dplyr::group_by(group1, group2) %>%
-  adjust_pvalue(method = "fdr") %>%
-  add_significance(p.col="p", output.col = "p.signif",
-                   cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
-                   symbols = c("****", "***", "**", "*", "ns")) %>%
-  merge(stat_test_coord, by=c("gene", ".y.", "group1", "group2")) %>%
-  dplyr::select(-p.adj) %>%
-  dplyr::mutate(p=round(p, 6))
-
-stat_test %>%
-  mutate(y.position = c(7.42688, 7.42688, 7.42688, 8.60988, 8.60988, 8.60988, 10.82388, 10.82388, 10.82388, 13.57988, 13.57988, 13.57988)) -> stat_test
-
-stat_test_pre_v_post_surgery <- stat_test
-
-## Build the plot ----
-ggpaired(data=pdata2, id = "Number.x", x = "collection_type2", y = "log2_conc", fill = "collection_type2", palette = "jco", line.color = "grey", line.size = 0.4) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(length(unique(pdata2$`collection_type2`)), "Dark2"), labels = c("Ctrl", "pre-surgery", "post-surgery")) +
-  # stat_compare_means(paired = TRUE) +
-  # ggnewscale::new_scale_color() +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
-  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
-  # scale_color_manual(label="Mean Conc.", values="red") +
-  # ggtitle("Pre-Surgery vs Post-Surgery") +
-  labs(x="", y="log2( Concentration )") +
-  facet_wrap(~gene, scales = "free_y") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="vertical",) +
-  guides(fill=guide_legend(title="EDTA Blood\nCollection Type", ncol=3),
-         col=guide_legend(title="Healthy Control Baseline")) -> p2
-
-p2 +
-  stat_pvalue_manual(stat_test, label = "{p.signif} p = {p}", step.group.by = "gene", step.increase = 0.15, hide.ns = T, size=4, bracket.size=1) +
-  geom_point(data=data.frame(Manifestation=c("pre-\nsurgery", "pre-\nsurgery", "pre-\nsurgery", "pre-\nsurgery"), gene=c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9" ), y=c(10, 15, 9, 12)),
-             aes(x=Manifestation, y=y, group=gene), alpha=0) -> p2
-
-supp_figure_pre_v_post <- p2
-
-## Build the table ----
-stat_test %>%
-  dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
-  dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
-  dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
-  tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p") %>%
-  gt() %>%
-  tab_header(title = md("**Statistical Comparisons Between Population Groups per Genes**")) %>%
-  cols_label(
-    Comparison = md("**Comparison**"),
-    `Statistical Test` = md("**Statistical Test**"),
-    ASAH1 = md("**ASAH1**"),
-    GPNMB = md("**GPNMB**"),
-    `MMP-9` = md("**MMP-9**"),
-    SYNEMIN = md("**SYNEMIN**"),
-  ) %>% 
-  tab_footnote(
-    footnote = md("Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples."),
-    locations = cells_column_labels("Statistical Test")
-  ) %>% 
-  tab_footnote(
-    footnote = md("Values in cells represent the non-adjusted p-value from the corresponding statiscal test."),
-    locations = cells_column_labels(c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9"))
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(ASAH1),
-      rows = ASAH1 < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(GPNMB),
-      rows = GPNMB < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(`MMP-9`),
-      rows = `MMP-9` < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(SYNEMIN),
-      rows = SYNEMIN < 0.05
-    )
-  ) -> stat_pre_post_surgery_comp_table
-
-# Build a master table ----
-rbind(stat_test_igm_v_rgm %>%
-        dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
-        dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
-        dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
-        tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p"),
-      stat_test_pre_v_post_surgery %>%
-        dplyr::mutate(method = plyr::mapvalues(method, from = c("perm.test", "paired.perm.test"), to = c("Permutation Test", "Paired Permutation Test"))) %>%
-        dplyr::select(Gene=gene, group1, group2, `Statistical Test`=method, p) %>%
-        dplyr::mutate(Comparison = sprintf("%s vs %s", group1, group2)) %>%
-        tidyr::pivot_wider(id_cols = c("Comparison", "Statistical Test"), names_from = c(Gene), values_from = "p")) %>%
-  gt() %>%
-  tab_row_group(
-    label = md("**Control vs Initial GBM vs Recurrent GBM Plasma Samples**"),
-    rows = 1:3
-  ) %>%
-  tab_row_group(
-    label = md("**Control vs Pre-Surgery vs Post-Surgery Plasma Samples**"),
-    rows = 4:6
-  ) %>%
-  tab_header(title = md("**Statistical Comparisons Between Population Groups per Genes**")) %>%
-  cols_label(
-    Comparison = md("**Comparison**"),
-    `Statistical Test` = md("**Statistical Test**"),
-    ASAH1 = md("**ASAH1**"),
-    GPNMB = md("**GPNMB**"),
-    `MMP-9` = md("**MMP-9**"),
-    SYNEMIN = md("**SYNEMIN**"),
-  ) %>% 
-  tab_footnote(
-    footnote = md("Permutation test using t-test statisics for non-paired samples, or paired permutation test using pair t-test statistics for paired samples."),
-    locations = cells_column_labels("Statistical Test")
-  ) %>% 
-  tab_footnote(
-    footnote = md("Values in cells represent the non-adjusted p-value from the corresponding statistical test."),
-    locations = cells_column_labels(c("ASAH1", "SYNEMIN", "GPNMB", "MMP-9"))
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(ASAH1),
-      rows = ASAH1 < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(GPNMB),
-      rows = GPNMB < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(`MMP-9`),
-      rows = `MMP-9` < 0.05
-    )
-  ) %>%
-  tab_style(
-    style = list(
-      cell_fill(color = "#F9E3D6"),
-      cell_text(style = "italic")
-    ),
-    locations = cells_body(
-      columns = c(SYNEMIN),
-      rows = SYNEMIN < 0.05
-    )
-  ) -> stat_comp_table
-
-# Generate plot of rGBM to iGBM Ratios ----
-
-data_long %>%
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::add_count() %>%
-  dplyr::filter(n==2) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>%
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) -> data_long_igm_rgm_ratios
-
-data_long_igm_rgm_ratios %>%
-  dplyr::filter(sample_comparison!="ctrl") %>%
-  dplyr::filter(sample_comparison!="pre-surgery/post-surgery") %>%
-  dplyr::group_by(gene, sample_comparison) %>%
-  dplyr::mutate(myaxis = paste0(sample_comparison, "\n", "n=", n())) %>%
-  ggplot() +
-  # geom_violin(aes(x=as.factor(sample_comparison), y=log2(ratio), fill=as.factor(sample_comparison)), width=0.9) +
-  geom_boxplot(aes(x=as.factor(gene), y=log2(ratio), fill=as.factor(gene)), width=0.3, alpha=0.9) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Dark2")) +
-  # ggnewscale::new_scale_color() +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc+sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc-sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
-  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
-  # scale_color_manual(label="Mean Conc.", values="red") +
-  geom_jitter(aes(x=as.factor(gene), y=log2(ratio), col=as.factor(Number.x)), alpha=0.7) +
-  # ggtitle("iGBM / rGBM Ratios") +
-  labs(x="", y=expression(paste("log2 ", bgroup("(", frac('initial GBM', 'recurrent GBM'), ")"), " Ratio"))) +
-  # facet_wrap(~gene) +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="vertical") +
-  guides(fill="none", col="none") -> p3
-
-supp_figure_ratios <- p3
-
-data_long %>%
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::add_count() %>%
-  dplyr::filter(n==2) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>%
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Second manifestation"]
-      .d_second <- .d[manifestation_type=="Initial manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBM/iGBM", "pre-surgery/post-surgery"))) -> data_long_rgm_igm_ratios
-
-data_long_rgm_igm_ratios %>%
-  dplyr::filter(sample_comparison!="ctrl") %>%
-  dplyr::filter(sample_comparison!="pre-surgery/post-surgery") %>%
-  dplyr::group_by(gene, sample_comparison) %>%
-  dplyr::mutate(myaxis = paste0(sample_comparison, "\n", "n=", n())) %>%
-  ggplot() +
-  # geom_violin(aes(x=as.factor(sample_comparison), y=log2(ratio), fill=as.factor(sample_comparison)), width=0.9) +
-  geom_boxplot(aes(x=as.factor(gene), y=log2(ratio), fill=as.factor(gene)), width=0.3, alpha=0.9) +
-  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Dark2")) +
-  # ggnewscale::new_scale_color() +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc, col="ctrl"), linetype="dashed") +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc+sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
-  # geom_hline(data=mean_ctrl, aes(yintercept=mean_conc-sd_conc, col="ctrl"), linetype="solid", alpha=0.4) +
-  # geom_rect(data=mean_ctrl, aes(xmin=-Inf, xmax=Inf, ymin=mean_conc-sd_conc, ymax=mean_conc+sd_conc), fill="red", alpha=0.1) +
-  # scale_color_manual(label="Mean Conc.", values="red") +
-  geom_jitter(aes(x=as.factor(gene), y=log2(ratio), col=as.factor(Number.x)), alpha=0.7) +
-  # ggtitle("rGBM / iGBM Ratios") +
-  labs(x="", y=expression(paste("log2 ", bgroup("(", frac('recurrent GBM', 'initial GBM'), ")"), " Ratio"))) +
-  # facet_wrap(~gene) +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="vertical") +
-  guides(fill="none", col="none") -> p3
-
-supp_figure_ratios2 <- p3
-
-# Master figure ----
-# subfigure_6 <- ggarrange(p1, p2, p3, ncol = 3)
-subfigure_6 <- p1
+## Save figure ----
+main_fig_a <- p1
 # subfigure_6
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.png", plot = subfigure_6, width = 21, height = 11, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.svg", plot = subfigure_6, width = 21, height = 11, dpi = 200)
-
-subfigure_6_org <- p1_org
-
-p1_avg_ttr
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a_avg.png", plot = p1_avg_ttr, width = 21, height = 11, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a_avg.svg", plot = p1_avg_ttr, width = 21, height = 11, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.png", plot = main_fig_a, width = 11, height = 9, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_a.svg", plot = main_fig_a, width = 11, height = 9, dpi = 200)
+
+supp_figure_1_a <- p1_org
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_1_a_avg.png", plot = supp_figure_1_a, width = 11, height = 9, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_1_a_avg.svg", plot = supp_figure_1_a, width = 11, height = 9, dpi = 200)
 ```
 
-<!-- # Dimensionality Reduction -->
-
+<!-- # Latency and TTR Data for KM plots-->
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
+# LAtency data ----
+latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
+latency_data <- fread(latency_data_file)
+# Clean names
+latency_data <- rstatix::make_clean_names(latency_data)
+latency_data %>%
+  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
+                SYNM.ng.ml = log2(SYNM.ng.ml+1),
+                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
+                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
+  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
+                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
+                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
+                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data
+
+### ASAH1 ----
+fit <- survfit(Surv(Latency..days., status) ~ ASAH1, data = latency_data)
+print(fit)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$ASAH1)
+# Change color, linetype by strata, risk.table color by strata
+surv_p1 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Latency Time (Days)")
+
+surv_p1 <- surv_p1$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_1.png", plot = surv_p1, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_1.svg", plot = surv_p1, width = 11, height = 8.5, dpi = 200)
+
+### SYNM ----
+fit <- survfit(Surv(Latency..days., status) ~ SYNM, data = latency_data)
+print(fit)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$SYNM)
+# Change color, linetype by strata, risk.table color by strata
+surv_p2 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Latency Time (Days)")
+
+surv_p2 <- surv_p2$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_2.png", plot = surv_p2, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_2.svg", plot = surv_p2, width = 11, height = 8.5, dpi = 200)
+
+### GPNMB ----
+fit <- survfit(Surv(Latency..days., status) ~ GPNMB, data = latency_data)
+print(fit)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$GPNMB)
+# Change color, linetype by strata, risk.table color by strata
+surv_p3 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Latency Time (Days)")
+
+surv_p3 <- surv_p3$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_3.png", plot = surv_p3, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_3.svg", plot = surv_p3, width = 11, height = 8.5, dpi = 200)
+
+### MMP9 ----
+fit <- survfit(Surv(Latency..days., status) ~ MMP9, data = latency_data)
+print(fit)
+n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$MMP9)
+# Change color, linetype by strata, risk.table color by strata
+surv_p4 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(1600, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Latency Time (Days)")
+
+surv_p4 <- surv_p4$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_4.png", plot = surv_p4, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_a_4.svg", plot = surv_p4, width = 11, height = 8.5, dpi = 200)
+
+elisa_supp_subfigure_a <- ggarrange(surv_p1 + rremove("ylab") + rremove("xlab"), 
+                                    surv_p2 + rremove("ylab") + rremove("xlab"),
+                                    surv_p3 + rremove("ylab") + rremove("xlab"),
+                                    surv_p4 + rremove("ylab") + rremove("xlab"),
+                                    nrow=1, labels = NULL, align = "hv")
+
+elisa_supp_subfigure_a <- annotate_figure(elisa_supp_subfigure_a, left = textGrob("Survival probability", rot = 90, hjust=0.4, vjust = 0.8, gp = gpar(cex = 1.3)),
+                    bottom = textGrob("Latency Time (Days)", gp = gpar(cex = 1.3)))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_a.png", plot = elisa_supp_subfigure_a, width = 25, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_a.svg", plot = elisa_supp_subfigure_a, width = 25, height = 8.5, dpi = 200)
+
+### ASAH1 + SYN ----
+fit <- survfit(Surv(Latency..days., status) ~ ASAH1 + SYNM, data = latency_data)
+print(fit)
+# Change color, linetype by strata, risk.table color by strata
+surv_p5 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(1600, 0.75), conf.int = F,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
+          ) +
+  labs(x="Latency Time (Days)")  +
+  guides(col=guide_legend(nrow=4,byrow=TRUE))
+
+surv_p5 <- surv_p5$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.6, 0.87))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_c_2.png", plot = surv_p5, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_c_2.svg", plot = surv_p5, width = 11, height = 8.5, dpi = 200)
+
+# survival model with time to recurrence; combined group of all 4 proteins ----
+ttr_surv_data <- merge(latency_data, patient_meta_data[, .(LAB_ID, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `Time-to-reccurence (days)`)], by.x="lab_id", by.y="LAB_ID", all.x=T)
+ttr_surv_data$`Time-to-reccurence (days)`[ ttr_surv_data$`Time-to-reccurence (days)`=="/" ] = ""
+ttr_surv_data$`Time-to-reccurence (days)` <- as.numeric(ttr_surv_data$`Time-to-reccurence (days)`)
+ttr_surv_data <- rstatix::make_clean_names(ttr_surv_data)
+
+### ASAH1 ----
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1, data = ttr_surv_data)
+print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$ASAH1)
+
+ttr_data_km_data_asah1 <- broom::tidy(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p1 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+ttr_surv_p1 <- ttr_surv_p1$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b_1.png", plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b_1.svg", plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
+
+### SYNM ----
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data %>% dplyr::select(-MMP.9.ng.ml))
+print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$SYNM)
+
+ttr_data_km_data_synm <- broom::tidy(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p2 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+ttr_surv_p2 <- ttr_surv_p2$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b_2.png", plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b_2.svg", plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
+
+elisa_main_subfigure_b <- ggarrange(ttr_surv_p1 + rremove("ylab") + rremove("xlab"), 
+                                    ttr_surv_p2 + rremove("ylab") + rremove("xlab"),
+                                    ncol=2, labels = NULL, align = "hv")
+
+elisa_main_subfigure_b <- annotate_figure(elisa_main_subfigure_b, left = textGrob("Time to Recurrence probability", rot = 90, hjust=0.4, vjust = 0.8, gp = gpar(cex = 1.3)),
+                    bottom = textGrob("Time to Recurrence (Days)", gp = gpar(cex = 1.3)))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = elisa_main_subfigure_b, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = elisa_main_subfigure_b, width = 11, height = 8.5, dpi = 200)
+
+### GPNMB ----
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ GPNMB, data = ttr_surv_data)
+print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$GPNMB)
+
+ttr_data_km_data_gpnmb <- broom::tidy(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p3 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+ttr_surv_p3 <- ttr_surv_p3$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b_1.png", plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b_1.svg", plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
+
+### MMP9 ----
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ MMP9, data = ttr_surv_data)
+print(fit)
+n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$MMP9)
+
+ttr_data_km_data_mmp9 <- broom::tidy(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p4 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          palette = "Dark2",
+          legend.title = "",
+          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
+
+ttr_surv_p4 <- ttr_surv_p4$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.8, 0.9))
+
+# ttr_data_km_data <- rbindlist(list(ASAH1=ttr_data_km_data_asah1, SYNM=ttr_data_km_data_synm, GPNMB=ttr_data_km_data_gpnmb, MMP9=ttr_data_km_data_mmp9), idcol = "Protein")
+# 
+# ttr_data_km_data[, strata2 := gsub("^\\w+=(\\w+)$", "\\1", strata)]
+# 
+# ggplot() +
+#   geom_step(data=filter(ttr_data_km_data, Protein %in% c("ASAH1", "SYNM")), aes(x=time, y=estimate, group=strata, col=Protein, linetype=strata2)) +
+#   
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b_2.png", plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b_2.svg", plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
+
+elisa_supp_subfigure_b <- ggarrange(ttr_surv_p3 + rremove("ylab") + rremove("xlab"), 
+                                    ttr_surv_p4 + rremove("ylab") + rremove("xlab"),
+                                    ncol=2, labels = NULL, align = "hv")
+
+elisa_supp_subfigure_b <- annotate_figure(elisa_supp_subfigure_b, left = textGrob("Time to Recurrence probability", rot = 90, hjust=0.4, vjust = 0.8, gp = gpar(cex = 1.3)),
+                    bottom = textGrob("Time to Recurrence (Days)", gp = gpar(cex = 1.3)))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b.png", plot = elisa_supp_subfigure_b, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_b.svg", plot = elisa_supp_subfigure_b, width = 11, height = 8.5, dpi = 200)
+
+### ASAH1 + SYNM ----
+fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1 + SYNM, data = ttr_surv_data)
+print(fit)
+
+# Change color, linetype by strata, risk.table color by strata
+ttr_surv_p5 <- ggsurvplot(fit,
+          pval = TRUE, pval.coord = c(640, 0.75), conf.int = F,
+          risk.table = FALSE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw(), # Change ggplot2 theme
+          legend.title = "",
+          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)\n", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)\n", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)\n", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
+          ) +
+  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
+  guides(col=guide_legend(ncol=1,byrow=TRUE))
+
+ttr_surv_p5 <- ttr_surv_p5$plot +
+  theme(axis.title = element_text(size=16),
+        axis.text = element_text(size=15),
+        legend.text = element_text(size=12),
+        legend.spacing.x = unit(0.25, 'cm'),
+        legend.position = c(0.6, 0.87))
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_c_1.png", plot = ttr_surv_p5, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_2_c_1.svg", plot = ttr_surv_p5, width = 11, height = 8.5, dpi = 200)
+
+elisa_supp_subfigure_c <- ggarrange(ttr_surv_p5 , 
+                                    surv_p5,
+                                    ncol=2, labels = NULL, align = "hv")
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_c.png", plot = elisa_supp_subfigure_c, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_c.svg", plot = elisa_supp_subfigure_c, width = 11, height = 8.5, dpi = 200)
+
+## Final Supp Figure 2 ----
+elisa_supp_figure_2 <- ggarrange(elisa_supp_subfigure_a, ggarrange(elisa_supp_subfigure_b, elisa_supp_subfigure_c, ncol = 2, labels=c("B", "C")),
+          labels = c("A"), nrow = 2)
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure_2.png", plot = elisa_supp_figure_2, width = 27, height = 18, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure_2.svg", plot = elisa_supp_figure_2, width = 27, height = 18, dpi = 200)
+
+```
+
+<!-- # sPLS-DA for conc data and Ratio data -->
 ```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
 # Dimensionality Reduction ----
-## Data wrnagling ----
+## Protein Concentration
+#### Data wrnagling ----
 data_long %>% 
@@
 
-## Only use proteins for prediction ----
+#### Only use proteins for prediction ----
 data_wide %>%
@@
 
-## sPLS-DA ----
+#### sPLS-DA ----
 train_df <- data_wide
@@
 
-## Predict boundary planes of separation ----
+#### Predict boundary planes of separation ----
 background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")
 
-## Prepare data frame for plotting ----
+#### Prepare data frame for plotting ----
 plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=as.factor(train_df[, 1]), as.data.frame(MyResult.splsda$variates$X))
@@
 
-## Get average trajectories of iGBM to rGBM in low dimension ----
+#### Get average trajectories of iGBM to rGBM in low dimension ----
 plsda_data[patient_id!="Ctrl" ] %>%
@@
 
-## Prepare decision boundaries dataframe fo plotting ----
+#### Prepare decision boundaries dataframe fo plotting ----
 for(i in 1:length(background))
@@
 
-## Build plot ----
+#### Build plot ----
 ### Average Trajectories 
@@
                           length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  geom_label(label="Ctrl", 
+             aes(x=-2, y=2), fill="#C2C2C2", color="white", size=6) +
+  geom_label(label="iGBM", 
+             aes(x=4, y=0), fill="#388ECC", color="white", size=6) +
+  geom_label(label="rGBM", 
+             aes(x=-2.5, y=-1), fill="#F68B33", color="white", size=6) +
+  geomtextpath::geom_labelpath(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id), label="Avg. iGBM to\nrGBM Trajectory",
+            arrow = arrow(type = "closed",
+                          length=unit(0.1, "inches")), size=5, alpha=0.8, 
+             fill="red", color="white", size=6) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
   guides(shape="none",
+         color_new="none",
          col_org=guide_legend(title="g", ncol=1),
-         col=guide_legend(title="Trajectory", ncol=1),
+         col="none",
          col_new=guide_legend(title="g", ncol=1))
@@
 
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-
-### Individual trajectories
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_b.png", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_b.svg", plot = splsda_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+
+## Ratios ----
+#### Get ratio data ----
+data_long_tmp <- median_centre_recurrent_data(data_long)
+train_df <- get_ratio_data(data_long_tmp, already_log_transformed = T)
+#### sPLS-DA ----
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7)], train_df$class, keepX = c(4,4))
+background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")
+
+plsda_data <- cbind( patient_id=train_df$Number.x, class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+plsda_data$patient_id <- as.factor(plsda_data$patient_id)
+plsda_data %>%
+  dplyr::arrange(class) -> plsda_data
+plsda_data <- as.data.table(plsda_data)
+
+plsda_data %>%
+  dplyr::group_by(class) %>%
+  dplyr::add_count() %>%
+  dplyr::group_by(class, n) %>%
+  dplyr::summarise(avg_comp1=sum(comp1)/n,
+                   avg_comp2=sum(comp2)/n) %>%
+  dplyr::ungroup() %>%
+  unique() %>%
+  dplyr::mutate(patient_id="avg_all") %>%
+  as.data.table() -> average_comp
+
+slope <- (average_comp$avg_comp2[2] - average_comp$avg_comp2[1])/(average_comp$avg_comp1[2] - average_comp$avg_comp1[1])
+b = -0.25
+
+for(i in 1:length(background))
+{
+  if(!is.null(background[[i]]))
+    background[[i]]=data.frame(id=i,col=names(background)[i],
+                               background[[i]])
+}
+
+background = do.call(rbind,background)
+
 p <- ggplot() +
@@
                                      fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
-  scale_fill_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
+  scale_fill_manual(values=c("#388ECC", "#F68B33")) +
   # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
@@
   scale_shape_manual(values=c(15, 16, 17)) +
-  scale_color_manual(values=c( "#C2C2C2", "#388ECC", "#F68B33")) +
-  ggnewscale::new_scale_color() +
-  geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
-            arrow = arrow(type = "closed",
-                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
+                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
+  ) +
+  # ggnewscale::new_scale_color() +
+  # ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=paste(patient_id, `Time-to-reccurence (days)`, sep=": "))) +
+  # ggnewscale::new_scale_color() +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+  #           arrow = arrow(type = "closed", ends='both',
+  #                         length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  geom_label(label=expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")), 
+             aes(x=min(plsda_data[class=="rGBM"]$comp1)+0.5, y=min(plsda_data[class=="rGBM"]$comp2)+0.01), fill="#F68B33", color="white", size=6) +
+  geom_label(label=expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")), 
+             aes(x=max(plsda_data[class=="iGBM"]$comp1)-0.5, y=max(plsda_data[class=="iGBM"]$comp2)-0.01), fill="#388ECC", color="white", size=6) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
   guides(shape="none",
-         col_org=guide_legend(title="g", ncol=1),
-         col=guide_legend(title="Patient ID\nTrajectory", ncol=2),
-         col_new=guide_legend(title="g", ncol=1))
-
-splsda_ind_traj_fig <- p
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig2_subfigure_a.png", plot = splsda_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig2_subfigure_a.svg", plot = splsda_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-
-```
-
-<!-- # Ratios -->
-
-```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
-## Data wrangling ----
-### iGBM over rGBM ratio
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
-
-### rGBM over iGBM
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_second$conc / .d_initial$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
-  dplyr::filter(grepl(";", agg_lab_id)) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
-
-### Prepare final table with ratios
-pdata %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="iGBM") -> pdata
-
-pdata2 %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="rGBM") -> pdata2
-
-pdata <- rbindlist(list(pdata, pdata2))
-
-pdata <- merge(pdata, meta_dt, by="Number.x")
-
-### Pivot Table to wide
-pdata %>%
-  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, `Time-to-reccurence (days)`, `Latency (days)`, `Type of resection`) -> data_wide
-
-# Prepare final table for analysis
-train_df <- data_wide
-train_df <- na.omit(train_df)  
-train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
-# Make Valid Column Names 
-colnames(train_df) <- make.names(colnames(train_df))
-train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.table(train_df)
-train_df$group=0
-
-## Plot Protein Ratios Against Each Other ----
-p_pa_pb_ratio_scatter1 <- ggplot(train_df, aes(x=ASAH1, y=GPNMB, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="ASAH1 Ratio", y="GPNMB Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter2 <- ggplot(train_df, aes(x=ASAH1, y=MMP.9, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="ASAH1 Ratio", y="MMP-9 Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter3 <- ggplot(train_df, aes(x=ASAH1, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="ASAH1 Ratio", y="SYNEMIN Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter4 <- ggplot(train_df, aes(x=MMP.9, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="MMP-9 Ratio", y="SYNEMIN Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter5 <- ggplot(train_df, aes(x=GPNMB, y=SYNEMIN, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="GPNMB Ratio", y="SYNEMIN Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter6 <- ggplot(train_df, aes(x=GPNMB, y=MMP.9, col=class, shape=as.factor(Number.x))) +
-  geom_point(size=3) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  labs(x="GPNMB Ratio", y="MMP-9 Ratio") +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "bottom", legend.box="horizontal",) +
-  guides(shape=guide_legend(title="Patient ID", ncol=5),
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Class", ncol=2, byrow = T))
-
-p_pa_pb_ratio_scatter <- ggpubr::ggarrange(p_pa_pb_ratio_scatter1, p_pa_pb_ratio_scatter2, p_pa_pb_ratio_scatter3, 
-                                           p_pa_pb_ratio_scatter4, p_pa_pb_ratio_scatter5, p_pa_pb_ratio_scatter6,
-                                           ncol=3, nrow=2, common.legend = TRUE, legend="bottom")
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_a.png", plot = p_pa_pb_ratio_scatter, width = 15, height = 9, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_a.svg", plot = p_pa_pb_ratio_scatter, width = 15, height = 9, dpi = 200)
-
-## sPLS-DA of Ratios ----
-
-MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
+         col="none")
+  # guides(shape="none",
+  #        colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+  #        colour_new='none',
+  #        col=guide_legend(title="Trajectory", ncol=1))
+
+splsda_ratio_avg_traj_fig <- p
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+
+## Ratios unique samples ----
+#### Get ratio data ----
+data_long_tmp <- median_centre_recurrent_data(data_long)
+ratio_data <- get_ratio_data(data_long_tmp, already_log_transformed = T)
+train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=NULL, unique_random_train_patient_per_class=T)
+train_df <- rbindlist(list(train_test$train_df, train_test$test_df), use.names = T)
+#### sPLS-DA ----
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7)], train_df$class, keepX = c(4,4))
 background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")
 
-plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+plsda_data <- cbind( patient_id=train_df$Number.x, class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
 plsda_data$patient_id <- as.factor(plsda_data$patient_id)
@@
   ggnewscale::new_scale_color() +
-  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
-            arrow = arrow(type = "closed", ends='both',
-                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=paste0("id: ", patient_id))) +
+  # ggnewscale::new_scale_color() +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+  #           arrow = arrow(type = "closed", ends='both',
+  #                         length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  geom_label(label=expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")), 
+             aes(x=min(background$Var1)+0.8, y=min(background$Var2)+0.2), fill="#F68B33", color="white", size=6) +
+  geom_label(label=expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")), 
+             aes(x=max(background$Var1)-0.8, y=max(background$Var2)-0.2), fill="#388ECC", color="white", size=6) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
   guides(shape="none",
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),
-         col=guide_legend(title="Trajectory", ncol=1))
-
-splsda_ratio_avg_traj_fig <- p
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.png", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_b.svg", plot = splsda_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-
-p <- ggplot() +
-  geom_polygon(data = background,aes(x=Var1, y=Var2,
-                                     fill = col), inherit.aes = FALSE, show.legend=FALSE, alpha=0.2) +
-  scale_fill_manual(values=c( "#388ECC", "#F68B33")) +
-  # geom_point(data=plsda_data[patient_id=="Ctrl"], aes(x=comp1, y=comp2, shape=class), col="black", size=3) +
-  # geom_point(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
-  geom_point(data=plsda_data, aes(x=comp1, y=comp2, col=class, shape=class), size=3) +
-  scale_shape_manual(values=c(15, 16, 17)) +
-  scale_color_manual(values=c( "#388ECC", "#F68B33"), labels=c(expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")),
-                                                               expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))
-  ) +
-  ggnewscale::new_scale_color() +
-  geom_path(data=plsda_data[patient_id!="Ctrl"], aes(x=comp1, y=comp2, group = patient_id, col=patient_id),
-            arrow = arrow(type = "closed", ends='both',
-                          length=unit(0.1, "inches")), size=1.5, alpha=0.8) +
-  labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "right", legend.box="vertical",) +
-  guides(shape="none",
-         colour_new=guide_legend(title="Class", ncol=1, byrow = T),         
-         col=guide_legend(title="Patient ID\nTrajectory", ncol=2))
-
-splsda_ratio_ind_traj_fig <- p
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.png", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure3_subfigure_c.svg", plot = splsda_ratio_ind_traj_fig, width = 11, height = 8.5, dpi = 200)
-
-## Time to Recurrence using Ratios ----
-train_df_Sub <- train_df[class=="rGBM"] %>%
-  dplyr::arrange(Time.to.reccurence..days.)
-
-train_df_Sub %>%
-  tidyr::pivot_longer(cols = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9"), names_to = "gene", values_to = "ratio") %>%
-  dplyr::group_by(gene, Time.to.reccurence..days.) %>%
-  # dplyr::summarise(ratio=mean(ratio)) %>%
-  dplyr::mutate(Time.to.reccurence..days. = as.numeric(Time.to.reccurence..days.)) %>%
-  dplyr::ungroup() -> train_df_Sub_long
-
-p_ratio_ttr <- ggplot(train_df_Sub_long) +
-  geom_point(aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene, col=as.factor(Number.x)), size=3) +
-  ggnewscale::new_scale("color") +
-  # geom_line(aes(x=as.factor(Time.to.reccurence..days.), y=ratio, group=gene, col=gene)) +
-  geom_smooth(aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene, col=gene), method = "lm", alpha=0.2) +
-  stat_cor(method = "pearson", aes(x=as.numeric(Time.to.reccurence..days.), y=ratio, group=gene)) +
-  labs(x="Time to Recurrence (Days)", y=expression(paste("log2 ", bgroup("(", frac('recurrent GBM', 'initial GBM'), ")"), " Ratio"))) +
-  theme_classic() +
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15),
-        legend.position = "right", legend.box="vertical") +
-  guides(col=guide_legend(title="Gene"), 
-         colour_new=guide_legend(title="Patient ID", ncol = 2)) +
-  facet_wrap(~gene)
-
-# train_df_Sub <- train_df %>%
-#   dplyr::select(Number.x, Time.to.reccurence..days., class, ASAH1, SYNEMIN, GPNMB, MMP.9) %>%
-#   tidyr::pivot_wider(id_cols = c("Number.x", "Time.to.reccurence..days."), names_from = c("class"), values_from = c("ASAH1", "SYNEMIN", "GPNMB", "MMP.9")) %>%
-#   dplyr::arrange(Time.to.reccurence..days.)
-#   train_df_Sub <- na.omit(train_df_Sub)
-# 
-# train_df_Sub$Number.x <- as.factor(train_df_Sub$Number.x)
-# train_df_Sub$Time.to.reccurence..days. <- as.numeric(train_df_Sub$Time.to.reccurence..days.)
-# lm_ttr <- lmerTest::lmer(Time.to.reccurence..days.
-~ ASAH1_iGBM + ASAH1_rGBM + SYNEMIN_iGBM + SYNEMIN_rGBM + GPNMB_iGBM + GPNMB_rGBM + MMP.9_iGBM + MMP.9_rGBM + (1 | Type.of.resection), data=train_df_Sub )
-# summary(lm_ttr)
-# anova(lm_ttr)
-# 
-# lm_ttr <- lm(Time.to.reccurence..days. ~ ASAH1_iGBM + ASAH1_rGBM + SYNEMIN_iGBM + SYNEMIN_rGBM + GPNMB_iGBM + GPNMB_rGBM + MMP.9_iGBM + MMP.9_rGBM, data=train_df_Sub )
-# summary(lm_ttr)
-# car::Anova(lm_ttr, type="III")
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_e.png", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_e.svg", plot = p_ratio_ttr, width = 15, height = 9, dpi = 200)
-
-```
-
-# Subsetting and hold out patient for Validating Ratios
-
-I still think there is value in usiing the ratios, so I performed the sPLS-DA on the ratio data again, but instead of using all the ratios and inverse ratios from all patients, I do the following:
-
-1. Randomly sample a single patient to use as a test patient, to make predictions on
-2. Randomly sample n random patients from remaining patients and compute iGBM/rGBM ratios
-3. For the rest of the patients I compute rGBM/iGBM ratios
-
-This results in the following data used in the sPLS-DA model
-
-**Training Data**
-
-| Patient ID | Class | ASAH1      | SYNMEIN    | GPNMB      | MMP-9      |
-|------------|-------|------------|------------|------------|------------|
-| 6          | iGBM  | -4.3219281 | 2.6629410  | 0.3162515  | 3.8886097  |
-| 10         | rGBM  | 2.2103599  | -3.8779037 | 2.5628539  | -3.1538943 |
-| 16         | rGBM  | 1.6956066  | 0.8506559  | 0.1659252  | -1.9814540 |
-| 23         | rGBM  | -0.0257631 | -3.6121574 | 1.6151451  | -1.7166787 |
-| 26         | iGBM  | 1.7744403  | 0.4268049  | 4.2786146  | -1.5475206 |
-| 29         | rGBM  | 3.3645724  | 0.4959143  | -0.6652013 | 0.5074684  |
-| 31         | iGBM  | -1.6599246 | -2.8570919 | -1.9497226 | -1.3115642 |
-| 34         | iGBM  | -0.1768778 | 0.7865737  | 2.1875247  | 1.1604031  |
-
-**Test Data**
-
-| Patient ID | Class | ASAH1      | SYNMEIN    | GPNMB      | MMP-9      |
-|------------|-------|------------|------------|------------|------------|
-| 35          | rGBM  | -0.02599521 | -4.115453  | -2.892362  | -1.679008  |
-
-This way we do not duplicate patient IDs (i.e.
-having iGBM/rGBM and rGBM/iGBM for the same patient) and we have a single patient we can make a prediction on to see where in the PLS-DA dimension they would land.
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=21, include=FALSE}
-data_long2 <- data_long
-
-## Data wrangling ----
-# Get ratio data
-message("Get Ratio Data")
-ratio_data <- get_ratio_data(data_long2)
-message("Get Test Train Data")
-train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=NULL, unique_random_train_patient_per_class=T)
-train_df <- train_test$train_df
-test_df <- train_test$test_df
-test_patients <- test_df$Number.x
-
-## sPLS-DA of Ratios ----
-message("Perform sPLS-DA")
-MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4))
-
-pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
-message("Perform background predictions")
-background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist", xlim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))), ylim = c(min(c(MyResult.splsda$variates$X, pred_test$variates)), max(c(MyResult.splsda$variates$X, pred_test$variates))))
-message("Prepare data for plotting")
-plsda_data <- cbind( patient_id=train_df  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
+         colour_new='none',
+         col="none")
+  # guides(shape="none",
+  #        colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+  #        colour_new='none',
+  #        col=guide_legend(title="Trajectory", ncol=1))
+
+splsda_ratio_unique_patients <- p
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_d.png", plot = splsda_ratio_unique_patients, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_d.svg", plot = splsda_ratio_unique_patients, width = 11, height = 8.5, dpi = 200)
+
+## Shuffled ratio data ----
+#### Get ratio data and shuffle labels----
+data_long_tmp <- median_centre_recurrent_data(data_long)
+train_df <- get_ratio_data(data_long_tmp, already_log_transformed = T)
+org_class <- train_df$class
+#### I set the seed for the purpose of downstream plotting of the legends inside the plot
+set.seed(2018)
+train_df$class <- sample(train_df$class)
+#### sPLS-DA ----
+MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7)], train_df$class, keepX = c(4,4))
+background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")
+
+plsda_data <- cbind( patient_id=train_df$Number.x, class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
 plsda_data$patient_id <- as.factor(plsda_data$patient_id)
@@
 
-message("Gnerate sPLS-DA Plot")
 p <- ggplot() +
@@
   ) +
-  ggnewscale::new_scale_color() +
-  geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
-            arrow = arrow(type = "closed", ends='both',
-                          length=unit(0.1, "inches")), size=2, alpha=0.8) +
-  ggnewscale::new_scale_color() +
-  geom_point(data=data.frame(x=pred_test$variates[,1], y=pred_test$variates[,2]), aes(x=x, y=y, col="black"), size=3, shape=3) +
-  scale_color_manual(values=c( "black"), labels=c(expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")))) +
+  # ggnewscale::new_scale_color() +
+  # ggrepel::geom_label_repel(data=merge(plsda_data %>% dplyr::mutate(patient_id = as.numeric(as.character(patient_id))), unique(data_long[manifestation_type!='' & patient_state==1 & `Time-to-reccurence (days)`!='', .(Number.x, `Time-to-reccurence (days)`)]), by.x='patient_id', by.y="Number.x", all.x=T), aes(x=comp1, y=comp2, col=`Time-to-reccurence (days)`, label=paste0("id: ", patient_id))) +
+  # ggnewscale::new_scale_color() +
+  # geom_path(data=average_comp, aes(x=avg_comp1, y=avg_comp2, group = patient_id, col="Avg. iGBM\nto rGBM\nTrajectory"),
+  #           arrow = arrow(type = "closed", ends='both',
+  #                         length=unit(0.1, "inches")), size=2, alpha=0.8) +
+  geom_label(label=expression(paste(frac('recurrent GBM', 'initial GBM'), " Ratio")), 
+             aes(x=min(background$Var1)+0.8, y=min(background$Var2)+0.2), fill="#F68B33", color="white", size=6) +
+  geom_label(label=expression(paste(frac('initial GBM', 'recurrent GBM'), " Ratio")), 
+             aes(x=max(background$Var1)-0.8, y=max(background$Var2)-0.2), fill="#388ECC", color="white", size=6) +
   labs(x=sprintf("X-variate 1: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[1]*100)), y =sprintf("X-variate 2: %s%% explained variance", round(MyResult.splsda$prop_expl_var$X[2]*100))) +
@@
   guides(shape="none",
-         colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
-         colour_new=guide_legend(title="Trajectory", ncol=1, byrow = T),
-         col=guide_legend(title="Prediction on\nHeld Out Patients", ncol=1))
-
-splsda_better_ratio_avg_traj_fig <- p
-splsda_better_ratio_avg_traj_fig
-ggsave("../../figures/elisa_plasma_proteomics/elisa_splada_ratio_random.png", plot = splsda_better_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_splada_ratio_random.svg", plot = splsda_better_ratio_avg_traj_fig, width = 11, height = 8.5, dpi = 200)
+         colour_new='none',
+         col="none")
+  # guides(shape="none",
+  #        colour_new_new=guide_legend(title="Class", ncol=1, byrow = T),
+  #        colour_new='none',
+  #        col=guide_legend(title="Trajectory", ncol=1))
+
+splsda_ratio_shuffled_patients <- p
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_e.png", plot = splsda_ratio_shuffled_patients, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_subfigure_e.svg", plot = splsda_ratio_shuffled_patients, width = 11, height = 8.5, dpi = 200)
 
@@
 
-# Comparison of Using All patients or Subsetting for Unique Patients per Class
-
-Lets compare to see if it makes a difference whether we use all patients for both iGBM/rGBM and rGBM/iGBM classes vs only using a unique set of patients for iGBM/rGBM and rGBM/iGBM (i.e.
-no duplicate patient in each class)
-
-1. I perform a leave one out cross validation, leaving one patient out and using the rest of the data for training. 
-2. The left out patient gets using for assessing model performance accuracy. 
-3. The mean accuracy of each left out sample is computed
-4. Steps 1-3 is repeated 100 times to assess the variation in mean accuracy of the LOOCV
-5. 1-4 is performed for both cases of using all patients and using unique subset patients
-
-```{r echo=FALSE, fig.height=8.5, fig.width=11, message=FALSE, warning=FALSE}
-# Get ratio data
-ratio_data <- get_ratio_data(data_long2)
-# Get unique set of patient ids
-patient_ids <- unique(ratio_data$Number.x)
-
-plan(multisession(workers = 8))
-
-# Compute mean accuracy when using unique patients per class for 100 repeats
-mean_acc_rep_unique <- future.apply::future_replicate(100, splsda_loocv_perf(ratio_data, patient_ids, unique_random_train_patient_per_class=TRUE))
-# hist(mean_acc_rep_unique)
-message("Unique Patient Class Split: 95% Confidence Interval:")
-print(c(quantile(mean_acc_rep_unique, .025), quantile(mean_acc_rep_unique, .975)))
-
-# Compute mean accuracy when using all patients per class for 100 repeats
-mean_acc_rep_all <- future.apply::future_replicate(100, splsda_loocv_perf(ratio_data, patient_ids, unique_random_train_patient_per_class=FALSE))
-# hist(mean_acc_rep_all)
-message("All Patients per Class: 95% Confidence Interval:")
-print(c(quantile(mean_acc_rep_all, .025), quantile(mean_acc_rep_all, .975)))
-
-plan(sequential)
-
-df_plot <- rbind(data.frame(method="Unique Patient Class Split", mean_acc=mean_acc_rep_unique), data.frame(method="All Patients per Class", mean_acc=mean_acc_rep_all))
-df_plot$method <- factor(df_plot$method, levels = c("All Patients per Class", "Unique Patient Class Split"))
-
-ggplot(df_plot, aes(x=mean_acc, fill=method)) +
-  geom_histogram(binwidth = 0.15, colour = "black") +
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
-  scale_fill_brewer(palette = "Dark2") +
-  stat_bin(binwidth = 0.15, aes(label=..count..), vjust=-0.5, geom = "text") +
-  facet_wrap(~method) +
-  labs(x="Mean Accuracy (LOOCV)", y="Count") +
-  ggtitle("Mean Accuracy of LOOCV for 100 Repeats") +
-  theme_classic() + 
-  theme(plot.title = element_text(hjust = 0.5, size=17),
-        axis.title = element_text(size = 17),
-        axis.text = element_text(size=16),
-        strip.text = element_text(size=16),
-        legend.title = element_text(size=16),
-        legend.text = element_text(size=15)) 
-```
-
-<!-- # ROC Prediction Analysis -->
-
+<!-- # ROC prediction analyses -->
 ```{r fig.height=8, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}
@@
 
-# iGBM vs rGBM ratio  with clinical data ----
-# data_long %>% 
-#   dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-#   dplyr::group_by(gene, Number.x) %>%
-#   dplyr::group_map(function(.d, ...){
-#     .d <- as.data.table(.d)
-#     .d %>% 
-#       dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-#       dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-#                        agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-#     # print(.d)
-#     if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-#       .d_initial <- .d[manifestation_type=="Initial manifestation"]
-#       .d_second <- .d[manifestation_type=="Second manifestation"]
-#       tmp$ratio <- .d_initial$conc / .d_second$conc
-#     } else if (nrow(.d)==2) {
-#       .d_initial <- .d[collection_type=="E1"]
-#       .d_second <- .d[collection_type=="E2"]
-#       tmp$ratio <- .d_initial$conc / .d_second$conc
-#     } else {
-#       tmp$ratio <- .d$conc
-#     }
-#     return(tmp)
-#   }, .keep = TRUE) %>%
-#   rbindlist() %>%
-#   dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
-#   dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-#   dplyr::mutate(log2_conc = log2(ratio)) %>%
-#   # dplyr::group_by(gene, Number.x) %>%
-#   # dplyr::add_count() %>%
-#   # dplyr::filter(n==2) %>%
-#   dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
-# 
-# data_long %>% 
-#   dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-#   dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
-#   dplyr::group_by(gene, Number.x) %>%
-#   dplyr::group_map(function(.d, ...){
-#     .d <- as.data.table(.d)
-#     .d %>% 
-#       dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-#       dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-#                        agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-#     # print(.d)
-#     if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-#       .d_initial <- .d[manifestation_type=="Initial manifestation"]
-#       .d_second <- .d[manifestation_type=="Second manifestation"]
-#       tmp$ratio <- .d_second$conc / .d_initial$conc
-#     } else if (nrow(.d)==2) {
-#       .d_initial <- .d[collection_type=="E1"]
-#       .d_second <- .d[collection_type=="E2"]
-#       tmp$ratio <- .d_initial$conc / .d_second$conc
-#     } else {
-#       tmp$ratio <- .d$conc
-#     }
-#     return(tmp)
-#   }, .keep = TRUE) %>%
-#   rbindlist() %>%
-#   dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
-#   dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
-#   dplyr::filter(grepl(";", agg_lab_id)) %>%
-#   dplyr::mutate(log2_conc = log2(ratio)) %>%
-#   # dplyr::group_by(gene, Number.x) %>%
-#   # dplyr::add_count() %>%
-#   # dplyr::filter(n==2) %>%
-#   dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
-# 
-# pdata %>%
-#   dplyr::select(Number.x, gene, log2_conc) %>%
-#   tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-#   dplyr::mutate(class="iGBM") -> pdata
-# 
-# pdata2 %>%
-#   dplyr::select(Number.x, gene, log2_conc) %>%
-#   tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-#   dplyr::mutate(class="rGBM") -> pdata2
-# 
-# 
-# pdata <- rbindlist(list(pdata, pdata2))
-# 
-# pdata <- merge(pdata, patient_meta_data[, .(Number.x, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `Time-to-reccurence (days)`)], by="Number.x", all.x=T)
-# 
-# ## Pivot Table
-# pdata %>%
-#   dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`, Sex, `Tumor localization`, Sex, `Age at surgery (years)`, `Type of resection`, `Time-to-reccurence (days)`) -> data_wide
-# # data_wide %>%
-# #   dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
-# 
-# ## Prediction
-# train_df <- data_wide
-# train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
-# train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
-# ## Make Valid Column Names 
-# colnames(train_df) <- make.names(colnames(train_df))
-# train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
-# colnames(train_df) <- make.names(colnames(train_df))
-# train_df <- as.data.frame(train_df)
-# pred_obj_1.6 <- run_predictions(train_df, pred_col_prob = "rGBM")
-# 
-# ## ROC Figure 
-# g <- ggplot(pred_obj_1.6$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
-#   geom_roc(n.cuts=0) + 
-#   scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
-#   coord_equal() +
-#   style_roc(x="1 - Specificity", y="Sensitivity")
-# 
-# auc_tbl_1.6 <- as.data.table(calc_auc(g))
-# 
-# g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.6[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[1])
-# g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.6[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[2])
-# g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.6[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[3])
-# g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.6[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.6$master_lift$liftModelVar)), "Dark2")[4])
-# subfigure_1.6 <- g
-# 
-# pred_obj_1.6$roc_data %>%
-#   dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
-#   dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
-#   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.6_roc_data
-# 
-# ## Merged Plot
-# pred_obj_1.6$roc_data %>%
-#   dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
-#   dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
-#   dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.6_roc_data
+# iGBM vs rGBM with clinical data Only ----
+data_long %>% 
+  dplyr::group_by(gene, Number.x) %>%
+  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
+  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
+  dplyr::mutate(log2_conc = log2(conc+1)) %>%
+  dplyr::filter(manifestation_type=="Second manifestation" | manifestation_type=="Initial manifestation") %>%
+  dplyr::mutate(class = plyr::mapvalues(manifestation_type, from=c("Initial manifestation","Second manifestation"), to=c("iGBM", "rGBM"))) -> pdata
+
+## Pivot Table
+pdata %>%
+  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
+  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
+  dplyr::ungroup() -> data_wide
+
+## Only use proteins for prediction
+data_wide %>%
+  mutate(n = row_number()) %>%
+  dplyr::select(Number.x, class, n,  Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`) -> data_wide
+# data_wide %>%
+  # dplyr::mutate(`Time-to-reccurence (days)` = ifelse(class=="iGBM", 0, `Time-to-reccurence (days)`)) -> data_wide
+
+## Prediction
+train_df <- data_wide
+train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
+colnames(train_df) <- make.names(colnames(train_df))
+train_df <- as.data.frame(train_df)
+# train_df_proteins <- median_centre_data(train_df[, c(1:5)])
+# train_df <- cbind(train_df_proteins, train_df[, c(6:9)])
+pred_obj_1.41 <- run_predictions(train_df, pred_col_prob = "rGBM")
+
+# coef(pred_obj_1.5$model$finalModel, pred_obj_1.5$model$bestTune$lambda)
+# coef(pred_obj_1.4$model$finalModel, pred_obj_1.4$model$bestTune$lambda)
+
+## ROC Figure 
+g <- ggplot(pred_obj_1.41$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) + 
+  geom_roc(n.cuts=0) + 
+  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_1.4$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
+  coord_equal() +
+  style_roc(x="1 - Specificity", y="Sensitivity")
+
+auc_tbl_1.41 <- as.data.table(calc_auc(g))
+
+g <- g + annotate("text", x=0.70, y=0.45, label=paste("AUC =", round(auc_tbl_1.41[Model=="Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[1])
+g <- g + annotate("text", x=0.70, y=0.35, label=paste("AUC =", round(auc_tbl_1.41[Model=="Elastic Logit"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[2])
+g <- g + annotate("text", x=0.70, y=0.25, label=paste("AUC =", round(auc_tbl_1.41[Model=="KNN"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[3])
+g <- g + annotate("text", x=0.70, y=0.15, label=paste("AUC =", round(auc_tbl_1.41[Model=="RF"]$AUC, 4)), color=RColorBrewer::brewer.pal(length(unique(pred_obj_1.41$master_lift$liftModelVar)), "Dark2")[4])
+subfigure_1.41 <- g
+
+pred_obj_1.41$roc_data %>%
+  dplyr::rename(ctrl=iGBM, GBM=rGBM) %>%
+  dplyr::mutate(pred=plyr::mapvalues(pred, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) %>%
+  dplyr::mutate(obs=plyr::mapvalues(obs, from=c("iGBM", "rGBM"), to=c("ctrl", "GBM"))) -> tmp_pred_obj_1.41_roc_data
 
@@
 
-table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 13)
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 15)
 auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
@@
   style_roc(x="1 - Specificity", y="Sensitivity") +
-  annotation_custom(auc_grob, xmin=0.25, xmax=0.95, ymin=0.10, ymax=0.30) +
+  annotation_custom(auc_grob, xmin=0.25, xmax=0.985, ymin=0.10, ymax=0.30) +
   theme(axis.title = element_text(size=16),
@@
 
-table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))))
+table_theme <- ttheme_minimal(core = list(fg_params=list(col=(RColorBrewer::brewer.pal(5, "Dark2")))), base_size = 15)
 auc_grob <- tableGrob(auc_annotation_tbl, rows=NULL, theme = table_theme)
@@
         legend.title = element_text(size=14),
-        legend.text = element_text(size=13))
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_roc_controls.png", plot = g, width = 11, height = 8.5, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_roc_controls.svg", plot = g, width = 11, height = 8.5, dpi = 200)
+        legend.text = element_text(size=13)) +
+  guides(col="none")
+
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig1_c.png", plot = g, width = 11, height = 8.5, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_fig_c.svg", plot = g, width = 11, height = 8.5, dpi = 200)
 roc_controls <- g
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-# Generate table of prediction metrics ----
-pred_col_prob="GBM"
-
-pred_mods_dt <- rbindlist(list(`Ctrl vs iGBM`=pred_obj_1.1$model$pred, 
-                               `Ctrl vs rGBM`=pred_obj_1.2$model$pred, 
-                               `iGBM vs rGBM protein`=pred_obj_1.5$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM"))),
-                               `iGBM vs rGBM protein + clinical`=pred_obj_1.4$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM"))),
-                               `iGBM vs rGBM protein ratios`=pred_obj_1.3$model$pred %>% dplyr::rename(ctrl=iGBM, GBM=rGBM) %>% dplyr::mutate(pred=plyr::mapvalues(pred, c("iGBM", "rGBM"), c("ctrl", "GBM"))) %>% dplyr::mutate(obs=plyr::mapvalues(obs, c("iGBM", "rGBM"), c("ctrl", "GBM")))), idcol = "Comparison")
-
-conf_mats <-lapply(split(pred_mods_dt[, .(obs, pred)], f=pred_mods_dt$Comparison), function(.d){
-  caret::confusionMatrix(.d$obs, .d$pred, positive=pred_col_prob)
-})
-metric_table <- rbindlist(lapply(conf_mats, function(x){data.frame(metric=names(x$byClass), values=x$byClass)}), idcol = "Comparison")
-
-metric_table %>%
-  dplyr::group_by(Comparison, metric) %>% 
-  dplyr::filter(metric %in% c("Balanced Accuracy", "Sensitivity", "Specificity", "Pos Pred Value", "Neg Pred Value")) %>%
-  dplyr::mutate(values = sprintf("%s%%", round(values*100, 2))) %>%
-  dplyr::select(Comparison, metric, values) -> metric_table
-
-metric_table %>%
-  tidyr::pivot_wider(id_cols="Comparison", names_from = metric, values_from = values) -> performance_dt
-
-performance_dt %>% 
-  dplyr::arrange(Comparison) %>%
-  ungroup() %>%
-  gt() %>%
-  gt::tab_header(md("**Logistic Regression with Elastic Net Regularization Performance Metrics**")) %>%
-  cols_label(.list = list(Comparison=md("**Comparison**"), `Balanced Accuracy`=md("**Balanced Accuracy**"), Sensitivity=md("**Sensitivity**"), Specificity=md("**Specificity**"), `Pos Pred Value`=md("**Pos Pred Value**"), `Neg Pred Value`=md("**Neg Pred Value**"))) -> logit_net_comp_perf_met_table
+# Main Figure
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
+
+main_figure <- ggarrange(
+ggarrange(main_fig_a, elisa_main_subfigure_b, labels=c("A", "B"), ncol = 2),
+ggarrange(splsda_ratio_avg_traj_fig, subfigure_1, labels=c("C", "D"), ncol = 2),
+nrow=2
+)
+main_figure
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_figure.png", plot = main_figure, width = 23, height = 18, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_main_figure.svg", plot = main_figure, width = 23, height = 18, dpi = 200)
 ```
 
-<!-- # Prediction for Ctrl vs iGBM -->
-```{r fig.height=8, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}
-# iGBM vs Ctrl protein conc ----
-data_long %>% 
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
-  dplyr::mutate(log2_conc = log2(conc+1)) %>%
-  dplyr::filter(manifestation_type=="Initial manifestation" | sample_comparison=="ctrl") %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata
-
-## Pivot Table
-pdata %>%
-  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
-  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
-  dplyr::ungroup() -> data_wide
-
-## Only use proteins for prediction
-data_wide %>%
-  mutate(n = row_number()) %>%
-  dplyr::select(Number.x, class, n, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide
-
-## Prediction
-train_df <- data_wide
-train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.frame(train_df)
-pred_obj_2.1 <- run_predictions(train_df)
-
-## ROC Figure 
-g <- ggplot(pred_obj_2.1$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
-  geom_roc(n.cuts=0) + 
-  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_2.1$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
-  coord_equal() +
-  style_roc(x="1 - Specificity", y="Sensitivity")
-
-auc_tbl_2 <- as.data.table(calc_auc(g))
-
-auc_tbl_2 %>%
-  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
-  dplyr::select(AUC) -> auc_annotation_tbl
-
-table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
-auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)
-
-g +
-  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
-  theme(axis.title = element_text(size=14),
-        axis.text = element_text(size=13),
-        legend.title = element_text(size=14),
-        legend.text = element_text(size=13)) -> g
-g
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_a.png", plot = g, width = 9, height = 8, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_a.svg", plot = g, width = 9, height = 8, dpi = 200)
-
-subfigure_2 <- g
-
+# Supplemental Figure 1
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
+supp_figure_1 <- ggarrange(
+ggarrange(supp_figure_1_a, splsda_avg_traj_fig, roc_controls, labels=c("A", "B", "C"), ncol = 3),
+ggarrange(splsda_ratio_unique_patients, splsda_ratio_shuffled_patients, labels=c("D", "E"), ncol = 2),
+nrow=2
+)
+supp_figure_1
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure_1.png", plot = supp_figure_1, width = 25, height = 18, dpi = 200)
+ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure_1.svg", plot = supp_figure_1, width = 25, height = 18, dpi = 200)
 ```
 
-<!-- # Prediction for iGBM vs rGBM  -->
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
-# iGBM vs rGBM protein conc ----
-data_long %>% 
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBM/rGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="iGBM/rGBM" | sample_comparison=="ctrl") %>%
-  dplyr::mutate(log2_conc = log2(conc+1)) %>%
-  dplyr::filter(manifestation_type=="Second manifestation" | sample_comparison=="ctrl") %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(0,1), to=c("ctrl", "GBM"))) -> pdata
-
-## Pivot Table
-pdata %>%
-  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, log2_conc, sample_comparison, class) %>%
-  tidyr::pivot_wider(id_cols = c("Number.x",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Latency (days)', 'sample_comparison',  'class'), names_from = "gene", values_from = "log2_conc" ) %>%
-  dplyr::ungroup() -> data_wide
-
-## Only use proteins for prediction
-data_wide %>%
-  mutate(n = row_number()) %>%
-  dplyr::select(Number.x, class, n, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide
-
-## Prediction
-train_df <- data_wide
-train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, -n)
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.frame(train_df)
-pred_obj_3 <- run_predictions(train_df)
-
-## ROC Figure 
-g <- ggplot(pred_obj_3$roc_data, aes(m=GBM, d=factor(obs, levels = c("ctrl", "GBM")), color=Model)) + 
-  geom_roc(n.cuts=0) + 
-  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_3$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
-  coord_equal() +
-  style_roc(x="1 - Specificity", y="Sensitivity")
-
-auc_tbl_3 <- as.data.table(calc_auc(g))
-
-auc_tbl_3 %>%
-  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
-  dplyr::select(AUC) -> auc_annotation_tbl
-
-table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
-auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)
-
-g +
-  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
-  theme(axis.title = element_text(size=14),
-        axis.text = element_text(size=13),
-        legend.title = element_text(size=14),
-        legend.text = element_text(size=13)) -> g
-g
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_b.png", plot = g, width = 9, height = 8, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_b.svg", plot = g, width = 9, height = 8, dpi = 200)
-
-subfigure_3 <- g
+\blandscape
+
+# Supplemental Figure 2
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
+elisa_supp_figure_2
 ```
 
-<!-- # Prediction for iGBM vs rGBM ratio -->
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
-# iGBM vs rGBM ratio ----
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
-
-data_long %>% 
-  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
-  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
-  dplyr::group_by(gene, Number.x) %>%
-  dplyr::group_map(function(.d, ...){
-    .d <- as.data.table(.d)
-    .d %>% 
-      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>%
-      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
-                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
-    # print(.d)
-    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
-      .d_initial <- .d[manifestation_type=="Initial manifestation"]
-      .d_second <- .d[manifestation_type=="Second manifestation"]
-      tmp$ratio <- .d_second$conc / .d_initial$conc
-    } else if (nrow(.d)==2) {
-      .d_initial <- .d[collection_type=="E1"]
-      .d_second <- .d[collection_type=="E2"]
-      tmp$ratio <- .d_initial$conc / .d_second$conc
-    } else {
-      tmp$ratio <- .d$conc
-    }
-    return(tmp)
-  }, .keep = TRUE) %>%
-  rbindlist() %>%
-  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
-  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
-  dplyr::filter(grepl(";", agg_lab_id)) %>%
-  dplyr::mutate(log2_conc = log2(ratio)) %>%
-  # dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2
-
-pdata %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="iGBM") -> pdata
-
-pdata2 %>%
-  dplyr::select(Number.x, gene, log2_conc) %>%
-  tidyr::pivot_wider(id_cols = "Number.x", names_from = "gene", values_from = "log2_conc") %>%
-  dplyr::mutate(class="rGBM") -> pdata2
-
-pdata <- rbindlist(list(pdata, pdata2))
-
-## Pivot Table
-pdata %>%
-  dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide
-
-## Prediction
-train_df <- data_wide
-train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x, )
-train_df <- train_df[ !is.infinite(train_df$ASAH1) , ]
-## Make Valid Column Names 
-colnames(train_df) <- make.names(colnames(train_df))
-train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.frame(train_df)
-pred_obj_4 <- run_predictions(train_df, pred_col_prob = "rGBM")
-
-## ROC Figure 
-g <- ggplot(pred_obj_4$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
-  geom_roc(n.cuts=0) + 
-  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_4$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
-  coord_equal() +
-  style_roc(x="1 - Specificity", y="Sensitivity")
-
-auc_tbl_4 <- as.data.table(calc_auc(g))
-
-auc_tbl_4 %>%
-  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
-  dplyr::select(AUC) -> auc_annotation_tbl
-
-table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
-auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)
-
-g +
-  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
-  theme(axis.title = element_text(size=14),
-        axis.text = element_text(size=13),
-        legend.title = element_text(size=14),
-        legend.text = element_text(size=13)) -> g
-g
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_d.png", plot = g, width = 9, height = 8, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_d.svg", plot = g, width = 9, height = 8, dpi = 200)
-
-subfigure_4 <- g
+\elandscape
+
+# Extra COX-PH Model
+## Checking Assumptions
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
+# fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
+# print(summary(fit))
+
+library(finalfit)
+
+dependent_var <- "Surv(Time.to.reccurence..days., status)"
+explanatory_vars <- c("ASAH1.ng.ml", "SYNM.ng.ml", "GPNMB.ng.ml", "MMP.9.ng.ml")
+
+ttr_surv_data %>% 
+    coxphmulti(dependent_var, explanatory_vars) %>% 
+    cox.zph() %>% 
+    {zph_result <<- .}
+
+par(mfrow = c(2, 2), mai = c(0.65, 0.8, 0.1, 0.1)) # 2-by-2 grid of plots
+
+plot(zph_result, var=1)
+plot(zph_result, var=2)
+plot(zph_result, var=3)
+plot(zph_result, var=4)
+
+zph_result 
 ```
 
-<!-- # Protein Measurements and Clinical Data -->
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
-data_long %>%
-  dplyr::filter(!is.na(manifestation_type) | patient_state==0) %>%
-  dplyr::mutate(log2_conc = log2(conc+1)) %>%
-  dplyr::group_by(gene, Number.x) %>%
-  # dplyr::add_count() %>%
-  # dplyr::filter(n==2) %>%
-  dplyr::mutate(Manifestation = plyr::mapvalues(manifestation_type, from=c("Initial manifestation", "Second manifestation"), to=c("iGBM", "rGBM"))) %>%
-  dplyr::mutate(Manifestation = ifelse(is.na(Manifestation), "Ctrl", Manifestation)) %>%
-  dplyr::mutate(Manifestation = factor(Manifestation, levels = c("Ctrl", "iGBM", "rGBM"))) %>%
-  dplyr::filter(Manifestation!="Ctrl") %>%
-  dplyr::select(Number.x, gene, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, log2_conc, Manifestation) %>%
-  tidyr::pivot_wider(id_cols = c("Number.x", "Manifestation",'Age at surgery (years)', 'Sex', 'Tumor localization', 'Type of resection', 'MGMT promotor methylation', 'p53 accumulation', 'Ki67-Li', 'EGFR vIII', 'Treatment', 'Time-to-reccurence (days)', 'Manifestation'), names_from = "gene", values_from = "log2_conc" ) %>%
-  dplyr::ungroup() -> data_wide #'Latency (days)'
-
-# data_wide$`Age at surgery (years)` <- factor(data_wide$`Age at surgery (years)`, levels = unique(data_wide$`Age at surgery (years)`))
-data_wide$`Tumor localization` <- factor(data_wide$`Tumor localization`, levels = unique(data_wide$`Tumor localization`))
-data_wide$`Tumor localization` <- as.numeric(data_wide$`Tumor localization`)
-data_wide$`Type of resection` <- factor(data_wide$`Type of resection`, levels = unique(data_wide$`Type of resection`))
-data_wide$`Type of resection`  <- as.numeric(data_wide$`Type of resection` )
-data_wide$`MGMT promotor methylation` <- factor(data_wide$`MGMT promotor methylation`, levels = unique(data_wide$`MGMT promotor methylation`))
-data_wide$`MGMT promotor methylation` <- as.numeric(data_wide$`MGMT promotor methylation`)
-data_wide$`p53 accumulation` <- factor(data_wide$`p53 accumulation`, levels = unique(data_wide$`p53 accumulation`))
-data_wide$`p53 accumulation` <- as.numeric(data_wide$`p53 accumulation` )
-data_wide$`Ki67-Li` <- factor(data_wide$`Ki67-Li`, levels = unique(data_wide$`Ki67-Li`))
-data_wide$`Ki67-Li` <- as.numeric(data_wide$`Ki67-Li` )
-data_wide$`EGFR vIII` <- factor(data_wide$`EGFR vIII`, levels = unique(data_wide$`EGFR vIII`))
-data_wide$`EGFR vIII` <- as.numeric(data_wide$`EGFR vIII` )
-data_wide$Treatment <- factor(data_wide$Treatment, levels = unique(data_wide$Treatment))
-data_wide$Treatment <- as.numeric(data_wide$Treatment)
-# data_wide$`Time-to-reccurence (days)` <- factor(data_wide$`Time-to-reccurence (days)`, levels = unique(data_wide$`Time-to-reccurence (days)`))
-# data_wide$`Latency (days)` <- factor(data_wide$`Latency (days)`, levels = unique(data_wide$`Latency (days)`))
-
-# Encode Category Sex
-data_wide$Sex <- plyr::mapvalues(data_wide$Sex, from = c("w", "m"), to = c(0, 1))
-data_wide$Sex <- as.factor(data_wide$Sex)
-data_wide$Sex <- as.numeric(data_wide$Sex)
-
-data_wide  %>%
-  dplyr::select(Number.x, -`Age at surgery (years)`, -Sex, `Tumor localization`, -`Type of resection`, -`MGMT promotor methylation`, -`p53 accumulation`, -`Ki67-Li`, -`EGFR vIII`, -Treatment, -`Time-to-reccurence (days)`, Manifestation, ASAH1, SYNEMIN,GPNMB,`MMP-9`) -> data_wide
-
-# Prediction
-train_df <- data_wide
-train_df <- na.omit(train_df)  %>% dplyr::select(-Number.x)
-# Make Valid Column Names 
-colnames(train_df) <- make.names(colnames(train_df))
-train_df %>% dplyr::rename(class=Manifestation) -> train_df
-train_df$class <- factor(train_df$class, levels=c("iGBM", "rGBM"))
-colnames(train_df) <- make.names(colnames(train_df))
-train_df <- as.data.frame(train_df)
-pred_obj_5 <- run_predictions(train_df, pred_col_prob = "rGBM")
-
-# ROC
-g <- ggplot(pred_obj_5$roc_data, aes(m=rGBM, d=factor(obs, levels = c("iGBM", "rGBM")), color=Model)) +
-  geom_roc(n.cuts=0) + 
-  scale_color_manual(values = RColorBrewer::brewer.pal(length(unique(pred_obj_5$master_lift$liftModelVar)), "Dark2"), labels=c("Logit", "Elastic Logit", "KNN", "RF")) +
-  coord_equal() +
-  style_roc(x="1 - Specificity", y="Sensitivity")
-
-auc_tbl_5 <- as.data.table(calc_auc(g))
-
-auc_tbl_5 %>%
-  dplyr::mutate(AUC = sprintf("AUC: %s%%", round(AUC*100, 0))) %>%
-  dplyr::select(AUC) -> auc_annotation_tbl
-
-table_theme <- ttheme_minimal(base_size = 14, core = list(fg_params=list(col=(RColorBrewer::brewer.pal(4, "Dark2")))))
-auc_grob <- tableGrob(auc_annotation_tbl, cols = NULL, rows=NULL, theme = table_theme)
-
-g +
-  annotation_custom(auc_grob, xmin=0.70, xmax=0.90, ymin=0.10, ymax=0.30) +
-  theme(axis.title = element_text(size=14),
-        axis.text = element_text(size=13),
-        legend.title = element_text(size=14),
-        legend.text = element_text(size=13)) -> g
-g
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_c.png", plot = g, width = 9, height = 8, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_figure1_subfigure_c.svg", plot = g, width = 9, height = 8, dpi = 200)
-
-subfigure_5 <- g
+Checking assumptions of no time trend of the residuals, it seems like only MMP-9 might violate this assumption.
+
+\blandscape
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=7.5}
+
+ttr_surv_data %>%
+  finalfit(dependent_var, explanatory_vars) %>%
+  gt()
+
+ttr_surv_data %>%
+  hr_plot(dependent_var, explanatory_vars, dependent_label="Time to Recurrence")
+
 ```
 
-<!-- # Latency Data -->
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
-latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
-latency_data <- fread(latency_data_file)
-# Clean names
-latency_data <- rstatix::make_clean_names(latency_data)
-latency_data %>%
-  dplyr::mutate(ASAH1.ng.ml = log2(ASAH1.ng.ml+1),
-                SYNM.ng.ml = log2(SYNM.ng.ml+1),
-                GPNMB.ng.ml = log2(GPNMB.ng.ml+1),
-                MMP.9.ng.ml = log2(MMP.9.ng.ml+1)) %>%
-  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(ASAH1.ng.ml), "high", "low"),
-                SYNM = ifelse(SYNM.ng.ml>=mean(SYNM.ng.ml), "high", "low"),
-                GPNMB = ifelse(GPNMB.ng.ml>=mean(GPNMB.ng.ml), "high", "low"),
-                MMP9 = ifelse(MMP.9.ng.ml>=mean(MMP.9.ng.ml, na.rm=T), "high", "low")) -> latency_data
-
-fit <- survfit(Surv(Latency..days., status) ~ ASAH1, data = latency_data)
-print(fit)
-n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$ASAH1)
-# Change color, linetype by strata, risk.table color by strata
-surv_p1 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Latency Time (Days)")
-
-surv_p1 <- surv_p1$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'))
-
-fit <- survfit(Surv(Latency..days., status) ~ SYNM, data = latency_data)
-print(fit)
-n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$SYNM)
-# Change color, linetype by strata, risk.table color by strata
-surv_p2 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Latency Time (Days)")
-
-surv_p2 <- surv_p2$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'))
-
-fit <- survfit(Surv(Latency..days., status) ~ GPNMB, data = latency_data)
-print(fit)
-n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$GPNMB)
-# Change color, linetype by strata, risk.table color by strata
-surv_p3 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Latency Time (Days)")
-
-surv_p3 <- surv_p3$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'))
-
-fit <- survfit(Surv(Latency..days., status) ~ MMP9, data = latency_data)
-print(fit)
-n_data <- table(latency_data[!is.na(latency_data$Latency..days.)]$MMP9)
-# Change color, linetype by strata, risk.table color by strata
-surv_p4 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Latency Time (Days)")
-
-surv_p4 <- surv_p4$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'))
-
-fit <- survfit(Surv(Latency..days., status) ~ ASAH1 + SYNM, data = latency_data)
-print(fit)
-# Change color, linetype by strata, risk.table color by strata
-surv_p5 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = F,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
-          ) +
-  labs(x="Latency Time (Days)")  +
-  guides(col=guide_legend(nrow=4,byrow=TRUE))
-
-surv_p5 <- surv_p5$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-surv_p5_with_ci <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(1700, 0.95), conf.int = T,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
-          ) +
-  labs(x="Latency Time (Days)")  +
-  guides(col=guide_legend(nrow=2,byrow=TRUE))
-
-surv_p5_with_ci <- surv_p5_with_ci$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-survival_plot_grid <- ggarrange(surv_p1, surv_p2, surv_p3, surv_p4, ncol = 4)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.png", plot = survival_plot_grid, width = 18, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c.svg", plot = survival_plot_grid, width = 18, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.png", plot = surv_p1, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c1.svg", plot = surv_p1, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.png", plot = surv_p2, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c2.svg", plot = surv_p2, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.png", plot = surv_p3, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c3.svg", plot = surv_p3, width = 11, height = 7, dpi = 200)
-
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.png", plot = surv_p4, width = 11, height = 7, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_main_subfigure_c4.svg", plot = surv_p4, width = 11, height = 7, dpi = 200)
-
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_.png", plot = surv_p5$plot, width = 11, height = 7, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_.svg", plot = surv_p5$plot, width = 11, height = 7, dpi = 200)
-# 
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_with_ci.png", plot = surv_p5_with_ci$plot, width = 11, height = 7, dpi = 200)
-# ggsave("../../figures/elisa_plasma_proteomics/elisa_supp_subfigure_with_ci.svg", plot = surv_p5_with_ci$plot, width = 11, height = 7, dpi = 200)
-
-# survival model with time to recurrence; combined group of all 4 proteins
-ttr_surv_data <- merge(latency_data, patient_meta_data[, .(LAB_ID, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `Time-to-reccurence (days)`)], by.x="lab_id", by.y="LAB_ID", all.x=T)
-ttr_surv_data$`Time-to-reccurence (days)`[ ttr_surv_data$`Time-to-reccurence (days)`=="/" ] = ""
-ttr_surv_data$`Time-to-reccurence (days)` <- as.numeric(ttr_surv_data$`Time-to-reccurence (days)`)
-ttr_surv_data <- rstatix::make_clean_names(ttr_surv_data)
-
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1, data = ttr_surv_data)
-print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$ASAH1)
-# Change color, linetype by strata, risk.table color by strata
-ttr_surv_p1 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high\n(n=%s)", n_data[1]), sprintf("ASAH1=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
-
-ttr_surv_p1 <- ttr_surv_p1$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ SYNM, data = ttr_surv_data)
-print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$SYNM)
-# Change color, linetype by strata, risk.table color by strata
-ttr_surv_p2 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("SYNM=high\n(n=%s)", n_data[1]), sprintf("SYNM=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
-
-ttr_surv_p2 <- ttr_surv_p2$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ GPNMB, data = ttr_surv_data)
-print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$GPNMB)
-# Change color, linetype by strata, risk.table color by strata
-ttr_surv_p3 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("GPNMB=high\n(n=%s)", n_data[1]), sprintf("GPNMB=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
-
-ttr_surv_p3 <- ttr_surv_p3$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ MMP9, data = ttr_surv_data)
-print(fit)
-n_data <- table(ttr_surv_data[!is.na(ttr_surv_data$Time.to.reccurence..days.)]$MMP9)
-# Change color, linetype by strata, risk.table color by strata
-ttr_surv_p4 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = TRUE,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          palette = "Dark2",
-          legend.title = "",
-          legend.labs = c(sprintf("MMP9=high\n(n=%s)", n_data[1]), sprintf("MMP9=low\n(n=%s)", n_data[2]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")
-
-ttr_surv_p4 <- ttr_surv_p4$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-fit <- survfit(Surv(Time.to.reccurence..days., status) ~ ASAH1 + SYNM, data = ttr_surv_data)
-print(fit)
-# Change color, linetype by strata, risk.table color by strata
-ttr_surv_p5 <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = F,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high, SYNM=high (n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low (n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
-  guides(col=guide_legend(nrow=4,byrow=TRUE))
-
-ttr_surv_p5 <- ttr_surv_p5$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-
-ttr_surv_p5_with_ci <- ggsurvplot(fit,
-          pval = TRUE, pval.coord = c(650, 0.85), conf.int = T,
-          risk.table = FALSE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw(), # Change ggplot2 theme
-          legend.title = "",
-          legend.labs = c(sprintf("ASAH1=high, SYNM=high\n(n=%s)", fit$n.risk[1]), sprintf("ASAH1=high, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1]), sprintf("ASAH1=low, SYNM=high\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]), sprintf("ASAH1=low, SYNM=low\n(n=%s)", fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]+fit$n.risk[fit$n.risk[1]+1+fit$n.risk[fit$n.risk[1]+1]]]))
-          ) +
-  labs(x="Time to Recurrence (Days)", y="Time to Recurrence probability")  +
-  guides(col=guide_legend(nrow=2,byrow=TRUE))
-
-ttr_surv_p5_with_ci <- ttr_surv_p5_with_ci$plot +
-  theme(axis.title = element_text(size=16),
-        axis.text = element_text(size=15),
-        legend.text = element_text(size=12),
-        legend.spacing.x = unit(0.25, 'cm'),
-        legend.position = "bottom")
-```
-
-\blandscape
-
-# Master Figure
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
-# mataster_figure <- ggarrange(ggarrange(subfigure_6, 
-#                                        splsda_ratio_avg_traj_fig, ncol=2, labels = c("A", "B")),
-#                              ggarrange(survival_plot_grid, ncol=1, labels = c("C")), 
-#                              ggarrange(subfigure_1, 
-#                                        p_ratio_ttr, ncol=2, labels = c("D", "E")),
-#                              ncol = 1, nrow=3)
-
-mataster_figure <- ggarrange(ggarrange(subfigure_6, 
-                                       ggarrange(ttr_surv_p1, ttr_surv_p2, ncol=2), ncol=2, labels = c("A", "B")),
-                             ggarrange(subfigure_1, 
-                                       splsda_better_ratio_avg_traj_fig, ncol=2, labels = c("C", "D")),
-                             ncol = 1, nrow=2)
-mataster_figure
-ggsave("../../figures/elisa_plasma_proteomics/elisa_master_figure.png", plot = mataster_figure, width = 25, height = 21, dpi = 200)
-ggsave("../../figures/elisa_plasma_proteomics/elisa_master_figure.svg", plot = mataster_figure, width = 25, height = 21, dpi = 200)
-
-```
-
-<!-- **Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA).
-(**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM populations.
-Note, a patient was omitted due to being an outlier, see Supp Figure X for full set of patients.
-(**b**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
-(**c**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of longer survival.
-(**d**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
-(**e**) The potential of recurrent GBM / initial GBM protein ratios for predicting time to recurrence.
-Note, each point represents a single patient, so we may not be able to directly state a specific ratio can be used to infer to time to recurrence.
-However, we can still state given a specific patients ratio, we can predict that patients time to recurrence.
+Holding other variables constant, a reduction in ASAH1 reduces the time to recurrence by a factor of 0.55, or 45 percent.
+This seems to be inline with the KM plot for ASAH1, however, SYNM seems to be a different story..
+
+\elandscape
+
+<!-- # Text -->
+
+<!-- Based on our observations of variations in protein concentrations (ASAH1, SYNEMIN, and MMP-9) between the initial serum measurement and recurrent serum measurements (as shown in Figure xa), as well as the classification of low and high concentrations of ASAH1 and SYNEMIN with respect to the time to recurrence (Figure xb), we further investigated the use of interpersonal longitudinal information from each patient.
+To this end, we calculated individual patient ratios for each protein, i.e., the initial serum measurement divided by the recurrent serum measurement and vice versa.
+To represent this information in a lower dimensional space, we used sparse partial least squares discriminant analysis (sPLS-DA) to categorize patients based on their initial serum/recurrent serum ratio against their recurrent serum/initial serum ratio (Figure xc).
+The resulting separating plane suggests that patients start at the border of the classifier plane and move outward as their state of recurrence changes.
+This way, repeated serum measurements can be used to locate a patient's position in this lower dimensional space and determine their proximity to the initial post-surgery state or a recurrent state.
+Additionally, we created an elastic net logistic regression classification model to predict a patient's state, either initial manifestation or recurrent manifestation (Figure xd).
+The results indicate that the protein measurements provide most of the predictive power, while the addition of clinical parameters reduces the classification performance.
+However, the best performing classification was achieved when using the ratios (AUC 89%), suggesting that the interpersonal longitudinal information is more informative than the protein concentrations themselves.
 -->
-**Figure 3: A Promising Plasma Proteomics Signature for Stratification and Prediction of Recurrent Glioblastoma Multiforme.** Concentration of four proteins identified from TMT-MS tissue proteomics, were measured in patients' plasma samples using Enzyme-Linked Immunosorbent Assay (ELISA).
-(**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM populations.
-Note, a patient was omitted due to being an outlier, see Supp Figure X for full set of patients.
-(**b**)  Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of tumor recurrence occurring at a later date.
-(**c**) The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating each population, driven mostly from information contained in the proteins.
-(**d**) Reducing the data of the four proteins ratios demonstrates an average trajectory and separation of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$.
-One patient was randomly left out as a test patient (black + point), 4 samples were randomly sampled to represent $\frac{initial \quad GBM}{recurrent \quad GBM}$ (blue points) and remaining 4 samples were used to represent $\frac{recurrent \quad GBM}{initial \quad GBM}$.
-
-\elandscape
-
-\pagebreak
-
-# Supplemental Figures and Tables
-
-```{r message=FALSE, warning=FALSE, include=FALSE}
-fig_counter <- 1
-tab_counter <- 1
-```
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8.5, fig.width=11}
-
-supp_figure_0 <- ggarrange(ggarrange(subfigure_6_org, supp_figure_pre_v_post, ncol=1, nrow=2, labels=c("A", "B")), ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), ncol=2, nrow=1, labels=c("", "C"))
-supp_figure_0
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_0, width = 18, height = 16, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_0, width = 18, height = 16, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_6_org, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_6_org, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = supp_figure_pre_v_post, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot =  ggarrange(supp_figure_ratios, supp_figure_ratios2, nrow=2, ncol=1), width = 9, height = 8, dpi = 200)
-```
-**Supplemental Figure** `r knitr::asis_output(md(sprintf("**%s**", fig_counter)))` `r fig_counter <- fig_counter + 1`**: Distribution of Pre- vs Post-Surgery Plasma Proteomics and Ratios of Initial and Recurrent GBM Plasma Proteomics.**  (**a**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, initial GBM and paired recurrent GBM population for all patients.
-(**b**) The log2 transformed protein concentration measurements demonstrates the difference in log2 concentration distributions between control, pre-surgery and post-surgery populations.
-(**c**) Demonstrates the log transformed $\frac{initial \quad GBM}{recurrent \quad GBM}$ ratio (top) as well as the inverse ratio (bottom) for the potential use of identifying a patients state to recurrence using these ratios.
-
-\pagebreak
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
-supp_figure_1 <- ggarrange(ggarrange(roc_controls, subfigure_2, subfigure_3, ncol=3, nrow=1, labels=c("A", "B", "C")), ggarrange(subfigure_5, subfigure_4, labels=c("D", "E")), ncol=1, nrow=2)
-
-supp_figure_1
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_1, width = 18, height = 18, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_1, width = 18, height = 18, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = roc_controls, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = roc_controls, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = subfigure_3, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.png", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_d.svg", fig_counter), plot = subfigure_5, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_e.png", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_e.svg", fig_counter), plot = subfigure_4, width = 9, height = 8, dpi = 200)
-```
-**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Variation in Predictive Performance using Different Models.** Separation on Controls and testing parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms on different prediction tasks.
-(**a**)  The receiver operating characteristic (ROC) shows the potential of using plasma proteomics for separating controls (health patients) from patients with GBM using an elastic logistic regression.
-(**b**) Most models have AUCs greater than 60% when predicting control vs initial GBM.
-(**c**) Most models have comparable AUCs, except for k-nearest neighbours when predicting for initial GBM vs recurrent GBM.
-(**d**) Including clinical data for *Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection* reduces predictive performance for all models.
-(**e**) All models, except random forest, have AUCs above 80% when predicting initial GBM vs recurrent GBM using protein ratios.
-
-\pagebreak
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=8.5, dpi = 200}
-supp_figure_2 <- ggarrange(splsda_avg_traj_fig, splsda_ind_traj_fig, ncol=1, nrow=2, labels=c("A", "B"))
-
-supp_figure_2
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_2, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_2, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = splsda_avg_traj_fig, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = splsda_avg_traj_fig, width = 9, height = 8, dpi = 200)
-
-# ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-# ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = subfigure_2, width = 9, height = 8, dpi = 200)
-
-```
-**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Individual Patient Trajectory in Low Dimensional Space.** (**a**) A sparse partial least squares discriminant analysis shows separation between the different population groups, as well as an overall trajectory of a initial GBM patient going towards a recurrent GBM patient in a low dimensional space.
-(**b**) Individual trajectories demonstrate the individual variance of patients path to recurrence in a low-dimensional space.
-
-\pagebreak
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21, dpi=200}
-supp_figure_3 <- ggarrange( p_pa_pb_ratio_scatter, splsda_ratio_ind_traj_fig, ncol=1, nrow=2, labels=c("A", "B"), heights = c(1.5,1))
-
-supp_figure_3
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_3, width = 18, height = 18, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_3, width = 18, height = 18, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = p_pa_pb_ratio_scatter, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = p_pa_pb_ratio_scatter, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = splsda_ratio_ind_traj_fig, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = splsda_ratio_ind_traj_fig, width = 9, height = 8, dpi = 200)
-```
-**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Information Contained in Plasma Proteomic Ratios for Initial GBM vs Recurrent GBM Stratification.** (**a**) Pairwise protein ratios shows promise of individual protein-pair ratios being able to stratify initial vs recurrent plasma GBM samples.
-(**b**) Individual trajectories demonstrate the individual variance of patients path to recurrence in a low-dimensional space using ratios.
-
-\pagebreak
-
-\blandscape
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8.5, fig.width=11, dpi = 200}
-
-ttr_survival_plot_grid <- ggarrange(surv_p1, surv_p2, surv_p3, surv_p4, ncol = 4, labels="A")
-bottom_plots <- ggarrange(ggarrange(ttr_surv_p3, ttr_surv_p4, ncol=2), ggarrange(surv_p5, ttr_surv_p5, ncol=2), labels=c("B", "C"))
-
-supp_figure_4 <- ggarrange(ttr_survival_plot_grid, bottom_plots, nrow=2, ncol=1)
-supp_figure_4
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.png", fig_counter), plot = supp_figure_4, width = 25, height = 15, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s.svg", fig_counter), plot = supp_figure_4, width = 25, height = 15, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.png", fig_counter), plot = ttr_survival_plot_grid, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a.svg", fig_counter), plot = ttr_survival_plot_grid, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.png", fig_counter), plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_1.svg", fig_counter), plot = ttr_surv_p1, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.png", fig_counter), plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_2.svg", fig_counter), plot = ttr_surv_p2, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.png", fig_counter), plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_3.svg", fig_counter), plot = ttr_surv_p3, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.png", fig_counter), plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_a_4.svg", fig_counter), plot = ttr_surv_p4, width = 11, height = 8.5, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.png", fig_counter), plot = surv_p5, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_b.svg", fig_counter), plot = surv_p5, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.png", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c.svg", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
-
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.png", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
-ggsave(sprintf("../../figures/elisa_plasma_proteomics/supp_figure_%s_subfigure_c_with_ci.svg", fig_counter), plot = ttr_surv_p5, width = 9, height = 8, dpi = 200)
-```
-**Supplemental Figure `r fig_counter` `r fig_counter <- fig_counter + 1`: Survival and Time to Recurrence Analysis.** (**a**) Kaplan-Meyer plots showing demonstrating high concentration of ASAH1 and high concentration of SYNEMIN contributing to a higher probability of tumor recurrence occurring at a later date.
-(**b**) Kaplan-Meyer plots demonstrating the combination of both ASAH1 and SYNEMIN and their contribution of survival probability and time to recurrence.
-
-\elandscape
-
-\pagebreak
-
-\blandscape
-
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-logit_net_comp_perf_met_table %>%
-  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))
-logit_net_comp_perf_met_table %>%
-  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))
-
-logit_net_comp_perf_met_table
-```
-
-**Supplemental Table `r as.roman(tab_counter)` `r tab_counter <- tab_counter + 1`: Logistic Regression with Elastic Net Regularization Performance Metrics**
-
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-stat_comp_table %>%
-  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.rtf", tab_counter))
-stat_comp_table %>%
-  gtsave(sprintf("../../tables/elisa_plasma_proteomics/supp_table_%s.tex", tab_counter))
-
-stat_comp_table
-```
-
-**Supplemental Table `r as.roman(tab_counter)` `r tab_counter <- tab_counter + 1`: Statistical Comparisons Between Population Groups per Genes**
-
-\elandscape
-
-\pagebreak
-
-# Methods
-
-## Statistical Test for Comparison of Log2 Distributions
-
-Log2 transformed distributions of protein concentrations were compared pairwise for control, initial GBM, and recurrent GBM samples for each protein, ASAH1, GNMPB, MMP-9 and SYNEMIN.
-Statistical p-values were computed using a permutation test, that utilizes t-test statistics but performs random permutations of the data for computing a p-value.
-Note, for paired sample data, initial GBM vs recurrent GBM, a paired permutation test was performed instead.
-Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382 ), permutation tests were performed using broman (v0.80).
-
-## Dimensionaly Reduction
-
-To visualize the samples plasma proteomics on a different dimension, we applied a sparse Partial Least Squares Discriminant Analysis (sPLS-DA), which maximizes the covariance between the explanatory and response variable(s).
-This was performed on the log2 transformed protein concentrations, as well as the log2 transformed ratios ($\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$) of each protein.
-Individual trajectories are displayed as vector arrows connecting initial GBM patient samples to their corresponding recurrent GBM patient samples, in the sPLS-DA space using the first two components that contain the most amount of explained variance.
-An average trajectory direction can be computed by taking the average of the individual trajectory vectors, to visualize an overall trajectory of initial GBM state to recurrent GBM state.
-To visualize decision boundaries for class separation (control, initial GBM and recurrent GBM samples), we applied a maximum distance approach.
-[1] Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382), sPLS-DA analysis was performed using mixOmics (v6.18.1).
-
-## Prediction Analysis
-
-Prediction tasks were performed using parametric (logistic regression, logistic regression with elastic net regularization), non-parametric (k-nearest neighbours) and ensemble (random forest) supervised learning algorithms.
-General predictive tasks were performed on: controls vs initial GBM samples, controls vs recurrent GBM samples, initial GBM samples vs recurrent GBM samples.
-Data used in prediction tasks are log2 transformed protein concentrations, log2 transformed protein concentrations supplemented with clinical data (*Tumor localization*, *Sex*, *Age at surgery (years)*, *Type of resection*), and log2 transformed ratios of $\frac{initial \quad GBM}{recurrent \quad GBM}$ vs $\frac{recurrent \quad GBM}{initial \quad GBM}$ protein measurements.
-Model training and prediction performances are performed on the entire data set using leave-one-out cross-validation due to the small sample size.
-Visualization and statistical analyses were performed in R (v4.1.2) and RStudio (v2021.09.2+382), prediction analyses were performed using caret (v6.0-93).
-
-# References
-
-1.
-Rohart, F, B Gautier, A Singh, and K-A Lê Cao.
-2017.
-“MixOmics: An R Package for ‘Omics Feature Selection and Multiple Data Integration.” PLoS Computational Biology 13 (11).
-Cold Spring Harbor Labs Journals.

### 2023-02-14 reports/gbm_recurrence_general_proteomics_manuscript_report.qmd
proteomics report clean

---
title: "GBM recurrence - Shotgun proteomics analysis"
author: "Miguel Cosenza-Contreras"
format:
  gfm: 
    toc-depth: 4 # default is 3
    toc-title: Contents
    toc-location: left # default is right, option is 'body'
    number-sections: true
    number-depth: 3
    embed-resources: true
always_allow_html: true
editor: visual
---

```{r knirt_helper_funcs}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))
```

```{r load_packages}
## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)
library(ggrepel)
library(extrafont)
```

```{r theme_set}
theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 
                                         'mm'),
                legend.key.width = unit(3, 
                                        'mm'),
                legend.position = "bottom"))
```

The following report holds the code, results and interpretation of the proteomics data analysis of Initial-Recurrent glioblastoma (GBM) samples of matched patients.

# Sample annotation info

```{r echo=TRUE}
sample_annotation <- read_csv(here("data/sample_annotation.csv"))

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",
                         patient,
                         sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_")) %>%
  filter(recurrence %in% c("prim", 
                           "rec"))

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")
```

## Sample annotation

### Brief note on sample annotation: {.unnumbered .unlisted}

Twenty-two patient matched samples (corresponding to 11 patients) were processed for protein extraction, TMT-labelled and submitted via high-pH fractionation and further analyzed via LC-MS/MS.

```{r echo=FALSE, error=TRUE}
kbl(sample_annotation, booktabs = TRUE) %>%
  kable_styling(font_size = 14, 
                bootstrap_options = c("striped", 
                                      "hover"))
```

# Brief note on data analysis and processing:

Three proteomics data analysis approaches were used to explore the spectral data and will be briefly described:

1.  General large scale proteomics.
2.  Proteolytic processing.
3.  Proteogenomics

The FragPipe bioinformatic pipeline was used for each of these approaches with varying parameters depending on the type of explorative approach.

**This report holds the code, analysis, visualizations and results for the General large scale proteomics approach.**

**1. General large scale proteomics analyses** : In order to explore the differential abundance of proteins between Recurrent and Initial tumor, a fully-tryptic search was performed against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides. A minimum percentage of %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

**2. Proteolytic processing**: The large scale evaluation of differential proteolytic activity between recurrent vs primary tumor was executed via a semi-tryptic search against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides. A minimum percentage of %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

**3. Proteogenomics**: The explorative proteogenomics analysis started with the generation of a GBM-specific database from publicly available RNA-seq sequencing data from tumor specimens. Fastq files were obtain from the SRA BioProject PRJNA627121 (Cell Rep, 2021 Mar 2;34(9):108787). Eight samples containing paired reads were selected, based on: 1) the higher number of reads per file; 2) having a similar proportion of males and females and 3) not bearing isocitrate-dehyderogenase (IDH)-1 mutation (IDH WT). None of these patients undergo previous treatment, therefore potential variants in the sequencing data are not expected to be associated to treatment-related mutagenic processes. Spectral data was then searched against this custom database using a fully-tryptic approac. No minimum intensity threshold was set for reporter ion intensity for quantitation.

Exceptionally when stated differently within each sub-approach described above, all these searches were executed using the following parameters: MSFragger was used as a search engine, setting a fully or semi-tryptic specificity allowing for 1 missed cleavages. Precursor mass tolerance was set to -20/20 and fragment mass tolerance of 20 ppm with mass calibration and parameter optimization resulting in 5 ppm. Peptide N-terminal acetylation and peptide N-terminal TMT labeling were set as variable modifications. TMT labelling at K and carbamidomethylation of C were set as fixed modifications. MSBooster was used for deep-learning based predictions of retention time and spectra. Predicted features were used by Percolator for post-processing scoring and false discovery rate (FDR) control via target-decoy competition of all peptide-to-spectrum matches (PSMs) obtained from MSFragger search. ProteinProphet was used as a protein inference algorithm keeping all PSMs with probability score bigger than 0.9. Final report and FDR estimation was based on the filtered PSM and protein lists before protein scoring.

Relative quantitation of identified peptides within each sample was performed via their reported ion intensities using TMT-integrator. Only PSMs coming from unique peptides with a minimum probability of 0.9 and an isotopic purity of at least 50% were considered for quantitation. Quantitative values were normalized via median centering and summarized by protein/gene and peptide using a virtual reference channel, to finally generate protein and peptide normalized abundance matrices of abundance used for further statistical processing in combination with the sample annotation.

# Initial data loading and wrangling {.unnumbered .unlisted}

```{r message=FALSE}
## tmt-integrator output loading 
tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
  janitor::clean_names()

identified_peptides <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_peptide.tsv")) %>%
  janitor::clean_names()

fasta_ident_prots <- read.fasta(here("data/specific_search_fragpipe17/specific_no_ptms_2/mix_2/protein.fas"), seqtype = "AA", as.string = TRUE)
```

```{r}
# wide expression matrix with NAs
expr_matrix <- dplyr::select(tmt_protdata_frag,
                             Protein = index, 
                             starts_with("x"))

long_mat <- pivot_longer(expr_matrix,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", 
                                      "recurrence"),
                         values_to = "Abundance")

long_mat_2 <- long_mat %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_"))

paired_annot_uniq <- long_mat_2 %>%
  dplyr::select(patient, 
                recurrence, 
                paired_id) %>%
  distinct()

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x", 
                         patient, 
                         sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_")) %>%
  filter(recurrence %in% c("prim", 
                           "rec"))

## Annotated abundance data in long format ----

quant_annotated <- left_join(long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id", 
                                    "patient",
                                    "recurrence")) %>%
  mutate(Channel_mix = paste(mixture, 
                             channel,  
                             sep = "_"))

## Wide xpression matrices without NAs
expr_matnona <- column_to_rownames(expr_matrix,
                                   "Protein") %>%
          as.matrix() %>%
          na.omit()
```

# ID Coverage before sparicity reduction and filtering

These numbers are based on the results from the output of FragPipe's TMT-integrator, before sparicity reduction and filtering.

```{r}
ids_summary <- quant_annotated %>% 
  dplyr::select(Protein, mixture, Abundance) %>% 
  distinct(Protein, mixture, .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "In all three",
                   `Nr of IDs` = nrow(expr_matnona))) %>%
  bind_rows(.,
            tibble(mixture = "Overall",
                   `Nr of IDs` = nrow(expr_matrix)))
```

-   Number of identified proteins: `r nrow(expr_matrix)`

-   Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`

```{r fig.height=4, fig.width=7, warning=FALSE}
ggplot(data = ids_summary,
       aes(y = `Nr of IDs`, x = mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 5, 
            position = position_stack(vjust = 0.5), 
            aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Proteins by Mixture") +
  ylab("Nr of IDed and Quant Proteins") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0.1, 
                                   size = 10, 
                                   angle = 0),
        axis.text.y = element_text(hjust = 0.5, 
                                   vjust = 0.1, 
                                   size = 10, 
                                   angle = 0),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey"),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 1.5),
        axis.title = element_text(size = 12,
                                face = "bold"))
```

# Exploratory analysis

## Visualization of missing values

```{r fig.height=7, fig.width=8, warning=FALSE}
vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values") + 
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill=NA, 
                                            size=1.5),
                axis.title = element_text(size = 12,
                                        face = "bold"))
```

There is an uneven distribution of missingness between TMT-batches.

## Plots of normalized abundance distribution

### Normalized ion reporter abundance distribution by TMT channel/sample {.unnumbered}

```{r fig.height=4, fig.width=7, warning=FALSE}
ggplot(quant_annotated,
                 mapping = aes(x = Channel_mix, 
                               y = Abundance, 
                               fill = recurrence)) +
          geom_boxplot() + 
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title = element_text(size = 12,
                                         face = "bold"))
```

The median-centering after TMT-integrator seem to show a slight mixture-dependent centered median per sample. The paired distribution of initial and recurrent tumor allows the biological effect to be more important than any potential TMT-mixture-based batch effect.

The different abundance distribution can also be explained by a different distribution of missing values.

PCA and sPLS-DA analysis allow to confirm this.

### Normalized ion reporter abundance distribution by protein {.unnumbered}

Select/sample *N* proteins to observe their abundance distributio.

```{r}
which_prots <- sample_proteins(x = quant_annotated$Protein,
                               size = 50,
                               seed = 156)
```

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=4, fig.width=7, warning=FALSE}
ggplot(quant_annotated %>% 
          filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,
                                           Abundance,
                                           na.rm = TRUE), 
                               y = Abundance)) +
          geom_boxplot() + 
          geom_jitter(color = "black", 
                      size = 0.4, 
                      alpha = 0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
               x = "Protein ID ") + 
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

```

### Normalized ion reporter abundance distribution by protein and batch {.unnumbered}

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=4, fig.width=9, warning=FALSE}
ggplot(quant_annotated %>% 
                         filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,
                                           Abundance,
                                           na.rm = TRUE), 
                               y = Abundance, 
                               fill = mixture)) +
          geom_boxplot() + 
          geom_jitter(color = "black", 
                      size = 0.4, 
                      alpha = 0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
                       x = "Protein ID ") + 
          facet_grid(.~mixture, 
                     scales = "free") +
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title = element_text(size = 12,
                                          face = "bold"))
```

The abundance distribution of individual proteins can be considered as normaly distributed in general, which makes then suitable for inferential statistics via linear models.

## Exploratory PCA

```{r echo=FALSE}
## Get predictors for mixtures
expr_matrix2 <- expr_matrix %>%
  dplyr::select(Protein, 
                sample_annotation2$paired_id)

paired_id <- sample_annotation2$paired_id
recurrence <- sample_annotation2$recurrence
patient <- sample_annotation2$patient
mixture <- sample_annotation2$mixture

## get predictor for mixtures wo patient 6
expr_matrixwo6 <- expr_matrix %>%
  dplyr::select(Protein, 
                sample_annotationwo6$paired_id)

paired_idwo6 <- sample_annotationwo6$paired_id
recurrencewo6 <- sample_annotationwo6$recurrence
patientwo6 <- sample_annotationwo6$patient
mixturewo6 <- sample_annotationwo6$mixture

## Prep matrix without missing values for PCA

expr_matnona <- column_to_rownames(expr_matrix2,
                                   "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnona) <- paste(sample_annotation2$paired_id, 
                                sample_annotation2$mixture,
                                sep = "_")

t_expr_matnona <- t(expr_matnona)

expr_mat <- column_to_rownames(expr_matrix2,
                               "Protein") %>%
          as.matrix()

t_expr_mat <- t(expr_mat)

## Matrix without patient 6

expr_matnonawo6 <- column_to_rownames(expr_matrixwo6,
                                      "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnonawo6) <- paste(sample_annotationwo6$paired_id, 
                                   sample_annotationwo6$mixture,
                                   sep = "_")

t_expr_matnonawo6 <- t(expr_matnonawo6)

expr_matwo6 <- column_to_rownames(expr_matrixwo6,
                                  "Protein") %>%
          as.matrix()

t_expr_matwo6 <- t(expr_matwo6)
```

## Percentage of explained variance per component {.unnumbered .unlisted}

The PCA was applied on the expression matrix without missing values

```{r}
pca_res <- pca(t_expr_matnona, 
               ncomp = 10, 
               center = TRUE, 
               scale = TRUE)
```

```{r fig.width=4.7, fig.height=3.5}
plot(pca_res)
```

\~45% of variance based on protein abundance can be explained by components 1 and 2 after PCA.

## PCA plot on samples based on protein abundance

```{r}
# preprocess pca results 
pca_variates <- pca_res$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", 
                    "recurrence", 
                    "mixture"),
           sep = "\\_", 
           extra = "merge") %>%
  mutate(paired_id = paste(sample, 
                           recurrence, 
                           sep = "_")) %>%
  dplyr::select(-c(mixture, 
                   recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id") %>%
  mutate(Stage = if_else(condition = recurrence == "prim", 
                         true = "Initial", 
                         false = "Recurrent"),
         Mix = case_when(mixture == "tmt_1" ~ "Mix1",
                         mixture == "tmt_2" ~ "Mix2",
                         mixture == "tmt_3" ~ "Mix3"))

pca_variateswopat6 <- pca_variates %>% 
  filter(patient != "x6")
```

```{r fig.width=5, fig.height=4, warning=FALSE}
ggplot(data = pca_variates,
       aes(x = PC1, 
           y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, 
                           color = recurrence)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins") +
  xlab(paste("PC1", 
             round(pca_res$prop_expl_var$X[1]*100), 
             "% var explained")) + 
  ylab(paste("PC2", 
             round(pca_res$prop_expl_var$X[2]*100), 
             "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 10, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, 
                                    size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

There is no evident batch effect after PCA analysis.

There is a good number of individual patients that show important differences in their PCA variates values in recurrent vs primary.

```{r fig.width=5.8, fig.height=4, warning=FALSE}
ggplot(data = pca_variateswopat6,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, 
                           color = recurrence)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins (patient 6 excluded)") +
  xlab(paste("PC1", 
             round(pca_res$prop_expl_var$X[1]*100), 
             "% var explained")) + 
  ylab(paste("PC2", 
             round(pca_res$prop_expl_var$X[2]*100), 
             "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 10, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, 
                                    size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r pca_proteomics_figure}
pca_plot <- ggplot(data = pca_variateswopat6,
       aes(x = PC1, 
           y = PC2)) +
  geom_point(size = 4, aes(shape = Mix, 
                           color = Stage)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 4.5), 
            size = 2) +
  xlab(paste("PC1", 
             round(pca_res$prop_expl_var$X[1]*100), 
             "% var explained")) + 
  ylab(paste("PC2", 
             round(pca_res$prop_expl_var$X[2]*100), 
             "% var explained")) +
theme(axis.text.x = element_text(hjust = 0.5, 
                                 vjust = 0, 
                                 size = 7,
                                 angle = 360),
      axis.text.y = element_text(hjust = 0.95, 
                                 vjust = 0.2,
                                 size = 7),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(colour = "black", 
                                  fill=NA, 
                                  size=0.5),
      axis.title = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 8),
      legend.key.height= unit(3, 'mm'),
      legend.key.width= unit(3, 'mm'),
      legend.position="bottom")

ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

## Exploratory sPLS-DA

```{r}
plsda_rest1 <- plsda(X = t_expr_matnona,
                     Y = recurrence,
                     ncomp = 3)
```

```{r fig.width=5, fig.height=4}
plotIndiv(plsda_rest1,
          nd.names = TRUE, 
          ellipse = TRUE, 
          legend = FALSE, 
          title = "sPLS-DA_Rec-Prim", 
          size.title = rel(1.5), 
          size.xlabel = rel(1.5), 
          size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), 
          point.lwd = 0.8, 
          cex = 3)  
```

Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.

Therefore we decided to exclude patient 6 from inferential analyses.

### sPLS-DA excluding 'patient 6' {.unnumbered .unlisted}

```{r}
plsda_restwo6 <- plsda(X = t_expr_matnonawo6,
                       Y = recurrencewo6,
                       ncomp = 3)
```

```{r fig.width=5, fig.height=4}
plotIndiv(plsda_restwo6,
          nd.names = TRUE, 
          ellipse = TRUE, 
          legend = FALSE, 
          title = "sPLS-DA_Rec-Prim wo Pat6", 
          size.title = rel(1.5), 
          size.xlabel = rel(1.5), 
          size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), 
          point.lwd = 0.8, 
          cex = 3)  
```

### GGplot-based visualization of Exploratory sPLS-DA {.unnumbered .unlisted}

```{r}
# preprocess pca results 
plsda_variates <- plsda_restwo6$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", 
                    "recurrence", 
                    "mixture"),
           sep = "\\_", 
           extra = "merge") %>%
  mutate(paired_id = paste(sample, 
                           recurrence, 
                           sep = "_")) %>%
  dplyr::select(-c(mixture, 
                   recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id") 
```

```{r fig.width=5.5, fig.height=4.5, warning=FALSE}
ggplot(data = plsda_variates,
       aes(x = comp1, 
           y = comp2)) +
  geom_point(size = 4, aes(shape = mixture, 
                           color = recurrence)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 2)) +
  labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
  xlab(paste("sPLS-DA Comp-1", 
             round(plsda_restwo6$prop_expl_var$X[1]*100), 
             "% var explained")) + 
  ylab(paste("sPLS-DA Comp-2", 
             round(plsda_restwo6$prop_expl_var$X[2]*100), 
             "% var explained")) + 
  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 10, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill=NA, 
                                    size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

The difference between initial and recurrent tumor is better resolved in the sPLS-DA plot when excluding patient 6.

## Sparcity reduction

Only proteins that were identified and quantified in 2 out of 3 TMT mixtures were kept for the inferential analysis. The proteins missing in 1 mixture were imputed with `missForest`.

```{r}
# sel_proteins_missing function is listed in the helper functions
proteins2exclude <- sel_proteins_missing(quant_annotated,
                                         threshold = 1)
```

Exclude proteins from initial abundance matrix.

```{r}
expr_matrix_filt <- expr_matrix2 %>%
  filter(!Protein %in% proteins2exclude)
```

### Missing values before and after filtering {.unnumbered .unlisted}

```{r}
nofiltnas_plot <- vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("All quant proteins. Total =", 
                                   nrow(expr_matrix))) + 
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 12,
                                        face = "bold"))

filtnas_plot <- vis_miss(expr_matrix_filt) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("Only present in 2/3 TMT mixtures. Total =",
                                   nrow(expr_matrix_filt))) + 
          theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0.1, 
                                           size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 12,
                                        face = "bold"))
```

```{r fig.height=6, fig.width=10}
cowplot::plot_grid(nofiltnas_plot, 
                   filtnas_plot, 
                   ncol = 2)
```

## Missing value imputation (`missForest`)

```{r}
mat_filt <- expr_matrix_filt %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt <- t(mat_filt)

if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){

 t_imp_mat_filt <- missForest::missForest(t_mat_filt)
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

 saveRDS(t_imp_mat_filt,
         file = here("rds/missforest_imp_mat_filt.Rds"))

} else {

 t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

}
```

# Differential abundance analysis (DAA)

**Protein to Gene annotation**

```{r}
prot2gene <- identified_proteins %>%
  dplyr::select(Protein = protein_id, Gene = gene,
                Description = description)
```

**Prep protein abundance matrix**

```{r}
mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")
```

## Set up design matrix

Initial vs recurrent samples are patient-matched, therefore the design matrix is setup so the linear model can account for the patient random effect.

```{r}
design_wo6 <- model.matrix(~ patientwo6 + recurrencewo6)

rownames(design_wo6) <- sample_annotationwo6$paired_id
```

Check design matrix:

```{r}
print(design_wo6[1:4,])
```

Check expression matrix:

```{r}
mat_filt_impwo6[1:4,1:6]
```

## Fit linear model (`limma`)

```{r}
source(here("scr/fit_limmawo6.R"))
```

```{r, warning=FALSE}
# fit limma
limma_tab_wo6 <- fit_limmawo6(mat_filt_impwo6, 
                         design_wo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect",
                         prot2gene = prot2gene)

#get proteins increased in recurrence
increased_in_rec <- limma_tab_wo6 %>%
  filter(logFC > 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)

#get proteins decreased in recurrence
decreased_in_rec <- limma_tab_wo6 %>%
  filter(logFC < 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)
```

## Volcano plot (DAA visualization)

```{r fig.width=6.5, fig.height=5.5, warning=FALSE}
our_volcano(limma_tab_wo6, 
            FC_cutoff = 0, 
            pval_cutoff = 0.05, 
            color_diffex = "red", 
            color_nondifex = "#2a9d8f", 
            interesting_proteins = NULL, 
            vert_line_col = "red",
            hline_col = "red", 
            hline_pos = 0.05, 
            linetype = "dashed",
            increased_in = "Recurrent", 
            comparison_title = "Recurrent vs Initial") +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 10, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

### Generate manuscript-ready volcano-plots {.unnumbered .unlisted}

```{r generate_volcano, fig.width=8, fig.height=6, echo=FALSE, warning=FALSE}
protlabels <- limma_tab_wo6 %>%
      filter(logFC > 0 | logFC < 0,
             adj.P.Val < 0.05)

signi_hits <- nrow(protlabels)

signi_up <- limma_tab_wo6 %>%
      filter(logFC > 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

signi_down <- limma_tab_wo6 %>%
      filter(logFC < 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

size <- 0.4

volcano <- ggplot(data = limma_tab_wo6,
                  mapping = aes(x = logFC, 
                                y = -log10(adj.P.Val))) +
      geom_point(data = limma_tab_wo6 %>% 
                 filter(logFC > 0,
                        adj.P.Val < 0.05),
                 mapping = aes(x = logFC, 
                               y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = limma_tab_wo6 %>% 
                   filter(logFC < -0,
                          adj.P.Val < 0.05),
                 mapping = aes(x = logFC, 
                               y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% 
                   filter(logFC > 0,
                          adj.P.Val > 0.05),
                 mapping = aes(x = logFC, 
                               y = -log10(adj.P.Val)), 
                 color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% 
                   filter(logFC < -0,
                          adj.P.Val > 0.05),
                 mapping = aes(x = logFC, 
                               y = -log10(adj.P.Val)), 
                 color = "#2a9d8f",
                 size = size) +
      ggrepel::geom_text_repel(data = protlabels,
                               aes(label = Gene), 
                               size = 1.5) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", 
                 linetype = "dashed") +
      xlab("logFC - Recurrent / Primary")

```

```{r echo=FALSE}
ggsave(plot = volcano, 
       filename = here::here("figures/volcano_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano, 
       filename = here::here("figures/volcano_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

```{r fig.width=8, fig.height=6, echo=FALSE}
protlabels <- limma_tab_wo6 %>%
      filter(logFC > 0 | logFC < 0,
             adj.P.Val < 0.05,
             Gene %in% c("ASAH1", "GPNMB", "SYNM", "CD14"))

size <- 0.4

volcano2 <- ggplot(data = limma_tab_wo6,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size)+
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      ggrepel::geom_text_repel(data = protlabels,
                               aes(label = Gene), size = 3.5) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Primary")

```

```{r echo=FALSE}
ggsave(plot = volcano2, 
       filename = here::here("figures/volcano2_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano2, 
       filename = here::here("figures/volcano2_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

### Write tabular results (for supplementary tables) {.unnumbered .unlisted}

```{r}
write_tsv(x = limma_tab_wo6, 
          file = here("suppl_tables/diff_abundance_gbm_rec_vs_init_proteomics.tsv"))
```

## Boxplots specific proteins

```{r}
quant_anwgene <- left_join(quant_annotated, 
                           prot2gene)

tab_2_paired_boxplot <-  quant_anwgene %>%
  filter(Gene %in% c("ASAH1", 
                     "SYNM", 
                     "GPNMB",
                     "MPO"))
# prep data
slim_data <- pivot_longer(expr_matrix_filt, 
                          cols = 2:length(names(expr_matrix_filt)), 
                          names_to = c("patient",
                                       "condition"),
                          values_to = "Intensity",
                          names_sep = "\\_") %>% 
  group_by(Protein)

slim_data2 <- pivot_wider(slim_data, 
                          values_from = Intensity, 
                          names_from = condition) %>%
  filter(patient != "x6")

slim_data3 <- slim_data2 %>% 
  mutate(Abs_diff = rec-prim) %>%
  mutate(rank = row_number(Abs_diff)) %>% 
  dplyr::rename(Initial = prim, 
                Recurrent = rec) %>%
  ungroup() %>%
  filter(Protein %in% c("Q13510", 
                        "Q14956", 
                        "O15061",
                        "P05164")) %>%
  mutate(Gene = case_when(Protein == "Q13510" ~ "ASAH1",
                          Protein == "Q14956" ~ "GPNMB",
                          Protein == "O15061" ~ "SYNM",
                          Protein == "P05164" ~ "MPO"))
```

```{r fig.width=7.5, fig.height=5.9}
ggpaired(slim_data3,
         cond1 = "Initial",
         cond2 = "Recurrent", 
         fill = "condition",
         ylab = "Normalized Abundance",
         label = NULL,
         repel = TRUE,
         facet.by = "Gene", 
         point.size = 0.2, 
         line.size = 0.1) 
```

```{r echo=FALSE}
size <- 5

paired_boxplots <- ggpaired(slim_data3,
                            cond1 = "Initial",
                            cond2 = "Recurrent", 
                            fill = "condition",
                            ylab = "Normalized Abundance",
                            label = NULL,
                            repel = TRUE,
                            facet.by = "Gene", 
                            point.size = 0.2, 
                            line.size = 0.1) +
      theme(axis.text.x = element_text(hjust = 0.5, 
                                       vjust = 0, 
                                       size = 5, 
                                       angle = 360),
           axis.text.y = element_text(hjust = 0.95, 
                                      vjust = 0.2, 
                                      size = 5),
           panel.background = element_blank(),
           panel.grid.major = element_blank(),
           panel.border = element_rect(colour = "black", 
                                       fill = NA, 
                                       size = 0.5),
           axis.title = element_text(size = size),
           legend.position="none",
           strip.text = element_text(hjust = 0.5, 
                                     vjust = 0, 
                                     size = 4, 
                                     angle = 360,
                                     margin = margin(0.4,0,0.4,0, "mm"),
                                     lineheight = 0.3)) 
```

```{r echo=FALSE}
ggsave(plot = paired_boxplots, 
       filename = here::here("figures/paired_box_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 61,
       height = 45)

ggsave(plot = paired_boxplots, 
       filename = here::here("figures/paired_box_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 61,
       height = 45)
```

#### ASAH1 vs MPO abundance (Rec vs Init)

```{r}
tab_2_mporatio <- tab_2_paired_boxplot %>%
  filter(Gene %in% c("ASAH1", "MPO")) %>%
  dplyr::select(Gene,
                recurrence,
                Abundance,
                patient,
                sample_id) %>%
  mutate(Condition = case_when(recurrence == "prim" ~ "Initial", 
                               recurrence == "rec" ~ "Recurrent")) %>%
  dplyr::select(-recurrence)

mpo_ratio_prep <- tab_2_mporatio %>% 
  pivot_wider(values_from = Abundance,
              names_from = Gene) %>%
  filter(patient != "x6") %>%
  mutate(Ratio_MPO_vs_ASAH1 = ASAH1 - MPO) %>%
  dplyr::select(-c(MPO, ASAH1)) %>%
  pivot_wider(values_from = Ratio_MPO_vs_ASAH1,
              names_from = Condition) %>% 
  mutate(measure = "Delta(ASAH1 - MPO)")

mpo_ratio_prep2 <- tab_2_mporatio %>% 
  pivot_wider(values_from = Abundance,
              names_from = Gene) %>%
  filter(patient != "x6") 
```

```{r}
paired_boxplots_asah1mporatio <- ggpaired(mpo_ratio_prep,
                                          cond1 = "Initial",
                                          cond2 = "Recurrent", 
                                          fill = "condition",
                                          ylab = "Abundance Delta(ASAH1 - MPO)",
                                          label = NULL,
                                          repel = TRUE,
                                          point.size = 0.2, 
                                          line.size = 0.1) +
      stat_compare_means(paired = TRUE, 
                         vjust = 2, 
                         size = 1.5) +
      theme(axis.text.x = element_text(hjust = 0.5, 
                                       vjust = 0, 
                                       size = 5, 
                                       angle = 360),
           axis.text.y = element_text(hjust = 0.95, 
                                      vjust = 0.2, 
                                      size = 5),
           panel.background = element_blank(),
           panel.grid.major = element_blank(),
           panel.border = element_rect(colour = "black", 
                                       fill = NA, 
                                       size = 0.5),
           axis.title = element_text(size = size),
           legend.position="none",
           strip.text = element_text(hjust = 0.5, 
                                     vjust = 0, 
                                     size = 4, 
                                     angle = 360,
                                     margin = margin(0.4,
                                                     0,
                                                     0.4,
                                                     0, "mm"),
                                     lineheight = 0.3))  
```

##### Compare Deltas of abundance (Boxplot)

```{r fig.width=4.8, fig.height=3.6}
print(paired_boxplots_asah1mporatio) 
```

```{r echo=FALSE}
ggsave(plot = paired_boxplots_asah1mporatio, 
       filename = here::here("figures/paired_box_delta_asah1_mpo_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 60,
       height = 45)

ggsave(plot = paired_boxplots_asah1mporatio, 
       filename = here::here("figures/paired_box_delta_asah1_mpo_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 60,
       height = 45)
```

##### Scatter plot ASAH1 vs MPO

```{r}
scatter_asah1mpo <- ggplot(data = mpo_ratio_prep2,
                           aes(x = ASAH1,
                               y = MPO)) + 
                    geom_point(mapping = aes(color = Condition)) + 
                    geom_smooth(method = lm, 
                                se = FALSE, 
                                color = "black",
                                size = 0.5) 
```

```{r fig.width=4.8, fig.height=3.6}
print(scatter_asah1mpo)
```

```{r echo=FALSE}
ggsave(plot = scatter_asah1mpo, 
       filename = here::here("figures/scatterplot_asah1_mpo_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 60,
       height = 60)

ggsave(plot = scatter_asah1mpo, 
       filename = here::here("figures/scatterplot_asah1_mpo_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 60,
       height = 60)
```

## Differences in abundance of ASAH1 by type of resection

Load the patient clinical data

```{r echo=FALSE}
patient_table <- readxl::read_xlsx(path = here("data/Patients_information_FINAL.xlsx"))
```

```{r}
patients_proteomics <- patient_table %>%
  filter(`Mass Spectrometry` == "X") %>%
  dplyr::select(Sex, 
                `Age at surgery (years)`, 
                `Time-to-reccurence (days)`,
                `Type of resection`,
                `Tumor localization`, 
                Treatment,
                sample_id = `LAB-ID`) %>%
  mutate(`Time-to-reccurence (days)` = as.numeric(`Time-to-reccurence (days)`))
```

```{r}
sample2patient_id <- mpo_ratio_prep2 %>% 
  dplyr::select(sample_id, 
                patient) %>% 
  distinct()

patients_proteomics <- left_join(patients_proteomics, 
                                 sample2patient_id) %>%
  dplyr::select(-matches("sample"))
```

```{r}
resection_diff_asah1 <- left_join(mpo_ratio_prep2,
                                  patients_proteomics) %>% 
  dplyr::select(-MPO) %>%
  dplyr::rename(abundance_ASAH1 = ASAH1,
                type_resection = `Type of resection`) %>%
  na.exclude() %>%
  distinct()
```

```{r}
resection_diff_asah1plot <- ggplot(data = resection_diff_asah1, 
                                   mapping = aes(x = Condition,
                                                 y = abundance_ASAH1,
                                                 fill = type_resection)) + 
  geom_boxplot()
```

```{r}
resection_diff_asah1plot
```

```{r}
twowayanova_resection <- aov(abundance_ASAH1 ~ Condition * type_resection, 
                             data = resection_diff_asah1)
```

```{r}
summary(twowayanova_resection)
```

```{r}
TukeyHSD(twowayanova_resection)
```

There is no significant difference in ASAH1 abundance based on type of resection.

# ID coverage after sparcity reduction

```{r}
ids_summary2 <- quant_annotated %>% 
  dplyr::select(Protein, 
                mixture, 
                Abundance) %>% 
  distinct(Protein, 
           mixture, 
           .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "Quant_in_all",
                   `Nr of IDs` = nrow(expr_matnona))) %>%
  bind_rows(.,
            tibble(mixture = "Quant_2/3_mixtures",
                   `Nr of IDs` = nrow(mat_filt_impwo6))) %>%
  bind_rows(.,
            tibble(mixture = "Overall",
                   `Nr of IDs` = nrow(expr_matrix)))

ids_summary2 <- mutate(ids_summary2,
                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
                                           mixture == "tmt_2" ~ "Mix_2",
                                           mixture == "tmt_3" ~ "Mix_3",
                                           mixture == "Quant_in_all" ~ "Quant. in all",
                                           mixture == "Overall" ~ "Overall",
                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
  mutate(Mixture = factor(Mixture, levels = c("Overall","Quant. in 2/3", "Quant. in all",
                                               "Mix_3", "Mix_2", "Mix_1")))
```

Number of identified proteins: `r nrow(expr_matrix)`

Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`

Number of identified and quantified proteins 2 out of 3 batches: `r nrow(mat_filt_impwo6)`

```{r fig.height=4, fig.width=7, warning=FALSE}
ggplot(data = ids_summary2,
       aes(y = `Nr of IDs`, 
           x = mixture)) +
  geom_col(fill = "#008683") +

  geom_text(size = 5, 
            position = position_stack(vjust = 0.5), 
            aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Proteins by Mixture") +
  ylab("Nr of IDed and Quant Proteins") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0.1, 
                                   size = 14, 
                                   angle = 0),
        axis.text.y = element_text(hjust = 0.5, 
                                   vjust = 0.1, 
                                   size = 12, 
                                   angle = 0),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey"),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 1.5),
        axis.title = element_text(size = 12, 
                                  face = "bold"))
```

```{r echo=FALSE}
barplot_cover <- ggplot(data = ids_summary2,
       aes(y = `Nr of IDs`, 
           x = Mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 2, 
            position = position_stack(vjust = 0.5), 
            aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 7, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 7),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r echo=FALSE}
ggsave(plot = barplot_cover, 
       filename = here::here("figures/barplot_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = barplot_cover, 
       filename = here::here("figures/barplot_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

# Exploratory analysis after sparcity reduction

## Plots of normalized abundance distribution

```{r}
expr_matrix_filtwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6"))

long_mat_f <- pivot_longer(expr_matrix_filtwo6,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_3 <- long_mat_f %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq2 <- long_mat_3 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

paired_annot_uniq2$paired_id
## Annotated abundance data in long format ----

quant_annotated_f <- left_join(long_mat_3,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"),
         paired_id = factor(paired_id, levels = paired_annot_uniq2$paired_id)) %>%
  mutate(Sample = str_replace_all(paired_id, pattern = "prim", replacement = "init"),
         Stage = if_else(recurrence == "prim",
                         true = "Initial",
                         false = "Recurrent"))

```

```{r}
expr_matrix_filtwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6"))

long_mat_f <- pivot_longer(expr_matrix_filtwo6,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_3 <- long_mat_f %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq2 <- long_mat_3 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

paired_annot_uniq2$paired_id
## Annotated abundance data in long format ----

quant_annotated_f <- left_join(long_mat_3,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"),
         paired_id = factor(paired_id, levels = paired_annot_uniq2$paired_id)) %>%
  mutate(Sample = str_replace_all(paired_id, pattern = "prim", replacement = "init"),
         Stage = if_else(recurrence == "prim",
                         true = "Initial",
                         false = "Recurrent"))

```

### Normalized ion reporter abundance distribution by TMT channel/sample {.unnumbered}

```{r warning=FALSE}
qcplot_f <- ggplot(quant_annotated_f,
                   mapping = aes(x = Sample, 
                                 y = Abundance, 
                                 fill = Stage)) +
                   geom_boxplot(outlier.size=0.1, size = 0.25) + 
                   theme(axis.text.x = element_text(hjust = 0.5, 
                                                    vjust = 0, 
                                                    size = 7, 
                                                    angle = 90),
                         axis.text.y = element_text(hjust = 0.5, 
                                                    vjust = 0, 
                                                    size = 6),
                         panel.background = element_blank(),
                         panel.grid.major = element_blank(),
                         panel.border = element_rect(colour = "black", 
                                                     fill = NA, 
                                                     size = 0.5),
                         axis.title=element_text(size = 8),
                         axis.title.x = element_blank(),
                         legend.text = element_text(size = 7),
                         legend.title = element_text(size = 8),
                         legend.key.height= unit(3, 'mm'),
                         legend.key.width= unit(3, 'mm'),
                         legend.position="bottom")
```

```{r fig.height=4, fig.width=7, warning=FALSE, echo=FALSE}
print(qcplot_f)
```

```{r echo=FALSE}
ggsave(plot = qcplot_f, 
       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = qcplot_f, 
       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

# Over-representation analyses/tests (ORAs)

## Prep tabular info for ORAs {.unnumbered .unlisted}

```{r}
tab_sig_prots_tryptic <- bind_rows(tibble(protein = increased_in_rec,
                                  characteristic = "up-regulated"),
                           tibble(protein = decreased_in_rec,
                                  characteristic = "down-regulated"))

limma_tab_wo62_tryptic <- limma_tab_wo6

unip2symb_tryptic <- bitr(limma_tab_wo62_tryptic$Protein, 
                  fromType = "UNIPROT", 
                  toType = c("SYMBOL", "ENTREZID"),
                  OrgDb = org.Hs.eg.db) %>%
  dplyr::rename(protein = UNIPROT) %>%
  mutate(Protein = protein)

tab_sig_prots_tryptic <- left_join(tab_sig_prots_tryptic, unip2symb_tryptic)

unip2symbIDed_tryptic <- bitr(identified_proteins$protein_id, 
                  fromType = "UNIPROT", 
                  toType = c("SYMBOL", "ENTREZID"),
                  OrgDb = org.Hs.eg.db)

limma_tab_wo62_tryptic <- left_join(limma_tab_wo62_tryptic, unip2symb_tryptic)

geneList <- limma_tab_wo62_tryptic$logFC
names(geneList) <- limma_tab_wo62_tryptic$Protein
geneList <-  sort(geneList, decreasing = TRUE)

geneListr <- limma_tab_wo62_tryptic$logFC
names(geneListr) <- limma_tab_wo62_tryptic$ENTREZID
geneListr <-  sort(geneListr, decreasing = TRUE)
```

## Reactome

```{r}
group_comparison_react <- compareCluster(ENTREZID~characteristic, 
                                              data=tab_sig_prots_tryptic, 
                                              fun="enrichPathway",
                                              organism = "human",
                                              pvalueCutoff = 0.05,
                                              pAdjustMethod = "BH",
                                              qvalueCutoff = 0.2,
                                              universe = unip2symbIDed_tryptic$ENTREZID,
                                              minGSSize = 10,
                                              maxGSSize = 1000,
                                              readable = TRUE)
```

### Dotplot Reactome

```{r warning=FALSE, message=FALSE}
dotplot_react <- enrichplot::dotplot(group_comparison_react, 
                                     x = "characteristic") + 
          xlab("Quant in Recurrent status") +
  scale_color_continuous(low = "red", 
                         high = "blue",
            guide=guide_colorbar(reverse = TRUE),
            breaks = c(0.01, 
                       0.04)) +
  scale_size_continuous(breaks = c(0.10, 
                                   0.25)) +
  theme(axis.text.x = element_text(hjust = 0.4, 
                                   vjust = 0.1, 
                                   size = 6, 
                                   angle = 340),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 6),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.height= unit(2, 'mm'),
        legend.key.width= unit(2, 'mm'),
        legend.position="bottom")
```

```{r}
print(dotplot_react)
```

```{r}
ggsave(plot = dotplot_react, 
       filename = here::here("figures/dotplot_reactome_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 91,
       height = 90)

ggsave(plot = dotplot_react, 
       filename = here::here("figures/dotplot_reactome_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 91,
       height = 90)
```

### Cnet plot Reactome

```{r}
cnet_plot <- enrichplot::cnetplot(group_comparison_react) + 
  theme(legend.text = element_text(size = 7),
        text = element_text(size = 4),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 'mm'),
                legend.key.width = unit(3, 'mm'),
                legend.position = "bottom")
```

```{r}
print(cnet_plot)
```

```{r}
ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 110,
       height = 110)

ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 110,
       height = 110)

ggsave(plot = cnet_plot, 
       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.svg"), 
       device = "svg",
       units = "mm",
       width = 110,
       height = 110)
```

# Protein coverage plots

```{r}
source(here("scr/protein_peptide_coverage_plot.R"))
source(here("scr/annotate_peptides.R"))
```

## Annotate peptides from interesting proteins {.unnumbered .unlist}

```{r}
# get protein to peptide pairs
pept2prot_inter <- identified_peptides %>%
  filter(gene %in% c("ASAH1", 
                     "SYNM", 
                     "GPNMB")) %>%
  dplyr::select(Peptide = sequence, 
                Genes = protein_id)

# map peptides to protein sequence
annotated_peptides <- annotate_peptides(expr_mat = pept2prot_inter,
                                        fasta = fasta_ident_prots)

annotated_peptides2 <- annotated_peptides %>%
  mutate(peptide_length = str_length(Peptide))
```

## ASAH1

```{r}
feat_asah1 <- drawProteins::get_features("Q13510")

feat_asah1df <- drawProteins::feature_to_dataframe(feat_asah1)
#Get peptides as coverage features for plotting

cov_feat_asah1 <- get_coverage(annotated_peptides = annotated_peptides2, 
                               id = "Q13510")
#Merge coverage feature info with protein sequence features info:

fullfeat_asah1 <- add_peptides(feat_asah1df,
                               peptide_coverage_data = cov_feat_asah1)
# drawCanvas
canvas <- draw_canvas(fullfeat_asah1)

# draw the protein chain
wchain <- draw_chains(canvas, fullfeat_asah1)

# draw protein domains
wdomainsasah1 <- draw_domains(wchain, data = fullfeat_asah1)
#pasa <- drawProteins::draw_folding(wdomainsasah1, fullfeat_asah1)
cov_feat_asah1$order <- 1.4

pepcov_asah1 <- draw_peptides(wdomainsasah1, cov_feat_asah1)
```

```{r}
print(pepcov_asah1)
```

```{r}
ggsave(plot = pepcov_asah1, 
       filename = here::here("figures/asah1_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_asah1, 
       filename = here::here("figures/asah1_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

## SYMN

```{r}
feat_SYNM <- drawProteins::get_features("O15061")
feat_SYNMdf <- drawProteins::feature_to_dataframe(feat_SYNM)
#Get peptides as coverage features for plotting

cov_feat_SYNM <- get_coverage(annotated_peptides = annotated_peptides2, 
                               id = "O15061")
#Merge coverage feature info with protein sequence features info:

fullfeat_SYNM <- add_peptides(feat_SYNMdf,
                               peptide_coverage_data = cov_feat_SYNM)
# drawCanvas
canvasSYNM <- draw_canvas(fullfeat_SYNM)

# draw the protein chain
wchainSYNM <- draw_chains(canvasSYNM, fullfeat_SYNM)
cov_feat_SYNM$order <- 1.4

pepcov_SYNM <- draw_peptides(wchainSYNM, cov_feat_SYNM)
```

```{r}
print(pepcov_SYNM)
```

```{r}
ggsave(plot = pepcov_SYNM, 
       filename = here::here("figures/SYNM_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_SYNM, 
       filename = here::here("figures/SYNM_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

# Associate protein expression with Time to Recurrence

```{r}
library(readxl)
```

```{r echo=FALSE}
latency_data <- read_excel(path = here("data/ELISA_LATENCY_KM.xlsx"), 
                           skip = 1
                           ) %>%
  clean_names()
```

```{r echo=FALSE}
patient_table <- readxl::read_xlsx(path = here("data/Patients_information_FINAL.xlsx"))
```

```{r echo=FALSE}
latency_data %>%
  dplyr::mutate(ASAH1.ng.ml = log2(asah1_ng_ml_6+1),
                SYNM.ng.ml = log2(synm_ng_ml_7+1),
                GPNMB.ng.ml = log2(gpnmb_ng_ml_8+1),
                MMP.9.ng.ml = log2(mmp_9_ng_ml_9+1)) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1.ng.ml>=mean(asah1_ng_ml_6), "high", "low"),
                SYNM = ifelse(SYNM.ng.ml>=mean(synm_ng_ml_7), "high", "low"),
                GPNMB = ifelse(GPNMB.ng.ml>=mean(gpnmb_ng_ml_8), "high", "low"),
                MMP9 = ifelse(MMP.9.ng.ml>=mean(mmp_9_ng_ml_9, na.rm=T), "high", "low")) -> latency_data
```

```{r}
patients_proteomics <- patient_table %>%
  filter(`Mass Spectrometry` == "X") %>%
  dplyr::select(Sex, 
                `Age at surgery (years)`, 
                `Time-to-reccurence (days)`,
                `Type of resection`,
                `Tumor localization`, 
                Treatment,
                sample_id = `LAB-ID`) %>%
  mutate(`Time-to-reccurence (days)` = as.numeric(`Time-to-reccurence (days)`))
```

## What's the association of ASAH1/SYNM/GPNNMB/MMP9 at Initial stage with time to recurrence?

```{r}
quant_annot_intr_init <- quant_annotated %>%
  filter(Protein %in% c("Q13510", 
                        "Q14956", 
                        "O15061",
                        "P14780"),
         recurrence == "prim") %>%
  mutate(Gene = case_when(Protein == "Q13510" ~ "ASAH1a",
                          Protein == "Q14956" ~ "GPNMBa",
                          Protein == "O15061" ~ "SYNMa",
                          Protein == "P14780" ~ "MMP9a")) %>% 
  left_join(patients_proteomics) %>% 
  dplyr::select(sample_id,
                patient,
                Gene,
                Abundance, 
                `Time-to-reccurence (days)`) %>%
  pivot_wider(names_from = Gene, 
              values_from = Abundance) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1a >= mean(ASAH1a), 
                               "high", 
                               "low"),
                SYNM = ifelse(SYNMa >= mean(SYNMa), 
                              "high", 
                              "low"),
                GPNMB = ifelse(GPNMBa >= mean(GPNMBa), 
                               "high", 
                               "low"),
                MMP9 = ifelse(MMP9a >= mean(MMP9a, 
                                            na.rm=T), 
                              "high", 
                              "low"))

labid2patientn <- quant_annot_intr_init %>% 
  dplyr::select(patient, sample_id) %>% 
  dplyr::distinct()

patients_proteomics <- left_join(labid2patientn, 
                                 patients_proteomics)

```

```{r}
library(survival)
library(survminer)
```

```{r}
fit1 <- survfit(Surv(`Time-to-reccurence (days)`) ~ ASAH1, 
               data = quant_annot_intr_init)
print(fit1)
# Change color, linetype by strata, risk.table color by strata
surv_p1 <- ggsurvplot(fit1,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit2 <- survfit(Surv(`Time-to-reccurence (days)`) ~ SYNM, 
               data = quant_annot_intr_init)
print(fit2)
# Change color, linetype by strata, risk.table color by strata
surv_p2 <- ggsurvplot(fit2,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit3 <- survfit(Surv(`Time-to-reccurence (days)`) ~ GPNMB, 
               data = quant_annot_intr_init)
print(fit3)
# Change color, linetype by strata, risk.table color by strata
surv_p3 <- ggsurvplot(fit3,
          pval = TRUE, 
          conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit4 <- survfit(Surv(`Time-to-reccurence (days)`) ~ MMP9, 
               data = quant_annot_intr_init)
print(fit4)
# Change color, linetype by strata, risk.table color by strata
surv_p4 <- ggsurvplot(fit4,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")
```

```{r}
surv_p1
surv_p2
surv_p3
surv_p4
```

## What's the association of ASAH1/SYNM/GPNNMB/MMP9 at Recurrent stage with time to recurrence?

```{r}
quant_annot_intr_rec <- quant_annotated %>%
  filter(Protein %in% c("Q13510", 
                        "Q14956", 
                        "O15061",
                        "P05164"),
         recurrence == "rec") %>%
  mutate(Gene = case_when(Protein == "Q13510" ~ "ASAH1a",
                          Protein == "Q14956" ~ "GPNMBa",
                          Protein == "O15061" ~ "SYNMa",
                          Protein == "P05164" ~ "MMP9a"))  %>% 
  left_join(patients_proteomics, by = "patient") %>% 
  dplyr::select(patient,
                Gene,
                Abundance, 
                `Time-to-reccurence (days)`) %>%
  pivot_wider(names_from = Gene, 
              values_from = Abundance) %>%
  dplyr::mutate(ASAH1 = ifelse(ASAH1a >= mean(ASAH1a), 
                               "high", 
                               "low"),
                SYNM = ifelse(SYNMa >= mean(SYNMa), 
                              "high", 
                              "low"),
                GPNMB = ifelse(GPNMBa >= mean(GPNMBa), 
                               "high", 
                               "low"),
                MMP9 = ifelse(MMP9a >= mean(MMP9a, 
                                            na.rm=T), 
                              "high", 
                              "low")) 
```

```{r}
fit12 <- survfit(Surv(`Time-to-reccurence (days)`) ~ ASAH1, 
               data = quant_annot_intr_rec)
print(fit12)
# Change color, linetype by strata, risk.table color by strata
surv_p12 <- ggsurvplot(fit12,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit22 <- survfit(Surv(`Time-to-reccurence (days)`) ~ SYNM, 
               data = quant_annot_intr_rec)
print(fit22)
# Change color, linetype by strata, risk.table color by strata
surv_p22 <- ggsurvplot(fit22,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit32 <- survfit(Surv(`Time-to-reccurence (days)`) ~ GPNMB, 
               data = quant_annot_intr_rec)
print(fit32)
# Change color, linetype by strata, risk.table color by strata
surv_p32 <- ggsurvplot(fit32,
          pval = TRUE, 
          conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")

fit42 <- survfit(Surv(`Time-to-reccurence (days)`) ~ MMP9, 
               data = quant_annot_intr_rec)
print(fit42)
# Change color, linetype by strata, risk.table color by strata
surv_p42 <- ggsurvplot(fit42,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x="Time to Recurrence (Days)", y = "Time to Recurrence Probability")
```

```{r}
surv_p12
surv_p22
surv_p32
surv_p42
```

### 2023-02-14 reports/gbm_recurrence_general_proteomics_manuscript_report.qmd
proteomics report - v2

@@
     toc-depth: 4 # default is 3
-    toc-title: Contents
-    toc-location: left # default is right, option is 'body'
     number-sections: true
-    number-depth: 3
-    embed-resources: true
+    execute:
+      output: true
 always_allow_html: true

### 2023-02-14 reports/gbm_recurrence_general_proteomics_manuscript_report.qmd
proteomics report add toc

@@
   gfm:
+    toc: true 
     toc-depth: 4 # default is 3
@@
 always_allow_html: true
-editor: visual
+editor: source
 ---

### 2023-02-16 reports/gbm_recurrence_large_scale_proteolysis_manuscript_report.qmd
proteolysis_report_clean_up

---
title: "GBM recurrence - Shot gun analysis of proteolytic products"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---
```{r knirt_helper_funcs}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))
```

```{r load_packages}
## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)
library(ggrepel)
library(extrafont)
extrafont::loadfonts(device = "win")
```

```{r theme_set}
theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 'mm'),
                legend.key.width = unit(3, 'mm'),
                legend.position = "bottom",
                text = element_text(family = "Helvetica")))
```

The following report holds the code, results and interpretation of the proteomics data analysis of Initial-Recurrent glioblastoma (GBM) samples of matched patients.

# Sample annotation info

```{r echo=TRUE}
sample_annotation <- read_csv(here("data/sample_annotation.csv"))

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",
                         patient,
                         sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_")) %>%
  filter(recurrence %in% c("prim", 
                           "rec"))

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")
```

## Sample annotation

### Brief note on sample annotation: {.unnumbered .unlisted}

Twenty-two patient matched samples (corresponding to 11 patients) were processed for protein extraction, TMT-labelled and submitted via high-pH fractionation and further analyzed via LC-MS/MS.

```{r echo=FALSE, error=TRUE}
kbl(sample_annotation, booktabs = TRUE) %>%
  kable_styling(font_size = 14, 
                bootstrap_options = c("striped", 
                                      "hover"))
```

# Brief note on data analysis and processing:

Three proteomics data analysis approaches were used to explore the spectral data and will be briefly described:

1. General large scale proteomics.
2. Proteolytic processing.
3. Proteogenomics

The FragPipe bioinformatic pipeline was used for each of these approaches with varying parameters depending on the type of explorative approach.

**This report holds the code, analysis, visualizations and results for large scale analysis of proteolytic processing.**

__1. General large scale proteomics analyses__ : In order to explore the differential abundance of proteins between Recurrent and Initial tumor, a fully-tryptic search was performed against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides. A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__2. Proteolytic processing__: The large scale evaluation of differential proteolytic activity between recurrent vs primary tumor was executed via a semi-tryptic search against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides.  A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__3. Proteogenomics__: The explorative proteogenomics analysis started with the generation of a GBM-specific database from publicly available RNA-seq sequencing data from tumor specimens. Fastq files were obtain from the SRA BioProject PRJNA627121 (Cell Rep, 2021 Mar 2;34(9):108787). Eight samples containing paired reads were selected, based on: 1) the higher number of reads per file; 2) having a similar proportion of males and females and 3) not bearing isocitrate-dehyderogenase (IDH)-1 mutation (IDH WT). None of these patients undergo previous treatment, therefore potential variants in the sequencing data are not expected to be associated to treatment-related mutagenic processes. Spectral data was then searched against this custom database using a fully-tryptic approac. No minimum intensity threshold was set for reporter ion intensity for quantitation.

Exceptionally when stated differently within each sub-approach described above, all these searches were executed using the following parameters: MSFragger was used as a search engine, setting a fully or semi-tryptic specificity allowing for 1 missed cleavages. Precursor mass tolerance was set to -20/20 and fragment mass tolerance of 20 ppm with mass calibration and parameter optimization resulting in 5 ppm. Peptide N-terminal acetylation and peptide N-terminal TMT labeling were set as variable modifications. TMT labelling at K and carbamidomethylation of C were set as fixed modifications. MSBooster was used for deep-learning based predictions of retention time and spectra. Predicted features were used by Percolator for post-processing scoring and false discovery rate (FDR) control via target-decoy competition of all peptide-to-spectrum matches (PSMs) obtained from MSFragger search. ProteinProphet was used as a protein inference algorithm keeping all PSMs with probability score bigger than 0.9. Final report and FDR estimation was based on the filtered PSM and protein lists before protein scoring.

Relative quantitation of identified peptides within each sample was performed via their reported ion intensities using TMT-integrator. Only PSMs coming from unique peptides with a minimum probability of 0.9 and an isotopic purity of at least 50% were considered for quantitation. Quantitative values were normalized via median centering and summarized by protein/gene and peptide using a virtual reference channel, to finally generate protein and peptide normalized abundance matrices of abundance used for further statistical processing in combination with the sample annotation.

# Initial data loading and wrangling {.unnumbered .unlisted}

```{r}
# semi-specific search
tmt_peptdata_frag <- read_tsv(here("data/semi_specific_search_fragpipe17/tmt-report/abundance_peptide_MD.tsv")) %>% 
  janitor::clean_names()

pept_ids_mix1 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_1/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("empty")))

pept_ids_mix2 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_2/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("x"), 
                           starts_with("empty"), 
                           "master"))

pept_ids_mix3 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_3/peptide.tsv")) %>%
                    clean_names() %>%
          dplyr::select(-c(starts_with("x"), 
                           starts_with("empty"), 
                           "master"))

identified_proteins_semi <- read_tsv(here("data/semi_specific_search_fragpipe17/combined_protein.tsv")) %>% 
  janitor::clean_names()

#specific search
# tmt-integrator output loading 
tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
  janitor::clean_names()

identified_peptides <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_peptide.tsv")) %>%
  janitor::clean_names()

```

```{r}
prot2gene <- identified_proteins_semi %>%
  dplyr::select(Protein = protein_id, Gene = gene,
                Description = description)

```

```{r}
# wide expression matrix with NAs
pept_matrix <- dplyr::select(tmt_peptdata_frag,
                             index, 
                             starts_with("x"))

pept_long_mat <- pivot_longer(pept_matrix,
                         cols = starts_with("x"),
                         names_to = c("patient"),
                         values_to = "Abundance") %>%
  mutate(recurrence = str_extract(patient, "[^0-9]+$")) %>%
  mutate(patient = str_remove_all(patient, "[^0-9]+$"))

pept_long_mat_2 <- pept_long_mat %>%
  mutate(paired_id_semi = paste(patient, 
                                recurrence, 
                                sep = "_"))

sample_annotation2 <- sample_annotation2 %>%
  mutate(paired_id_semi = paste(patient, 
                                recurrence, 
                                sep = "_"))

## Annotated abundance data in long format ----

pept_quant_annotated <- left_join(pept_long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id_semi", 
                                    "patient",
                                    "recurrence")) %>%
  mutate(Channel_mix = paste(mixture,
                             channel,  
                             sep = "_"))

## Wide xpression matrices without NAs
pept_expr_matnona <- column_to_rownames(pept_matrix,
                                        "index") %>%
          as.matrix() %>%
          na.omit()
```

```{r}
# wide expression matrix with NAs
expr_matrix <- dplyr::select(tmt_protdata_frag,
                             Protein = index, starts_with("x"))

long_mat <- pivot_longer(expr_matrix,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_2 <- long_mat %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq <- long_mat_2 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",patient, sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
  filter(recurrence %in% c("prim", "rec"))

## Annotated abundance data in long format ----

quant_annotated <- left_join(long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

## Wide xpression matrices without NAs
expr_matnona <- column_to_rownames(expr_matrix,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()
```

Wrangle and merge peptide ids for further annotation based on they proteolytic specificity

```{r}
peptide_joined1 <- full_join(pept_ids_mix1, 
                             pept_ids_mix2) %>%
                    mutate(duplicated = duplicated(peptide))

peptide_joined <- full_join(peptide_joined1, 
                            pept_ids_mix3) %>%
                    mutate(duplicated = duplicated(peptide))

peptide_ids_filtered <- filter(peptide_joined,
                               peptide %in% peptide_joined$peptide) %>%
                    distinct(peptide,
                             assigned_modifications,
                             observed_modifications, 
                             .keep_all = TRUE) %>%
                    dplyr::rename(Peptide = peptide) %>%
                    dplyr::mutate(Genes = protein_id)
```

# Sparcity reduction

Only peptides that were identified and quantified in all 3 mixtures would be used for further inferential analyses for proteolytic processing.

```{r}
pept_matrix_nona <- pept_matrix %>% 
  na.omit()

pept_nona <- pept_matrix_nona %>%
  dplyr::select(index) %>%
  separate(col = index, 
           sep = "_", 
           into = c("protein", 
                    "peptide"))

pept_matrix_nonawop6 <- pept_matrix %>% 
  dplyr::select(-starts_with("x6")) %>%
  na.omit()

```

# Coverage information

```{r}
ids_summarypep <- pept_quant_annotated %>% 
  dplyr::select(index, 
                mixture, 
                Abundance) %>% 
  distinct(index, 
           mixture, 
           .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "In all three",
                   `Nr of IDs` = nrow(pept_matrix_nona)))

ids_summarypep <- mutate(ids_summarypep,
                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
                                           mixture == "tmt_2" ~ "Mix_2",
                                           mixture == "tmt_3" ~ "Mix_3",
                                           mixture == "In all three" ~ "Quant. in all",
                                           mixture == "Overall" ~ "Overall",
                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
  mutate(Mixture = factor(Mixture, 
                          levels = c("Overall", 
                                     "Quant. in all",
                                     "Mix_3", 
                                     "Mix_2", 
                                     "Mix_1")))
```

```{r warning=FALSE, message=FALSE}
barplot_pept_cover <- ggplot(data = ids_summarypep,
                             aes(y = `Nr of IDs`, 
                                 x = Mixture)) +
                        geom_col(fill = "#008683") + 
                        geom_text(size = 5, 
                                  position = position_stack(vjust = 0.5), 
                                  aes(label = `Nr of IDs`)) +
                        coord_cartesian(ylim = c(0, 
                                                 5000))+
                        coord_flip() +
                        labs(title = "Number of Identified and quantified Peptides by Mixture") +
                        ylab("Nr of IDed and Quant Peptides") + 
                        xlab("TMT Mixture") + 
                        theme(axis.text.x = element_text(hjust = 0.5, 
                                                         vjust = 0, 
                                                         size = 7, 
                                                         angle = 360),
                              axis.text.y = element_text(hjust = 0.95,
                                                         vjust = 0.2, 
                                                         size = 7),
                              panel.background = element_blank(),
                              panel.grid.major = element_blank(),
                              panel.border = element_rect(colour = "black", 
                                                          fill = NA, 
                                                          size = 0.5),
                              axis.title.x = element_blank(),
                              axis.title.y = element_blank(),
                              legend.text = element_text(size = 7),
                              legend.title = element_text(size = 8),
                              legend.key.height = unit(3, 'mm'),
                              legend.key.width = unit(3, 'mm'),
                              legend.position = "bottom")
```

```{r fig.height=4, fig.width=5.5, warning=FALSE}
print(barplot_pept_cover)
```

```{r}
ggsave(plot = barplot_pept_cover, 
       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 38,
       height = 50)

ggsave(plot = barplot_pept_cover, 
       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 38,
       height = 50)
```

# Annotate peptides with cleavage information

In this section we annotate peptides by their specificity and cleavage information.

```{r}
if(!file.exists(here("data/semi_specific_search_fragpipe17/protein_combined.fas"))){
  source(here("scr/prepare_combined_fasta.R"))
}

fasta <- read.fasta(file = here("data/semi_specific_search_fragpipe17/protein_combined.fas"),
                    seqtype = "AA", 
                    as.string = TRUE)
```

```{r}
if(!file.exists(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))){

  cleavage_annoated_peptides <- annotate_peptides(expr_mat = peptide_ids_filtered, 
                                                  fasta = fasta,
                                                  decoy_tag = "rev_")

  write_tsv(cleavage_annoated_peptides,
            here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))

} else {

  cleavage_annoated_peptides <- read_tsv(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))

}

cleavage_annoated_peptides <- dplyr::mutate(cleavage_annoated_peptides,
                                            Genes = protein_id)

peptides_annotated <- left_join(peptide_ids_filtered, 
                                cleavage_annoated_peptides,
                                by = c("Peptide", 
                                       "protein_id"))
```

# Annotate N-terminal modifications of peptides

```{r echo=TRUE}
nterannot <- annotate_nterm(peptides_annotated,
                            tmtmass = 229.1629) %>%
  clean_names()
```

```{r}
exp_nterannot <- nterannot %>%
  mutate(ntercount = str_count(string = assigned_modifications, 
                               pattern = "N-term"))
```

## Merge peptide annotation with TMT integrator quant report

```{r}
# all quantified peptides
tmt_reprt_annot <- left_join(tmt_peptdata_frag, 
                             nterannot)

check_na <- filter(tmt_reprt_annot, 
                   is.na(semi_type))

excluded_tmt_reprt_annot <- anti_join(nterannot, 
                                      tmt_reprt_annot) %>% 
                    clean_names()

#only those present in all mixtures
tmt_peptdata_fragnona <- tmt_peptdata_frag %>%
  filter(peptide %in% pept_nona$peptide)

tmt_reprt_annotnona <- left_join(tmt_peptdata_fragnona, 
                                 nterannot) 
```

```{r echo=TRUE}
source(here("scr/summarize_peptide_counts.R"))
```

```{r echo=TRUE}
summary_count <- summarize_peptide_counts(tmt_reprt_annot) 
```

## Barplot

These are the ones that would be used for inferential statistics.

```{r echo=TRUE}
to_count_info <- tmt_reprt_annot %>% 
   filter(index %in% pept_matrix_nona$index) %>%
                      dplyr::select(index, 
                                    specificity, 
                                    nterm, 
                                    semi_type, 
                                    tmt_tag,
                                    aa_after, 
                                    aa_before, 
                                    following_10_resid, 
                                    previous_10_resid, 
                                    start_position, 
                                    end_position) %>%
  mutate(specificity = if_else(str_detect(semi_type, 
                                          "unspecific"), 
                               true = "semi_specific",
                               false = specificity),
         duplicated = duplicated(index)) %>%
  distinct()

to_count_info_semi <- to_count_info %>%
                           dplyr::select(index,
                                         specificity) %>%
                           distinct() %>%
  mutate(duplicated = duplicated(index))

false_duplication <- to_count_info_semi %>%
  filter(duplicated) %>%
  pull(index)

to_count_info_semi <- to_count_info_semi %>%
  mutate(specificity = if_else(index %in% false_duplication, 
                               true = "semi_specific",
                               false = specificity)) %>%
  dplyr::select(-duplicated) %>%
  distinct()

  n_semi <- dplyr::count(to_count_info_semi, 
                         specificity) %>% 
                      dplyr::rename(feature_type = specificity) %>%
                      dplyr::mutate(category = "specificity")

  n_total <- tibble(feature_type = "Total",
                    n = length(unique(to_count_info$index)),
                    category = "Total")

  n_term <- dplyr::count(to_count_info, 
                         nterm) %>% 
                      dplyr::rename(feature_type = nterm) %>%
                      dplyr::mutate(category = "N-term")

  summary_count_pept_nona <- bind_rows(n_semi,
                                       n_term,
                                       n_total)

total_sumcount <- summary_count_pept_nona %>%
  filter(feature_type == "Total") %>%
  pull(n)

summary_countnona_no_tot <- summary_count_pept_nona %>%
  dplyr::filter(category != "Total") %>%
  dplyr::rename(`Feature category` = category)
```

```{r echo=FALSE, warning=FALSE}
summary_pept_annot <- ggplot(summary_countnona_no_tot,
                             aes(x = feature_type, 
                                 y = n, 
                                 fill = `Feature category`)) + 
        coord_flip()+ 
        geom_col() +
        geom_text(aes(label = n), 
                  hjust = -0.2, 
                  size = 2.5) + 
        scale_y_continuous(limits = c(0, 
                                      17000), 
                           breaks = c(0, 
                                      7000, 
                                      15000)) + 
        labs(y = "Number of IDs", 
             x = "Feature type") +
        facet_wrap(~`Feature category`, 
                   nrow = 1, 
                   scales = "free") +
        theme(axis.text.x = element_text(hjust = 0.5, 
                                         vjust = 0, 
                                         size = 7, 
                                         angle = 360),
              axis.text.y = element_text(hjust = 0.95, 
                                         vjust = 0.2, 
                                         size = 7),
              panel.background = element_blank(),
              panel.grid.major = element_blank(),
              panel.border = element_rect(colour = "black", 
                                          fill = NA, 
                                          size = 0.5),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              legend.text = element_text(size = 7),
              legend.title = element_text(size = 8),
              legend.key.height= unit(3, 'mm'),
              legend.key.width= unit(3, 'mm'),
              legend.position="bottom", 
              strip.background = element_blank(),
              strip.text = element_blank())
```

```{r fig.width=5, fig.height=2, warning=FALSE}
print(summary_pept_annot)
```

```{r}
ggsave(plot = summary_pept_annot, 
       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 76,
       height = 50)

ggsave(plot = summary_pept_annot, 
       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 76,
       height = 50)
```

# Differential abundance analysis (DAA)

## Prep design matrix {.unnumbered .unlisted}

```{r}
paired_id <- sample_annotation2$paired_id
recurrence <- sample_annotation2$recurrence
patient <- sample_annotation2$patient
mixture <- sample_annotation2$mixture

paired_idwo6 <- sample_annotationwo6$paired_id
recurrencewo6 <- sample_annotationwo6$recurrence
patientwo6 <- sample_annotationwo6$patient
mixturewo6 <- sample_annotationwo6$mixture
```

```{r}
design_limmawo6 <- model.matrix(~ patientwo6 + recurrencewo6)

sample_annotationwo6 <- sample_annotationwo6 %>%
  mutate(paired_id_semi = paste(patient, 
                                recurrence, 
                                sep = ""))

rownames(design_limmawo6) <- sample_annotationwo6$paired_id_semi
```

## Prep abundance matrices {.unnumbered .unlisted}

```{r}
pep_matwo6 <- dplyr::select(pept_matrix_nonawop6,
                             index, rownames(design_limmawo6))  %>%
  column_to_rownames("index") %>%
  as.matrix()
```

```{r}
colnames(pep_matwo6)
```

```{r}
rownames(design_limmawo6)
```

The column names in our peptide abundance matrix have the same order as the row names in our design matrix.

## Fitting linear model {.unlisted .unnumbered}

```{r echo=TRUE, warning=FALSE, message=FALSE}
fit_mod_peptwo6 <- fit_limmapeptwo6(pep_matwo6, 
                         design_limmawo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect")

compar_tab_peptwo6 <-  fit_mod_peptwo6 %>%
                    separate(Protein, 
                             into = c("protein", 
                                      "peptide"), 
                             remove = FALSE) %>%
  dplyr::rename(index = Protein)
```

## Extract coefficients {.unlisted .unnumbered}

```{r}
# load function for volcano plot
source(here("scr/plot_volcano.R"))
```

```{r}
features1 <- tmt_reprt_annotnona %>%
                    dplyr::select(peptide, index, specificity, nterm, 
                                  semi_type, is_terminal)

compar_tab_peptwo6_annn <- left_join(compar_tab_peptwo6,
                                     features1)

increased_recwo6_pept_annn <- compar_tab_peptwo6_annn %>%
                    filter(logFC > 0,
                           adj.P.Val < 0.05,
                           specificity == "semi_specific",
                           is_terminal == "not_terminal") %>% 
  pull(peptide)

decreased_recwo6_pept_annn <- compar_tab_peptwo6_annn %>%
                    filter(logFC < 0,
                           adj.P.Val < 0.05,
                           specificity == "semi_specific",
                           is_terminal == "not_terminal") %>% 
  pull(peptide)
```

## Volcano plot

```{r}
recwo6_pept22 <- compar_tab_peptwo6_annn %>%
  filter(adj.P.Val < 0.05,
         specificity == "semi_specific",
         is_terminal == "not_terminal")

recwo6_pept_no <- compar_tab_peptwo6_annn %>%
  filter(adj.P.Val < 0.05,
         specificity == "specific",
         is_terminal == "not_terminal")

compar_tab_peptwo6_annn2 <- compar_tab_peptwo6_annn %>%
  dplyr::rename(Protein = protein) %>%
 mutate(Feature = case_when(adj.P.Val < 0.05 & logFC > 0 & specificity == "semi_specific" & is_terminal == "not_terminal" ~ "Up-regulated Semi-tryptic",
                            adj.P.Val < 0.05 & logFC < 0 & specificity == "semi_specific" & is_terminal == "not_terminal" ~ "Down-regulated Semi-tryptic",
                            adj.P.Val < 0.05 & specificity == "specific" ~ "Tryptic",
                            TRUE ~ "No change")) %>%
 mutate(Feature = factor(Feature, 
                         levels = c("Up-regulated Semi-tryptic", 
                                    "Down-regulated Semi-tryptic",
                                    "No change",
                                    "Tryptic"))) %>%
  left_join(.,prot2gene)

gpnbm_pept_semi_lim <- compar_tab_peptwo6_annn2 %>%
  filter(Gene == "GPNMB")

diff_abund_semi <- compar_tab_peptwo6_annn2 %>% 
  filter(Feature %in% c("Up-regulated Semi-tryptic" , 
                        "Down-regulated Semi-tryptic"))

size <- 0.5

volcano_semi_gbm <- ggplot(data = compar_tab_peptwo6_annn2,
                      mapping = aes(x = logFC, 
                                    y = -log10(adj.P.Val),
                                    color = Feature)) + 
      geom_point(data = filter(compar_tab_peptwo6_annn2,
                               !Feature %in% c("Up-regulated Semi-tryptic" , 
                                               "Down-regulated Semi-tryptic")),
                 size = size) +
      geom_point(data = filter(compar_tab_peptwo6_annn2, 
                               Feature %in% c("Up-regulated Semi-tryptic" , 
                                              "Down-regulated Semi-tryptic")),
                 size = 1) + 
      scale_color_manual(values = c("Up-regulated Semi-tryptic" = "#ff6150",
                                    "Down-regulated Semi-tryptic" = "#54d2d2",
                                    "Tryptic" = "grey",
                                    "No change" = "darkgrey"
                                    )) +            
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
     theme(strip.text = element_text(vjust = 1))
```

```{r fig.width=5.5, fig.height=4, warning=FALSE}
print(volcano_semi_gbm)
```

```{r}
ggsave(plot = volcano_semi_gbm, 
       filename = here::here("figures/volcano_proteolysis_gbm_2.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 63)

ggsave(plot = volcano_semi_gbm, 
       filename = here::here("figures/volcano_proteolysis_gbm_2.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 63)
```

```{r}
# supplementary results table for proteolysis results
summary_semi_limma <- compar_tab_peptwo6_annn2 %>% 
  filter(Feature %in% c("Up-regulated Semi-tryptic" , 
                        "Down-regulated Semi-tryptic",
                        "No change"))

write_tsv(x = summary_semi_limma,
          file = here::here("results/semi_tryptic/summary_proteolytic_processing_limma_results.tsv"))
```

# Checking specific hits

## GPNMB - distribution of abundances for semi-specific peptides

We identified a group of differentially abundant semi-specific peptides from GPNMB.

```{r}
pept_quant_annotated2 <- pept_quant_annotated %>%
  separate(col = index,
           into = c("Protein", 
                    "peptide"))

quant_anpeptwgene <- left_join(pept_quant_annotated2, 
                               prot2gene)

tab_2_peptpaired_boxplot <-  quant_anpeptwgene %>%
  filter(Protein %in% "Q14956")

gpmnb_pept_2annnot <- tab_2_peptpaired_boxplot %>%
  dplyr::select(Peptide = peptide,
                Genes = Protein) %>% 
  distinct()

# prep data
slim_datap <- pivot_longer(pept_matrix, 
                           cols = 2:length(names(pept_matrix)), 
                           names_to = c("patient","condition"),
                           values_to = "Intensity",
                           names_sep = "\\_") %>% 
  mutate(condition = case_when(str_detect(patient, "rec") ~ "rec",
                               str_detect(patient, "prim") ~ "prim"),
         patient = paste0("x",parse_number(patient))) %>%
  filter(str_detect(patient, "x6", negate = TRUE)) %>% 
  separate(col = index,
           into = c("Protein", "peptide"), 
           sep = "_") %>%
  left_join(.,sample_annotation2) %>%
    filter(Protein %in% "Q14956")

slim_datap2 <- pivot_wider(slim_datap, 
                           values_from = Intensity, 
                            names_from = condition)

slim_datap3 <- slim_datap2 %>% 
  mutate(Abs_diff = rec-prim) %>%
  mutate(rank = row_number(Abs_diff)) %>% 
  dplyr::rename(Initial = prim, Recurrent = rec) 
```

```{r}
source(here("scr/annotate_peptides.R"))
```

```{r}
annotated_gpnmb_peptides <- annotate_peptides(gpmnb_pept_2annnot,
                                              fasta = fasta) %>%
  dplyr::select(-previous_all_resid) %>% 
  filter(specificity == "semi_specific")

slim_datap4 <- slim_datap3 %>%
  dplyr::filter(peptide %in% annotated_gpnmb_peptides$Peptide)
```

```{r}
gpnmb_semis <- ggpaired(slim_datap4,
                        cond1 = "Initial",
                        cond2 = "Recurrent", 
                        fill = "condition",
                        ylab = "Normalized Abundance",
                        label = NULL,
                        repel = TRUE,
                        facet.by = "peptide",
                        point.size = 0.2, 
                        line.size = 0.1, 
                        line.color = "mixture") +
      theme(axis.text.x = element_text(hjust = 0.5, 
                                       vjust = 0, 
                                       size = 7, 
                                       angle = 360),
            axis.text.y = element_text(hjust = 0.95, 
                                       vjust = 0.2, 
                                       size = 7),
            panel.background = element_blank(),
            panel.grid.major = element_blank(),
            panel.border = element_rect(colour = "black", 
                                        fill = NA, 
                                        size = 0.5),
            axis.title = element_text(size = 7),
            legend.text = element_text(size = 7),
            legend.title = element_text(size = 7),
            legend.key.height= unit(6, 'mm'),
            legend.key.width= unit(6, 'mm'),
            legend.position="bottom",
            strip.text = element_text(hjust = 0.5, 
                                      vjust = 0, 
                                      size = 7, 
                                      angle = 360)) 
```

```{r fig.width=8, fig.height=6.5}
print(gpnmb_semis)
```

```{r}
ggsave(plot = gpnmb_semis, 
       filename = here::here("figures/gpnmb_proteolysis_gbm_boxplots.tiff"), 
       device = "tiff",
       units = "mm",
       width = 190,
       height = 180)

ggsave(plot = gpnmb_semis, 
       filename = here::here("figures/gpnmb_proteolysis_gbm_boxplots.eps"), 
       device = "eps",
       units = "mm",
       width = 190,
       height = 180)
```

## Coverage of GBPNMB

```{r}
source(here("scr/protein_peptide_coverage_plot.R"))
```

```{r}
peptides_GPNMB <- cleavage_annoated_peptides %>% 
  filter(protein_id %in% "Q14956") %>%
  mutate(peptide_length = str_length(string = Peptide))
```

```{r}
feat_gpnmb <- drawProteins::get_features("Q14956")

feat_gpnmbdf <- drawProteins::feature_to_dataframe(feat_gpnmb)
#Get peptides as coverage features for plotting

cov_feat_gpnmbdf <- get_coverage(annotated_peptides = peptides_GPNMB, 
                                 id = "Q14956")

# get fold-changes of peptides to plot for coverage
fcs_cov_gpnmbdf <- compar_tab_peptwo6_annn2 %>%
  filter(peptide %in% cov_feat_gpnmbdf$description) %>%
  dplyr::select(description = peptide, logFC)

cov_feat_gpnmbdf2 <- left_join(cov_feat_gpnmbdf, 
                               fcs_cov_gpnmbdf)

#Merge coverage feature info with protein sequence features info:

fullfeat_gpnmb <- add_peptides(feat_gpnmbdf,
                               peptide_coverage_data = cov_feat_gpnmbdf)
# drawCanvas
canvas <- draw_canvas(fullfeat_gpnmb)

# draw the protein chain
wchain <- draw_chains(canvas, 
                      fullfeat_gpnmb)

# draw protein domains
wdomainsgpnmb <- draw_domains(wchain, 
                              data = fullfeat_gpnmb)
cov_feat_gpnmbdf$order <- 1.4

pepcov_gpnmbdf <- wchain + 
  ggplot2::geom_rect(data = cov_feat_gpnmbdf2,
                         mapping=ggplot2::aes(xmin=begin,
                                              xmax=end,
                                              ymin=order-0.05,
                                              ymax=order+0.05,
                                              fill = logFC)) +
  scale_fill_gradient(low = "#2C9C8E",
                      high = "#FC0000") +
    annotate("text", x = -50, y = 1.4, label = "Peptides") +
    theme_bw(base_size = 12) + # white background
    theme(panel.grid.minor=element_blank(),
          panel.grid.major=element_blank()) +
    theme(axis.ticks = element_blank(),
          axis.text.y = element_blank()) +
    theme(panel.border = element_blank(),
          legend.position = "bottom")
```

```{r warning=FALSE}
pepcov_gpnmbdf
```

```{r}
ggsave(plot = pepcov_gpnmbdf, 
       filename = here::here("figures/gpnmbdf_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_gpnmbdf, 
       filename = here::here("figures/gpnmbdf_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

## GFAP - distribution of abundances for semi-specific peptides

We identified a group of differentially abundant semi-specific peptides from GPNMB.

```{r}
tab_2_peptpaired_gfap_boxplot <-  quant_anpeptwgene %>%
  filter(Gene %in% "GFAP")

gfap_pept_2annnot <- tab_2_peptpaired_gfap_boxplot %>%
  dplyr::select(Peptide = peptide,
                Genes = Protein) %>% 
  distinct()

# prep data
slim_datap_gfap <- pivot_longer(pept_matrix, 
                           cols = 2:length(names(pept_matrix)), 
                           names_to = c("patient","condition"),
                           values_to = "Intensity",
                           names_sep = "\\_") %>% 
  mutate(condition = case_when(str_detect(patient, "rec") ~ "rec",
                               str_detect(patient, "prim") ~ "prim"),
         patient = paste0("x",parse_number(patient))) %>%
  filter(str_detect(patient, "x6", negate = TRUE)) %>% 
  separate(col = index,
           into = c("Protein", "peptide"), 
           sep = "_") %>%
  left_join(.,sample_annotation2) %>%
      filter(Protein %in% "P14136")

slim_datap_gfap <- pivot_wider(slim_datap_gfap, 
                           values_from = Intensity, 
                            names_from = condition)

slim_datap_gfap3 <- slim_datap_gfap %>% 
  mutate(Abs_diff = rec-prim) %>%
  mutate(rank = row_number(Abs_diff)) %>% 
  dplyr::rename(Initial = prim, Recurrent = rec) 
```

```{r}
annotated_gfap_peptides <- annotate_peptides(gfap_pept_2annnot,
                                              fasta = fasta) %>%
  dplyr::select(-previous_all_resid) %>% 
  filter(specificity == "semi_specific")

slim_datap_gfap4 <- slim_datap_gfap3 %>%
  dplyr::filter(peptide %in% unique(annotated_gfap_peptides$Peptide))
```

```{r}
gfap_semis <- ggpaired(slim_datap_gfap4,
                        cond1 = "Initial",
                        cond2 = "Recurrent", 
                        fill = "condition",
                        ylab = "Normalized Abundance",
                        label = NULL,
                        repel = TRUE,
                        facet.by = "peptide",
                        point.size = 0.2, 
                        line.size = 0.1, 
                        line.color = "mixture") +
      theme(axis.text.x = element_text(hjust = 0.5, 
                                       vjust = 0, 
                                       size = 7, 
                                       angle = 360),
            axis.text.y = element_text(hjust = 0.95, 
                                       vjust = 0.2, 
                                       size = 7),
            panel.background = element_blank(),
            panel.grid.major = element_blank(),
            panel.border = element_rect(colour = "black", 
                                        fill = NA, 
                                        size = 0.5),
            axis.title = element_text(size = 7),
            legend.text = element_text(size = 7),
            legend.title = element_text(size = 7),
            legend.key.height= unit(6, 'mm'),
            legend.key.width= unit(6, 'mm'),
            legend.position="bottom",
            strip.text = element_text(hjust = 0.5, 
                                      vjust = 0, 
                                      size = 7, 
                                      angle = 360)) 
```

```{r fig.width=8, fig.height=6.5}
print(gfap_semis)
```

## Coverage of GFAP

```{r}
source(here("scr/protein_peptide_coverage_plot.R"))
```

```{r}
peptides_GFAP <- cleavage_annoated_peptides %>% 
  filter(protein_id %in% "P14136",
         specificity == "semi_specific") %>%
  mutate(peptide_length = str_length(string = Peptide))
```

```{r}
feat_GFAP <- drawProteins::get_features("P14136")

feat_GFAPf <- drawProteins::feature_to_dataframe(feat_GFAP)
#Get peptides as coverage features for plotting

cov_feat_GFAPf <- get_coverage(annotated_peptides = peptides_GFAP, 
                                 id = "P14136")

# get fold-changes of peptides to plot for coverage
fcs_cov_GFAPf <- compar_tab_peptwo6_annn2 %>%
  filter(peptide %in% cov_feat_GFAPf$description) %>%
  dplyr::select(description = peptide, logFC)

cov_feat_GFAP2 <- left_join(cov_feat_GFAPf, 
                            fcs_cov_GFAPf)

#Merge coverage feature info with protein sequence features info:

fullfeat_GFAP <- add_peptides(feat_GFAPf,
                              peptide_coverage_data = cov_feat_GFAPf)
# drawCanvas
canvas <- draw_canvas(fullfeat_GFAP)

# draw the protein chain
wchain <- draw_chains(canvas, 
                      fullfeat_GFAP)

# draw protein domains
wdomainsgfap <- draw_domains(wchain, 
                             data = fullfeat_GFAP)

cov_feat_GFAP2$order <- 1.4

pepcov_GFAP <- wchain + 
  ggplot2::geom_rect(data = cov_feat_GFAP2,
                         mapping=ggplot2::aes(xmin=begin,
                                              xmax=end,
                                              ymin=order-0.05,
                                              ymax=order+0.05,
                                              fill = logFC)) +
  scale_fill_gradient(low = "#2C9C8E",
                      high = "#FC0000") +
    annotate("text", x = -50, y = 1.4, label = "Peptides") +
    theme_bw(base_size = 12) + # white background
    theme(panel.grid.minor=element_blank(),
          panel.grid.major=element_blank()) +
    theme(axis.ticks = element_blank(),
          axis.text.y = element_blank()) +
    theme(panel.border = element_blank(),
          legend.position = "bottom")
```

```{r warning=FALSE}
pepcov_GFAP
```

```{r}
ggsave(plot = pepcov_GFAP, 
       filename = here::here("figures/gfap_coverage_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 150,
       height = 100)

ggsave(plot = pepcov_GFAP, 
       filename = here::here("figures/gfap_coverage_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 150,
       height = 100)
```

# Proportional abundance of proteolytic peptides

__Wrangle data__

```{r}
pept_semi_annot <- peptides_annotated %>%
  dplyr::select(Peptide, 
                specificity, 
                is_terminal)

pept_quant_to_summary <- pept_quant_annotated %>%
  separate(col = index, 
           into = c("Protein", "Peptide"), 
           sep = "\\_", 
           remove = FALSE) %>%
  left_join(.,pept_semi_annot)

pep_quant_presummary <- pept_quant_to_summary %>%
  dplyr::select(Protein, 
                Peptide, 
                recurrence, 
                paired_id, 
                Abundance, 
                specificity, 
                is_terminal) 
```

__Prepare data-frame with abudance percentage of semi-specific peptides per sample__

```{r}
pept_summary_int_all <- pep_quant_presummary %>%
  group_by(paired_id, 
           recurrence) %>%
  summarise(Sum_All = sum(Abundance, 
                          na.rm = TRUE))

pept_summary_int_semi <- pep_quant_presummary %>%
  filter(specificity == "semi_specific",
         is_terminal == "not_terminal") %>%
  group_by(paired_id, 
           recurrence) %>%
  summarise(Sum_Semi = sum(Abundance, 
                           na.rm = TRUE))

pept_sum_summary <- left_join(pept_summary_int_all, 
                              pept_summary_int_semi) %>%
  mutate(Percentage = Sum_Semi/Sum_All * 100,
         Stage = if_else(recurrence == "prim",
                         true = "Initial",
                         false = "Recurrence"))
```

## Plot of % of abundance of semi-specific peptides

```{r}
plotSumEM <- ggplot(pept_sum_summary, 
                    aes(x = Stage, 
                        y = Percentage, 
                        fill = Stage, 
                        cex.axis = 1.5)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", 
               stackdir = "center", 
               dotsize = 0.5) +
  # Box plot with jittered points
  # 0.2 : degree of jitter in x direction
  ylab("Proportional intensity of semi-tryptic peptides [%]") +
  geom_signif(comparisons = list(c("Initial",
                                   "Recurrence")),
              map_signif_level = TRUE) + 
  stat_compare_means(method = "t.test") +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 6, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height = unit(3, 'mm'),
        legend.key.width = unit(3, 'mm'),
        legend.position ="bottom")

```

```{r warning=FALSE}
print(plotSumEM)
```

```{r}
ggsave(plot = plotSumEM, 
       filename = here::here("figures/box_perc_semispec_pepts.tiff"), 
       device = "tiff",
       units = "mm",
       width = 40,
       height = 50)

ggsave(plot = plotSumEM, 
       filename = here::here("figures/box_perc_semispec_pepts.eps"), 
       device = "eps",
       units = "mm",
       width = 40,
       height = 50)
```

## Comparative analysis of semi-specific peptides vs protein abundance

### Get log2FC Summary of abundances of semi-specific peptides {.unnumbered .unlisted}

```{r}
#taking output from limma semi-specific
log2_semi_tryptic <- dplyr::select(compar_tab_peptwo6_annn2,
                                   logFC, 
                                   Protein, 
                                   peptide, 
                                   specificity, 
                                   is_terminal,
                                   Gene) %>%
  filter(specificity == "semi_specific",
         is_terminal == "not_terminal") %>%
  dplyr::rename(logFC_semi_peptide = logFC) %>%
  dplyr::select(-c(specificity, 
                   is_terminal))
```

### Get log2 Summary of protein abundances based only on fully-tryptic peptides {.unnumbered .unlisted}

```{r}
med_prot_quant_specific <- pep_quant_presummary %>%
  group_by(Protein, 
           paired_id) %>%
  summarise(`median(Abundance)` = median(Abundance, 
                                         na.rm = TRUE)) %>%
  ungroup()
```

```{r}
proteins2exclude <- sel_proteins_missing(quant_annotated,
                                         threshold = 1)
```

```{r}
# prep wide matrix for median proteins abundances (arousing only from fully-tryptic peptides)
wide_premat_prot_only_tryp <- med_prot_quant_specific %>%
  pivot_wider(id_cols = c("Protein"),
              values_from = `median(Abundance)`, 
              names_from = paired_id) %>%
  # keep only proteins present in 2 / 3 of the samples
  filter(!Protein %in% proteins2exclude)

mat_prot_only_tryp <- wide_premat_prot_only_tryp %>%
  column_to_rownames("Protein") %>%
  as.matrix()
```

### Run limma on this new data to generate log2 FCs of proteins from only fully-tryptic peptides {.unnumbered .unlisted}

__missForest-based imputation__

```{r}
t_mat_prot_only_tryp <- t(mat_prot_only_tryp)

if(!file.exists(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))){

 imp_prot_only_tryp <- missForest::missForest(t_mat_prot_only_tryp)
 t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp

 mat_prot_only_tryp <- t(t_mat_prot_only_tryp)

 saveRDS(imp_prot_only_tryp,
         file = here("rds/missforest_imp_mat_prot_only_tryp.Rds"))

} else {

 imp_prot_only_tryp <- readRDS(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))
 t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp

 mat_imp_prot_only_tryp <- t(t_mat_imp_prot_only_tryp)

}
```

```{r}
split <- str_split_fixed(colnames(mat_imp_prot_only_tryp), pattern = "\\_", n = 2)

patient_tryp <- split[,1]
recurrence <- split[,2]

design_tryp <- model.matrix(~patient_tryp+recurrence)

rownames(design_tryp) <- colnames(mat_imp_prot_only_tryp)
```

```{r}
rownames(design_tryp) 
```

```{r}
colnames(mat_prot_only_tryp)
```

```{r warning=FALSE}
limma_tab_wo6tryp <- fit_limmawo6_tryp(mat_imp_prot_only_tryp, 
                         design_tryp, 
                         method = 'robust', 
                         Limma = "Robust - Proteins w only tryptic peptides",
                         prot2gene = prot2gene)

```

```{r}
log2_fully_tryptic <- dplyr::select(limma_tab_wo6tryp,
                                   logFC, 
                                   Protein, 
                                   Gene) %>%
  dplyr::rename(logFC_fully_tryp_protein = logFC) 
```

### Merge log2FCs of semi-tryptic peptides vs proteins with fully tryptic peptides {.unnumbered .unlisted}

```{r warning=FALSE}
expr_matrix2 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotation2$paired_id)

expr_matrix_filt <- expr_matrix2 %>%
  filter(!Protein %in% proteins2exclude)

mat_filt <- expr_matrix_filt %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt <- t(mat_filt)

if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){

 t_imp_mat_filt <- missForest::missForest(t_mat_filt)
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

 saveRDS(t_imp_mat_filt,
         file = here("rds/missforest_imp_mat_filt.Rds"))

} else {

 t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

}

mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

source(here("scr/fit_limmawo6.R"))

# fit limma
limma_tab_wo6 <- fit_limmawo6(mat_filt_impwo6, 
                         design_limmawo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect",
                         prot2gene = prot2gene)

#get proteins increased in recurrence
increased_in_rec <- limma_tab_wo6 %>%
  filter(logFC > 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)

#get proteins decreased in recurrence
decreased_in_rec <- limma_tab_wo6 %>%
  filter(logFC < 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)

tab_sig_prots_tryptic <- bind_rows(tibble(protein = increased_in_rec,
                                   characteristic = "up-regulated"),
                           tibble(protein = decreased_in_rec,
                                  characteristic = "down-regulated"))

```

```{r}
log2semipept2_log2protein_spec <- left_join(log2_fully_tryptic, 
                                            log2_semi_tryptic) %>%
  na.omit() %>%
  mutate(DA_Protein = if_else(Protein %in% tab_sig_prots_tryptic$protein,
                              true = TRUE,
                              false = FALSE),
         DA_peptide = if_else(peptide %in% recwo6_pept22$peptide,
                              true = TRUE,
                              false = FALSE))
```

### log2FC Proteins vs Semi-tryp peptides (scatter plot)

```{r}
check_increased_pept <- cleavage_annoated_peptides %>%
  filter(Peptide %in% increased_recwo6_pept_annn)
```

```{r}
inter_scater <- log2semipept2_log2protein_spec %>%
  filter(Protein %in% check_increased_pept$protein_id)

inter_scater$Gene[duplicated(inter_scater$Gene)] <- NA

scatter_proteolysis_plot <- ggplot(log2semipept2_log2protein_spec, 
                                   aes(x = logFC_fully_tryp_protein, 
                                       y = logFC_semi_peptide)) + 
  geom_smooth(method = lm, 
              se = FALSE, 
              linetype = "dashed", 
              size = 0.5, 
              color = "black") + 
  geom_point(aes(color = DA_Protein, 
                 shape = DA_peptide), 
             size = 0.6) +
  scale_color_manual(values = c("#2a9d8f", 
                                "red")) +
  xlab("log2(FC) - Protein abundances") + 
  ylab("log2(FC) - Semi-specific peptides") + 
  ggrepel::geom_text_repel(data = inter_scater,
                           aes(label = Gene), 
                           size = 1,
                           box.padding = 0.5,
                           max.overlaps = 25) +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 8, 
                                   angle = 360),
        axis.text.y = element_text(hjust = 0.95, 
                                   vjust = 0.2, 
                                   size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", 
                                    fill = NA, 
                                    size = 0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

```{r}
cor.test(log2semipept2_log2protein_spec$logFC_fully_tryp_protein,
    log2semipept2_log2protein_spec$logFC_semi_peptide, 
    method = "pearson")
```
```{r}
ggsave(plot = scatter_proteolysis_plot, 
       filename = here::here("figures/scatter_proteolysis_plot_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 63)

ggsave(plot = scatter_proteolysis_plot, 
       filename = here::here("figures/scatter_proteolysis_plot_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 63)
```

# Analysis of differential amino acid usage

```{r}
source(here("scr/get_cleavage_area.R"))
```

```{r}
cleave_areas <- get_cleave_area(cleavage_annoated_peptides)
```

```{r}
library(dagLogo)
```

Used the identified protein sequences as background proteome.

```{r}
proteome_ided <- prepareProteome(fasta = here("data/semi_specific_search_fragpipe17/protein_combined.fas"), 
                                 species = "Homo sapiens")
```

```{r}
increased_Rec <- filter(cleave_areas$cleave_area20,
                            Peptide %in% increased_recwo6_pept_annn)

decreased_Rec <- filter(cleave_areas$cleave_area20,
                            Peptide %in% decreased_recwo6_pept_annn)

increased_Rec_4ice <- filter(cleave_areas$cleave_area20,
                            Peptide %in% increased_recwo6_pept_annn) %>%
                    pull(cleave_area20)

decreased_Rec_4ice <- filter(cleave_areas$cleave_area20,
                            Peptide %in% decreased_recwo6_pept_annn) %>%
                    pull(cleave_area20)
```

```{r}
write_tsv(x = increased_Rec,
          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))

write_tsv(x = decreased_Rec,
          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))

```

**Format peptide sequences**

```{r}
if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))){

  form_peptidesincreased_4ice <- formatSequence(increased_Rec_4ice, 
                                  proteome = proteome_ided)

  write_rds(form_peptidesincreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))

} else {

  form_peptidesincreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))

}
```

```{r}
if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))){

  form_peptidesdecreased_4ice <- formatSequence(decreased_Rec_4ice, 
                                  proteome = proteome_ided)

  write_rds(form_peptidesdecreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))

} else {

  form_peptidesdecreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))

}
```

## DAU increased

```{r}
bg_mod_ztest_increased <- buildBackgroundModel(form_peptidesincreased_4ice,
                                           proteome = proteome_ided,
                           background = "wholeProteome",
                           testType = "ztest")
```

```{r}
dau_nogroup_increased <- testDAU(form_peptidesincreased_4ice, 
                           dagBackground = bg_mod_ztest_increased)
```

```{r}
dagHeatmap(dau_nogroup_increased) 
```

## DAU decreased

```{r}
bg_mod_ztest_decreased <- buildBackgroundModel(form_peptidesdecreased_4ice,
                                           proteome = proteome_ided,
                           background = "wholeProteome",
                           testType = "ztest")
```

```{r}
dau_nogroup_decreased <- testDAU(form_peptidesdecreased_4ice, 
                           dagBackground = bg_mod_ztest_decreased)
```

```{r}
dagHeatmap(dau_nogroup_decreased) 
```

# Heatmaps of positional abundance of amino acids of proteolytic products

```{r}
cleavage_area_list <- get_cleave_area(cleavage_annoated_peptides)
```

```{r}
increased <- compar_tab_peptwo6_annn %>%
                filter(logFC > 0,
                       adj.P.Val < 0.05,
                       specificity == "semi_specific",
                       is_terminal == "not_terminal") %>% 
                    pull(peptide)

decreased <- compar_tab_peptwo6_annn %>%
                filter(logFC < 0,
                       adj.P.Val < 0.05,
                       specificity == "semi_specific",
                       is_terminal == "not_terminal") %>% 
                    pull(peptide)

list_result <- list(output_limma = compar_tab_peptwo6_annn,
                    increased = increased,
                    decreased = decreased)
```

```{r}
prelogo <- prep_peptides(cleavage_area_list$cleave_area20)
```

## Increased Semi-specific peptides (TMT-labelled) (simple counts-no stats)

```{r}
order_of_aas <- read_csv(file = here::here("data/ordered_amino_acids.csv"))
```

```{r}
increased_pept_mat <- prelogo$form_peptidesincreased_4ice@peptides
```

```{r}
positions <- c(paste0("P",10:1),paste0("P",1:10,"'"))
```

```{r}
colnames(increased_pept_mat) <- positions
```

```{r}
increased_pept_mat_df <- increased_pept_mat %>%
                    as_tibble() 
```

```{r}
long_incr_ <- pivot_longer(increased_pept_mat_df,
                           cols = everything(),
                           names_to = "position",
                           values_to = "AA")

```

```{r}
count_incr <- long_incr_ %>%
                    dplyr::count(position, AA)
```

```{r}
wide_count_incr <- count_incr %>%
                    pivot_wider(id_cols = AA,
                                names_from = position,
                                values_from = n) %>%
                    dplyr::select(AA, all_of(positions)) %>%
                    dplyr::arrange(factor(AA, 
                                          levels = order_of_aas$...1))
```

```{r}
wide_count_incr_mat <- wide_count_incr %>%
                    dplyr::select(AA, paste0("P",5:1),paste0("P",1:5,"'")) %>%
                    column_to_rownames("AA") %>%
                    as.matrix()
```

```{r}
pheatmap::pheatmap(wide_count_incr_mat, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "AA Counts - Increased Semi-specific pept in Recurrence")
```

```{r}
pheatmap::pheatmap(wide_count_incr_mat, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE)
```

# Heatmaps of positional abundance of amino acids of proteolytic products from GPNBM

```{r}
cleavage_annoated_peptides_gpnbm <- cleavage_annoated_peptides %>%
  filter(Peptide %in% gpmnb_pept_2annnot$Peptide)

cleavage_area_list_gpnbm <- get_cleave_area(cleavage_annoated_peptides_gpnbm)
```

```{r}
prelogo_gpnbm <- prep_peptides_min(cleavage_area_list_gpnbm$cleave_area20$cleave_area20, 
                                   proteome_ided = proteome_ided)
```

## Heatmap

```{r}
gpnbm_pept_mat <- prelogo_gpnbm$form_peptides_4ice@peptides
```

```{r}
positions <- c(paste0("P",10:1),
               paste0("P",1:10,"'"))
```

```{r}
colnames(gpnbm_pept_mat) <- positions
```

```{r}
gpnbm_pept_mat_df <- gpnbm_pept_mat %>%
                    as_tibble() 
```

```{r}
long_gpnbm_pept <- pivot_longer(gpnbm_pept_mat_df,
                           cols = everything(),
                           names_to = "position",
                           values_to = "AA")
```

```{r}
count_gpnbm_pept <- long_gpnbm_pept %>%
                    dplyr::count(position, AA)
```

```{r}
amino_acids_order <- data.frame(AA = order_of_aas$...1)

vect_aa_order <- amino_acids_order$AA
```

```{r}
wide_count_gpnbm <- count_gpnbm_pept %>%
                    pivot_wider(id_cols = AA,
                                names_from = position,
                                values_from = n) %>%
                    dplyr::select(AA, all_of(positions)) %>%
                    full_join(., amino_acids_order) %>%
                    dplyr::arrange(factor(AA, 
                                          levels = vect_aa_order))
```

```{r}
wide_count_mat <- wide_count_gpnbm %>%
                    dplyr::select(AA, paste0("P",5:1),paste0("P",1:5,"'")) %>%
                    column_to_rownames("AA") %>%
                    as.matrix()
```

```{r}
pheatmap::pheatmap(wide_count_mat, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "AA Counts - Semi-specific peptides GPNBM")
```

```{r}
pheatmap::pheatmap(wide_count_incr_mat, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE)
```

# Heatmaps of positional abundance of amino acids of proteolytic products from GFAP

```{r}
gfap_semi_pepts <- diff_abund_semi %>% 
  filter(Gene == "GFAP") %>% 
  pull(peptide)

gfap_semi <- diff_abund_semi %>% 
  filter(Gene == "GFAP") 
```

```{r}
cleavage_annoated_peptides_gfap <- cleavage_annoated_peptides %>%
  filter(Peptide %in% gfap_semi_pepts)

cleavage_area_list_gfap <- get_cleave_area(cleavage_annoated_peptides_gfap)
```

```{r}
prelogo_gfap <- prep_peptides_min(cleavage_area_list_gfap$cleave_area20$cleave_area20, 
                                   proteome_ided = proteome_ided)
```

```{r}
gfap_semis_fcs_annt <- left_join(gfap_semi %>% dplyr::rename(Peptide = peptide),
                                 cleavage_area_list_gfap$cleave_area20)
```

## Heatmap

```{r}
gfap_pept_mat <- prelogo_gfap$form_peptides_4ice@peptides
```

```{r}
positions <- c(paste0("P",10:1),paste0("P",1:10,"'"))
```

```{r}
colnames(gfap_pept_mat) <- positions
```

```{r}
gfap_pept_mat_df <- gfap_pept_mat %>%
                    as_tibble() 
```

```{r}
long_gfap_pept <- pivot_longer(gfap_pept_mat_df,
                           cols = everything(),
                           names_to = "position",
                           values_to = "AA")
```

```{r}
count_gfap_pept <- long_gfap_pept %>%
                    dplyr::count(position, AA)
```

```{r}
wide_count_gfap <- count_gfap_pept %>%
                    pivot_wider(id_cols = AA,
                                names_from = position,
                                values_from = n) %>%
                    dplyr::select(AA, all_of(positions)) %>%
                    full_join(., amino_acids_order) %>%
                    dplyr::arrange(factor(AA, 
                                          levels = order_of_aas$...1))
```

```{r}
wide_count_mat <- wide_count_gfap %>%
                    dplyr::select(AA, paste0("P",5:1),paste0("P",1:5,"'")) %>%
                    column_to_rownames("AA") %>%
                    as.matrix()
```

```{r}
pheatmap::pheatmap(wide_count_mat, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   main = "AA Counts - Semi-specific peptides GFAP")
```

# Genearate Supplementary Table of Annotated Semi-specific peptides

This table would contain the list of differentially abundant semi-specific peptides with their respective annotation of extended cleavage sites. A different table will be added for the GPMBN peptides.

It will also contain the peptides identified as associated to GPMBN.

```{r}
cleavage_annoated_peptides_diffabund <- cleavage_annoated_peptides %>%
  filter(Peptide %in% diff_abund_semi$peptide)

cleavage_area_list_diffabund <- get_cleave_area(cleavage_annoated_peptides_diffabund)
```

```{r}
diffabund_semis_fcs_annt <- left_join(diff_abund_semi %>% dplyr::rename(Peptide = peptide),
                                 cleavage_area_list_diffabund$cleave_area20) %>%
  left_join(., cleavage_area_list_gpnbm$cleave_area20) %>%
  dplyr::select(-c(prev_10_pad,
                   following_10_pad,
                   len_20,
                   len_clearea,
                   len_follow,
                   len_prev,
                   len_followpad,
                   len_prevpad))
```

```{r}
write_tsv(x = diffabund_semis_fcs_annt,
          file = here("suppl_tables/Supplementary_Table_3_differential_abundant_semispepecific_peptides.tsv"))
```

```{r}
gpnbm_supp_table <- cleavage_area_list_gpnbm$cleave_area20 %>%
  dplyr::select(-c(prev_10_pad,
                   following_10_pad,
                   len_20,
                   len_clearea,
                   len_follow,
                   len_prev,
                   len_followpad,
                   len_prevpad))
```

```{r}
write_tsv(x = gpnbm_supp_table,
          file = here("suppl_tables/Supplementary_Table_4_gpnbm_semispepecific_peptides.tsv"))
```

### 2023-02-16 scr/helper_functions.R

@@ -153,3 +153,172 @@ our_volcano <- function(dataarg,
   return(volcano)
   
 }
+
+## Function to select proteins to exclude because are missing in 2 out of 3 batches ----
+
+sel_proteins_missing <- function(long_matrix,
+                                 threshold) {
+  
+  na_count <- group_by(long_matrix,
+                       Protein, mixture) %>%
+    summarise(na_count = sum(is.na(Abundance)),
+              total = n()) %>% 
+    ungroup() %>% 
+    mutate(NA_fraction = na_count/total)
+  
+  na_count_perbatch <- na_count %>%
+    group_by(Protein) %>%
+    summarise(na_per_batch = sum(NA_fraction)) %>%
+    ungroup()
+  
+  proteins2exclude <- na_count_perbatch %>%
+    filter(na_per_batch > threshold) %>%
+    pull(Protein)
+  
+  return(proteins2exclude)
+}  
+
+## Label N-terminal peptide modifications
+
+annotate_nterm <- function(peptidestsv, # peptide.tsv table
+                           tmtmass = 304.2072, # either 304.2072 for 16plex or 229.1629 for 10/11plex
+                           protease_specificity = "R|K") # or R for argc 
+{
+  
+  require(dplyr)
+  require(stringr)
+  
+  if (tmtmass == 304.2072){
+    
+    nterm_tmt <- "N-term\\(304.2072\\)"
+    ktmt <- "K\\(304.2072\\)"
+    
+  } else if (tmtmass == 229.1629){
+    
+    nterm_tmt <- "N-term\\(229.1629\\)"
+    ktmt <- "K\\(229.1629\\)"                    
+    
+  } else {
+    
+    error("Please check your tmtmass argument input.")
+    
+  }
+  
+  nterm_annotated <- peptidestsv %>% 
+    mutate(nterm = case_when(str_detect(assigned_modifications, 
+                                        nterm_tmt) ~ "TMT-labelled",
+                             str_detect(assigned_modifications, 
+                                        "N-term\\(42.0106\\)") ~ "acetylated",
+                             TRUE ~ "free")) %>%
+    mutate(tmt_tag = case_when(str_detect(assigned_modifications, 
+                                          nterm_tmt) ~ "nterm",
+                               str_detect(assigned_modifications, 
+                                          ktmt) ~ "lysine",
+                               str_detect(assigned_modifications, 
+                                          ktmt,
+                                          negate = TRUE) & str_detect(assigned_modifications, 
[…]

### 2023-02-16 scr/helpers_gbm_analysis.R

@@ -1,0 +1,89 @@
+## function for feature-specific FDR correction
+## used specifically in this case to apply FDR correction only on semi-specific peptides
+
+feature_fdr_correction_mst <- function(toptable,
+                                       features_table,
+                                       interesting_features_table){
+
+                    # merge limma output with feature annotation  ----
+
+                    tab_limma_feature_annot <- left_join(toptable,
+                                                         features_table,
+                                                         by = c("peptide"))
+
+                    # filter limma table based on interesting feature ----
+
+                    tab_limma_subsetin <- filter(tab_limma_feature_annot,
+                                                 peptide %in% interesting_features_table$peptide) %>%
+                                        # apply FDR correction only on the subset
+                                        mutate(adj.pvalue = p.adjust(p = pvalue,
+                                                                    method = "BH")) %>%
+                                        mutate(fdr_correction = 'feature-specific')
+
+                    # create a table of non-interesting features containing the globally adjusted p-values
+                    tablimma_subsetout <- filter(tab_limma_feature_annot,
+                                                 !peptide %in% interesting_features_table$peptide) %>%
+                                        mutate(fdr_correction = 'global')
+
+                    output_limma3 <- bind_rows(tab_limma_subsetin,
+                                               tablimma_subsetout)
+
+                    return(output_limma3)
+
+}
+
+get_da_peptides <- function(tabular_dea_result,
+                            type) {
+
+                    if (type == "limma"){
+
+                                        increased <- tabular_dea_result %>%
+                                                            filter(logFC > 0,
+                                                                   adj.P.Val < 0.05) %>%
+                                                            pull(peptide)
+
+                                        decreased <- tabular_dea_result %>%
+                                                            filter(logFC < 0,
+                                                                   adj.P.Val < 0.05) %>%
+                                                            pull(peptide)
+
+                                        peptides_da <- list(increased = increased,
+                                                            decreased = decreased)
+
+                    } else if (type == "msstats"){
+
+                                        increased <- tabular_dea_result %>%
+                                                            filter(log2FC > 0,
+                                                                   adj.pvalue < 0.05) %>%
+                                                            pull(peptide)
+
+                                        decreased <- tabular_dea_result %>%
+                                                            filter(log2FC < 0,
+                                                                   adj.pvalue < 0.05) %>%
+                                                            pull(peptide)
+
+                                        peptides_da <- list(increased = increased,
+                                                            decreased = decreased)
+
+                    } else if (type == "msstats_limma") {
+
[…]

### 2023-02-16 scr/prepare_combined_fasta.R

@@ -1,0 +1,35 @@
+# merge fasta of identified sequences from semi-specific search into a single fasta.
+
+fasta1 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_1/protein.fas"),
+                             seqtype = "AA", 
+                             as.string = TRUE)
+
+fasta2 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_2/protein.fas"),
+                             seqtype = "AA", 
+                             as.string = TRUE)
+
+fasta3 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_3/protein.fas"),
+                             seqtype = "AA", 
+                             as.string = TRUE)
+
+fasta22 <- append(fasta1, 
+                  fasta2)
+
+fasta <- append(fasta22, 
+                fasta3)
+
+fasta <- discard(fasta,
+                 duplicated(names(fasta)))
+
+attr(fasta[[1]], "Annot")
+
+extr_annt <- function(x){
+  ext <- attr(x, "Annot")
+}
+
+names <- purrr::map_chr(fasta, 
+                        .f = extr_annt)
+
+seqinr::write.fasta(fasta,
+                    file.out = here("data/semi_specific_search_fragpipe17/protein_combined.fas"),
+                    names = names)

### 2023-02-16 scr/protein_peptide_coverage_plot.R

@@ -15,6 +15,7 @@ get_coverage <- function(annotated_peptides, id){
     dplyr::mutate(order = 1)
   
 }
+
 add_peptides <- function(feature_df,
                          peptide_coverage_data){
   
@@ -33,6 +34,7 @@ add_peptides <- function(feature_df,
   return(wpept_features)
   
 }
+
 draw_peptides <- function(p, data){
   data$order <- 1.4
   p + ggplot2::geom_rect(data = data,

### 2023-02-16 scr/visualize_semi_peptides.R

@@ -1,0 +1,3 @@
+# visualize protein coverage by peptide identification
+
+# 

### 2023-02-16 scr/feature_specific_fdr_correction.R
proteolysis_cleanup2

@@ -1,33 +1,0 @@
-## function for feature-specific FDR correction
-## used specifically in this case to apply FDR correction only on semi-specific peptides
-
-feature_fdr_correction <- function(toptable,
-                                   features_table,
-                                   interesting_features_table){
-  
-  # merge limma output with feature annotation  ----
-  
-  tab_limma_feature_annot <- left_join(toptable, 
-                                       features_table,
-                                       by = c("peptide","index"))
-  
-  # filter limma table based on interesting feature ----
-  
-  tab_limma_subsetin <- filter(tab_limma_feature_annot,
-                               peptide %in% interesting_features_table$peptide) %>%
-    # apply FDR correction only on the subset
-    mutate(adj.P.Val = p.adjust(p = P.Value, 
-                                method = "BH")) %>%
-    mutate(fdr_correction = 'feature-specific')
-  
-  # create a table of non-interesting features containing the globally adjusted p-values
-  tablimma_subsetout <- filter(tab_limma_feature_annot,
-                               !peptide %in% interesting_features_table$peptide) %>%
-    mutate(fdr_correction = 'global')
-  
-  output_limma3 <- bind_rows(tab_limma_subsetin,
-                             tablimma_subsetout) 
-  
-  return(output_limma3)
-  
-}

### 2023-02-16 reports/gbm_recurrence_proteogenomics_manuscript_report.qmd
proteogenomics_clean_up

--- 2022-01-21 scr/gbm_proteomics_data_analysis_proteolytic_products_prep_figures_v3_gh_format.Rmd
@@
 ---
-title: "GBM Proteomics and Proteolytic processing data analysis - Reproducible report"
-author: "Miguel Cosenza"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: 
-  github_document
+title: "GBM Recurrence Proteogenomics"
+author: "Miguel Cosenza-Contreras"
+format:
+  gfm:
+    toc: true 
+    toc-depth: 4 # default is 3
+    number-sections: true
+    execute:
+      output: true
 always_allow_html: true
+editor: source
 ---
@@
                       warning = FALSE)
-```
-
-```{r results='hide'}
+
+source(here::here("scr/helper_functions.R"))
+
 ## Required packages ----
@@
 library(ggpubr)
-
-source(here("scr/helper_functions.R"))
-
-theme_set(theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, 
-                                           angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0, size = 6),
+library(ggrepel)
+library(extrafont)
+extrafont::loadfonts(device = "win")
+
+theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
+                                           vjust = 0, 
+                                           size = 6, 
+                                           angle = 360),
+                axis.text.y = element_text(hjust = 0.5, 
+                                           vjust = 0, 
+                                           size = 6),
                 panel.background = element_blank(),
                 panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=8),
-                legend.text = element_text(size = 7),
+                panel.border = element_rect(colour = "black", 
+                                            fill = NA, 
+                                            size = 1.5),
+                axis.title=element_text(size = 8),
                 legend.title = element_text(size = 8),
-                legend.key.height= unit(3, 'mm'),
-                legend.key.width= unit(3, 'mm'),
-                legend.position="bottom"))
-```
-
-The following report holds the code, results and interpretation of the proteomics data analysis of Initial-Recurrent glioblastoma samples of matched patients.
-
-# Sample annotation info
-
-```{r echo=TRUE}
-sample_annotation <- read_csv(here("data/sample_annotation.csv"))
-
-# correct annotation
-
-sample_annotation2 <- sample_annotation %>%
-  mutate(patient = paste("x",patient, sep = ""),
-         recurrence = case_when(recurrence == "initial" ~ "prim",
-                                recurrence == "recurrent" ~ "rec",
-                                TRUE ~ recurrence)) %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
-  filter(recurrence %in% c("prim", "rec"))
-
-sample_annotationwo6 <- sample_annotation2 %>%
-  filter(patient != "x6")
-```
-
-## Sample annotation
-
-### Brief note on sample annotation:
-
-Twenty-two patient matched samples (corresponding to 11 patients) were processed for protein extraction, TMT-labelled and submitted via high-pH fractionation and further analyzed via LC-MS/MS.
-
-```{r echo=FALSE, error=TRUE}
-kbl(sample_annotation, booktabs = TRUE) %>%
-  kable_styling(font_size = 14, bootstrap_options = c("striped", "hover"))
-```
-
-# Brief note on data analysis and processing:
-
-Three proteomics data analysis approaches were used to explore the spectral data and will be briefly described:
-
-1. General large scale proteomics.
-2. Proteolytic processing.
-3. Proteogenomics
-
-The FragPipe bioinformatic pipeline was used for each of these approaches with varying parameters depending on the type of explorative approach.
-
-__1.
-General large scale proteomics analyses__ : In order to explore the differential abundance of proteins between Recurrent and Initial tumor, a fully-tryptic search was performed against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides.
-A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.
-
-__2.
-Proteolytic processing__: The large scale evaluation of differential proteolytic activity between recurrent vs primary tumor was executed via a semi-tryptic search against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides.
-A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.
-
-__3.
-Proteogenomics__: The explorative proteogenomics analysis started with the generation of a GBM-specific database from publicly available RNA-seq sequencing data from tumor specimens.
-Fastq files were obtain from the SRA BioProject PRJNA627121 (Cell Rep, 2021 Mar 2;34(9):108787).
-Eight samples containing paired reads were selected, based on: 1) the higher number of reads per file; 2) having a similar proportion of males and females and 3) not bearing isocitrate-dehyderogenase (IDH)-1 mutation (IDH WT).
-None of these patients undergo previous treatment, therefore potential variants in the sequencing data are not expected to be associated to treatment-related mutagenic processes.
-Spectral data was then searched against this custom database using a fully-tryptic approac.
-No minimum intensity threshold was set for reporter ion intensity for quantitation.
-
-Exceptionally when stated differently within each sub-approach described above, all these searches were executed using the following parameters: MSFragger was used as a search engine, setting a fully or semi-tryptic specificity allowing for 1 missed cleavages.
-Precursor mass tolerance was set to -20/20 and fragment mass tolerance of 20 ppm with mass calibration and parameter optimization resulting in 5 ppm.
-Peptide N-terminal acetylation and peptide N-terminal TMT labeling were set as variable modifications.
-TMT labelling at K and carbamidomethylation of C were set as fixed modifications.
-MSBooster was used for deep-learning based predictions of retention time and spectra.
-Predicted features were used by Percolator for post-processing scoring and false discovery rate (FDR) control via target-decoy competition of all peptide-to-spectrum matches (PSMs) obtained from MSFragger search.
-ProteinProphet was used as a protein inference algorithm keeping all PSMs with probability score bigger than 0.9.
-Final report and FDR estimation was based on the filtered PSM and protein lists before protein scoring.
-
-Relative quantitation of identified peptides within each sample was performed via their reported ion intensities using TMT-integrator.
-Only PSMs coming from unique peptides with a minimum probability of 0.9 and an isotopic purity of at least 50% were considered for quantitation.
-Quantitative values were normalized via median centering and summarized by protein/gene and peptide using a virtual reference channel, to finally generate protein and peptide normalized abundance matrices of abundance used for further statistical processing in combination with the sample annotation.
-
-# General proteomics analysis
-
-## Initial data loading and wrangling
-
-Load id/quant data and prep coverage information
-
-```{r message=FALSE}
-## tmt-integrator output loading 
-tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
-  janitor::clean_names()
-
-identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
-  janitor::clean_names()
-
-identified_peptides <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_peptide.tsv")) %>%
-  janitor::clean_names()
-
-fasta_ident_prots <- read.fasta(here("data/specific_search_fragpipe17/specific_no_ptms_2/mix_2/protein.fas"), seqtype = "AA", as.string = TRUE)
-
-```
-
-Wrangle and merge protein ID/quant with sample annotation
-
-```{r}
-# wide expression matrix with NAs
-expr_matrix <- dplyr::select(tmt_protdata_frag,
-                             Protein = index, starts_with("x"))
-
-long_mat <- pivot_longer(expr_matrix,
-                         cols = starts_with("x"),
-                         names_sep = "_",
-                         names_to = c("patient", "recurrence"),
-                         values_to = "Abundance")
-
-long_mat_2 <- long_mat %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_"))
-
-paired_annot_uniq <- long_mat_2 %>%
-  dplyr::select(patient, recurrence, paired_id) %>%
-  distinct()
-
-# correct annotation
-
-sample_annotation2 <- sample_annotation %>%
-  mutate(patient = paste("x",patient, sep = ""),
-         recurrence = case_when(recurrence == "initial" ~ "prim",
-                                recurrence == "recurrent" ~ "rec",
-                                TRUE ~ recurrence)) %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
-  filter(recurrence %in% c("prim", "rec"))
-
-## Annotated abundance data in long format ----
-
-quant_annotated <- left_join(long_mat_2,
-                             sample_annotation2, 
-                             by = c("paired_id", "patient","recurrence")) %>%
-  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))
-
-## Wide xpression matrices without NAs
-expr_matnona <- column_to_rownames(expr_matrix,
-                            "Protein") %>%
-          as.matrix() %>%
-          na.omit()
-```
-
-## ID coverage
-
-### Coverage plot
-
-```{r}
-ids_summary <- quant_annotated %>% 
-  dplyr::select(Protein, mixture, Abundance) %>% 
-  distinct(Protein, mixture, .keep_all = TRUE) %>%
-  group_by(mixture) %>%
-  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
-  bind_rows(.,
-            tibble(mixture = "In all three",
-                   `Nr of IDs` = nrow(expr_matnona))) %>%
-  bind_rows(.,
-            tibble(mixture = "Overall",
-                   `Nr of IDs` = nrow(expr_matrix)))
-```
-
-* Number of identified proteins: `r nrow(expr_matrix)`
-
-* Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`
-
-The initial analysis with MaxQuant yielded 2090 IDs in all samples (without missing values) and 5176 overall.
-
-```{r fig.height=4, fig.width=7}
-ggplot(data = ids_summary,
-       aes(y = `Nr of IDs`, x = mixture)) +
-  geom_col(fill = "#008683") + 
-  geom_text(size = 5, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
-  coord_cartesian(ylim = c(0, 5000))+
-  coord_flip() +
-  labs(title = "Number of Identified and quantified Proteins by Mixture") +
-  ylab("Nr of IDed and Quant Proteins") + 
-  xlab("TMT Mixture") + 
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
-        axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
-        panel.background = element_blank(),
-        panel.grid.major = element_line(color = "grey"),
-        panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-        axis.title=element_text(size=12,face="bold"))
-```  
-## Exploratory analysis and visualizations
-
-### Visualization of missing values
-
-```{r fig.height=7, fig.width=8}
-vis_miss(expr_matrix) + 
-          ggtitle("Distribution of missing values") + 
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-```
-
-### QC plots after TMT integrator pre-processing and normalization (Protein level)  {.tabset .tabset-fade}
-
-#### By TMT channel/sample {.unnumbered}
-
-##### General Abudance distribution by TMT channel/sample {.unlisted .unnumbered}
-
-```{r fig.height=6, fig.width=9, warning=FALSE}
-qcplot <- ggplot(quant_annotated,
-                 mapping = aes(x = Channel_mix, 
-                               y = Abundance, 
-                               fill = recurrence)) +
-          geom_boxplot() + 
-          #facet_grid(.~recurrence, scales = "free") +
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-
-print(qcplot)
-
-```
-
-The median-centering after TMT-integrator seem to show a slight mixture-dependent centered median per sample.
-The paired distribution of initial and recurrent tumor allows the biological effect to be more important than any potential TMT-mixture-based batch effect.
-
-The different abundance distribution can also be explained by a different distribution of missing values.
-
-PCA and sPLS-DA analysis allow to confirm this.
-
-#### By Protein (General)
-
-```{r}
-# select/sample N proteins to observe their abundance distribution  
-which_prots <- sample_proteins(x = quant_annotated$Protein,
-                               size = 50,
-                               seed = 156)
-```
-
-`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.
-
-```{r fig.height=5.5, fig.width=8, warning=FALSE}
-
-qcplot_prots <- ggplot(quant_annotated %>% 
-                         filter(Protein %in% which_prots),
-                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
-                               y = Abundance)) +
-          geom_boxplot() + 
-          geom_jitter(color="black", size=0.4, alpha=0.9) +
-          labs(subtitle = "Protein IDs arranged by median abudance",
-               x = "Protein ID ") + 
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-
-print(qcplot_prots)
-```
-
-#### By Protein and Batch {.unnumbered}
-
-`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.
-
-```{r fig.height=5, fig.width=10, warning=FALSE}
-qcplot_prots2 <- ggplot(quant_annotated %>% 
-                         filter(Protein %in% which_prots),
-                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
-                               y = Abundance, 
-                               fill = mixture)) +
-          geom_boxplot() + 
-          geom_jitter(color="black", size=0.4, alpha=0.9) +
-          labs(subtitle = "Protein IDs arranged by median abudance",
-                       x = "Protein ID ") + 
-          facet_grid(.~mixture, scales = "free") +
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-
-print(qcplot_prots2)
-```
-
-The abundance distribution of individual proteins can be considered as normaly distributed in general, which makes then suitable for inferential statistics via linear models.
-
-### Exploratory PCA analisys
-
-```{r echo=FALSE}
-## Get predictors for mixtures
-expr_matrix2 <- expr_matrix %>%
-  dplyr::select(Protein, sample_annotation2$paired_id)
-
-paired_id <- sample_annotation2$paired_id
-recurrence <- sample_annotation2$recurrence
-patient <- sample_annotation2$patient
-mixture <- sample_annotation2$mixture
-
-## get predictor for mixtures wo patient 6
-expr_matrixwo6 <- expr_matrix %>%
-  dplyr::select(Protein, sample_annotationwo6$paired_id)
-
-paired_idwo6 <- sample_annotationwo6$paired_id
-recurrencewo6 <- sample_annotationwo6$recurrence
-patientwo6 <- sample_annotationwo6$patient
-mixturewo6 <- sample_annotationwo6$mixture
-
-## Prep matrix without missing values for PCA
-
-expr_matnona <- column_to_rownames(expr_matrix2,
-                            "Protein") %>%
-          as.matrix() %>%
-          na.omit()
-
-colnames(expr_matnona) <- paste(sample_annotation2$paired_id, 
-                                sample_annotation2$mixture,
-                                sep = "_")
-
-t_expr_matnona <- t(expr_matnona)
-
-expr_mat <- column_to_rownames(expr_matrix2,
-                               "Protein") %>%
-          as.matrix()
-
-t_expr_mat <- t(expr_mat)
-
-## Matrix without patient 6
-
-expr_matnonawo6 <- column_to_rownames(expr_matrixwo6,
-                            "Protein") %>%
-          as.matrix() %>%
-          na.omit()
-
-colnames(expr_matnonawo6) <- paste(sample_annotationwo6$paired_id, 
-                                sample_annotationwo6$mixture,
-                                sep = "_")
-
-t_expr_matnonawo6 <- t(expr_matnonawo6)
-
-expr_matwo6 <- column_to_rownames(expr_matrixwo6,
-                               "Protein") %>%
-          as.matrix()
-
-t_expr_matwo6 <- t(expr_matwo6)
-```
-
-#### General PCA
-
-##### Percentage of explained variance per component
-
-The PCA was applied on the expression matrix without missing values
-
-```{r}
-pca_res = pca(t_expr_matnona, ncomp = 10, center = TRUE, scale = TRUE)
-```
-
-```{r fig.width=7, fig.height=5}
-plot(pca_res)
-```
-
-~45% of variance based on protein abundance can be explained by components 1 and 2 after PCA.
-
-##### PCA plot on samples based on protein abundance
-
-```{r}
-# preprocess pca results 
-pca_variates <- pca_res$variates$X %>% 
-  as.data.frame() %>%
-  rownames_to_column("Sample") %>%
-  separate(col = "Sample",
-           into = c("sample", "recurrence", "mixture"),
-           sep = "\\_", extra = "merge") %>%
-  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
-  dplyr::select(-c(mixture, recurrence)) %>%
-  left_join(. , sample_annotation2,
-            by = "paired_id") %>%
-  mutate(Stage = if_else(condition = recurrence == "prim", 
-                         true = "Initial", 
-                         false = "Recurrent"),
-         Mix = case_when(mixture == "tmt_1" ~ "Mix1",
-                         mixture == "tmt_2" ~ "Mix2",
-                         mixture == "tmt_3" ~ "Mix3"))
-
-pca_variateswopat6 <- pca_variates %>% 
-  filter(patient != "x6")
-
-```
-
-```{r fig.width=7.3, fig.height=5}
-ggplot(data = pca_variates,
-       aes(x = PC1, y = PC2)) +
-  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
-  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
-  labs(title = "PCA plot of samples\nBased on the abundance values of proteins") +
-  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
-  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 10),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-There is no evident batch effect after PCA analysis.
-
-There is a good number of individual patients that show important differences in their PCA variates values in recurrent vs primary.
-
-```{r fig.width=7.3, fig.height=5}
-ggplot(data = pca_variateswopat6,
-       aes(x = PC1, y = PC2)) +
-  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
-  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
-  labs(title = "PCA plot of samples\nBased on the abundance values of proteins (patient 6 excluded)") +
-  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
-  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 10),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-```{r}
-pca_plot <- ggplot(data = pca_variateswopat6,
-       aes(x = PC1, y = PC2)) +
-  geom_point(size = 4, aes(shape = Mix, color = Stage)) +  
-  geom_text(aes(label = patient), position = position_nudge(x = 4.5), size = 2) +
-  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
-  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) +
-theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-      axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-      panel.background = element_blank(),
-      panel.grid.major = element_blank(),
-      panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-      axis.title = element_text(size = 7),
-      legend.text = element_text(size = 7),
-      legend.title = element_text(size = 8),
-      legend.key.height= unit(3, 'mm'),
-      legend.key.width= unit(3, 'mm'),
-      legend.position="bottom")
-```
-
-```{r}
-ggsave(plot = pca_plot, 
-       filename = here::here("figures/pca_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = pca_plot, 
-       filename = here::here("figures/pca_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-### Exploratory sPLS-DA
-
-```{r}
-plsda_rest1 <- plsda(X = t_expr_matnona,
-                     Y = recurrence,
-                     ncomp = 3)
-```
-
-```{r fig.width=6, fig.height=5}
-plotIndiv(plsda_rest1,
-          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
-          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
-          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)
-
-```
-
-Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.
-
-Therefore we decided to exclude patient 6 from further analyses.
-
-```{r}
-plsda_rest2 <- plsda(X = t_expr_matnona,
-                     Y = mixture,
-                     ncomp = 3)
-```
-
-```{r fig.width=6, fig.height=5}
-plotIndiv(plsda_rest2,
-          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
-          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
-          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)
-
-```
-
-We can say that the biological differences between initial and recurrent status are stronger than the differences that could come from potential batch effects.
-
-### Exploratory sPLS-DA without patient 6
-
-```{r}
-plsda_restwo6 <- plsda(X = t_expr_matnonawo6,
-                     Y = recurrencewo6,
-                     ncomp = 3)
-```
-
-```{r}
-plotIndiv(plsda_restwo6,
-          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim wo Pat6", 
-          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
-          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)
-
-```
-
-#### GGplot-based visualization of Exploratory sPLS-DA
-
-```{r}
-# preprocess pca results 
-plsda_variates <- plsda_restwo6$variates$X %>% 
-  as.data.frame() %>%
-  rownames_to_column("Sample") %>%
-  separate(col = "Sample",
-           into = c("sample", "recurrence", "mixture"),
-           sep = "\\_", extra = "merge") %>%
-  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
-  dplyr::select(-c(mixture, recurrence)) %>%
-  left_join(. , sample_annotation2,
-            by = "paired_id")
-
-plsda_variateswpat6 <- plsda_rest1$variates$X %>% 
-  as.data.frame() %>%
-  rownames_to_column("Sample") %>%
-  separate(col = "Sample",
-           into = c("sample", "recurrence", "mixture"),
-           sep = "\\_", extra = "merge") %>%
-  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
-  dplyr::select(-c(mixture, recurrence)) %>%
-  left_join(. , sample_annotation2,
-            by = "paired_id")
-
-```
-
-```{r fig.width=7.3, fig.height=5}
-ggplot(data = plsda_variateswpat6,
-       aes(x = comp1, y = comp2)) +
-  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
-  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
-  labs(title = "sPLS-DA plot of samples\nBased on the abundance values of proteins") +
-  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
-  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
-  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 10),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.
-
-Therefore we decided to exclude patient 6 from further analyses.
-
-```{r fig.width=7.3, fig.height=5}
-ggplot(data = plsda_variates,
-       aes(x = comp1, y = comp2)) +
-  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
-  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
-  labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
-  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
-  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
-  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 10),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-The difference between initial and recurrent tumor is better resolved in the sPLS-DA plot when excluding patient 6.
-
-## Sparcity reduction
-
-Only proteins that are identified and quantified in 2 out of 3 TMT mixtures will be kept for the inferential analysis. The proteins missing in 1 mixture will be imputed with `missForest`.
-
-__Function to select proteins to exclude:__
-
-```{r}
-sel_proteins_missing <- function(long_matrix,
-                                 threshold) {
-
-na_count <- group_by(long_matrix,
-                     Protein, mixture) %>%
-    summarise(na_count = sum(is.na(Abundance)),
-              total = n()) %>% 
-    ungroup() %>% 
-    mutate(NA_fraction = na_count/total)
-
-na_count_perbatch <- na_count %>%
-  group_by(Protein) %>%
-  summarise(na_per_batch = sum(NA_fraction)) %>%
-  ungroup()
-
-proteins2exclude <- na_count_perbatch %>%
-  filter(na_per_batch > threshold) %>%
-  pull(Protein)
-
-return(proteins2exclude)
-}
-
-```
-
-__Excluded proteins:__
-
-```{r}
-proteins2exclude <- sel_proteins_missing(quant_annotated,
-                                         threshold = 1)
-```
-
-Exclude proteins from initial abundance matrix.
-
-```{r}
-expr_matrix_filt <- expr_matrix2 %>%
-  filter(!Protein %in% proteins2exclude)
-```
-
-### Visualize missing values (before and after filtering)
-
-```{r}
-nofiltnas_plot <- vis_miss(expr_matrix) + 
-          ggtitle("Distribution of missing values",
-                  subtitle = paste("All quant proteins. Total =", nrow(expr_matrix))) + 
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-
-filtnas_plot <- vis_miss(expr_matrix_filt) + 
-          ggtitle("Distribution of missing values",
-                  subtitle = paste("Only present in 2/3 TMT mixtures. Total =", nrow(expr_matrix_filt))) + 
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
-                panel.background = element_blank(),
-                panel.grid.major = element_line(color = "grey"),
-                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-                axis.title=element_text(size=12,face="bold"))
-```
-
-```{r fig.height=6, fig.width=12}
-cowplot::plot_grid(nofiltnas_plot, filtnas_plot, ncol = 2)
-```
-
-## `missForest` imputation:
-
-```{r}
-mat_filt <- expr_matrix_filt %>%
-  column_to_rownames("Protein") %>%
-  as.matrix()
-
-t_mat_filt <- t(mat_filt)
-
-if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){
-
- t_imp_mat_filt <- missForest::missForest(t_mat_filt)
- t_imp_filt_mat <- t_imp_mat_filt$ximp
-
- mat_filt_imp <- t(t_imp_filt_mat)
-
- saveRDS(t_imp_mat_filt,
-         file = here("rds/missforest_imp_mat_filt.Rds"))
-
-} else {
-
- t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
- t_imp_filt_mat <- t_imp_mat_filt$ximp
-
- mat_filt_imp <- t(t_imp_filt_mat)
-
-}
-
-```
-
-## Inferential analysis
-
-__Protein to Gene annotation__
-
-```{r}
-prot2gene <- tmt_protdata_frag %>%
-  dplyr::select(Protein = index, Gene = gene)
-```
-
-__Prep protein abundance matrix__
-
-```{r}
-mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
-  rownames_to_column("Protein") %>%
-  dplyr::select(-starts_with("x6")) %>%
-  column_to_rownames("Protein") %>%
-  as.matrix()
-
-sample_annotationwo6 <- sample_annotation2 %>%
-  filter(patient != "x6")
-```
-
-### Set up design matrix
-
-Initial vs recurrent samples are patient-matched, therefore the design matrix is setup so the linear model can account for the patient random effect.
-
-```{r}
-design_wo6 <- model.matrix(~patientwo6+recurrencewo6)
-
-rownames(design_wo6) <- sample_annotationwo6$paired_id
-```
-
-Check design matrix:
-
-```{r}
-print(design_wo6)
-```
-
-```{r}
-mat_filt_impwo6[1:3,1:12]
-```
-
-### Fit limma
-
-```{r}
-source(here("scr/fit_limmawo6.R"))
-```
-
-```{r}
-# fit limma
-limma_tab_wo6 <- fit_limmawo6(mat_filt_impwo6, 
-                         design_wo6, 
-                         method = 'robust', 
-                         Limma = "Robust - w Patient effect wo Pat6",
-                         prot2gene = prot2gene)
-
-#get proteins increased in recurrence
-increased_in_rec <- limma_tab_wo6 %>%
-  filter(logFC > 0,
-         adj.P.Val < 0.05) %>%
-  pull(Protein)
-
-#get proteins decreased in recurrence
-decreased_in_rec <- limma_tab_wo6 %>%
-  filter(logFC < 0,
-         adj.P.Val < 0.05) %>%
-  pull(Protein)
-```
-
-### Volcano plots
-
-```{r fig.width=8, fig.height=6}
-our_volcano(limma_tab_wo6, 
-            FC_cutoff = 0, 
-            pval_cutoff = 0.05, 
-            color_diffex = "red", 
-            color_nondifex = "#2a9d8f", 
-            interesting_proteins = NULL, 
-            vert_line_col = "red",
-            hline_col = "red", 
-            hline_pos = 0.05, 
-            linetype = "dashed",
-            increased_in = "Recurrent", 
-            comparison_title = "Recurrent vs Initial") +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 10),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom") 
-```
-
-```{r fig.width=8, fig.height=6, echo=FALSE}
-protlabels <- limma_tab_wo6 %>%
-      filter(logFC > 0 | logFC < 0,
-             adj.P.Val < 0.05)
-
-signi_hits <- nrow(protlabels)
-
-signi_up <- limma_tab_wo6 %>%
-      filter(logFC > 0 ,
-             adj.P.Val < 0.05) %>%
-  nrow()
-
-signi_down <- limma_tab_wo6 %>%
-      filter(logFC < 0 ,
-             adj.P.Val < 0.05) %>%
-  nrow()
-
-size <- 0.4
-
-volcano <- ggplot(data = limma_tab_wo6,
-                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
-                 color = "red",
-                 size = size)+
-      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
-                 size = size) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      ggrepel::geom_text_repel(data = protlabels,
-                               aes(label = Gene), size = 1.5) +
-      geom_hline(yintercept = -log10(0.05),
-                 color = "red", linetype = "dashed") +
-      xlab("logFC - Recurrent / Primary")+
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-```
-
-```{r echo=FALSE}
-ggsave(plot = volcano, 
-       filename = here::here("figures/volcano_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = volcano, 
-       filename = here::here("figures/volcano_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-```{r fig.width=8, fig.height=6, echo=FALSE}
-protlabels <- limma_tab_wo6 %>%
-      filter(logFC > 0 | logFC < 0,
-             adj.P.Val < 0.05,
-             Gene %in% c("ASAH1", "GPNMB", "SYNM"))
-
-signi_hits <- nrow(protlabels)
-
-size <- 0.4
-
-volcano2 <- ggplot(data = limma_tab_wo6,
-                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
-                 color = "red",
-                 size = size)+
-      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
-                 size = size) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC > 0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      geom_point(data = limma_tab_wo6 %>% filter(logFC < -0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      ggrepel::geom_text_repel(data = protlabels,
-                               aes(label = Gene), size = 3.5) +
-      geom_hline(yintercept = -log10(0.05),
-                 color = "red", linetype = "dashed") +
-      xlab("logFC - Recurrent / Primary")+
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-```
-
-```{r echo=FALSE}
-ggsave(plot = volcano2, 
-       filename = here::here("figures/volcano2_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = volcano2, 
-       filename = here::here("figures/volcano2_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-### Boxplots specific proteins
-
-```{r}
-quant_anwgene <- left_join(quant_annotated, prot2gene)
-
-tab_2_paired_boxplot <-  quant_anwgene %>%
-  filter(Gene %in% c("ASAH1", "SYNM", "GPNMB"))
-# prep data
-  slim_data <- pivot_longer(expr_matrix_filt, 
-                          cols = 2:length(names(expr_matrix_filt)), 
-                          names_to = c("patient","condition"),
-                          values_to = "Intensity",
-                          names_sep = "\\_") %>% 
-  group_by(Protein)
-
-  slim_data2 <- pivot_wider(slim_data, 
-                          values_from = Intensity, 
-                          names_from = condition) %>%
-    filter(patient != "x6")
-
-  slim_data3 <- slim_data2 %>% 
-    mutate(Abs_diff = rec-prim) %>%
-    mutate(rank = row_number(Abs_diff)) %>% 
-    dplyr::rename(Initial = prim, Recurrent = rec) %>%
-    ungroup() %>%
-    filter(Protein %in% c("Q13510", "Q14956", "O15061")) %>%
-    mutate(Gene = case_when(Protein == "Q13510" ~ "ASAH1",
-                              Protein == "Q14956" ~ "GPNMB",
-                              Protein == "O15061" ~ "SYNM"))
-
-```
-
-```{r fig.width=7, fig.height=3.5}
-ggpaired(slim_data3,
-         cond1 = "Initial",
-         cond2 = "Recurrent", fill = "condition",
-         ylab = "Normalized Abundance",
-         label = NULL,
-         repel = TRUE,
-         facet.by = "Gene", point.size = 0.2, line.size = 0.1) +
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom",
-        strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) 
-```
-
-```{r echo=FALSE}
-paired_boxplots <- ggpaired(slim_data3,
-         cond1 = "Initial",
-         cond2 = "Recurrent", fill = "condition",
-         ylab = "Normalized Abundance",
-         label = NULL,
-         repel = TRUE,
-         facet.by = "Gene", point.size = 0.2, line.size = 0.1) +
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 6),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.height= unit(2, 'mm'),
-        legend.key.width= unit(2, 'mm'),
-        legend.position="bottom",
-        strip.text = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360)) 
-```
-
-```{r echo=FALSE}
-ggsave(plot = paired_boxplots, 
-       filename = here::here("figures/paired_box_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 61,
-       height = 45)
-
-ggsave(plot = paired_boxplots, 
-       filename = here::here("figures/paired_box_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 61,
-       height = 45)
-```
-
-## ID coverage (final)
-
-### Coverage plot
-
-```{r}
-ids_summary2 <- quant_annotated %>% 
-  dplyr::select(Protein, mixture, Abundance) %>% 
-  distinct(Protein, mixture, .keep_all = TRUE) %>%
-  group_by(mixture) %>%
-  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
-  bind_rows(.,
-            tibble(mixture = "Quant_in_all",
-                   `Nr of IDs` = nrow(expr_matnona))) %>%
-  bind_rows(.,
-            tibble(mixture = "Quant_2/3_mixtures",
-                   `Nr of IDs` = nrow(mat_filt_impwo6))) %>%
-  bind_rows(.,
-            tibble(mixture = "Overall",
-                   `Nr of IDs` = nrow(expr_matrix)))
-
-ids_summary2 <- mutate(ids_summary2,
-                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
-                                           mixture == "tmt_2" ~ "Mix_2",
-                                           mixture == "tmt_3" ~ "Mix_3",
-                                           mixture == "Quant_in_all" ~ "Quant. in all",
-                                           mixture == "Overall" ~ "Overall",
-                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
-  mutate(Mixture = factor(Mixture, levels = c("Overall","Quant. in 2/3", "Quant. in all",
-                                               "Mix_3", "Mix_2", "Mix_1")))
-```
-
-Number of identified proteins: `r nrow(expr_matrix)`
-
-Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`
-
-Number of identified and quantified proteins 2 out of 3 batches: `r nrow(mat_filt_impwo6)`
-
-```{r fig.width=7, fig.height=4}
-ggplot(data = ids_summary2,
-       aes(y = `Nr of IDs`, x = mixture)) +
-  geom_col(fill = "#008683") +
-
-  geom_text(size = 5, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
-  coord_cartesian(ylim = c(0, 5000))+
-  coord_flip() +
-  labs(title = "Number of Identified and quantified Proteins by Mixture") +
-  ylab("Nr of IDed and Quant Proteins") + 
-  xlab("TMT Mixture") + 
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 14, angle = 0),
-        axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 12, angle = 0),
-        panel.background = element_blank(),
-        panel.grid.major = element_line(color = "grey"),
-        panel.border = element_rect(colour = "black", fill=NA, size=1.5),
-        axis.title=element_text(size=12, face="bold"))
-```
-```{r echo=FALSE}
-barplot_cover <- ggplot(data = ids_summary2,
-       aes(y = `Nr of IDs`, x = Mixture)) +
-  geom_col(fill = "#008683") + 
-  geom_text(size = 2, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
-  coord_cartesian(ylim = c(0, 5000))+
-  coord_flip() +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-```{r echo=FALSE}
-ggsave(plot = barplot_cover, 
-       filename = here::here("figures/barplot_coverage_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = barplot_cover, 
-       filename = here::here("figures/barplot_coverage_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-## Median abundance distribution of filtered and inputed samples
-
-```{r}
-expr_matrix_filtwo6 <- as.data.frame(mat_filt_imp) %>%
-  rownames_to_column("Protein") %>%
-  dplyr::select(-starts_with("x6"))
-
-long_mat_f <- pivot_longer(expr_matrix_filtwo6,
-                         cols = starts_with("x"),
-                         names_sep = "_",
-                         names_to = c("patient", "recurrence"),
-                         values_to = "Abundance")
-
-long_mat_3 <- long_mat_f %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_"))
-
-paired_annot_uniq2 <- long_mat_3 %>%
-  dplyr::select(patient, recurrence, paired_id) %>%
-  distinct()
-
-paired_annot_uniq2$paired_id
-## Annotated abundance data in long format ----
-
-quant_annotated_f <- left_join(long_mat_3,
-                             sample_annotation2, 
-                             by = c("paired_id", "patient","recurrence")) %>%
-  mutate(Channel_mix = paste(mixture,channel,  sep = "_"),
-         paired_id = factor(paired_id, levels = paired_annot_uniq2$paired_id)) %>%
-  mutate(Sample = str_replace_all(paired_id, pattern = "prim", replacement = "init"),
-         Stage = if_else(recurrence == "prim",
-                         true = "Initial",
-                         false = "Recurrent"))
-
-```
-
-### By TMT channel/sample {.unnumbered}
-
-#### General Abudance distribution by TMT channel/sample {.unlisted .unnumbered}
-
-```{r}
-qcplot_f <- ggplot(quant_annotated_f,
-                 mapping = aes(x = Sample, 
-                               y = Abundance, 
-                               fill = Stage)) +
-          geom_boxplot(outlier.size=0.1, size = 0.25) + 
-          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 90),
-                axis.text.y = element_text(hjust = 0.5, vjust = 0, size = 6),
-                panel.background = element_blank(),
-                panel.grid.major = element_blank(),
-                panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-                axis.title=element_text(size=8),
-                axis.title.x = element_blank(),
-                legend.text = element_text(size = 7),
-                legend.title = element_text(size = 8),
-                legend.key.height= unit(3, 'mm'),
-                legend.key.width= unit(3, 'mm'),
-                legend.position="bottom")
-
-```
-
-```{r fig.height=4, fig.width=5, warning=FALSE, echo=FALSE}
-print(qcplot_f)
-```
-```{r echo=FALSE}
-ggsave(plot = qcplot_f, 
-       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = qcplot_f, 
-       filename = here::here("figures/boxplot_abund_dist_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-## Over-representation analyses/tests (ORAs)
-
-### Prep tabular info for ORAs
-
-```{r}
-tab_sig_prots_tryptic <- bind_rows(tibble(protein = increased_in_rec,
-                                  characteristic = "up-regulated"),
-                           tibble(protein = decreased_in_rec,
-                                  characteristic = "down-regulated"))
-
-limma_tab_wo62_tryptic <- limma_tab_wo6
-
-unip2symb_tryptic <- bitr(limma_tab_wo62_tryptic$Protein, 
-                  fromType = "UNIPROT", 
-                  toType = c("SYMBOL", "ENTREZID"),
-                  OrgDb = org.Hs.eg.db) %>%
-  dplyr::rename(protein = UNIPROT) %>%
-  mutate(Protein = protein)
-
-tab_sig_prots_tryptic <- left_join(tab_sig_prots_tryptic, unip2symb_tryptic)
-
-unip2symbIDed_tryptic <- bitr(identified_proteins$protein_id, 
-                  fromType = "UNIPROT", 
-                  toType = c("SYMBOL", "ENTREZID"),
-                  OrgDb = org.Hs.eg.db)
-
-limma_tab_wo62_tryptic <- left_join(limma_tab_wo62_tryptic, unip2symb_tryptic)
-
-geneList <- limma_tab_wo62_tryptic$logFC
-names(geneList) <- limma_tab_wo62_tryptic$Protein
-geneList <-  sort(geneList, decreasing = TRUE)
-
-geneListr <- limma_tab_wo62_tryptic$logFC
-names(geneListr) <- limma_tab_wo62_tryptic$ENTREZID
-geneListr <-  sort(geneListr, decreasing = TRUE)
-```
-
-### Reactome
-
-```{r}
-group_comparison_react <- compareCluster(ENTREZID~characteristic, 
-                                              data=tab_sig_prots_tryptic, 
-                                              fun="enrichPathway",
-                                              organism = "human",
-                                              pvalueCutoff = 0.05,
-                                              pAdjustMethod = "BH",
-                                              qvalueCutoff = 0.2,
-                                              universe = unip2symbIDed_tryptic$ENTREZID,
-                                              minGSSize = 10,
-                                              maxGSSize = 1000,
-                                              readable = TRUE)
-```
-
-#### Dotplot Reactome
-
-```{r}
-enrichplot::dotplot(group_comparison_react, x = "characteristic") + 
-          xlab("Quant in Recurrent status") +
-  scale_color_continuous(low="red", high="blue",
-            guide=guide_colorbar(reverse=TRUE),breaks = c(0.01, 0.04)) +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-
-```{r}
-dotplot_react <- enrichplot::dotplot(group_comparison_react, x = "characteristic") + 
-          xlab("Quant in Recurrent status") +
-  scale_color_continuous(low="red", high="blue",
-            guide=guide_colorbar(reverse=TRUE),breaks = c(0.01, 0.04)) +
-  scale_size_continuous(breaks = c(0.10, 0.25)) +
-  theme(axis.text.x = element_text(hjust = 0.4, vjust = 0.1, size = 6, angle = 340),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 6),
-        panel.background = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.height= unit(2, 'mm'),
-        legend.key.width= unit(2, 'mm'),
-        legend.position="bottom")
-```
-```{r}
-ggsave(plot = dotplot_react, 
-       filename = here::here("figures/dotplot_reactome_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 91,
-       height = 90)
-
-ggsave(plot = dotplot_react, 
-       filename = here::here("figures/dotplot_reactome_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 91,
-       height = 90)
-```
-
-#### Cnet plot Reactome
-
-```{r}
-enrichplot::cnetplot(group_comparison_react) + 
-  theme(legend.text = element_text(size = 7),
-                legend.title = element_text(size = 8),
-                legend.key.height= unit(3, 'mm'),
-                legend.key.width= unit(3, 'mm'),
-                legend.position="bottom")
-```
-
-```{r}
-cnet_plot <- enrichplot::cnetplot(group_comparison_react) + 
-  theme(legend.text = element_text(size = 7),
-        text = element_text(size = 4),
-                legend.title = element_text(size = 8),
-                legend.key.height= unit(3, 'mm'),
-                legend.key.width= unit(3, 'mm'),
-                legend.position="bottom")
-```
-
-```{r}
-ggsave(plot = cnet_plot, 
-       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 110,
-       height = 110)
-
-ggsave(plot = cnet_plot, 
-       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 110,
-       height = 110)
-
-ggsave(plot = cnet_plot, 
-       filename = here::here("figures/cnet_plot_reactome_proteomics_gbm.svg"), 
-       device = "svg",
-       units = "mm",
-       width = 110,
-       height = 110)
-```
-
-## Protein coverage plots
-
-```{r}
-source(here("scr/protein_peptide_coverage_plot.R"))
-```
-
-### Annotate peptides from interesting proteins
-
-```{r}
-source(here("scr/annotate_peptides.R"))
-```
-
-```{r}
-# get protein to peptide pairs
-pept2prot_inter <- identified_peptides %>%
-  filter(gene %in% c("ASAH1", "SYNM", "GPNMB")) %>%
-  dplyr::select(Peptide = sequence, Genes = protein_id)
-
-annotated_peptides <- annotate_peptides(expr_mat = pept2prot_inter,
-                                        fasta = fasta_ident_prots)
-
-annotated_peptides2 <- annotated_peptides %>%
-  mutate(peptide_length = str_length(Peptide))
-```
-
-### ASAH1
-
-```{r}
-feat_asah1 <- drawProteins::get_features("Q13510")
-
-feat_asah1df <- drawProteins::feature_to_dataframe(feat_asah1)
-#Get peptides as coverage features for plotting
-
-cov_feat_asah1 <- get_coverage(annotated_peptides = annotated_peptides2, 
-                               id = "Q13510")
-#Merge coverage feature info with protein sequence features info:
-
-fullfeat_asah1 <- add_peptides(feat_asah1df,
-                               peptide_coverage_data = cov_feat_asah1)
-# drawCanvas
-canvas <- draw_canvas(fullfeat_asah1)
-
-# draw the protein chain
-wchain <- draw_chains(canvas, fullfeat_asah1)
-
-# draw protein domains
-wdomainsasah1 <- draw_domains(wchain, data = fullfeat_asah1)
-#pasa <- drawProteins::draw_folding(wdomainsasah1, fullfeat_asah1)
-cov_feat_asah1$order <- 1.4
-
-pepcov_asah1 <- draw_peptides(wdomainsasah1, cov_feat_asah1)
-pepcov_asah1
-```
-
-```{r}
-ggsave(plot = pepcov_asah1, 
-       filename = here::here("figures/asah1_coverage_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 150,
-       height = 100)
-
-ggsave(plot = pepcov_asah1, 
-       filename = here::here("figures/asah1_coverage_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 150,
-       height = 100)
-```
-
-### SYNM coverage plot
-Get features from Uniprot
-
-```{r}
-feat_SYNM <- drawProteins::get_features("O15061")
-feat_SYNMdf <- drawProteins::feature_to_dataframe(feat_SYNM)
-#Get peptides as coverage features for plotting
-
-cov_feat_SYNM <- get_coverage(annotated_peptides = annotated_peptides2, 
-                               id = "O15061")
-#Merge coverage feature info with protein sequence features info:
-
-fullfeat_SYNM <- add_peptides(feat_SYNMdf,
-                               peptide_coverage_data = cov_feat_SYNM)
-# drawCanvas
-canvasSYNM <- draw_canvas(fullfeat_SYNM)
-
-# draw the protein chain
-wchainSYNM <- draw_chains(canvasSYNM, fullfeat_SYNM)
-cov_feat_SYNM$order <- 1.4
-
-pepcov_SYNM <- draw_peptides(wchainSYNM, cov_feat_SYNM)
-pepcov_SYNM
-```
-
-```{r}
-ggsave(plot = pepcov_SYNM, 
-       filename = here::here("figures/SYNM_coverage_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 150,
-       height = 100)
-
-ggsave(plot = pepcov_SYNM, 
-       filename = here::here("figures/SYNM_coverage_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 150,
-       height = 100)
-```
-
-# Analysis of proteolytic processing
-
-## Load semi-specific search results
-
-```{r}
-tmt_peptdata_frag <- read_tsv(here("data/semi_specific_search_fragpipe17/tmt-report/abundance_peptide_MD.tsv")) %>% 
-  janitor::clean_names()
-
-pept_ids_mix1 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_1/peptide.tsv")) %>%
-                    clean_names() %>%
-          dplyr::select(-c(starts_with("empty")))
-
-pept_ids_mix2 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_2/peptide.tsv")) %>%
-                    clean_names() %>%
-          dplyr::select(-c(starts_with("x"), starts_with("empty"), "master"))
-
-pept_ids_mix3 <- read_delim(here("data/semi_specific_search_fragpipe17/mix_3/peptide.tsv")) %>%
-                    clean_names() %>%
-          dplyr::select(-c(starts_with("x"), starts_with("empty"), "master"))
-
-identified_proteins_semi <- read_tsv(here("data/semi_specific_search_fragpipe17/combined_protein.tsv")) %>% 
-  janitor::clean_names()
-
-```
-
-## Wrangle and merge peptide ID/quant with sample annotation
-
-```{r}
-# wide expression matrix with NAs
-pept_matrix <- dplyr::select(tmt_peptdata_frag,
-                             index, starts_with("x"))
-
-pept_long_mat <- pivot_longer(pept_matrix,
-                         cols = starts_with("x"),
-                         names_to = c("patient"),
-                         values_to = "Abundance") %>%
-  mutate(recurrence = str_extract(patient, "[^0-9]+$")) %>%
-  mutate(patient = str_remove_all(patient, "[^0-9]+$"))
-
-pept_long_mat_2 <- pept_long_mat %>%
-  mutate(paired_id_semi = paste(patient, recurrence, sep = "_"))
-
-sample_annotation2 <- sample_annotation2 %>%
-  mutate(paired_id_semi = paste(patient, recurrence, sep = "_"))
-
-## Annotated abundance data in long format ----
-
-pept_quant_annotated <- left_join(pept_long_mat_2,
-                             sample_annotation2, 
-                             by = c("paired_id_semi", "patient","recurrence")) %>%
-  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))
-
-## Wide xpression matrices without NAs
-pept_expr_matnona <- column_to_rownames(pept_matrix,
-                            "index") %>%
-          as.matrix() %>%
-          na.omit()
-```
-
-Wrangle and merge peptide ids for further annotation based on they proteolytic specificity
-
-```{r}
-peptide_joined1 <- full_join(pept_ids_mix1, pept_ids_mix2) %>%
-                    mutate(duplicated = duplicated(peptide))
-
-peptide_joined <- full_join(peptide_joined1, pept_ids_mix3) %>%
-                    mutate(duplicated = duplicated(peptide))
-
-peptide_ids_filtered <- filter(peptide_joined,
-                               peptide %in% peptide_joined$peptide) %>%
-                    distinct(peptide, assigned_modifications,observed_modifications, .keep_all = TRUE) %>%
-                    dplyr::rename(Peptide = peptide) %>%
-                    dplyr::mutate(Genes = protein_id)
-
-```
-
-### Annotate peptides with cleavage information
-
-```{r}
-fasta1 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_1/protein.fas"),
-                            seqtype = "AA", as.string = TRUE)
-
-fasta2 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_2/protein.fas"),
-                             seqtype = "AA", as.string = TRUE)
-
-fasta3 <- seqinr::read.fasta(here("data/semi_specific_search_fragpipe17/mix_3/protein.fas"),
-                             seqtype = "AA", as.string = TRUE)
-
-fasta22 <- append(fasta1, fasta2)
-
-fasta <- append(fasta22, fasta3)
-
-fasta <- discard(fasta,
-                 duplicated(names(fasta)))
-
-attr(fasta[[1]], "Annot")
-
-extr_annt <- function(x){
-  ext <- attr(x, "Annot")
-}
-
-names <- purrr::map_chr(fasta, .f = extr_annt)
-
-seqinr::write.fasta(fasta,
-                    file.out = here("data/semi_specific_search_fragpipe17/protein_combined.fas"),
-                    names = names)
-
-if(!file.exists(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))){
-  cleavage_annoated_peptides <- annotate_peptides(expr_mat = peptide_ids_filtered, 
-                                                  fasta = fasta,
-                                                  decoy_tag = "rev_")
-
-  write_tsv(cleavage_annoated_peptides, here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))
-} else {
-  cleavage_annoated_peptides <- read_tsv(here("results/semi_tryptic/cleavage_annoated_peptides_final.tsv"))
-}
-
-cleavage_annoated_peptides <- dplyr::mutate(cleavage_annoated_peptides,
-                                            Genes = protein_id)
-
-peptides_annotated <- left_join(peptide_ids_filtered, cleavage_annoated_peptides,
-                                by = c("Peptide", "protein_id"))
-
-```
-
-## Sparcity reduction
-
-Only peptides that were identified and quantified in all 3 mixtures would be used for further inferential analyses for proteolytic processing.
-
-```{r}
-pept_matrix_nona <- pept_matrix %>% 
-  na.omit()
-
-pept_nona <- pept_matrix_nona %>%
-  dplyr::select(index) %>%
-  separate(col = index, 
-           sep = "_", into = c("protein", "peptide"))
-
-pept_matrix_nonawop6 <- pept_matrix %>% 
-  dplyr::select(-starts_with("x6")) %>%
-  na.omit()
-
-```
-
-## Coverage information (peptide-level)
-
-```{r}
-ids_summarypep <- pept_quant_annotated %>% 
-  dplyr::select(index, mixture, Abundance) %>% 
-  distinct(index, mixture, .keep_all = TRUE) %>%
-  group_by(mixture) %>%
-  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
-  bind_rows(.,
-            tibble(mixture = "In all three",
-                   `Nr of IDs` = nrow(pept_matrix_nona)))# %>%
-  #bind_rows(.,
-  #          tibble(mixture = "Overall",
-  #                `Nr of IDs` = nrow(pept_matrix)))
-
-ids_summarypep <- mutate(ids_summarypep,
-                       Mixture = case_when(mixture == "tmt_1" ~ "Mix_1",
-                                           mixture == "tmt_2" ~ "Mix_2",
-                                           mixture == "tmt_3" ~ "Mix_3",
-                                           mixture == "In all three" ~ "Quant. in all",
-                                           mixture == "Overall" ~ "Overall",
-                                           mixture == "Quant_2/3_mixtures" ~ "Quant. in 2/3")) %>%
-  mutate(Mixture = factor(Mixture, levels = c("Overall", "Quant. in all",
-                                               "Mix_3", "Mix_2", "Mix_1")))
-```
-
-```{r fig.height=4, fig.width=7}
-ggplot(data = ids_summarypep,
-       aes(y = `Nr of IDs`, x = Mixture)) +
-  geom_col(fill = "#008683") + 
-  geom_text(size = 5, position = position_stack(vjust = 0.5), 
-            aes(label = `Nr of IDs`)) +
-  coord_cartesian(ylim = c(0, 5000))+
-  coord_flip() +
-  labs(title = "Number of Identified and quantified Peptides by Mixture") +
-  ylab("Nr of IDed and Quant Peptides") + 
-  xlab("TMT Mixture") + 
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```
-```{r}
-barplot_pept_cover <- ggplot(data = ids_summarypep,
-       aes(y = `Nr of IDs`, x = Mixture)) +
-  geom_col(fill = "#008683") + 
-  geom_text(size = 2, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
-  coord_flip() +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-```  
-```{r}
-ggsave(plot = barplot_pept_cover, 
-       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 38,
-       height = 50)
-
-ggsave(plot = barplot_pept_cover, 
-       filename = here::here("figures/barplot_coverage_pept_proteomics_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 38,
-       height = 50)
-```
-
-## Label N-terminal peptides
-
-```{r echo=TRUE}
-source(here("scr/annotate_nterm.R"))
-```
-
-```{r echo=TRUE}
-nterannot <- annotate_nterm(peptides_annotated,
-                            tmtmass = 229.1629) %>%
-  clean_names()
-```
-
-### Merge peptide annotation with TMT integrator quant report
-
-```{r}
-# all quantified peptides
-tmt_reprt_annot <- left_join(tmt_peptdata_frag, nterannot)
-
-check_na <- filter(tmt_reprt_annot, is.na(semi_type))
-
-excluded_tmt_reprt_annot <- anti_join(nterannot, tmt_reprt_annot) %>% 
-                    clean_names()
-
-#only those present in all mixtures
-tmt_peptdata_fragnona <- tmt_peptdata_frag %>%
-  filter(peptide %in% pept_nona$peptide)
-
-tmt_reprt_annotnona <- left_join(tmt_peptdata_fragnona, nterannot) 
-```
-
-## Summarize feature counts per feature
-
-```{r echo=TRUE}
-source(here("scr/summarize_peptide_counts.R"))
-```
-
-```{r echo=TRUE}
-summary_count <- summarize_peptide_counts(tmt_reprt_annot) 
-```
-
-### Barplot
-
-These are the ones that would be used for inferential statistics.
-
-```{r echo=TRUE}
-to_count_info <- tmt_reprt_annot %>% 
-   filter(index %in% pept_matrix_nona$index) %>%
-                      dplyr::select(index, specificity, nterm, semi_type, tmt_tag,
-                                    aa_after, aa_before, following_10_resid, previous_10_resid, start_position, end_position) %>%
-  mutate(specificity = if_else(str_detect(semi_type, "unspecific"), 
-                               true = "semi_specific",
-                               false = specificity),
-         duplicated = duplicated(index)) %>%
-  distinct()
-
-to_count_info_semi <- to_count_info %>%
-                           dplyr::select(index,specificity) %>%
-                           distinct() %>%
-  mutate(duplicated = duplicated(index))
-
-false_duplication <- to_count_info_semi %>%
-  filter(duplicated) %>%
-  pull(index)
-
-to_count_info_semi <- to_count_info_semi %>%
-  mutate(specificity = if_else(index %in% false_duplication, 
-                               true = "semi_specific",
-                               false = specificity)) %>%
-  dplyr::select(-duplicated) %>%
-  distinct()
-
-  n_semi <- dplyr::count(to_count_info_semi, specificity) %>% 
-                      dplyr::rename(feature_type = specificity) %>%
-                      dplyr::mutate(category = "specificity")
-
-  n_total <- tibble(feature_type = "Total",
-                    n = length(unique(to_count_info$index)),
-                    category = "Total")
-
-  n_term <- dplyr::count(to_count_info, nterm) %>% 
-                      dplyr::rename(feature_type = nterm) %>%
-                      dplyr::mutate(category = "N-term")
-
-  summary_count_pept_nona <- bind_rows(n_semi,
-                                       n_term,
-                                       n_total)
-
-total_sumcount <- summary_count_pept_nona %>%
-  filter(feature_type == "Total") %>%
-  pull(n)
-
-summary_countnona_no_tot <- summary_count_pept_nona %>%
-  dplyr::filter(category != "Total") %>%
-  dplyr::rename(`Feature category` = category)
-```
-
-```{r fig.width=5, fig.height=2, echo=FALSE}
-ggplot(summary_countnona_no_tot,
-       aes(x = feature_type, y = n, fill = `Feature category`)) + 
-  coord_flip()+ 
-  geom_col() +
-  geom_text(aes(label = n), hjust = -0.2, size = 2.5) + 
-  scale_y_continuous(limits = c(0, 17000), breaks = c(0, 7000, 15000)) + 
-  labs(y = "Number of IDs", x = "Feature type") +
-  facet_wrap(~`Feature category`, nrow = 1, scales="free") +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom", 
-        strip.background = element_blank(),
-        strip.text = element_blank())
-```
-
-```{r echo=FALSE}
-summary_pept_annot <- ggplot(summary_countnona_no_tot,
-       aes(x = feature_type, y = n, fill = `Feature category`)) + 
-  coord_flip()+ 
-  geom_col() +
-  geom_text(aes(label = n), hjust = -0.2, size = 2) + 
-  scale_y_continuous(limits = c(0, 20000), breaks = c(0, 7000, 15000)) + 
-  labs(y = "Number of IDs", x = "Feature type", 
-       caption = paste0("Total = ",total_sumcount)) +
-  facet_wrap(~`Feature category`, nrow = 1, scales="free") +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.text = element_text(size = 7),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom", 
-        strip.background = element_blank(),
-        strip.text = element_blank())
-
-```
-```{r}
-ggsave(plot = summary_pept_annot, 
-       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 76,
-       height = 50)
-
-ggsave(plot = summary_pept_annot, 
-       filename = here::here("figures/barplot_coverage_peptidesfeatures_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 76,
-       height = 50)
-```
-
-## Inferential analysis (peptide-level)
-
-### Prep design matrix
-
-```{r}
-design_limmawo6 <- model.matrix(~patientwo6+recurrencewo6)
-
-sample_annotationwo6 <- sample_annotationwo6 %>%
-  mutate(paired_id_semi = paste(patient, recurrence, sep = ""))
-
-rownames(design_limmawo6) <- sample_annotationwo6$paired_id_semi
-```
-
-### Prep abundance matrices
-
-```{r}
-pep_matwo6 <- dplyr::select(pept_matrix_nonawop6,
-                             index, rownames(design_limmawo6))  %>%
-  column_to_rownames("index") %>%
-  as.matrix()
-
-```
-
-```{r}
-colnames(pep_matwo6)
-```
-```{r}
-rownames(design_limmawo6)
-```
-
-The column names in our peptide abundance matrix have the same order as the row names in our design matrix.
-
-### Peptide-level model fit
-
-```{r}
-fit_limmapeptwo6 <- function(mat, design, method,Limma){
-  limmafit <- lmFit(mat, design, method = method)
-  limmafit <- eBayes(limmafit)
-  limma_tab <- topTable(limmafit, coef = "recurrencewo6rec", number = Inf, adjust.method = "BH") %>%
-  mutate(Protein = rownames(.),
-         Limma = Limma)
-
-  return(limma_tab)
-}
-```
-
-```{r echo=TRUE, warning=FALSE, message=FALSE}
-fit_mod_peptwo6 <- fit_limmapeptwo6(pep_matwo6, 
-                         design_limmawo6, 
-                         method = 'robust', 
-                         Limma = "Robust - w Patient effect")
-
-compar_tab_peptwo6 <-  fit_mod_peptwo6 %>%
-                    separate(Protein, 
-                             into = c("protein", "peptide"), 
-                             remove = FALSE) %>%
-  dplyr::rename(index = Protein)
-
-```
-
-### Extract coefficients and feature specific FDR correction
-
-#### Interesting features
-
-```{r}
-features1 <- tmt_reprt_annotnona %>%
-                    dplyr::select(peptide, index, specificity, nterm, 
-                                  semi_type, is_terminal)
-
-interesting_features1 <- features1 %>%
-                    filter(specificity == "semi_specific",
-                           is_terminal == "not_terminal") 
-```
-
-```{r}
-# load function for feature specific fdr correction
-source(here("scr/feature_specific_fdr_correction.R"))
-```
-
-```{r}
-# load function for volcano plot
-source(here("scr/plot_volcano.R"))
-```
-
-```{r}
-# excluding patient 6
-compar_tab_correctedwo6_pept <- feature_fdr_correction(compar_tab_peptwo6,
-                                               features1,
-                                               interesting_features1)
-
-compar_tab_interestingwo6_pept <- compar_tab_correctedwo6_pept %>% 
-  dplyr::rename(Protein = protein) %>%
-  filter(fdr_correction == 'feature-specific') %>%
-  left_join(.,prot2gene)
-
-increased_recwo6_pept <- compar_tab_interestingwo6_pept %>%
-                    filter(logFC>0,
-                           adj.P.Val < 0.05) %>% 
-  pull(peptide)
-
-decreased_recwo6_pept <- compar_tab_interestingwo6_pept %>%
-                    filter(logFC<0,
-                           adj.P.Val < 0.05) %>% 
-  pull(peptide)
-
-check_increased_pept <- cleavage_annoated_peptides %>%
-  filter(Peptide %in% increased_recwo6_pept)
-
-```
-
-```{r}
-length(increased_recwo6_pept)
-```
-
-```{r}
-length(decreased_recwo6_pept)
-```
-
-`r length(increased_recwo6_pept)` semi-specific peptides are increased in Recurrent tumor and `r length(decreased_recwo6_pept)` are decreased in Recurrent tumor.
-
-### Volcano peptides
-
-```{r fig.width=8, fig.height=6}
-recwo6_pept22 <- compar_tab_interestingwo6_pept %>%
-  filter(adj.P.Val < 0.05)
-
-size <- 0.4
-
-volcano_pept <- ggplot(data = compar_tab_interestingwo6_pept,
-                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
-      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC > 0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
-                 color = "red",
-                 size = size)+
-      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC < -0,
-                                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
-                 size = size) +
-      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC > 0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      geom_point(data = compar_tab_interestingwo6_pept %>% filter(logFC < -0,
-                                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
-                 size = size) +
-      geom_hline(yintercept = -log10(0.05),
-                 color = "red", linetype = "dashed") +
-      xlab("logFC - Recurrent / Primary") + 
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-```
-
-```{r}
-ggsave(plot = volcano_pept, 
-       filename = here::here("figures/volcano_proteolysis_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 50)
-
-ggsave(plot = volcano_pept, 
-       filename = here::here("figures/volcano_proteolysis_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 50)
-```
-
-## Proportional abundance of proteolytic peptides
-
-### Prepare data
-
-__Wrangle data__
-
-```{r}
-pept_semi_annot <- peptides_annotated %>%
-  dplyr::select(Peptide, specificity, is_terminal)
-
-pept_quant_to_summary <- pept_quant_annotated %>%
-  separate(col = index, 
-           into = c("Protein", "Peptide"), 
-           sep = "\\_", 
-           remove = FALSE) %>%
-  left_join(.,pept_semi_annot)
-
-pep_quant_presummary <- pept_quant_to_summary %>%
-  dplyr::select(Protein, Peptide, recurrence, paired_id, 
-                Abundance, specificity, is_terminal)
-
-```
-
-__Prepare data-frame with abudance percentage of semi-specific peptides per sample__
-
-```{r}
-pept_summary_int_all <- pep_quant_presummary %>%
-  group_by(paired_id, recurrence) %>%
-  summarise(Sum_All = sum(Abundance, na.rm = TRUE))
-
-pept_summary_int_semi <- pep_quant_presummary %>%
-  filter(specificity == "semi_specific",
-         is_terminal == "not_terminal") %>%
-  group_by(paired_id, recurrence) %>%
-  summarise(Sum_Semi = sum(Abundance, na.rm = TRUE))
-
-pept_sum_summary <- left_join(pept_summary_int_all, pept_summary_int_semi) %>%
-  mutate(Percentage = Sum_Semi/Sum_All * 100,
-         Stage = if_else(recurrence == "prim",
-                         true = "Initial",
-                         false = "Recurrence"))
-
-```
-
-### Generate plot % of abundance of semi-specific peptides
-
-```{r}
-plotSumEM <- ggplot(pept_sum_summary, 
-                    aes(x = Stage, 
-                        y = Percentage, fill = Stage, 
-                        cex.axis = 1.5)) +
-  geom_boxplot() +
-  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5) +
-  # Box plot with jittered points
-  # 0.2 : degree of jitter in x direction
-  # geom_jitter(shape=16, position=position_jitter(0.2))
-  ylab("Proportional intensity of semi-tryptic peptides [%]") +
-  geom_signif(
-    comparisons = list(c("Initial", "Recurrence")),
-    map_signif_level = TRUE
-  ) + 
-  stat_compare_means(method="t.test") +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-```
-
-```{r}
-plotSumEM
-```
-
-```{r}
-ggsave(plot = plotSumEM, 
-       filename = here::here("figures/box_perc_semispec_pepts.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 40,
-       height = 50)
-
-ggsave(plot = plotSumEM, 
-       filename = here::here("figures/box_perc_semispec_pepts.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 40,
-       height = 50)
-```
-
-## Comparative analysis of semi-specific peptides vs protein abundance
-
-### Prepare data
-
-#### Get log2FC Summary of abundances of semi-specific peptides
-
-```{r}
-#taking output from limma semi-specific
-log2_semi_tryptic <- dplyr::select(compar_tab_interestingwo6_pept,
-                                   logFC, Protein, peptide, specificity, 
-                                   is_terminal,
-                                   Gene) %>%
-  filter(specificity == "semi_specific",
-         is_terminal == "not_terminal") %>%
-  dplyr::rename(logFC_semi_peptide = logFC) %>%
-  dplyr::select(-c(specificity, is_terminal))
-
-```
-
-#### Get log2 Summary of protein abundances based only on fully-tryptic peptides
-
-```{r}
-med_prot_quant_specific <- pep_quant_presummary %>%
-  group_by(Protein, paired_id) %>%
-  summarise(`median(Abundance)` = median(Abundance, na.rm = TRUE)) %>%
-  ungroup()
-```
-
-##### Prep wide matrix of median protein abundances from only fully-tryptic peptides
-
-```{r}
-# prep wide matrix for median proteins abundances (arousing only from fully-tryptic peptides)
-wide_premat_prot_only_tryp <- med_prot_quant_specific %>%
-  pivot_wider(id_cols = c("Protein"),
-              values_from = `median(Abundance)`, 
-              names_from = paired_id) %>%
-  # keep only proteins present in 2 / 3 of the samples
-  filter(!Protein %in% proteins2exclude)
-
-mat_prot_only_tryp <- wide_premat_prot_only_tryp %>%
-  column_to_rownames("Protein") %>%
-  as.matrix()
-```
-
-__missForest-based imputation__
-
-```{r}
-t_mat_prot_only_tryp <- t(mat_prot_only_tryp)
-
-if(!file.exists(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))){
-
- imp_prot_only_tryp <- missForest::missForest(t_mat_prot_only_tryp)
- t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp
-
- mat_prot_only_tryp <- t(t_mat_prot_only_tryp)
-
- saveRDS(imp_prot_only_tryp,
-         file = here("rds/missforest_imp_mat_prot_only_tryp.Rds"))
-
-} else {
-
- imp_prot_only_tryp <- readRDS(here("rds/missforest_imp_mat_prot_only_tryp.Rds"))
- t_mat_imp_prot_only_tryp <- imp_prot_only_tryp$ximp
-
- mat_imp_prot_only_tryp <- t(t_mat_imp_prot_only_tryp)
-
-}
-
-```
-
-__Run limma on this new data to generate log2 FCs of proteins from only fully-tryptic peptides__
-
-```{r}
-split <- str_split_fixed(colnames(mat_imp_prot_only_tryp), pattern = "\\_", n = 2)
-
-patient_tryp <- split[,1]
-recurrence <- split[,2]
-
-design_tryp <- model.matrix(~patient_tryp+recurrence)
-
-rownames(design_tryp) <- colnames(mat_imp_prot_only_tryp)
-```
-
-```{r}
-rownames(design_tryp) 
-```
-
-```{r}
-colnames(mat_prot_only_tryp)
-```
-
-```{r}
-fit_limmawo6_tryp <- function(mat, design, method,Limma, prot2gene){
-  limmafit <- lmFit(mat, design, method = method)
-  limmafit <- eBayes(limmafit)
-  limma_tab <- topTable(limmafit, coef = "recurrencerec", number = Inf, adjust.method = "BH") %>%
-  mutate(Protein = rownames(.),
-         Limma = Limma)
-
-  limma_tab <- left_join(limma_tab, prot2gene)
-
-  return(limma_tab)
-}
-```
-
-```{r}
-limma_tab_wo6tryp <- fit_limmawo6_tryp(mat_imp_prot_only_tryp, 
-                         design_tryp, 
-                         method = 'robust', 
-                         Limma = "Robust - Proteins w only tryptic peptides",
-                         prot2gene = prot2gene)
-
-```
-
-```{r}
-log2_fully_tryptic <- dplyr::select(limma_tab_wo6tryp,
-                                   logFC, Protein, Gene) %>%
-  dplyr::rename(logFC_fully_tryp_protein = logFC) 
-```
-
-#### Merge log2FCs of semi-tryptic peptides vs proteins with fully tryptic peptides
-
-```{r}
-log2semipept2_log2protein_spec <- left_join(log2_fully_tryptic, log2_semi_tryptic) %>%
-  na.omit() %>%
-  mutate(DA_Protein = if_else(Protein %in% tab_sig_prots_tryptic$protein,
-                              true = TRUE,
-                              false = FALSE),
-         DA_peptide = if_else(peptide %in% recwo6_pept22$peptide,
-                              true = TRUE,
-                              false = FALSE))
-
-```
-
-#### log2FC Proteins vs Semi-tryp peptides (scatter plot)
-
-```{r}
-inter_scater <- log2semipept2_log2protein_spec %>%
-  filter(Protein %in% check_increased_pept$protein_id)
-
-inter_scater$Gene[duplicated(inter_scater$Gene)] <- NA
-
-#geom_point(data = dataarg %>% filter(logFC < -FC_cutoff,
-#                                           adj.P.Val < pval_cutoff),
-#                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red") +
-#      geom_point(data = dataarg %>% filter(logFC > FC_cutoff,
-#                                           adj.P.Val > pval_cutoff),
-#                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f") +
-
-scatter_proteolysis_plot <- ggplot(log2semipept2_log2protein_spec, 
-                                   aes(x = logFC_fully_tryp_protein, 
-                                       y = logFC_semi_peptide)) + 
-  geom_smooth(method=lm, se = FALSE, linetype="dashed", size = 0.5, 
-              color = "black") + 
-  geom_point(aes(color = DA_Protein, shape = DA_peptide), size = 0.6) +
-  scale_color_manual(values = c("#2a9d8f", "red")) +
-  xlab("log2(FC) - Protein abundances") + 
-  ylab("log2(FC) - Semi-specific peptides") + 
-  ggrepel::geom_text_repel(data = inter_scater,
-                               aes(label = Gene), 
-                           size = 1,
-                           box.padding = 0.5,
-                           max.overlaps = 25) +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
-        panel.background = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
-        axis.title = element_text(size = 8),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-cor.test(log2semipept2_log2protein_spec$logFC_fully_tryp_protein,
-    log2semipept2_log2protein_spec$logFC_semi_peptide, method = "pearson")
-```
-
-```{r}
-ggsave(plot = scatter_proteolysis_plot, 
-       filename = here::here("figures/scatter_proteolysis_plot_gbm.tiff"), 
-       device = "tiff",
-       units = "mm",
-       width = 63,
-       height = 63)
-
-ggsave(plot = scatter_proteolysis_plot, 
-       filename = here::here("figures/scatter_proteolysis_plot_gbm.eps"), 
-       device = "eps",
-       units = "mm",
-       width = 63,
-       height = 63)
-```
-
-## Analysis of differential amino acid usage (iceLogo)
-
-```{r}
-source(here("scr/get_cleavage_area.R"))
-```
-
-```{r}
-cleave_areas <- get_cleave_area(cleavage_annoated_peptides)
-```
-
-```{r}
-library(dagLogo)
-```
-
-**Load background proteome**
-
-Used the identified protein sequences as background proteome.
-
-```{r}
-proteome_ided <- prepareProteome(fasta = here("data/semi_specific_search_fragpipe17/protein_combined.fas"), 
-                                 species = "Homo sapiens")
-```
-
-```{r}
-increased_Rec <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% increased_recwo6_pept)
-
-decreased_Rec <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% decreased_recwo6_pept)
-
-write_tsv(x = increased_Rec,
-          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))
-
-write_tsv(x = decreased_Rec,
-          file = here("results/semi_tryptic/proteolytic_products_increased_in_Recurrent.tsv"))
-
-increased_Rec_4ice <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% increased_recwo6_pept) %>%
-                    pull(cleave_area20)
-
-decreased_Rec_4ice <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% decreased_recwo6_pept) %>%
-                    pull(cleave_area20)
-```
-
-**Format peptide sequences**
-
-```{r}
-if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))){
-
-  form_peptidesincreased_4ice <- formatSequence(increased_Rec_4ice, 
-                                  proteome = proteome_ided)
-
-  write_rds(form_peptidesincreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))
-} else {
-
-  form_peptidesincreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))
-
-}
-```
-
-```{r}
-if(!file.exists(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))){
-
-  form_peptidesdecreased_4ice <- formatSequence(decreased_Rec_4ice, 
-                                  proteome = proteome_ided)
-
-  write_rds(form_peptidesdecreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))
-} else {
-
-  form_peptidesdecreased_4ice <- read_rds(here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))
-
-}
-```
-
-#### DAU increased
-
-```{r}
-bg_mod_ztest_increased <- buildBackgroundModel(form_peptidesincreased_4ice,
-                                           proteome = proteome_ided,
-                           background = "wholeProteome",
-                           testType = "ztest")
-```
-
-```{r}
-dau_nogroup_increased <- testDAU(form_peptidesincreased_4ice, 
-                           dagBackground = bg_mod_ztest_increased)
-```
-
-```{r}
-dagHeatmap(dau_nogroup_increased) 
-```
-
-#### DAU decreased
-
-```{r}
-bg_mod_ztest_decreased <- buildBackgroundModel(form_peptidesdecreased_4ice,
-                                           proteome = proteome_ided,
-                           background = "wholeProteome",
-                           testType = "ztest")
-```
-
-```{r}
-dau_nogroup_decreased <- testDAU(form_peptidesdecreased_4ice, 
-                           dagBackground = bg_mod_ztest_decreased)
-```
-
-```{r}
-dagHeatmap(dau_nogroup_decreased) 
-```
-
-# Proteogenomics
-
-## Initial data loading and wrangling
+                legend.key.height = unit(3, 
+                                         'mm'),
+                legend.key.width = unit(3, 
+                                        'mm'),
+                legend.position = "bottom",
+                text = element_text(family = "Helvetica")))
+
+```
+
+# Initial data loading and wrangling
 
@@
 
+index2genepeptide <- tmt_peptdata_frag_progen %>%
+  dplyr::select(index, gene, peptide, protein_id)
+```
+
+```{r}
+sample_annotation <- read_csv(here("data/sample_annotation.csv"))
+
+# correct annotation
+
+sample_annotation2 <- sample_annotation %>%
+  mutate(patient = paste("x",
+                         patient, 
+                         sep = ""),
+         recurrence = case_when(recurrence == "initial" ~ "prim",
+                                recurrence == "recurrent" ~ "rec",
+                                TRUE ~ recurrence)) %>%
+  mutate(paired_id = paste(patient, 
+                           recurrence, 
+                           sep = "_")) %>%
+  filter(recurrence %in% c("prim", 
+                           "rec"))
+
+sample_annotationwo6 <- sample_annotation2 %>%
+  filter(patient != "x6")
 ```
@@
                          values_to = "Abundance") %>%
-  mutate(recurrence = str_extract(paired_id_semi, "[^0-9]+$")) %>%
-  mutate(patient = str_remove_all(paired_id_semi, "[^0-9]+$"))
+  mutate(recurrence = str_extract(paired_id_semi, 
+                                  "[^0-9]+$")) %>%
+  mutate(patient = str_remove_all(paired_id_semi, 
+                                  "[^0-9]+$"))
 
 long_mat_progen_2 <- long_mat_progen %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_"))
+  mutate(paired_id = paste(patient, 
+                           recurrence, 
+                           sep = "_"))
 
 paired_annot_uniq_progen <- long_mat_progen_2 %>%
-  dplyr::select(patient, recurrence, paired_id) %>%
+  dplyr::select(patient, 
+                recurrence, 
+                paired_id) %>%
   distinct()
@@
 expr_matrix_pept_progen <- dplyr::select(tmt_peptdata_frag_progen,
-                             index, peptide, starts_with("x")) %>%
+                             index,
+                             peptide,
+                             starts_with("x")) %>%
   dplyr::select(!starts_with("x6"))
@@
                          values_to = "Abundance") %>%
-  mutate(recurrence = str_extract(paired_id_semi, "[^0-9]+$")) %>%
-  mutate(patient = str_remove_all(paired_id_semi, "[^0-9]+$"))
+  mutate(recurrence = str_extract(paired_id_semi, 
+                                  "[^0-9]+$")) %>%
+  mutate(patient = str_remove_all(paired_id_semi, 
+                                  "[^0-9]+$"))
 
 long_mat_pept_progen_2 <- long_mat_pept_progen %>%
-  mutate(paired_id = paste(patient, recurrence, sep = "_"))
+  mutate(paired_id = paste(patient, 
+                           recurrence, 
+                           sep = "_"))
 
 paired_annot_uniq_pept_progen <- long_mat_pept_progen_2 %>%
-  dplyr::select(patient, recurrence, paired_id) %>%
+  dplyr::select(patient, 
+                recurrence, 
+                paired_id) %>%
   distinct()
@@
                              sample_annotation2) %>%
-  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))
+  mutate(Channel_mix = paste(mixture,
+                             channel, 
+                             sep = "_"))
 
@@
           separate(col = protein,
-                   into = c("sp", "Genes", "ensembl"),
+                   into = c("sp", 
+                            "Genes", 
+                            "ensembl"),
                    sep = "\\|") %>%
-          dplyr::select(Genes, Peptide) %>%
+          dplyr::select(Genes, 
+                        Peptide) %>%
           distinct() %>%
-          mutate(Peptide = str_remove_all(Peptide, "\\[.+?\\]")) %>%
-          mutate(Peptide = str_remove_all(Peptide, "^n"))
+          mutate(Peptide = str_remove_all(Peptide, 
+                                          "\\[.+?\\]")) %>%
+          mutate(Peptide = str_remove_all(Peptide, 
+                                          "^n"))
 
@@
 protein2peptide_tmtrept <- tmt_peptdata_frag_progen %>%
-  dplyr::select(index, protein_id, peptide)
-
-```
-
-## Localization of identified peptides and mapping to variant regions
+  dplyr::select(index, 
+                protein_id, 
+                peptide)
+```
+
+# Localization of identified peptides and mapping to variant regions
 
@@
 var_prots <- protein2peptideall %>%
-                    mutate(variant = if_else(str_detect(Genes, "\\_"),
+                    mutate(variant = if_else(str_detect(Genes, 
+                                                        "\\_"),
                            true = "variant",
@@
 
-### Localized peptides within protein sequence and confirm identification of SAAVs
+## Localize peptides within protein sequence and confirm identification of SAAVs
 
@@
           mutate(protein_id_var = protein_id,
-                 variant_type = if_else(str_detect(protein_id, "\\:"), 
+                 variant_type = if_else(str_detect(protein_id, 
+                                                   "\\:"), 
                                         true = "other",
                                         false = "saav")) %>%
-          mutate(protein_id = str_replace_all(protein_id, "\\.", "\\_")) %>%
-          mutate(protein_id = str_replace_all(protein_id, "\\:", "\\_")) %>%
+          mutate(protein_id = str_replace_all(protein_id, 
+                                              "\\.", 
+                                              "\\_")) %>%
+          mutate(protein_id = str_replace_all(protein_id, 
+                                              "\\:", 
+                                              "\\_")) %>%
           separate(protein_id, 
@@
                     na.omit() %>%
-          mutate(variant_type = if_else(condition = str_detect(variant_location, "[0-9]$"),
+          mutate(variant_type = if_else(condition = str_detect(variant_location, 
+                                                               "[0-9]$"),
                                         true = "other",
@@
 
-locat2prot <- str_split(annotated_w_variant_location$variant_location,
-                        pattern = "\\_")
-
-names(locat2prot) <- annotated_w_variant_location$ense_id %>%
-                    na.omit()
-
-locat2protdf <- list2DF(locat2prot)
+locat2protdf <- annotated_w_variant_location %>%
+  separate_rows(variant_location)
 
 long2prot <- locat2protdf %>%
-          pivot_longer(everything(),
-                       values_to = "variant_location",
-                       names_to = "ense_id") %>% 
-          distinct() %>%
           mutate(variant_num_location = readr::parse_number(variant_location))
@@
                                dplyr::select(annotated_w_variant_location,
-                                             protein_id_var, ense_id)) %>%
+                                             protein_id_var, 
+                                             ense_id)) %>%
   distinct()
@@
 
-protein2varlocat <- left_join(annotated_wo_variant_location, long2prot_worivar) %>%
+protein2varlocat <- left_join(annotated_wo_variant_location, 
+                              long2prot_worivar) %>%
           distinct() %>%
@@
                  end_num = as.numeric(end_position)) %>%
-          mutate(variant_peptide = variant_num_location >= as.numeric(start_position) & variant_num_location <= as.numeric(end_position)) %>% # evaluate if the falls on the position of a variant sequence
-          dplyr::select(protein_id_var, ense_id, protein_description, Peptide, start_position, 
-                        start_num,end_position,end_num, variant_location, 
-                        variant_num_location, variant_peptide, variant_type)
+  # evaluate if the SAAV falls on the position of a variant sequence
+          mutate(variant_peptide = variant_num_location >= as.numeric(start_position) & variant_num_location <= as.numeric(end_position)) %>% 
+          dplyr::select(protein_id_var, 
+                        ense_id, 
+                        protein_description, 
+                        Peptide, 
+                        start_position, 
+                        start_num,
+                        end_position,
+                        end_num, 
+                        variant_location, 
+                        variant_num_location, 
+                        variant_peptide, 
+                        variant_type)
 
 pept2variant <- protein2varlocat %>%
-  dplyr::select(Peptide, variant_peptide, variant_type) %>%
+  dplyr::select(Peptide, 
+                variant_peptide, 
+                variant_type,
+                protein_id_var,
+                ense_id,
+                variant_location,
+                peptide_start_position = start_position,
+                peptide_end_position = end_position,
+                protein_description) %>%
   distinct() %>%
   na.omit() %>%
-  filter(variant_type == "saav")
-```
-
-```{r echo=TRUE}
-table(pept2variant$variant_peptide)
-```  
-232 peptides were identified as mapping to single amino acid variant regions of their corresponding annotated proteins sequence.
-
-## Visualize quantitative features of identified SAAVs
-
-### Prep abundance matrices
+  filter(variant_type == "saav") %>% 
+  mutate(duplicated_peptide = duplicated(Peptide)) %>% 
+  relocate(duplicated_peptide,
+           .after = variant_type)
+```
+
+## Generate table of identified peptides mapping to SAAVs
+
+```{r}
+saavs_table <- pept2variant %>%
+  filter(variant_peptide == TRUE) %>% 
+  mutate(duplicated_peptide = duplicated(Peptide)) %>% 
+  dplyr::rename(peptide = Peptide) %>%
+  left_join(., index2genepeptide) %>%
+  mutate(gene_variant = paste(gene, 
+                              variant_location, 
+                              sep = "_"))
+
+```
+
+```{r}
+length(unique(saavs_table$Peptide))
+```
+
+```{r}
+length(unique(saavs_table$protein_id_var))
+```
+
+```{r}
+saavs_suppl_table <- saavs_table %>% 
+  dplyr::select(gene_variant,
+                variant_type,
+                variant_location,
+                peptide,
+                peptide_start_position,
+                peptide_end_position,
+                protein_id_var,
+                ense_id,
+                gene,
+                protein_description)
+```
+
+```{r}
+write_tsv(x = saavs_suppl_table,
+          file = here("suppl_tables/supplementary_table_5_proteogenomics_identified_saavs.tsv"))
+```
+
+# Visualize quantitative features of identified SAAVs
+
+## Prep abundance matrices
+
+```{r}
+recurrencewo6 <- sample_annotationwo6$recurrence
+patientwo6 <- sample_annotationwo6$patient
+```
+
+```{r echo=FALSE}
+design_limmawo6 <- model.matrix(~patientwo6+recurrencewo6)
+
+sample_annotationwo6 <- sample_annotationwo6 %>%
+  mutate(paired_id_semi = paste(patient, 
+                                recurrence, 
+                                sep = ""))
+
+rownames(design_limmawo6) <- sample_annotationwo6$paired_id_semi
+```
 
@@
 pep_prematwo6 <- dplyr::select(expr_matrix_pept_progen,
-                               index, peptide, rownames(design_limmawo6))
+                               index, 
+                               peptide, 
+                               rownames(design_limmawo6))
 
 pep_matwo6 <- dplyr::select(expr_matrix_pept_progen,
-                             index, rownames(design_limmawo6))  %>%
+                            index, 
+                            rownames(design_limmawo6))  %>%
   column_to_rownames("index") %>%
@@
   na.omit()
-
-```
-
-### Filter, merge and summarize SAAV abundances
+```
+
+## Filter, merge and summarize SAAV abundances
 
@@
 index2genepeptide <- tmt_peptdata_frag_progen %>% 
-  dplyr::select(index, peptide, gene, protein_id)
+  dplyr::select(index, 
+                peptide, 
+                gene, 
+                protein_id)
 
@@
 
-# data frame of identified SAAVs and their annotation
+# data frame of identified and TMT-integrator summarized SAAVs + annotation
 protein2varlocat_saav <- protein2varlocat %>%
@@
   filter(peptide %in% interesting_features1$peptide,
-         variant_peptide == TRUE) %>%
+         variant_peptide == TRUE,
+         variant_type == "saav") %>%
   left_join(., index2genepeptide) %>%
-  mutate(gene_variant = paste(gene, variant_location, sep = "_"))
+  mutate(gene_variant = paste(gene, 
+                              variant_location, 
+                              sep = "_"))
 
@@
 genevariant2index <- protein2varlocat_saav %>%
-  dplyr::select(index, gene_variant)
+  dplyr::select(index, 
+                gene_variant)
 
@@
   left_join(., genevariant2index) %>%
-  relocate(gene_variant)
+  relocate(gene_variant) %>%
+  mutate(duplicated_variant = duplicated(gene_variant)) %>%
+  relocate(duplicated_variant,
+           .after = peptide)
 
@@
   group_by(gene_variant) %>%
-  summarise_at(vars(starts_with("x")), .funs = function(x){median(x, na.rm = TRUE)}) %>%
+  summarise_at(vars(starts_with("x")), 
+               .funs = function(x){median(x, 
+                                          na.rm = TRUE)}) %>%
   ungroup()
@@
 full_info_sumgenvar <- pep_premat_sumgenarwo6 %>%
-  left_join(., protein2varlocat_saav) %>%
+  left_join(., 
+            protein2varlocat_saav) %>%
   filter(variant_type == "saav")
-
-write_tsv(protein2varlocat_saav,
-          here("results/proteogenomics/cellreps_2021/all_peptides_mapping_to_saavs_gbm.tsv"))
-
-write_tsv(pep_premat_sumgenarwo6,
-          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs.tsv"))
-
-write_tsv(full_info_sumgenvar,
-          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs_full.tsv"))
 
@@
 pep_matvariantswo6 <- pep_premat_sumgenarwo6 %>%
-  filter(str_detect(gene_variant, "CAP1_", 
+  filter(str_detect(gene_variant, 
+                    "CAP1_", 
                     negate = TRUE)) %>%
@@
 
-The variant-containing peptides identified are able to identify `r nrow(pep_premat_sumgenarwo6)` single amino acid mutations (SAAVs).
+### Generate summarized results tables
+
+```{r}
+write_tsv(protein2varlocat_saav,
+          here("results/proteogenomics/cellreps_2021/all_peptides_mapping_to_saavs_gbm.tsv"))
+
+write_tsv(pep_premat_sumgenarwo6,
+          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs.tsv"))
+
+write_tsv(full_info_sumgenvar,
+          here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs_full.tsv"))
+```
+
+The variant-containing peptides identified and accurately quantified and summarized are able to account for `r nrow(pep_premat_sumgenarwo6)` single amino acid mutations (SAAVs).
 
@@
 
-### Heatmap of all SAAVs observed
+## Heatmap of all SAAVs observed
 
@@
 patientstage <- colnames(pep_matvariantswo6nona) %>%
-  str_replace(., pattern = "prim", 
+  str_replace(., 
+              pattern = "prim", 
               replacement = "Init") %>%
-  str_replace(., pattern = "rec", 
+  str_replace(., 
+              pattern = "rec", 
               replacement = "Rec")
@@
 ```{r fig.height=20, fig.width=7}
-col_fun_or = colorRamp2(c(min(pep_matvariantswo6or_hm, na.rm = TRUE), 
-                       median(pep_matvariantswo6or_hm, na.rm = TRUE), 
-                       max(pep_matvariantswo6or_hm, na.rm = TRUE)), 
-                     c("#2a9d8f", "white", "red"))
+col_fun_or = colorRamp2(c(min(pep_matvariantswo6or_hm, 
+                              na.rm = TRUE), 
+                       median(pep_matvariantswo6or_hm, 
+                              na.rm = TRUE), 
+                       max(pep_matvariantswo6or_hm, 
+                           na.rm = TRUE)), 
+                     c("#2a9d8f", 
+                       "white", 
+                       "red"))
 
@@
         column_split = stage, 
+        row_names_gp = gpar(fontsize = 5),
+        column_names_gp = gpar(fontsize = 5),
+        cluster_rows = FALSE, cluster_columns = FALSE,
+        heatmap_legend_param = list(
+        title = "log2-\nNormalized\nAbundance"
+    ))
+```
+
+All identified SAAVs
+
+```{r fig.height=20, fig.width=7}
+col_fun_or = colorRamp2(c(min(pep_matvariantswo6or_hm, 
+                              na.rm = TRUE), 
+                       median(pep_matvariantswo6or_hm, 
+                              na.rm = TRUE), 
+                       max(pep_matvariantswo6or_hm, 
+                           na.rm = TRUE)), 
+                     c("#2a9d8f", 
+                       "white", 
+                       "red"))
+
+Heatmap(pep_matvariantswo6or_hm, 
+        col = col_fun_or,
+        column_split = patient, 
         row_names_gp = gpar(fontsize = 5.5),
@@
 
-### Heatmap of SAAVs observed on every sample
+## Heatmap of SAAVs observed on every sample
 
 ```{r fig.height=7.5, fig.width=7}
-col_fun_sel = colorRamp2(c(min(pep_matvariantswo6or_hm, na.rm = TRUE), 
-                       median(pep_matvariantswo6or_hm, na.rm = TRUE), 
-                       max(pep_matvariantswo6or_hm, na.rm = TRUE)), 
-                     c("#2a9d8f", "white", "red"))
+col_fun_sel = colorRamp2(c(min(pep_matvariantswo6or_hm, 
+                               na.rm = TRUE), 
+                       median(pep_matvariantswo6or_hm, 
+                              na.rm = TRUE), 
+                       max(pep_matvariantswo6or_hm, 
+                           na.rm = TRUE)), 
+                     c("#2a9d8f", 
+                       "white", 
+                       "red"))
 
@@
 
-### Limma on SAAVs
-
-### Prep design matrix
+# Limma on SAAVs
+
+## Prep design matrix
 
@@
 sample_annotationwo6 <- sample_annotationwo6 %>%
-  mutate(paired_id_semi = paste(patient, recurrence, sep = ""))
+  mutate(paired_id_semi = paste(patient, 
+                                recurrence, 
+                                sep = ""))
 
@@
 
-### Prep abundance matrices
+## Prep abundance matrices {.unnumbered .unlisted}
 
@@
 
-#### SAAV-level model fit
-
-```{r}
-fit_limmapeptwo6 <- function(mat, design, method,Limma){
-  limmafit <- lmFit(mat, design, method = method)
+## SAAV-level model fit {.unnumbered .unlisted}
+
+```{r echo=TRUE, warning=FALSE, message=FALSE}
+limmafit <- lmFit(pep_matvariantswo6nona, 
+                  design = design_limmawo6_saav, 
+                  method = 'robust')
+
 limmafit <- eBayes(limmafit)
-  limma_tab <- topTable(limmafit, coef = "stageRecurrent", number = Inf, adjust.method = "BH") %>%
+
+fit_mod_saavs <- topTable(limmafit, 
+                        coef = "stageRecurrent", 
+                        number = Inf, 
+                        adjust.method = "BH") %>%
   mutate(Protein = rownames(.),
-         Limma = Limma)
-
-  return(limma_tab)
-}
-```
-
-```{r echo=TRUE, warning=FALSE, message=FALSE}
-fit_mod_saavs <- fit_limmapeptwo6(pep_matvariantswo6nona, 
-                                  design_limmawo6_saav, 
-                                  method = 'robust', 
          Limma = "Robust - w Patient effect")
@@
 increased_recwo6_saavs <- compar_tab_saavs %>%
-                    filter(logFC>0,
+                    filter(logFC > 0,
                            adj.P.Val < 0.05) %>% 
@@
 decreased_recwo6_saavs <- compar_tab_saavs %>%
-                    filter(logFC<0,
+                    filter(logFC < 0,
                            adj.P.Val < 0.05) %>% 
@@
 ```
+
 ```{r}
@@
 
-#### Volcano SAAVs
-
-```{r fig.width=8, fig.height=6}
+Only 1 SAAV is observed as differentially abundant between Recurrent and Initial tumors.
+
+## Volcano SAAVs
+
+```{r}
 diff_saavs <- compar_tab_saavs %>%
@@
 volcano_saavs <- ggplot(data = compar_tab_saavs,
-                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
-      geom_point(data = compar_tab_saavs %>% filter(logFC > 0,
+                      mapping = aes(x = logFC, 
+                                    y = -log10(adj.P.Val))) +
+      geom_point(data = compar_tab_saavs %>% 
+                   filter(logFC > 0,
                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
+                 mapping = aes(x = logFC, 
+                               y = -log10(adj.P.Val)), 
                  color = "red",
                  size = size)+
-      geom_point(data = compar_tab_saavs %>% filter(logFC < -0,
+      geom_point(data = compar_tab_saavs %>% 
+                   filter(logFC < -0,
                           adj.P.Val < 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "red",
+                 mapping = aes(x = logFC, 
+                               y = -log10(adj.P.Val)), 
+                 color = "red",
                  size = size) +
-      geom_point(data = compar_tab_saavs %>% filter(logFC > 0,
+      geom_point(data = compar_tab_saavs %>% 
+                   filter(logFC > 0,
                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
+                 mapping = aes(x = logFC, 
+                               y = -log10(adj.P.Val)), 
+                 color = "#2a9d8f",
                  size = size) +
-      geom_point(data = compar_tab_saavs %>% filter(logFC < -0,
+      geom_point(data = compar_tab_saavs %>% 
+                   filter(logFC < -0,
                           adj.P.Val > 0.05),
-                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
+                 mapping = aes(x = logFC, 
+                               y = -log10(adj.P.Val)), 
+                 color = "#2a9d8f",
                  size = size) +
       geom_hline(yintercept = -log10(0.05),
-                 color = "red", linetype = "dashed") +
+                 color = "red", 
+                 linetype = "dashed") +
       ggrepel::geom_text_repel(data = diff_saavs,
-                               aes(label = gene_variant), size = 4) +
+                               aes(label = gene_variant), 
+                               size = 4) +
       xlab("logFC - Recurrent / Primary") + 
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
+      theme(axis.text.x = element_text(hjust = 0.5, 
+                                       vjust = 0, 
+                                       size = 6, 
+                                       angle = 360),
+            axis.text.y = element_text(hjust = 0.95, 
+                                       vjust = 0.2, 
+                                       size = 8),
             panel.background = element_blank(),
             panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
+            panel.border = element_rect(colour = "black", 
+                                        fill = NA, 
+                                        size = 0.5),
             axis.title = element_text(size = 8),
@@
             legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom") 
+            legend.key.height = unit(3, 
+                                    'mm'),
+            legend.key.width = unit(3, 
+                                   'mm'),
+            legend.position = "bottom") 
+```
+
+```{r fig.width=8, fig.height=6}
+volcano_saavs
 ```
@@
 
-### Boxplot of differentially abundant SAAV
+## Boxplot of differentially abundant SAAV
 
@@
                          values_to = "Abundance") %>%
-  mutate(Stage = str_extract(patient, "[^0-9]+$")) %>%
-  mutate(Patient = str_remove_all(patient, "[^0-9]+$")) %>%
+  mutate(Stage = str_extract(patient, 
+                             "[^0-9]+$")) %>%
+  mutate(Patient = str_remove_all(patient, 
+                                  "[^0-9]+$")) %>%
   filter(gene_variant == increased_recwo6_saavs)
@@
                     aes(x = Stage, 
-                        y = Abundance, fill = Stage, 
+                        y = Abundance, 
+                        fill = Stage, 
                         cex.axis = 1.5)) +
   geom_boxplot() +
-  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.2) +
+  geom_dotplot(binaxis = "y", 
+               stackdir = "center", 
+               dotsize = 0.2) +
   # Box plot with jittered points
@@
   geom_signif(
-    comparisons = list(c("Initial", "Recurrence")),
+    comparisons = list(c("Initial", 
+                         "Recurrence")),
     map_signif_level = TRUE
@@
   ggtitle(increased_recwo6_saavs) +
-  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
-        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
+  theme(axis.text.x = element_text(hjust = 0.5, 
+                                   vjust = 0, 
+                                   size = 6, 
+                                   angle = 360),
+        axis.text.y = element_text(hjust = 0.95, 
+                                   vjust = 0.2, 
+                                   size = 8),
         panel.background = element_blank(),
         panel.grid.major = element_blank(),
-        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
+        panel.border = element_rect(colour = "black", 
+                                    fill = NA, 
+                                    size = 0.5),
         axis.title = element_text(size = 8),
@@
         legend.title = element_text(size = 8),
-        legend.key.height= unit(3, 'mm'),
-        legend.key.width= unit(3, 'mm'),
-        legend.position="bottom")
-
-```
-
-### Boxplot DA SAAV
+        legend.key.height = unit(3, 
+                                'mm'),
+        legend.key.width = unit(3, 
+                               'mm'),
+        legend.position = "bottom")
+
+```
+
+## Boxplot DA SAAV
 
@@
                           values_from = Abundance, 
-                          names_from = Stage) #%>%
-  #dplyr::select(-patient)
+                          names_from = Stage)
 
@@
     mutate(rank = row_number(Abs_diff)) %>% 
-    dplyr::rename(Initial = prim, Recurrent = rec) %>%
+    dplyr::rename(Initial = prim, 
+                  Recurrent = rec) %>%
     ungroup()
@@
          label = NULL,
-         repel = TRUE, point.size = 0.2, line.size = 0.1) +
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
-            axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
+         repel = TRUE, 
+         point.size = 0.2, 
+         line.size = 0.1) +
+      theme(axis.text.x = element_text(hjust = 0.5, 
+                                       vjust = 0,
+                                       size = 5, 
+                                       angle = 360),
+            axis.text.y = element_text(hjust = 0.95, 
+                                       vjust = 0.2, 
+                                       size = 5),
             panel.background = element_blank(),
             panel.grid.major = element_blank(),
-            panel.border = element_rect(colour = "black", fill=NA, size=0.5),
+            panel.border = element_rect(colour = "black", 
+                                        fill = NA, 
+                                        size = 0.5),
             axis.title = element_text(size = 8),
@@
             legend.title = element_text(size = 8),
-            legend.key.height= unit(3, 'mm'),
-            legend.key.width= unit(3, 'mm'),
-            legend.position="bottom",
-            strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) + 
+            legend.key.height= unit(3, 
+                                    'mm'),
+            legend.key.width= unit(3, 
+                                   'mm'),
+            legend.position = "bottom",
+            strip.text = element_text(hjust = 0.5, 
+                                      vjust = 0, 
+                                      size = 8, 
+                                      angle = 360)) + 
   ggtitle(increased_recwo6_saavs)
@@
          label = NULL,
-         repel = TRUE, point.size = 0.2, line.size = 0.1) +
-      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 5, angle = 360),
-            axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 5),
+         repel = TRUE, 
+         point.size = 0.2, 
+         line.size = 0.1) +
+      theme(axis.text.x = element_text(hjust = 0.5, 
+                                       vjust = 0, 
+                                       size = 5, 
+                                       angle = 360),
+            axis.text.y = element_text(hjust = 0.95, 
+                                       vjust = 0.2, 
+                                       size = 5),
             panel.background = element_blank(),
             panel.grid.major = element_blank(),
-            panel.border = element_rect(colour = "black", fill=NA, size=0.5),
+            panel.border = element_rect(colour = "black", 
+                                        fill = NA, 
+                                        size = 0.5),
             axis.title = element_text(size = 8),
@@
             legend.title = element_text(size = 8),
-            legend.key.height= unit(3, 'mm'),
-            legend.key.width= unit(3, 'mm'),
-            legend.position="bottom",
-            strip.text = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360)) 
+            legend.key.height = unit(3, 
+                                    'mm'),
+            legend.key.width = unit(3, 
+                                   'mm'),
+            legend.position = "bottom",
+            strip.text = element_text(hjust = 0.5, 
+                                      vjust = 0, 
+                                      size = 8, 
+                                      angle = 360)) 
 ```
@@
 ```
-
-```{r}
-sessionInfo()
-```

### 2023-02-16 reports/gbm_recurrence_lipidomics_analysis.qmd
lipidomics_cleanup

---
title: "GBM recurrence - Lipidomics analysis"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r knirt_helper_funcs}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))
```

```{r load_packages}
## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(DT)
library(here)
library(janitor)
library(ggpubr)
library(ggrepel)
library(readxl)
library(extrafont)
library(clusterProfiler)
```

```{r theme_set}
theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 
                                         'mm'),
                legend.key.width = unit(3, 
                                        'mm'),
                legend.position = "bottom"))
```

# Initial data loading and wrangling

```{r}
# load raw data
lipidomics_raw <- readxl::read_excel(path = here("data/lipidomics/2022-07-26lipids_merged_foell_laloti.xlsx")) %>%
  clean_names()

sample_annotation <- read_csv(here("data/sample_annotation.csv"))

# correct annotation

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",
                         patient,
                         sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_")) %>%
  filter(recurrence %in% c("prim", 
                           "rec"))
```

```{r}
lipidomics_sample_prep <- readxl::read_excel(path = here("data/lipidomics/Lipidomics_sample_preparation_reformatted.xlsx")) %>%
  clean_names() %>%
  dplyr::rename(sample_name = patient) %>%
  mutate(patient = str_trim(str_remove(sample_name,
                              "(1°)|(R)")),
         condition = if_else(condition = str_detect(string = sample_name,
                                                    pattern = "R"), 
                             true = "recurrence",
                             false = "primary")) %>%
  mutate(name = paste0(condition, patient))
```

Selecting interesting columns from lipidomics data and prepare abundance matrix.

```{r}
lipid_1 <- lipidomics_raw %>%
  dplyr::select(m_meas, name, rt_min, 
                matches("primary"),
                matches("recurrence"))

lipid_pre_mat <- lipid_1 %>% 
  mutate(name = str_replace_all(string = name, 
                               pattern = "\\_",
                               replacement = " ")) %>%
   mutate(feature = paste(name, m_meas, rt_min,sep = "_")) %>%
   dplyr::select(-c(name, m_meas, rt_min)) %>%
   relocate(feature)

column_names <- colnames(lipid_pre_mat) %>%
  str_remove(., pattern = "_.*") %>% 
  str_extract(., pattern = "feature|primary.*|recurrence.*")

colnames(lipid_pre_mat) <- column_names
```

Create matrix

```{r}
# create matrix 
lipid_mat <- lipid_pre_mat %>%
  column_to_rownames("feature") %>%
  as.matrix()
```

Substitute zero values for `NA`s

```{r}
lipid_mat[lipid_mat == 0] <- NA
```

Log2-transformation and scaling

```{r}
lipid_log2_mat <- mutate_all(as.data.frame(lipid_mat), 
                             log2)

lipid_scaled_mat <- scale(lipid_log2_mat,
                     scale = F,
                     center = apply(lipid_log2_mat, 2, median, 
                                    na.rm = TRUE) - median(as.matrix(lipid_log2_mat), 
                                                           na.rm = TRUE))
```

Prep long matrix for visualizations

```{r}
lipid_long <- lipid_scaled_mat %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(cols = where(is.numeric), 
               names_to = "patient",
               values_to = "Abundance") %>%
  separate(col = feature, 
           into = c("name", "m_z", "rt"),
           remove = FALSE,
           sep = "_")
```

```{r}
lipid_long_annt <- lipid_long %>%
  dplyr::select(-c(patient, Abundance)) %>%
  distinct() %>%
  separate(col = name,
           into = c("identity", "definition"),
           sep = "\\;",
           remove = FALSE) %>%
  mutate(definition = case_when(str_detect(string = identity,
                                           pattern = "\\:") ~ definition,
                                TRUE ~ identity)) %>%
  mutate(identity = ifelse(str_detect(string = identity,
                                        pattern = "\\:",
                                        negate = TRUE),
                            yes = NA,
                            no = identity))

lipid_annotation <- lipid_long_annt %>%
  separate(col = identity,
           into = c("abbreviation", "composition"),
           sep = " ", 
           remove = FALSE,
           extra = "merge") %>%
  separate(col = composition,
           into = c("pre_fatty_acid_length", "n_doub_bonds"),
           sep = "\\:",
           remove = FALSE) %>%
  mutate(pre_fatty_acid_length = str_remove_all(string = pre_fatty_acid_length,
                                                pattern = "O-")) %>%
  mutate(main_fat_acd_length_num = parse_integer(pre_fatty_acid_length)) %>%
  mutate(main_fat_acd_length_fct = as.factor(main_fat_acd_length_num)) %>%
  mutate(definition = ifelse(test = is.na(identity),
                             yes = name,
                             no = abbreviation),
         definition2 = ifelse(test = is.na(identity),
                              yes = name,
                              no = identity))
```

```{r}
lipid_annotation2 <- lipid_annotation %>%
  mutate(name_lipid1 = case_when(str_detect(string = name,
                                            pattern = "; Dodeca") ~ str_remove_all(name, "; Dodecanoylcarnitine"),
                                 str_detect(string = name,
                                            pattern = "; Myra") ~ str_remove_all(name, "; Myristoylcarnitine"),
                                 str_detect(string = name,
                                            pattern = "; Palmit") ~ str_remove_all(name, "; Palmitoyl-L-carnitine"),
                                 str_detect(string = name,
                                            pattern = ";O2") ~ str_replace_all(name, ";O2", ""),
                                 str_detect(string = name,
                                            pattern = ";2O/") ~ str_replace_all(name, ";2O/", " "),
                                 str_detect(string = name,
                                            pattern = ";2O$") ~ str_remove_all(name, ";2O.*"),
                                 str_detect(string = name,
                                            pattern = ";O2$") ~ str_remove_all(name, ";O2.*"),
                                 str_detect(string = name,
                                            pattern = ";3O$") ~ str_remove_all(name, ";3O.*"),
                                 str_detect(string = name,
                                            pattern = "/N-") ~ str_replace_all(name, "/N-", " "),
                                 str_detect(string = name,
                                            pattern = "-SN1") ~ str_replace_all(name, "-SN1", ""),
                                 str_detect(string = name,
                                            pattern = "LPC O-") ~ str_replace_all(name, "O-", ""),
                                 str_detect(string = name,
                                            pattern = "PC O-") ~ str_replace_all(name, "O-", ""),
                                 str_detect(string = name,
                                            pattern = "LPE O-") ~ str_replace_all(name, "O-", ""),
                                 str_detect(string = name,
                                            pattern = "PE O-") ~ str_replace_all(name, "O-", ""),
                                 str_detect(string = name,
                                            pattern = "; PC") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = "; PE") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = "; PG") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = "; PI") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = "; PS") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = "; TG") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = ";O;S") ~ str_remove_all(name, ";.*"),
                                 str_detect(string = name,
                                            pattern = ";O$") ~ str_remove_all(name, ";O$"),
                                 TRUE ~ name)) %>%
  mutate(name_lipid3 = if_else(condition = str_detect(string = name_lipid1,
                                            pattern = "2OH"),
                               true = str_remove_all(name_lipid1, ";\\(.*"),
                               false = name_lipid1)) %>%
  dplyr::relocate(name_lipid1, name_lipid3, .before = identity)
```

```{r}
lipid_annotation3 <- lipid_annotation2 %>%
                    mutate(lipid = name)

lipid_long_annt2 <- left_join(lipid_long, lipid_annotation3)

```

## Boxplots after normalization

```{r}
boxplot(lipid_mat, 
        main = "Before scaling")
```

```{r}
boxplot(lipid_scaled_mat,
        main = "After scaling")
```

## Visualize missing values

```{r}
naniar::vis_miss(as.data.frame(lipid_scaled_mat),
                 cluster = TRUE, 
                 sort_miss = TRUE)
```

It doesn't seem to be an increased accumulation of missing values associated to a particular condition/stage.

# Data preparation for PCA and limma

I will run PCA for unsupervised analysis and `limma` for the differential abundance analysis.

**Prepare sample annotation table**

```{r}
sample_annotation <- tibble(name = colnames(lipid_mat),
                            condition = str_remove_all(string = colnames(lipid_mat),
                                                       pattern = "[:digit:]"),
                            patient = parse_number(colnames(lipid_mat)) %>%
                              as.factor())

sample_annotationlip <- sample_annotation %>%
  dplyr::rename(Sample = name)
```

**Define design matrix**

```{r}
patient_lip <- sample_annotation$patient

recurrence_lip <- sample_annotation$condition

design_lip <- model.matrix(~patient_lip+recurrence_lip)

rownames(design_lip) <- sample_annotation$name
```

`limma` does not allow for missing values... We need to apply sparsity reduction.

## Sparcity reduction for lipidomics data

```{r}
na_count <- group_by(lipid_long_annt2,
                     feature) %>%
    summarise(na_count = sum(is.na(Abundance)),
              total = n()) %>% 
    ungroup() %>% 
    mutate(NA_fraction = na_count/total)
```

Keep only features missing in less than 20% of the samples.

```{r}
# keep only features that are present in 80% of the samples
feature2include <- na_count %>%
  filter(NA_fraction < 0.2)

reduced_pre_mat_lip <- lipid_scaled_mat %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  filter(feature %in% feature2include$feature)

reduced_mat_lip <- reduced_pre_mat_lip %>%
  column_to_rownames("feature") %>%
  as.matrix()
```

## `missForest` imputation

```{r}
t_reduced_mat_lip <- t(reduced_mat_lip)

if(!file.exists( here("rds/missforest_imp_lipids.Rds"))){

  t_imp_lip <- missForest::missForest(t_reduced_mat_lip)

  write_rds(x = t_imp_lip,
           file = here("rds/missforest_imp_lipids.Rds"))

  imp_mat_lip <- t_imp_lip$ximp %>% t()

} else {

  t_imp_lip <- read_rds(file = here("rds/missforest_imp_lipids.Rds"))

  imp_mat_lip <- t_imp_lip$ximp %>% t()

}
```

## Save wide matrix

```{r}
imp_mat_lip[1:5, 1:5]
```

Prepare matrix for enrichment purposes

```{r}
df_imp_mat_lip <- imp_mat_lip %>% 
            as.data.frame() %>%
            rownames_to_column("feature") %>% 
  left_join(., lipid_annotation2) %>% 
  dplyr::select(name_lipid3, starts_with("prima"), starts_with("recurr"))
```

# Exploratory PCA

```{r}
pca_lips <- pca(t_imp_lip$ximp, 
                ncomp = 10, 
                center = TRUE, 
                scale = TRUE)
```

```{r fig.width=7, fig.height=5}
plot(pca_lips)
```

```{r}
# preprocess pca results 
pca_vari_lips <- pca_lips$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("name") %>%
  left_join(., 
            lipidomics_sample_prep,
            by = "name")

```

```{r fig.width=5, fig.height=4}
ggplot(data = pca_vari_lips,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = batch, 
                           color = condition)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 4)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of lipids") +
  xlab(paste("PC1", round(pca_vari_lips$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_vari_lips$prop_expl_var$X[2]*100), "% var explained")) 
```

```{r fig.width=5, fig.height=4}
ggplot(data = pca_vari_lips,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = condition, 
                           color = batch)) +  
  geom_text(aes(label = patient), 
            position = position_nudge(x = 4)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of lipids") +
  xlab(paste("PC1", round(pca_vari_lips$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_vari_lips$prop_expl_var$X[2]*100), "% var explained")) 
```

There is no batch effect associated to the sample prep day.

# `limma`

```{r}
limma_lip <- lmFit(object = imp_mat_lip, 
                  design = design_lip, 
                  method = "robust")

limma_lip <- eBayes(limma_lip)

limma_tab_lip <- topTable(limma_lip, 
                          coef = "recurrence_liprecurrence", 
                          number = Inf, 
                          adjust.method = "BH")

limma_tab_annotated <- limma_tab_lip %>%
  rownames_to_column("feature") %>%
  left_join(., lipid_annotation2)

ceramides_limma <- limma_tab_annotated %>%
  filter(str_detect(definition, "Cer"))

da_limma_lipids <- limma_tab_annotated %>% 
  filter(adj.P.Val < 0.05) %>%
  filter(!is.na(definition)) %>%
  filter(str_detect(definition, "NA", negate = TRUE))

defined_lipids <- lipid_annotation %>%
  filter(str_detect(definition, "NA", negate = TRUE),
         str_detect(feature, "zzz+",negate = TRUE))

def_imp_mat_lip <- imp_mat_lip %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  filter(feature %in% defined_lipids$feature) %>%
  column_to_rownames("feature") %>%
  as.matrix()
```

## Get tabular results of increased/decreased lipids

```{r}
increased_lipids <- limma_tab_annotated %>%
  filter(logFC > 0,
         adj.P.Val < 0.05) %>%
  filter(str_detect(name_lipid3, "NA", negate = TRUE))

decreased_lipids <- limma_tab_annotated %>%
  filter(logFC < 0,
         adj.P.Val < 0.05)  %>%
  filter(str_detect(name_lipid3, "NA", negate = TRUE))

background_lipids <- limma_tab_annotated %>%
  filter(str_detect(name_lipid3, "NA", negate = TRUE))

```

```{r}
write_csv(x = increased_lipids,
          file = here("data/lipidomics/increased_lipids.csv"))

write_csv(x = decreased_lipids,
          file = here("data/lipidomics/decreased_lipids.csv"))

write_csv(x = background_lipids,
          file = here("data/lipidomics/background_lipids.csv"))
```

## Volcano lipidomics

```{r}
size <- 1.3

volcano_lipids <- ggplot(data = limma_tab_annotated,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(data = limma_tab_annotated %>% filter(adj.P.Val < 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "#D9C5B2",
                 size = size) +
      geom_point(data = limma_tab_annotated %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_annotated %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), color = "#2a9d8f",
                 size = size) +
      geom_point(data = ceramides_limma,
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "red",
                 size = size) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Initial")+
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

```{r fig.height=5, fig.width=5}
volcano_lipids
```

```{r echo=FALSE}
ggsave(plot = volcano_lipids, 
       filename = here::here("figures/volcano_lipidomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = volcano_lipids, 
       filename = here::here("figures/volcano_lipidomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)

write_tsv(x = limma_tab_annotated,
          file = here("suppl_tables/ipidomics_limma_results.tsv"))
```

```{r}
inter_lip <- limma_tab_annotated %>%
  filter(name != "NA", 
         adj.P.Val < 0.05) 
```

## Volcano Ceramides

```{r}
volcano_lipids
```

## Volcano all identities

```{r}
      ggplot(data = limma_tab_annotated,
                      mapping = aes(x = logFC, y = -log10(adj.P.Val))) + 
      geom_point(data = limma_tab_annotated %>% filter(adj.P.Val < 0.05),
                     mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                     color = "#D9C5B2",
                     size = size) +
      geom_point(data = limma_tab_annotated %>% filter(logFC > 0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "#2a9d8f",
                 size = size) +
      geom_point(data = limma_tab_annotated %>% filter(logFC < -0,
                                           adj.P.Val > 0.05),
                 mapping = aes(x = logFC, y = -log10(adj.P.Val)), 
                 color = "#2a9d8f",
                 size = size) +
      geom_point(data = da_limma_lipids,
                 mapping = aes(x = logFC, y = -log10(adj.P.Val),
                               color = definition), 
                 size = size) +
      geom_hline(yintercept = -log10(0.05),
                 color = "red", linetype = "dashed") +
      xlab("logFC - Recurrent / Primary") +
      theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 8),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

## Boxplots per type of lipid

```{r}
lipid_long_annt3 <- lipid_long_annt2 %>% 
  filter(!is.na(definition)) %>%
  filter(str_detect(definition, "NA", negate = TRUE)) %>%
  mutate(stage = str_remove_all(string = patient,
                                pattern = "[:digit:]"),
         patient = parse_number(patient)) %>%
  mutate(stage = if_else(stage == "primary",
                         true = "primary",
                         false = "recurrence")) %>%
  pivot_wider(names_from = stage,
              values_from = Abundance)

```

```{r}
summarized_lipids <- lipid_long_annt3 %>%
  dplyr::select(feature, definition, primary, recurrence, patient) %>%
  filter(feature %in% feature2include$feature) %>%
  mutate(definition = str_remove(definition, "(.*);")) %>%
  mutate(definition = if_else(definition == "SPHINGOSYLPHOSPHORYL CHOLINE",
                              true = str_to_sentence(definition),
                              false = definition)) %>%
  group_by(definition, patient) %>%
  summarise(Initial = median(primary, 
                             na.rm = TRUE),
            Recurrent = median(recurrence, 
                                na.rm = TRUE)) %>%
  filter(str_detect(definition, "zzz", negate = TRUE)) 
```

```{r fig.width=8, fig.height=11}
ggpaired(summarized_lipids,
         cond1 = "Initial",
         cond2 = "Recurrent", 
         fill = "condition",
         ylab = "Scaled Abundance",
         label = NULL,
         repel = TRUE,
         facet.by = "definition", point.size = 0.2, line.size = 0.1) + 
  coord_cartesian(ylim = c(6,20)) +
      stat_compare_means(method = "t.test", 
                         paired = TRUE, 
                         label.y = 6.5, 
                         size = 3) 
```

# mixOmics PLS

We want to explore correlations between proteomics expression data and lipidomics, using projection of latent structure (PLS) as implemented in the mixOmics R package.

## Required data

**Proteomics data**

We'll use the proteomics data after filtering for proteins found in 2/3 mixtures, `missForest` imputation and exclusion of 'patient 6 (proteomics)'.

```{r}
identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
  janitor::clean_names()

prot2gene <- identified_proteins %>%
  dplyr::select(Protein = protein_id, Gene = gene,
                Description = description)
```

```{r}
## tmt-integrator output loading 
tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

expr_matrix <- dplyr::select(tmt_protdata_frag,
                             Protein = index, starts_with("x"))

## Get predictors for mixtures
expr_matrix2 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotation2$paired_id)

long_mat <- pivot_longer(expr_matrix,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", 
                                      "recurrence"),
                         values_to = "Abundance")

long_mat_2 <- long_mat %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_"))

## Annotated abundance data in long format ----

quant_annotated <- left_join(long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id", 
                                    "patient",
                                    "recurrence")) %>%
  mutate(Channel_mix = paste(mixture, 
                             channel,  
                             sep = "_"))

proteins2exclude <- sel_proteins_missing(quant_annotated,
                                         threshold = 1)

expr_matrix_filt <- expr_matrix2 %>%
  filter(!Protein %in% proteins2exclude)
```

```{r}
mat_filt <- expr_matrix_filt %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt <- t(mat_filt)

if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){

 t_imp_mat_filt <- missForest::missForest(t_mat_filt)
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

 saveRDS(t_imp_mat_filt,
         file = here("rds/missforest_imp_mat_filt.Rds"))

} else {

 t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

}

```

```{r}
mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

sample_annotationwo6 <- sample_annotation2 %>%
  filter(patient != "x6")

df_mat_filt_impwo6 <- mat_filt_impwo6 %>%
  as.data.frame() %>%
  rownames_to_column("Protein") %>%
  left_join(.,prot2gene) %>%
  relocate(Gene)

mat_filt_imp_gene <- df_mat_filt_impwo6 %>%
  dplyr::select(-c(Gene, Description)) %>%
  dplyr::relocate(ends_with("prim")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt_imp_gene_prot <- t(mat_filt_imp_gene)
```

```{r}
t_mat_filt_imp_gene_prot[1:5,1:5]
```

```{r}
dim(t_mat_filt_imp_gene_prot)
```

**Lipidomics data**

Only lipidomics features with an associated definition were used for the sPLS analysis.

```{r}
t_def_imp_mat_lip <- t(def_imp_mat_lip)
```

```{r}
t_def_imp_mat_lip[1:3,25:28]
```

```{r}
dim(def_imp_mat_lip)
```

# mixOmics PLS based on FCs between Proteomics and Lipidomics

## Create matrix of FCs per **protein** and patient

```{r}
# uniprot column names
pre_manual_fcs_prots <- df_mat_filt_impwo6 %>%
  dplyr::select(-Gene) %>% 
  pivot_longer(cols = where(is.numeric),
               names_to = "sample_id",
               values_to = "Abundance") %>%
  separate(col = sample_id,
           sep = "\\_", 
           into = c("patient_n", "stage"))

manual_fcs_prots <- pre_manual_fcs_prots %>%
  pivot_wider(names_from = stage,
              values_from = Abundance) %>%
  mutate(prot_pat_fc = rec - prim)

df_mat_fcs_prots <- manual_fcs_prots %>% 
  dplyr::select(Protein, patient_n,  prot_pat_fc) %>%
  pivot_wider(names_from = patient_n,
              values_from = prot_pat_fc)

mat_fcs_prots <- df_mat_fcs_prots %>%
  column_to_rownames("Protein") %>%
  as.matrix()

# gene symbol names
pre_manual_fcs_symb <- df_mat_filt_impwo6 %>%
  dplyr::select(-Protein) %>% 
  pivot_longer(cols = where(is.numeric),
               names_to = "sample_id",
               values_to = "Abundance") %>%
  separate(col = sample_id,
           sep = "\\_", 
           into = c("patient_n", "stage"))

manual_fcs_symb <- pre_manual_fcs_symb %>%
  pivot_wider(names_from = stage,
              values_from = Abundance) %>%
  mutate(prot_pat_fc = rec - prim)

df_mat_fcs_symb <- manual_fcs_symb %>% 
  dplyr::select(Gene, patient_n,  prot_pat_fc) %>%
  pivot_wider(names_from = patient_n,
              values_from = prot_pat_fc)

mat_fcs_symb <- df_mat_fcs_symb %>%
  column_to_rownames("Gene") %>%
  as.matrix()
```

```{r}
mat_fcs_prots[1:5, 1:5]
```

### Create matrix of FCs per **lipid feature** and patient

```{r}
pre_manual_fcs_lip <- def_imp_mat_lip %>%
  as.data.frame() %>%
  rownames_to_column("lipid_feature") %>%
  pivot_longer(cols = where(is.numeric),
               names_to = "sample_id",
               values_to = "Abundance") %>%
  mutate(patient_n = paste0("x",parse_number(sample_id)),
         stage = if_else(condition = str_detect(string = sample_id, 
                                                pattern = "primary"),
                         true = "prim",
                         false = "rec"))

manual_fcs_lip <- pre_manual_fcs_lip %>%
  dplyr::select(-sample_id) %>%
  pivot_wider(names_from = stage,
              values_from = Abundance) %>%
  mutate(lip_pat_fc = rec - prim)

df_mat_fcs_lip <- manual_fcs_lip %>% 
  dplyr::select(lipid_feature, patient_n,  lip_pat_fc) %>%
  pivot_wider(names_from = patient_n,
              values_from = lip_pat_fc)

mat_fcs_lip <- df_mat_fcs_lip %>%
  column_to_rownames("lipid_feature") %>%
  as.matrix()
```

```{r}
mat_fcs_lip[1:5, 1:3]
```

### Executing PLS with tunning (FC)

```{r}
t_mat_fcs_lip <- t(mat_fcs_lip)
t_mat_fcs_prots <- t(mat_fcs_symb)
```

```{r}
pls_res_fcs <- pls(X = t_mat_fcs_lip, 
                   Y = t_mat_fcs_prots, 
                   mode = "regression", 
                   ncomp = 5)
```

Performance measure:

```{r}
if(!file.exists(here("rds/performance_pls_mixomics_fcs_lip_prot.Rds"))){

   perf_fcs_pls <- perf(pls_res_fcs, 
                    folds = 5,
                    progressBar = TRUE, 
                    nrepeat = 10)

   write_rds(x = perf_fcs_pls,
             file = here("rds/performance_pls_mixomics_fcs_lip_prot.Rds"))

} else {

  perf_fcs_pls <- read_rds(here("rds/performance_pls_mixomics_fcs_lip_prot.Rds"))

}
```

```{r}
plot(perf_fcs_pls, criterion = 'Q2.total')
```

The Q2 criterion indicates that a component should be included in the model if the Q2 total \<= 0.0975... In this case, all components are smaller than this. This could be related to over-fitting or due to the inclusion of too many variables in the model (see [discussion in mixOmics forum](https://mixomics-users.discourse.group/t/q2-total-negative-in-perf-pls/138)).

We would work with 3 components, and keep 25 top proteins per 10 top lipids per component. This selection is arbitrary but was rationalized based on model interpretability.

```{r}
pls_res_fcs_tuned <- spls(X = t_mat_fcs_prots, 
                          Y = t_mat_fcs_lip, 
                          mode = "regression", 
                          ncomp = 4,
                          keepX = c(25, 25, 25, 25),
                          keepY = c(15, 15, 15, 15))
```

#### Circle plot

```{r}
plotVar(pls_res_fcs_tuned,
        cex = c(4,4),
        legend = TRUE)
```

```{r}
coordinatesfcs <- plotVar(pls_res_fcs_tuned, plot = FALSE)

coord2_fcs_circlgg <- coordinatesfcs %>% 
  mutate(`Component 1` = x,
         `Component 2` = y,
         Source = if_else(condition = Block == "X",
                          true = "Proteomics",
                          false = "Lipidomics")) %>% 
  as_tibble() %>% 
  dplyr::select(`Component 1`, 
                `Component 2`,
                Source,
                names)  %>%
  mutate(association = case_when(`Component 1` > 0.5 & Source == "Proteomics" ~ "Negative_association_LPCs",
                                 `Component 1` < -0.5 & Source == "Proteomics" ~ "Positive_association_LPCs",
                                 `Component 1` < -0.5 & Source == "Lipidomics" ~ "Positive_association_Proteins1", 
                                 `Component 2` < -0.5 & Source == "Lipidomics" ~ "Positive_association_Proteins2"),
         groups = case_when(`Component 1` < -0.5 & Source == "Proteomics" ~ "prot_g1",
                            `Component 2` < -0.5 & Source == "Proteomics" ~ "prot_g2",
                            `Component 2` > 0.5 & Source == "Proteomics" ~ "prot_g3",
                             `Component 1` > 0.5 & Source == "Proteomics" ~ "prot_g4",
                            `Component 1` < -0.5 & Source == "Lipidomics" ~ "lip_g1",
                            `Component 2` < -0.5 & Source == "Lipidomics" ~ "lip_g2"),
         names = str_remove(string = names, pattern = "_.*"))

proteomics_pls <- coord2_fcs_circlgg %>%
  filter(Source == "Proteomics")

```

```{r}
elipsize <- 0.3

circle_fcs_lipids <- ggplot(data = coord2_fcs_circlgg,
                  aes(x = `Component 1`, 
                      y = `Component 2`,
                      color = Source)) +
             geom_point(size = 0.8) +  
             geom_text_repel(aes(label = names), 
                             size = 0.8, 
                             max.overlaps = 30,
                             segment.size = 0.08) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                            filter(groups == "prot_g1"),
                          color = "#425F57",
                          linetype = "dashed",
                          size = elipsize) +
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "prot_g2"),
                          color = "#367E18",
                          linetype = "dashed",
                          size = elipsize) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "prot_g4"),
                          color = "#749F82",
                          linetype = "dashed",
                          size = elipsize) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "lip_g1"),
                          color = "#FF6D28",
                          size = elipsize) +
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "lip_g2"),
                          color = "#FF884B",
                          size = elipsize) + 
             scale_color_manual(values = c("Proteomics" = "#1A4D2E",
                                           "Lipidomics" = "#FF9F29")) + 
             theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
                   axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
                   panel.background = element_blank(),
                   panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                   axis.title = element_text(size = 10),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size = 8),
                   legend.key.height = unit(3, 'mm'),
                   legend.key.width = unit(3, 'mm'),
                   legend.position = "bottom")

```

```{r}
circle_fcs_lipids2 <- ggplot(data = coord2_fcs_circlgg,
                  aes(x = `Component 1`, 
                      y = `Component 2`,
                      color = Source)) +
             geom_point(size = 1.2) +  
             geom_text_repel(aes(label = names), 
                             size = 2.5, 
                             max.overlaps = 30, 
                             segment.size = 0.05) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                            filter(groups == "prot_g1"),
                          color = "#425F57",
                          linetype = "dashed",
                          size = 0.3) +
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "prot_g3"),
                          color = "#367E18",
                          linetype = "dashed",
                          size = 0.3) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "prot_g4"),
                          color = "#749F82",
                          linetype = "dashed",
                          size = 0.3) + 
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "lip_g1"),
                          color = "#FF6D28",
                          size = 0.3) +
             stat_ellipse(data = coord2_fcs_circlgg %>% 
                                 filter(groups == "lip_g2"),
                          color = "#FF884B",
                          size = 0.3) + 
             scale_color_manual(values = c("Proteomics" = "#1A4D2E",
                                           "Lipidomics" = "#FF9F29")) + 
             theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
                   axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
                   panel.background = element_blank(),
                   panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                   axis.title = element_text(size = 10),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size = 8),
                   legend.key.height = unit(3, 'mm'),
                   legend.key.width = unit(3, 'mm'),
                   legend.position = "bottom")
```

```{r fig.width=7, fig.height=7}
circle_fcs_lipids2
```

```{r}
ggsave(plot = circle_fcs_lipids, 
       filename = here::here("figures/spls_prot_fcslip_gbm2.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 63)

ggsave(plot = circle_fcs_lipids, 
       filename = here::here("figures/spls_prot_fcslip_gbm2.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 63)

write_tsv(x = coord2_fcs_circlgg,
          file = here::here("suppl_tables/proteomics_lipidomics_summary_spls_results.tsv"))
```

#### Enrichment analysis of proteomics signatures

```{r}
bitr_entrez_pls_fcs <- bitr(geneID = proteomics_pls$names, 
                            fromType = "SYMBOL", 
                            toType = "ENTREZID", 
                            OrgDb = "org.Hs.eg.db") %>% 
  dplyr::rename(names = SYMBOL)

proteomics_pls <- left_join(proteomics_pls, bitr_entrez_pls_fcs)

entrez_universe <- bitr(geneID = identified_proteins$gene %>% 
                                unique(), 
                            fromType = "SYMBOL", 
                            toType = "ENTREZID", 
                            OrgDb = "org.Hs.eg.db") 
```

```{r}
formula_pls_fcs_res <- compareCluster(names ~ groups, 
                              data = proteomics_pls, 
                              fun = "enrichGO",
                              OrgDb = "org.Hs.eg.db",
                              keyType = "SYMBOL",
                              ont = "MF",
                              pvalueCutoff = 0.05,
                              pAdjustMethod = "BH",
                              universe = identified_proteins$gene %>% 
                                unique(),
                              qvalueCutoff = 0.2,
                              minGSSize = 1,
                              maxGSSize = 500,
                              readable = TRUE,
                              pool = FALSE)

res_enrich_pls_fcs <- simplify(formula_pls_fcs_res)
```

```{r}
enrichplot::dotplot(res_enrich_pls_fcs)
```

```{r}
lipid_pls_enrich <- enrichplot::dotplot(res_enrich_pls_fcs) +
                    scale_color_continuous(low = "red", 
                                           high = "blue",
                                           guide = guide_colorbar(reverse=TRUE),
                                           breaks = c(0.02, 0.04)) +
                    scale_size_continuous(breaks = c(0.06, 0.15)) +
                    theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
                          axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
                          panel.background = element_blank(),
                          panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                          axis.title.x = element_blank(),
                          axis.title.y = element_blank(),
                          legend.text = element_text(size = 7),
                          legend.title = element_text(size = 8),
                          legend.key.height = unit(3, 'mm'),
                          legend.key.width = unit(3, 'mm'),
                          legend.position = "bottom",
                          text = element_text(family = "sans"))
```

```{r fig.width=6, fig.height=6}
lipid_pls_enrich
```

##### Visualize enrichment analysis of proteomics signatures

```{r}
pls_proteomics_feat_gomf_enrichment <- res_enrich_pls_fcs@compareClusterResult %>%
  mutate(`Group` = factor(Cluster,
                          levels = c("prot_g1", "prot_g2",
                                     "prot_g3", "prot_g4"))) %>%
    mutate(
      Description = if_else(
        condition = str_count(Description, '\\S+') > 3,
        true = str_c(word(Description,1,3), word(Description,4,str_count(Description,'\\S+')), sep = '\n'),
        false = Description
      )
    ) %>% 
  mutate(Description = factor(Description,
                              levels = rev(.$Description))) %>% 
  group_by(`Group`) %>%
  slice_min(order_by = p.adjust, 
            n = 5)

```

```{r}
ggplot(data = pls_proteomics_feat_gomf_enrichment,
       mapping = aes(y = -log10(p.adjust),
                     x = Description, 
                     fill = `Group`)) + 
  geom_col() + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "red") + 
  coord_flip() +  
  xlab(label = "-log10(Adjusted p-value") +
  ylab(label = "Enriched GO Cellular Compotent terms") +
   theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
         axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
         panel.background = element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=0.5),
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         legend.text = element_text(size = 7),
         legend.title = element_text(size = 8),
         legend.key.height= unit(3, 'mm'),
         legend.key.width= unit(3, 'mm'),
         legend.position="right",
         text = element_text(family = "sans"))

```

```{r}
barplot_enrich_prot_pls <- ggplot(data = pls_proteomics_feat_gomf_enrichment,
                                  mapping = aes(y = -log10(p.adjust),
                                                x = Description, 
                                                fill = `Group`)) + 
                           geom_col() + 
                           geom_hline(yintercept = -log10(0.05),
                                      linetype = "dashed",
                                      color = "red",
                                      size = 0.25) + 
                           coord_flip() +  
                           ylab(label = "-log10(Adjusted p-value)") +
                           xlab(label = "Enriched GO Cellular Compotent terms") +
                           theme(axis.text.x = element_text(hjust = 0.5, 
                                                            vjust = 0, 
                                                            size = 3, 
                                                            angle = 360),
                                  axis.text.y = element_text(hjust = 0.95, 
                                                             vjust = 0.2, 
                                                             size = 3),
                                  panel.background = element_blank(),
                                  panel.border = element_rect(colour = "black", 
                                                              fill=NA, 
                                                              size=0.5),
                                  axis.title.y = element_blank(),
                                  axis.title.x = element_text(size = 4),
                                  legend.text = element_text(size = 3),
                                  legend.title = element_text(size = 3),
                                  legend.key.height = unit(3, 'mm'),
                                  legend.key.width = unit(3, 'mm'),
                                  legend.position = "right")

```

```{r}
ggsave(plot = barplot_enrich_prot_pls, 
       filename = here::here("figures/barplot_gomf_pls_prot_lip_gbm2.tiff"), 
       device = "tiff",
       units = "mm",
       width = 100,
       height = 50)

ggsave(plot = barplot_enrich_prot_pls, 
       filename = here::here("figures/barplot_gomf_pls_prot_lip_gbm2.eps"), 
       device = "eps",
       units = "mm",
       width = 100,
       height = 50)
```

#### Enrichment analysis of lipidomics signatures after PLS

```{r}
lipids_g1 <- coord2_fcs_circlgg %>%
  filter(groups == "lip_g1") %>%
  dplyr::rename(feature = names) %>%
  left_join(., lipid_annotation3)

lipids_g2 <- coord2_fcs_circlgg %>%
  filter(groups == "lip_g2") %>%
  dplyr::rename(feature = names) %>%
  left_join(., lipid_annotation3)
```

**Save tables of features**

```{r}
write_csv(x = lipids_g1,
          file = here("data/lipidomics/lipids_g1_pls_fcs_left_group.csv"))

write_csv(x = lipids_g2,
          file = here("data/lipidomics/lipids_g2_pls_fcs_bottom_group.csv"))

```

##### Visualization of LION enrichment results

###### LION enrichment after limma

**Loading and preprocessing LION enrichment tabular results**

```{r message=FALSE}
lion_increased_volcano <- read_csv(file = here::here("data/lipidomics/LION-enrichment-increased_lipids.csv"))
```

```{r message=FALSE}
lion_decreased_volcano <- read_csv(file = here::here("data/lipidomics/LION-enrichment-decreased_lipids.csv"))
```

**Preprocessing**

```{r}
lion_increased_volcano <- lion_increased_volcano %>%
  mutate(`Association with Recurrence` = "Increased")

lion_decreased_volcano <- lion_decreased_volcano %>%
  mutate(`Association with Recurrence` = "Decreased")

lion_lipidom_limma <- bind_rows(lion_increased_volcano,
                                 lion_decreased_volcano) %>% 
  group_by(`Association with Recurrence`) %>% 
  slice_min(order_by = `FDR q-value`,
            n = 4) %>%
  dplyr::rename(Description = Discription) %>%
  ungroup()
```

```{r}
lion_lipidom_limma_2 <- lion_lipidom_limma %>%
  mutate(`Association with Protein FCs` = factor(`Association with Recurrence`,
                          levels = c("Increased", "Decreased"))) %>%
  dplyr::arrange(`Association with Recurrence`) %>% 
    mutate(
      Description = if_else(
        condition = str_count(Description, '\\S+') > 3,
        true = str_c(word(Description,1,3), word(Description,4,str_count(Description,'\\S+')), sep = '\n'),
        false = Description
      )
    ) %>% 
  mutate(Description = factor(Description,
                              levels = rev(.$Description))) 
```

**Visualize**

```{r}
ggplot(data = lion_lipidom_limma_2,
       mapping = aes(y = -log10(`FDR q-value`),
                     x = Description, 
                     fill = `Association with Protein FCs`)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "red") + 
  coord_flip() +  
  xlab(label = "-log10(Adjusted p-value") +
  ylab(label = "Enriched Lipid Ontology terms") +
   theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
         axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
         panel.background = element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=0.5),
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         legend.text = element_text(size = 7),
         legend.title = element_text(size = 8),
         legend.key.height= unit(3, 'mm'),
         legend.key.width= unit(3, 'mm'),
         legend.position="right")

```

```{r}
barplot_enr_lion_limma <- ggplot(data = lion_lipidom_limma_2,
                                     mapping = aes(y = -log10(`FDR q-value`),
                                                   x = Description, 
                                                   fill = `Association with Protein FCs`)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "red") + 
  coord_flip() +  
  ylab(label = "-log10(Adjusted p-value") +
  xlab(label = "Enriched Lipid Ontology terms") +
  theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 3, 
                                   angle = 360),
         axis.text.y = element_text(hjust = 0.95, 
                                    vjust = 0.2, 
                                    size = 3),
         panel.background = element_blank(),
         panel.border = element_rect(colour = "black", 
                                     fill=NA, 
                                     size=0.5),
         axis.title.y = element_blank(),
         axis.title.x = element_text(size = 4),
         legend.text = element_text(size = 3),
         legend.title = element_text(size = 3),
         legend.key.height = unit(3, 'mm'),
         legend.key.width = unit(3, 'mm'),
         legend.position = "right",
        text = element_text("sans"))

barplot_enr_lion_limma_helv <- barplot_enr_lion_limma + 
  theme(text = element_text("Helvetica"))
```

```{r}
ggsave(plot = barplot_enr_lion_limma, 
       filename = here::here("figures/barplot_lion_enrich_limma_lip_gbm2.tiff"), 
       device = "tiff",
       units = "mm",
       width = 80,
       height = 36)

ggsave(plot = barplot_enr_lion_limma_helv, 
       filename = here::here("figures/barplot_lion_enrich_limma_lip_gbm2.eps"), 
       device = "eps",
       units = "mm",
       width = 80,
       height = 36)
```

###### LION enrichment after FCs-based PLS

**Loading and preprocessing LION enrichment tabular results**

```{r message=FALSE}
lion_g1_pls <- read_csv(file = here::here("data/lipidomics/LION-enrichment-g1_lipids_fcsbased_pls.csv"))
```

```{r message=FALSE}
lion_g2_pls <- read_csv(file = here::here("data/lipidomics/LION-enrichment-g2_lipids_fcsbased_pls.csv"))
```

**Preprocessing**

```{r}
lion_g1_pls <- lion_g1_pls %>%
  mutate(`PLS group` = "G1")

lion_g2_pls <- lion_g2_pls %>%
  mutate(`PLS group` = "G2")

lion_lipidom_pls <- bind_rows(lion_g1_pls,
                                 lion_g2_pls) %>% 
  group_by(`PLS group`) %>% 
  slice_min(order_by = `FDR q-value`,
            n = 4) %>%
  dplyr::rename(Description = Discription) %>%
  ungroup()
```

```{r}
lion_lipidom_pls_2 <- lion_lipidom_pls %>%
  mutate(`PLS group` = factor(`PLS group`,
                          levels = c("G1", "G2"))) %>%
  dplyr::arrange(`PLS group`) %>% 
    mutate(
      Description = if_else(
        condition = str_count(Description, '\\S+') > 3,
        true = str_c(word(Description,1,3), word(Description,4,str_count(Description,'\\S+')), sep = '\n'),
        false = Description
      )
    ) %>% 
  mutate(Description = factor(Description,
                              levels = rev(.$Description))) 
```

**Visualize**

```{r}
ggplot(data = lion_lipidom_pls_2,
       mapping = aes(y = -log10(`FDR q-value`),
                     x = Description, 
                     fill = `PLS group`)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "red") + 
  coord_flip() +  
  xlab(label = "-log10(Adjusted p-value") +
  ylab(label = "Enriched Lipid Ontology terms") +
   theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 8, angle = 360),
         axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
         panel.background = element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=0.5),
         axis.title.x = element_blank(),
         axis.title.y = element_blank(),
         legend.text = element_text(size = 7),
         legend.title = element_text(size = 8),
         legend.key.height= unit(3, 'mm'),
         legend.key.width= unit(3, 'mm'),
         legend.position="right")

```

```{r}
barplot_enr_lion_pls <- ggplot(data = lion_lipidom_pls_2,
                                     mapping = aes(y = -log10(`FDR q-value`),
                                                   x = Description, 
                                                   fill = `PLS group`)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "red") + 
  coord_flip() +  
  xlab(label = "-log10(Adjusted p-value") +
  ylab(label = "Enriched Lipid Ontology terms") +
     theme(axis.text.x = element_text(hjust = 0.5, 
                                   vjust = 0, 
                                   size = 3, 
                                   angle = 360),
         axis.text.y = element_text(hjust = 0.95, 
                                    vjust = 0.2, 
                                    size = 3),
         panel.background = element_blank(),
         panel.border = element_rect(colour = "black", 
                                     fill=NA, 
                                     size=0.5),
         axis.title.y = element_blank(),
         axis.title.x = element_text(size = 4),
         legend.text = element_text(size = 3),
         legend.title = element_text(size = 3),
         legend.key.height = unit(3, 'mm'),
         legend.key.width = unit(3, 'mm'),
         legend.position = "right",
        text = element_text("sans"))

barplot_enr_lion_pls_helv <- barplot_enr_lion_pls + 
  theme(text = element_text("Helvetica"))
```

```{r}
ggsave(plot = barplot_enr_lion_pls, 
       filename = here::here("figures/barplot_lion_enrich_fcs_pls_lip_gbm2.tiff"), 
       device = "tiff",
       units = "mm",
       width = 80,
       height = 36)

ggsave(plot = barplot_enr_lion_pls_helv, 
       filename = here::here("figures/barplot_lion_enrich_fcs_pls_lip_gbm2.eps"), 
       device = "eps",
       units = "mm",
       width = 80,
       height = 36)
```

### 2023-02-16 reports/patient_summary_table.qmd
patient_info_cleanup

---
title: "GBM Proteomics - Patient summary"
author: "Miguel Cosenza-Contreras"
date: "`r Sys.Date()`"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

# Required packages {.unnumbered .unlisted}

```{r}
library(readxl)
library(gtsummary)
library(here)
library(tidyverse)
library(flextable)
```

# Load and prep the data {.unnumbered .unlisted}

```{r echo=FALSE}
patient_table <- readxl::read_xlsx(path = here("data/Patients_information_FINAL.xlsx"))
```

```{r}
patients_proteomics <- patient_table %>%
  filter(`Mass Spectrometry` == "X") %>%
  dplyr::select(Sex, 
                `Age at surgery (years)`, 
                `Time-to-reccurence (days)`,
                `Type of resection`,
                `Tumor localization`, 
                Treatment,
                sample_id = `LAB-ID`) %>%
  mutate(`Time-to-reccurence (days)` = as.numeric(`Time-to-reccurence (days)`))
```

```{r}
patients_clinical <- patient_table %>%
  mutate(`Time-to-reccurence (days)` = as.numeric(`Time-to-reccurence (days)`),
         `Age at surgery (years)` = as.numeric(`Age at surgery (years)`),
         `Latency (days)` = as.numeric(`Latency (days)`)) %>% 
  pivot_longer(cols = c("Mass Spectrometry",
                        "Lipidomics",
                        "RT-qPCR",
                        "IHC",
                        "ELISA"),
               names_to = "Experiment",
               values_to = "X") %>%
  filter(!is.na(X)) %>%
  dplyr::select(-c(`LAB-ID`, Number))
```

```{r}
patients_clinical_min <- patients_clinical %>%
  dplyr::select(Sex, 
                `Age at surgery (years)`, 
                `Time-to-reccurence (days)`,
                `Latency (days)`,
                `Type of resection`,
                Experiment)
```

# Summary table of patients (minimal information)

```{r}
clin_tb_min <- tbl_summary(data = patients_clinical_min,
                           by = Experiment) %>%
  add_n() %>% # add column with total number of non-missing observations
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels() 
```

```{r}
print(clin_tb_min)
```

```{r echo=FALSE}
clin_tb_min %>%
  as_flex_table() %>%
  flextable::save_as_docx(path = here("suppl_tables/minimal_clinical_summary_table.docx"))
```

# Summary table of patients (all clinical information)

```{r}
tbl_summary(data = patients_clinical,
            by = Experiment) %>%
  add_n() %>% # add column with total number of non-missing observations
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels() 
```

# Summary table of patients for proteomics

```{r}
tbl_summary(data = patients_proteomics) %>%
  add_n() %>% # add column with total number of non-missing observations
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels() 
```

### 2023-08-25 reports/gbm_recurrence_mine_single_cell_data_preprocessing.qmd
single_cell_asah1_expression_mining

---
title: "GBM Recurrence - Mine Single Cell data"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))

## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)
library(ggrepel)
library(extrafont)
extrafont::loadfonts(device = "win")

theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 
                                         'mm'),
                legend.key.width = unit(3, 
                                        'mm'),
                legend.position = "bottom",
                text = element_text(family = "Helvetica")))
```

# Initial data loading and wrangling

Load the matrix and annotation

```{r}
annot_1 <- read_csv(here::here("data/revision_single_cell_data/annot_Human_R_GBM_Full.csv"))

annot_2_ncbi <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972210_annot.Human.GBM.R1_2_3_4_4nc.csv.gz"))

# matrix data

matrix_ncbi <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972210_Human.GBM.R1_2_3_4_4nc.filtered.gene.bc.matrix.csv"))
```

```{r}
View(annot_1)
#View(annot_2_ncbi)
#View(matrix_ncbi)
```

```{r}
#head(str(matrix_ncbi))
#names(matrix_ncbi)[1]
```

# Normalize (CPM)

```{r}
library(scuttle)
```

# log2 transform

```{r}
matrix_ncbi <- matrix_ncbi %>%
  dplyr::rename('gene' = '...1') %>%
  column_to_rownames("gene") %>% 
  as.matrix()
```

```{r}
cpm_mat <- calculateCPM(matrix_ncbi)
mat_cpm_log2 <- log2(cpm_mat + 1)
rm(cpm_mat)
```

```{r}
mat_cpm_log2[1:15, 1:15]
```

# Wrangle

```{r}
mat_cpm_df <- mat_cpm_log2 %>% 
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  relocate(gene, everything()) 
```

```{r}
# ASAH1 expression
mat_log2_df_asah1 <- mat_cpm_df %>%
  filter(gene == "ASAH1")
```

```{r}
long_log2_df_asah1 <- mat_log2_df_asah1 %>%
  pivot_longer(cols = -gene, 
               names_to = "cell", 
               values_to = "asah1_expression") 
```

Merge with annotation

```{r}
long_log2_df_asah1_annot <- long_log2_df_asah1 %>%
  left_join(annot_1, by = c("cell" = "cell"))
```

```{r}
#View(long_log2_df_asah1_annot)
```

```{r}
saveRDS(long_log2_df_asah1_annot,
        here::here("rds/long_cpm_scrnaseq_asah1_annot.rds"))
```

### 2023-08-25 reports/gbm_recurrence_mine_single_cell_data_visualization.qmd
single_cell_asah1_expression_mining

---
title: "GBM Recurrence - Mine Single Cell data"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

## Required packages ----
library(tidyverse)
library(ggpubr)
library(ggrepel)
library(cowplot)

theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 12, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 12),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            linewidth = 1.5),
                axis.title=element_text(size = 12),
                legend.title = element_text(size = 12),
                legend.text = element_text(size = 11),
                legend.key.height = unit(4, 
                                         'mm'),
                legend.key.width = unit(4, 
                                        'mm'),
                legend.key.size = unit(10,
                                        'mm'),
                legend.position = "bottom"))
```

# Initial data loading and wrangling

Load the annotation with ASAH1 expression data

```{r}
annotated_with_asah1 <- readRDS(here::here("rds/long_cpm_scrnaseq_asah1_annot.rds"))
```

```{r}
View(annotated_with_asah1)
```

# Generate scatter plot

```{r}
umap_asah1_expr <- ggplot(annotated_with_asah1, 
       aes(x = x, 
           y = y, 
           color = asah1_expression)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "yellow", high = "red") +
  labs(x = "UMAP_1", 
       y = "UMAP_2", 
       color = "ASAH1 expression\n[log2(CPM)]") +
  theme(legend.position = "bottom")
```

```{r}
cell_type_centroid <- annotated_with_asah1 %>%
  group_by(cluster) %>%
  summarise(x = mean(x), 
            y = mean(y)) %>%
  ungroup()

  cell_type_centroid
```

```{r}
umap_cell_type <- ggplot(annotated_with_asah1, 
       aes(x = x, 
           y = y, 
           color = cluster)) +
  geom_point(alpha = 0.5) +
  geom_text(data = cell_type_centroid, 
            aes(x = x, 
                y = y, 
                label = cluster), 
            size = 5,
            color = "black") +
  #scale_color_gradient(low = "yellow", high = "red") +
  labs(x = "UMAP_1", 
       y = "UMAP_2", 
       color = "Cell type") +
  theme(legend.position = "bottom") 
```

```{r}
umap_cell_type
```

```{r}
umap_grid <- plot_grid(umap_cell_type, 
                       umap_asah1_expr, 
                       ncol = 2)
```

```{r fig.width=20, fig.height=10}
umap_grid
```

```{r}
ggsave(here::here("figures/umap_grid_asah1_expression_scRNAseq.pdf"), 
       umap_grid, 
       width = 34, 
       height = 34/2,
       units = "cm")
```

```{r}
ggsave(here::here("figures/umap_grid_asah1_expression_scRNAseq.png"), 
       umap_grid, 
       width = 34, 
       height = 34/2,
       units = "cm")
```

```{r}
dev.off()
```

### 2023-08-26 reports/gbm_recurrence_mine_single_cell_data_visualization.qmd
single_cell_asah1_expression_v2
sc_asah1_expression_violin

@@
 
-```{r}
-View(annotated_with_asah1)
-```
-
 # Generate scatter plot
@@
   ungroup()
-
-  cell_type_centroid
 ```
@@
   theme(legend.position = "bottom") 
-```
-
-```{r}
-umap_cell_type
 ```
@@
 
+# Violin plot
+
+```{r}
+annotated_with_asah14violin <- annotated_with_asah1 %>%
+  filter(asah1_expression > 2.5) 
+```
+
+```{r}
+violin_asahl_expr_celltype <- ggplot(annotated_with_asah14violin, 
+       aes(x = cluster, 
+           y = asah1_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "ASAH1 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_asahl_expr_celltype
+```
+
+```{r}
+umap_grid_w_violin <- plot_grid(umap_cell_type, 
+                                umap_asah1_expr, 
+                                violin_asahl_expr_celltype,
+                                ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_w_violin
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_asah1_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_asah1_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
 ```{r}

### 2023-08-26 reports/gbm_recurrence_cox_phm_incl_age_sex_revision.qmd
cox_phm_w_sex_age_covar_revision

---
title: "GBM Recurrence - Cox Proportional Hazards Model"
author: "Justin Sing - Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

## Required packages ----
library(tidyverse)
library(ggpubr)
library(ggrepel)
library(cowplot)
library(here)
library(survival)
library(gt)

theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 12, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 12),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            linewidth = 1.5),
                axis.title=element_text(size = 12),
                legend.title = element_text(size = 12),
                legend.text = element_text(size = 11),
                legend.key.height = unit(4, 
                                         'mm'),
                legend.key.width = unit(4, 
                                        'mm'),
                legend.key.size = unit(10,
                                        'mm'),
                legend.position = "bottom"))
```

# Initial data loading and wrangling

Load the annotation with ASAH1 expression data

```{r}
ttr_surv_data <- read_tsv(here::here("data/elisa_plasma_proteomics/ttr_survival_data.tsv"))
```

# Survival analysis

## Cox-PHM - including only sex (original analysis)

Fitting the model:

```{r}
## COXPH ----
#ttr_surv_data
fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex +  ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
```

Model summary:

```{r}
print(summary(fit))
```

Observation: Sex does not seem to represent a significant variable in the model.

### Genarate tabular and graphical summaries

```{r}
library(finalfit)

dependent_var <- "Surv(Time.to.reccurence..days., status)"
explanatory_vars <- c("ASAH1.ng.ml", "SYNM.ng.ml", "GPNMB.ng.ml", "MMP.9.ng.ml")

ttr_surv_data %>% 
    coxphmulti(dependent_var, explanatory_vars) %>% 
    cox.zph() %>% 
    {zph_result <<- .} %>% 
    plot(var=3)

#zph_result

ttr_surv_data %>%
  finalfit(dependent_var, explanatory_vars) %>%
  gt()

ttr_surv_data %>%
  hr_plot(dependent_var, explanatory_vars, 
          dependent_label="Time to Recurrence")

ttr_surv_data %>%
  ff_plot(dependent_var, explanatory_vars, 
          dependent_label="Time to Recurrence")
```

## Cox-PHM - including only sex and age (revision version analysis)

This section was created as part of the revision process, to check for the effect of age and sex in the model.

```{r}
## COXPH ----
#ttr_surv_data
fit2 <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + Age.at.surgery..years. + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
```

Model summary:

```{r}
print(summary(fit2))
```

```{r}
?coxphmulti
```

Observation: p-value for age is 0.0878 in the model; while ASAH1 concentration is stil below 0.05.

```{r}
dependent_var <- "Surv(Time.to.reccurence..days., status)"
explanatory_vars2 <- c("ASAH1.ng.ml", "SYNM.ng.ml", "GPNMB.ng.ml", "MMP.9.ng.ml",
                      "sex", "Age.at.surgery..years.") # includes sex and age

ttr_surv_data %>% 
    coxphmulti(dependent_var, explanatory_vars2) %>% 
    cox.zph() %>% 
    {zph_result2 <<- .} %>% 
    plot(var=3)

#zph_result2

ttr_surv_data %>%
  finalfit(dependent_var, explanatory_vars2) %>%
  gt()

ttr_surv_data %>%
  hr_plot(dependent_var, explanatory_vars2, dependent_label="Time to Recurrence")

ttr_surv_data %>%
  ff_plot(dependent_var, explanatory_vars2, dependent_label="Time to Recurrence")
```

When including Age and Sex in the model, ASAH1 concentrations are still significantly associated with the time to recurrence. Age and sex are not significant in the model.

```{r}
ggsave(here::here("figures/cox_phm_summary_rev_version_w_age_n_sex.pdf"), 
       ttr_surv_data %>%
  ff_plot(dependent_var, explanatory_vars2, dependent_label="Time to Recurrence"), 
       width = 20, 
       height = 20,
       units = "cm")
```

```{r}
ggsave(here::here("figures/cox_phm_summary_rev_version_w_age_n_sex.png"), 
       ttr_surv_data %>%
  ff_plot(dependent_var, explanatory_vars2, dependent_label="Time to Recurrence"), 
       width = 20, 
       height = 20,
       units = "cm")
```

```{r}
dev.off()
```

### 2023-08-26 reports/gbm_recurrence_proteogenomics_mindy_asso_w_ttr.qmd
exploration_assoc_mindy_saav_w_ttr

---
title: "GBM Recurrence Proteogenomics - MINDY assoc. with TTR"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))

## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)
library(ggrepel)
library(extrafont)
extrafont::loadfonts(device = "win")

theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 
                                         'mm'),
                legend.key.width = unit(3, 
                                        'mm'),
                legend.position = "bottom",
                text = element_text(family = "Helvetica")))

```

# Get MINDY expression data

Load SAAV abundance matrix

```{r}
saav_abund_mat <- read_tsv(here("results/proteogenomics/cellreps_2021/summarized_median_abundance_of_saavs.tsv"))
```

```{r}
saav_abund_long <- saav_abund_mat %>% 
    pivot_longer(cols = -gene_variant, 
                 names_to = "sample_label", 
                 values_to = "MINDY1_T433K") %>% 
    mutate(recurrence = case_when(str_detect(sample_label, "prim") ~ "prim",
                                  str_detect(sample_label, "rec") ~ "rec",
                                  TRUE ~ "NA"),
           patient = case_when(str_detect(sample_label, "prim") ~ str_replace(sample_label, "prim", ""),
                                 str_detect(sample_label, "rec") ~ str_replace(sample_label, "rec", ""),
                                 TRUE ~ "NA")) %>%
    mutate(paired_id = paste(patient, 
                             recurrence, 
                             sep = "_")) %>%
    dplyr::distinct() %>%
    filter(gene_variant %in% c("FAM63A_T433K")) 
```

# Associate MYNDY expression with Time to Recurrence

## Load and prepare patient data

```{r echo=FALSE}
patient_file <- here("data/Patients_information_FINAL.xlsx")
patient_table <- readxl::read_xlsx(path = patient_file)
```

```{r}
patients_proteomics <- patient_table %>%
  filter(`Mass Spectrometry` == "X") %>%
  mutate(latency = parse_number(`Latency (days)`), 
         status = case_when(str_detect(`Latency (days)`, "alive") ~ 1,
                            TRUE ~ 2)) %>%
  dplyr::select(Sex, 
                `Age at surgery (years)`, 
                `Time-to-reccurence (days)`,
                latency,
                status,
                `Type of resection`,
                `Tumor localization`, 
                Treatment,
                sample_id = `LAB-ID`) %>%
  mutate(`Time-to-reccurence (days)` = as.numeric(`Time-to-reccurence (days)`))
```

## Sample annotation

```{r}
sample_annotation <- read_csv(here("data/sample_annotation.csv"))

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x", 
                         patient, 
                         sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, 
                           recurrence, 
                           sep = "_")) %>%
  filter(recurrence %in% c("prim", 
                           "rec"))
```

## What's the association of MINDY1_T433K at Initial stage with time to recurrence?

```{r}
quant_annot_intr_init <- saav_abund_long %>%
    left_join(sample_annotation2) %>%
    left_join(patients_proteomics) %>%
    filter(!is.na(`Time-to-reccurence (days)`)) %>% 
    dplyr::mutate(MINDY1_T433K_abundance = ifelse(MINDY1_T433K >= mean(MINDY1_T433K), 
                                 "high", 
                                 "low"))

labid2patientn <- quant_annot_intr_init %>% 
  dplyr::select(patient, sample_id) %>% 
  dplyr::distinct()

patients_proteomics <- left_join(labid2patientn, 
                                 patients_proteomics)
```

```{r}
library(survival)
library(survminer)
```

```{r}
fit1 <- survfit(Surv(`Time-to-reccurence (days)`) ~ MINDY1_T433K_abundance, 
                data = quant_annot_intr_init)
print(fit1)
# Change color, linetype by strata, risk.table color by strata
surv_p1 <- ggsurvplot(fit1,
          pval = TRUE, conf.int = TRUE,
          risk.table = FALSE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = "Dark2"
          ) +
  labs(x = "Time to Recurrence (Days)", 
       y = "Time to Recurrence Probability")

```

```{r}
surv_p1
```

# Cox proportional hazards model on protein expression vs Time to Recurrence

## Checking Assumptions

```{r}
View(quant_annot_intr_init)
names(quant_annot_intr_init)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=18, fig.width=21}
# fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
# print(summary(fit))

library(finalfit)

quant_annot_intr_init2 <- quant_annot_intr_init %>%
  mutate(ttr_days = `Time-to-reccurence (days)`)

dependent_var <- "Surv(`Time-to-reccurence (days)`, status)"
explanatory_vars <- c("MINDY1_T433K", "Sex", "ttr_days")

dependent_var
explanatory_vars

quant_annot_intr_init2 %>% 
    coxphmulti(dependent_var, explanatory_vars) %>% 
    cox.zph() %>% 
    {zph_result <<- .}

```

## Model results

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=8}
library(globaltest)

quant_annot_intr_init2 %>%
  ff_plot(dependent_var, 
          explanatory_vars, 
          dependent_label = "Time to Recurrence")
```

```{r}
explanatory_vars_univar <- c("MINDY1_T433K")
```

```{r}
quant_annot_intr_init2 %>%
  ff_plot(dependent_var, 
          explanatory_vars_univar, 
          dependent_label = "Time to Recurrence")
```

```{r}
ggsave(here::here("figures/cox_phm_mindy_T433K_summary_rev_version_univariate.pdf"), 
       quant_annot_intr_init2 %>%
  ff_plot(dependent_var, 
          explanatory_vars_univar, 
          dependent_label = "Time to Recurrence"), 
       width = 20, 
       height = 10,
       units = "cm")
```

```{r}
ggsave(here::here("figures/cox_phm_mindy_T433K_summary_rev_version_w_sex_n_age.pdf"), 
       quant_annot_intr_init2 %>%
  ff_plot(dependent_var, 
          explanatory_vars, 
          dependent_label = "Time to Recurrence"), 
       width = 20, 
       height = 10,
       units = "cm")
```

### 2023-08-26 results/elisa_plasma_proteomics/manuscript_elisa_figure_analysis.Rmd
exploration_assoc_mindy_saav_w_ttr

@@
 ```{r message=FALSE, warning=FALSE, include=FALSE}
-patient_clinical_data <- "../../data/elisa_plasma_proteomics/Patients_information_FINAL.csv"
-patient_longitudinal_mapping_file <- "../../data/elisa_plasma_proteomics/Patients_for_ELISA_longitudinal_mapping.csv"
-elisa_data_file <- "../../data/elisa_plasma_proteomics/ELISA_data.csv"
+patient_clinical_data <- here::here("data/elisa_plasma_proteomics/Patients_information_FINAL.csv")
+
+patient_longitudinal_mapping_file <- here::here("data/elisa_plasma_proteomics/Patients_for_ELISA_longitudinal_mapping.csv")
+
+elisa_data_file <- here::here("data/elisa_plasma_proteomics/ELISA_data.csv")
 ```
@@
     do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "Manifestation")
-  }, mc.cores = 4) %>%
+  }) %>%
     rbindlist() %>%
     unique()
-}, mc.cores=3) %>% rbindlist()
+}) %>% rbindlist()
 
@@
     do_perm_test(data = .d_wide, gene = .d$gene, response_var = "log2_conc", group1 = comp[1], group2 = comp[2], method = method, grouping_column = "collection_type2")
-  }, mc.cores = 4) %>%
+  }) %>%
     rbindlist() %>%
     unique()
-}, mc.cores=3) %>% rbindlist()
+}) %>% rbindlist()
 
@@
   coord_equal() +
-  style_roc(x="1 - Specificity", y="Sensitivity")
+  style_roc(x = "1 - Specificity", 
+            y = "Sensitivity")
 
@@
   # dplyr::filter(n==2) %>%
-  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata
+  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to = c("iGBMoverrGBM"))) -> pdata
 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9, include=FALSE}
-latency_data_file <- "../../data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv"
+latency_data_file <- here::here("data/elisa_plasma_proteomics/ELISA_LATENCY_KM.csv")
 latency_data <- fread(latency_data_file)
@@
 ttr_surv_data <- rstatix::make_clean_names(ttr_surv_data)
+
+library(readr)
+write_tsv(ttr_surv_data,
+          here::here("data/elisa_plasma_proteomics/ttr_survival_data.tsv"))
 
@@
 ## COXPH ----
+ttr_surv_data
 fit <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
 print(summary(fit))
+
+library(finalfit)
+
+dependent_var <- "Surv(Time.to.reccurence..days., status)"
+explanatory_vars <- c("ASAH1.ng.ml", "SYNM.ng.ml", "GPNMB.ng.ml", "MMP.9.ng.ml")
+
+ttr_surv_data %>% 
+    coxphmulti(dependent_var, explanatory_vars) %>% 
+    cox.zph() %>% 
+    {zph_result <<- .} %>% 
+    plot(var=3)
+
+zph_result
+
+ttr_surv_data %>%
+  finalfit(dependent_var, explanatory_vars) %>%
+  gt()
+
+ttr_surv_data %>%
+  hr_plot(dependent_var, explanatory_vars, dependent_label="Time to Recurrence")
+
+ttr_surv_data %>%
+  surv_plot(dependent_var, explanatory_vars, xlab="Time (days)", pval=TRUE, legend="none")
+
+ttr_surv_data %>%
+  ff_plot(dependent_var, explanatory_vars, dependent_label="Time to Recurrence")
+```
+
+# Cox-PHM
+
+This section was created as part of the revision process, to check for the effect of age and sex in the model.
+
+```{r}
+?Surv
+?coxphmulti
+```
+
+```{r}
+## COXPH ----
+#ttr_surv_data
+fit2 <- coxph(Surv(Time.to.reccurence..days., status) ~ sex + Age.at.surgery..years. + ASAH1.ng.ml + SYNM.ng.ml + GPNMB.ng.ml + MMP.9.ng.ml, data = ttr_surv_data)
+print(summary(fit2))
 

### 2023-10-24 reports/gbm_recurrence_mine_single_cell_data_preprocessing.qmd
revision 2 commit

@@
 
+annot_2_nd <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972211_annot.Human.GBM.ND1_2_3_4_5_6_7.csv"))
+
 annot_2_ncbi <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972210_annot.Human.GBM.R1_2_3_4_4nc.csv.gz"))
@@
 # matrix data
+matrix_ncbi <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972210_Human.GBM.R1_2_3_4_4nc.filtered.gene.bc.matrix.csv"))
 
-matrix_ncbi <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972210_Human.GBM.R1_2_3_4_4nc.filtered.gene.bc.matrix.csv"))
+# matrix data init
+matrix_ncbi_nd <- read_csv(here::here("data/revision_single_cell_data/from_ncbi/GSM4972211_Human.GBM.ND1_2_3_4_5_6_7.filtered.gene.bc.matrix.csv"))
 ```
@@
 ```{r}
-View(annot_1)
+library(scuttle)
+```
+
+```{r}
+#View(annot_1)
 #View(annot_2_ncbi)
@@
 
-```{r}
-library(scuttle)
-```
+# log2 transform
 
-# log2 transform
+## rGBM
 
@@
 ```{r}
+if(!file.exists(here::here("rds/matrix_ncbi_scSeq_CPM_rgbm.rds"))) {
+
   cpm_mat <- calculateCPM(matrix_ncbi)
@@
   rm(cpm_mat)
+
+  saveRDS(mat_cpm_log2, 
+          here::here("rds/matrix_ncbi_scSeq_CPM_rgbm.rds"))
+} else {
+
+  mat_cpm_log2 <- readRDS(here::here("rds/matrix_ncbi_scSeq_CPM_rgbm.rds"))
+
+}
 ```
@@
 ```{r}
-mat_cpm_log2[1:15, 1:15]
+mat_cpm_log2[10:15, 10:15]
+```
+
+## iGBM (ND)
+
+```{r}
+matrix_ncbi_nd <- matrix_ncbi_nd %>%
+  dplyr::rename('gene' = '...1') %>%
+  column_to_rownames("gene") %>% 
+  as.matrix()
+```
+
+```{r}
+if(!file.exists(here::here("rds/matrix_ncbi_scSeq_CPM_igbm.rds"))) {
+
+  cpm_mat_nd <- calculateCPM(matrix_ncbi_nd)
+  mat_cpm_log2_nd <- log2(cpm_mat_nd + 1)
+  rm(cpm_mat_nd)
+
+  saveRDS(mat_cpm_log2_nd, 
+          here::here("rds/matrix_ncbi_scSeq_CPM_igbm.rds"))
+} else {
+
+  mat_cpm_log2 <- readRDS(here::here("rds/matrix_ncbi_scSeq_CPM_igbm.rds"))
+
+}
+```
+
+```{r}
+mat_cpm_log2_nd[10:15, 10:15]
 ```
@@
 # Wrangle
+
+## rGBM
 
@@
   relocate(gene, everything())
+
+mat_cpm_df_names <- mat_cpm_df %>%
+  dplyr::select(gene)
 ```
+
+## iGBM (ND)
+
+```{r}
+mat_cpm_df_nd <- mat_cpm_log2_nd %>% 
+  as.data.frame() %>%
+  rownames_to_column("gene") %>%
+  relocate(gene, everything())
+
+mat_cpm_df_names_nd <- mat_cpm_df_nd %>%
+  dplyr::select(gene)
+
+  #View(mat_cpm_df_nd)
+```
+
+# Filter for ASAH1
+
+## rGBM
 
@@
 ```{r}
-#View(long_log2_df_asah1_annot)
+View(long_log2_df_asah1_annot)
 ```
@@
 ```
+
+## iGBM (ND)
+
+```{r}
+# ASAH1 expression
+mat_log2_df_asah1_nd <- mat_cpm_df_nd %>%
+  filter(gene == "ASAH1")
+```
+
+```{r}
+long_log2_df_asah1_nd <- mat_log2_df_asah1_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "asah1_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_asah1_annot_nd <- long_log2_df_asah1_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+saveRDS(long_log2_df_asah1_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_asah1_annot_nd.rds"))
+```
+
+# Filter for MPO
+
+## rGBM
+
+```{r}
+# MPO expression
+mat_log2_df_mpo <- mat_cpm_df %>%
+  filter(gene == "MPO")
+```
+
+```{r}
+long_log2_df_mpo <- mat_log2_df_mpo %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "mpo_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_mpo_annot <- long_log2_df_mpo %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_mpo_annot)
+```
+
+```{r}
+saveRDS(long_log2_df_mpo_annot,
+        here::here("rds/long_cpm_scrnaseq_mpo_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# MPO expression
+
+mat_log2_df_mpo_nd <- mat_cpm_df_nd %>%
+  filter(gene == "MPO")
+```
+
+```{r}
+long_log2_df_mpo_nd <- mat_log2_df_mpo_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "mpo_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_mpo_annot_nd <- long_log2_df_mpo_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_mpo_annot_nd)
+```
+
+```{r}
+saveRDS(long_log2_df_mpo_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_mpo_annot_nd.rds"))
+```
+
+# Filter for CD14
+
+## rGBM
+
+```{r}
+# CD14 expression
+mat_log2_df_cd14 <- mat_cpm_df %>%
+  filter(gene == "CD14")
+```
+
+```{r}
+long_log2_df_cd14 <- mat_log2_df_cd14 %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "cd14_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_cd14_annot <- long_log2_df_cd14 %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_cd14_annot)
+```
+
+```{r}
+saveRDS(long_log2_df_cd14_annot,
+        here::here("rds/long_cpm_scrnaseq_cd14_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# CD14 expression
+mat_log2_df_cd14_nd <- mat_cpm_df_nd %>%
+  filter(gene == "CD14")
+```
+
+```{r}
+long_log2_df_cd14_nd <- mat_log2_df_cd14_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "cd14_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_cd14_annot_nd <- long_log2_df_cd14_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_cd14_annot_nd)
+```
+
+```{r}
+saveRDS(long_log2_df_cd14_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_cd14_annot_nd.rds"))
+```
+
+# Filter for IBA1
+
+## rGBM
+
+```{r}
+# IBA1 expression
+mat_log2_df_iba1 <- mat_cpm_df %>%
+  filter(gene == "AIF1")
+```
+
+```{r}
+long_log2_df_iba1 <- mat_log2_df_iba1 %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "iba1_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_iba1_annot <- long_log2_df_iba1 %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+saveRDS(long_log2_df_iba1_annot,
+        here::here("rds/long_cpm_scrnaseq_iba1_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# IBA1 expression
+mat_log2_df_iba1_nd <- mat_cpm_df_nd %>%
+  filter(gene == "AIF1")
+```
+
+```{r}
+long_log2_df_iba1_nd <- mat_log2_df_iba1_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "iba1_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_iba1_annot_nd <- long_log2_df_iba1_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+saveRDS(long_log2_df_iba1_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_iba1_annot_nd.rds"))
+```
+
+# Filter for lipocalin-2 (LCN2)
+
+## rGBM
+
+```{r}
+mat_cpm_df[1:5, 1:5]
+```
+
+```{r}
+# LCN2 expression
+mat_log2_df_lcn2 <- mat_cpm_df %>%
+  filter(gene == "LCN2")
+```
+
+```{r}
+long_log2_df_lcn2 <- mat_log2_df_lcn2 %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "lcn2_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_lcn2_annot <- long_log2_df_lcn2 %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_lcn2_annot)
+head(long_log2_df_lcn2_annot)
+```
+
+```{r}
+saveRDS(long_log2_df_lcn2_annot,
+        here::here("rds/long_cpm_scrnaseq_lcn2_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# LCN2 expression
+mat_log2_df_lcn2_nd <- mat_cpm_df_nd %>%
+  filter(gene == "LCN2")
+```
+
+```{r}
+long_log2_df_lcn2_nd <- mat_log2_df_lcn2_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "lcn2_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_lcn2_annot_nd <- long_log2_df_lcn2_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_lcn2_annot_nd)
+head(long_log2_df_lcn2_annot_nd)
+```
+
+```{r}
+saveRDS(long_log2_df_lcn2_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_lcn2_annot_nd.rds"))
+```
+
+# Filter for neutrophil elastase (ELANE)
+
+## rGBM
+
+```{r}
+# ELANE expression
+mat_log2_df_elane <- mat_cpm_df %>%
+  filter(gene == "ELANE")
+```
+
+```{r}
+long_log2_df_elane <- mat_log2_df_elane %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "elane_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_elane_annot <- long_log2_df_elane %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_elane_annot)
+head(long_log2_df_elane_annot)
+```
+
+```{r}
+saveRDS(long_log2_df_elane_annot,
+        here::here("rds/long_cpm_scrnaseq_elane_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# ELANE expression
+mat_log2_df_elane_nd <- mat_cpm_df_nd %>%
+  filter(gene == "ELANE")
+```
+
+```{r}
+long_log2_df_elane_nd <- mat_log2_df_elane_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "elane_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_elane_annot_nd <- long_log2_df_elane_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_elane_annot_nd)
+head(long_log2_df_elane_annot_nd)
+```
+
+```{r}
+saveRDS(long_log2_df_elane_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_elane_annot_nd.rds"))
+```
+
+# Filter for CD68
+
+## rGBM
+
+```{r}
+# CD68 expression
+mat_log2_df_cd68 <- mat_cpm_df %>%
+  filter(gene == "CD68")
+```
+
+```{r}
+long_log2_df_cd68 <- mat_log2_df_cd68 %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "cd68_expression") 
+```
+Merge with annotation
+
+```{r}
+long_log2_df_cd68_annot <- long_log2_df_cd68 %>%
+  left_join(annot_1, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_cd68_annot)
+head(long_log2_df_cd68_annot)
+```
+
+```{r}
+saveRDS(long_log2_df_cd68_annot,
+        here::here("rds/long_cpm_scrnaseq_cd68_annot.rds"))
+```
+
+## iGBM (ND)
+
+```{r}
+# CD68 expression
+mat_log2_df_cd68_nd <- mat_cpm_df_nd %>%
+  filter(gene == "CD68")
+```
+
+```{r}
+long_log2_df_cd68_nd <- mat_log2_df_cd68_nd %>%
+  pivot_longer(cols = -gene, 
+               names_to = "cell", 
+               values_to = "cd68_expression") 
+```
+
+Merge with annotation
+
+```{r}
+long_log2_df_cd68_annot_nd <- long_log2_df_cd68_nd %>%
+  left_join(annot_2_nd, by = c("cell" = "cell"))
+```
+
+```{r}
+#View(long_log2_df_cd68_annot_nd)
+head(long_log2_df_cd68_annot_nd)
+```
+
+```{r}
+saveRDS(long_log2_df_cd68_annot_nd,
+        here::here("rds/long_cpm_scrnaseq_cd68_annot_nd.rds"))
+```

### 2023-10-24 reports/gbm_recurrence_mine_single_cell_data_visualization.qmd
single_cell_data_visualization
revision 2 commit

@@
 ```{r}
-knitr::opts_chunk$set(echo = TRUE, 
+knitr::opts_chunk$set(echo = FALSE, 
                       message = FALSE, 
@@
 library(cowplot)
+library(scales)    # Install ggnewscale package
+library(ggnewscale)    
+library(ggrastr)
 
@@
 
-# Generate scatter plot
+# Generate scatter plot for ASAH1 expression
+
+## rGBM
+
+```{r}
+cell_type_centroid <- annotated_with_asah1 %>%
+  group_by(cluster) %>%
+  summarise(x = mean(x), 
+            y = mean(y)) %>%
+  ungroup()
+```
 
@@
            color = asah1_expression)) +
-  geom_point(alpha = 0.5) +
+  ggrastr::rasterize(geom_point(alpha = 0.5)) +
+  geom_text(data = cell_type_centroid, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") +
   scale_color_gradient(low = "yellow", high = "red") +
@@
        y = "UMAP_2", 
-       color = "ASAH1 expression\n[log2(CPM)]") +
+       color = "ASAH1 expression\n[log2(CPM)]",
+       title = "Recurrent GBM") +
   theme(legend.position = "bottom")
@@
            color = cluster)) +
-  geom_point(alpha = 0.5) +
+  ggrastr::rasterize(geom_point(alpha = 0.5)) +
   geom_text(data = cell_type_centroid, 
@@
        y = "UMAP_2", 
-       color = "Cell type") +
+       color = "Cell type",
+       title = "Recurrent GBM") +
   theme(legend.position = "bottom") 
@@
 ```{r fig.width=20, fig.height=10}
-umap_grid
+#umap_grid
 ```
@@
 
-# Violin plot
+## Violin plot
 
@@
 ```{r}
-violin_asahl_expr_celltype
+#violin_asahl_expr_celltype
 ```
@@
 
-```{r fig.width=20, fig.height=10}
+```{r fig.width=20, fig.height=20}
 umap_grid_w_violin
@@
 ```
+
+## iGBM
+
+```{r}
+annotated_with_asah1_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_asah1_annot_nd.rds"))
+```
+
+```{r}
+cell_type_centroid_nd <- annotated_with_asah1_nd %>%
+  group_by(cluster) %>%
+  summarise(x = mean(x), 
+            y = mean(y)) %>%
+  ungroup()
+```
+
+```{r}
+umap_cell_type_nd <- ggplot(annotated_with_asah1_nd, 
+       aes(x = x, 
+           y = y, 
+           color = cluster)) +
+  ggrastr::rasterize(geom_point(alpha = 0.5)) +
+  geom_text(data = cell_type_centroid_nd, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "Cell type",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom") 
+```
+
+# Generate scatter plot MPO expression
+
+```{r}
+annotated_with_mpo <- readRDS(here::here("rds/long_cpm_scrnaseq_mpo_annot.rds"))
+```
+
+```{r}
+umap_mpo_expr <- ggplot(annotated_with_mpo, 
+       aes(x = x, 
+           y = y, 
+           color = mpo_expression)) +
+  geom_point(alpha = 0.5) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "MPO expression\n[log2(CPM)]") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_mpo <- plot_grid(umap_cell_type, 
+                           umap_mpo_expr, 
+                           ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+#umap_grid_mpo
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_mpo_expression_scRNAseq.pdf"), 
+       umap_grid_mpo, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_mpo_expression_scRNAseq.png"), 
+       umap_grid_mpo, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+## Violin plot
+
+```{r}
+annotated_with_mpo4violin <- annotated_with_mpo %>%
+  filter(mpo_expression > 0) 
+```
+
+```{r}
+violin_mpo_expr_celltype <- ggplot(annotated_with_mpo4violin, 
+       aes(x = cluster, 
+           y = mpo_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "MPO expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+#violin_mpo_expr_celltype
+```
+
+```{r}
+umap_grid_w_violin_mpo <- plot_grid(umap_cell_type, 
+                                    umap_mpo_expr, 
+                                    violin_mpo_expr_celltype,
+                                    ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_mpo
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_mpo_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin_mpo, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_mpo_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin_mpo, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+# Generate scatter plot for CD14 expression
+
+## rGBM
+
+```{r}
+annotated_with_cd14 <- readRDS(here::here("rds/long_cpm_scrnaseq_cd14_annot.rds"))
+```
+
+```{r}
+umap_cd14_expr <- ggplot(annotated_with_cd14, 
+       aes(x = x, 
+           y = y, 
+           color = cd14_expression)) +
+  rasterize(geom_point(alpha = 0.5)) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD14 expression\n[log2(CPM)]") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_cd14 <- plot_grid(umap_cell_type, 
+                            umap_cd14_expr, 
+                            ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_cd14
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_scRNAseq.pdf"), 
+       umap_grid_cd14, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_scRNAseq.png"), 
+       umap_grid_cd14, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+### Violin plot
+
+```{r}
+annotated_with_cd144violin <- annotated_with_cd14 %>%
+  filter(cd14_expression > 0) 
+```
+
+```{r}
+violin_cd14_expr_celltype <- ggplot(annotated_with_cd144violin, 
+       aes(x = cluster, 
+           y = cd14_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "CD14 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+#violin_cd14_expr_celltype
+```
+
+```{r}
+umap_grid_w_violin_cd14 <- plot_grid(umap_cell_type, 
+                                     umap_cd14_expr, 
+                                     violin_cd14_expr_celltype,
+                                     ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_cd14
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin_cd14, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin_cd14, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+## iGBM
+
+```{r}
+annotated_with_cd14_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_cd14_annot_nd.rds"))
+```
+
+```{r}
+umap_cd14_expr_nd <- ggplot(annotated_with_cd14_nd, 
+       aes(x = x, 
+           y = y, 
+           color = cd14_expression)) +
+  rasterize(geom_point(alpha = 0.5)) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD14 expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_cd14_nd <- plot_grid(umap_cell_type_nd, 
+                               umap_cd14_expr_nd, 
+                               ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_cd14_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_scRNAseq_nd.pdf"), 
+       umap_grid_cd14_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_scRNAseq_nd.png"), 
+       umap_grid_cd14_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+### Violin plot
+
+```{r}
+annotated_with_cd144violin_nd <- annotated_with_cd14_nd %>%
+  filter(cd14_expression > 0) 
+```
+
+```{r}
+violin_cd14_expr_celltype_nd <- ggplot(annotated_with_cd144violin_nd, 
+       aes(x = cluster, 
+           y = cd14_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "CD14 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+#violin_cd14_expr_celltype_nd
+```
+
+```{r}
+umap_grid_w_violin_cd14_nd <- plot_grid(umap_cell_type, 
+                                        umap_cd14_expr_nd, 
+                                        violin_cd14_expr_celltype_nd,
+                                        ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_cd14_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_violin_scRNAseq_nd.pdf"), 
+       umap_grid_w_violin_cd14_nd, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd14_expression_violin_scRNAseq_nd.png"), 
+       umap_grid_w_violin_cd14_nd, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+# Generate scatter plot for IBA1 expression
+
+## rGBM
+
+```{r}
+annotated_with_iba1 <- readRDS(here::here("rds/long_cpm_scrnaseq_iba1_annot.rds"))
+```
+
+```{r}
+umap_iba1_expr <- ggplot(annotated_with_iba1, 
+       aes(x = x, 
+           y = y, 
+           color = iba1_expression)) +
+  geom_point(alpha = 0.5) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "IBA1 expression\n[log2(CPM)]") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_iba1 <- plot_grid(umap_cell_type, 
+                            umap_iba1_expr, 
+                            ncol = 2)
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_scRNAseq.pdf"), 
+       umap_grid_iba1, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_scRNAseq.png"), 
+       umap_grid_iba1, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+### Violin plot
+
+```{r}
+annotated_with_iba14violin <- annotated_with_iba1 %>%
+  filter(iba1_expression > 0)
+
+  head(annotated_with_iba14violin)
+```
+
+```{r}
+violin_iba1_expr_celltype <- ggplot(annotated_with_iba14violin, 
+       aes(x = cluster, 
+           y = iba1_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "IBA1 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+umap_grid_w_violin_iba1 <- plot_grid(umap_cell_type, 
+                                     umap_iba1_expr, 
+                                     violin_iba1_expr_celltype,
+                                     ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_iba1
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin_iba1, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin_iba1, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+## iGBM
+
+```{r}
+annotated_with_iba1_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_iba1_annot_nd.rds"))
+```
+
+```{r}
+umap_iba1_expr_nd <- ggplot(annotated_with_iba1_nd, 
+       aes(x = x, 
+           y = y, 
+           color = iba1_expression)) +
+  rasterize(geom_point(alpha = 0.5)) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "IBA1 expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_iba1_nd <- plot_grid(umap_cell_type_nd, 
+                               umap_iba1_expr_nd, 
+                               ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_iba1_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_scRNAseq_nd.pdf"), 
+       umap_grid_iba1_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_expression_scRNAseq_nd.png"), 
+       umap_grid_iba1_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+# Generate scatter plot for ELANE expression
+
+## rGBM
+
+```{r}
+annotated_with_elane <- readRDS(here::here("rds/long_cpm_scrnaseq_elane_annot.rds"))
+```
+
+```{r}
+umap_elane_expr <- ggplot(annotated_with_elane, 
+       aes(x = x, 
+           y = y, 
+           color = elane_expression)) +
+  geom_point(alpha = 0.5) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ELANE expression\n[log2(CPM)]") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_elane <- plot_grid(umap_cell_type, 
+                             umap_elane_expr, 
+                             ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+#umap_grid_elane
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_scRNAseq.pdf"), 
+       umap_grid_elane, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_scRNAseq.png"), 
+       umap_grid_elane, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+## Violin plot
+
+```{r}
+annotated_with_elane4violin <- annotated_with_elane %>%
+  filter(elane_expression > 0) 
+```
+
+```{r}
+violin_elane_expr_celltype <- ggplot(annotated_with_elane4violin, 
+       aes(x = cluster, 
+           y = elane_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "ELANE expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_elane_expr_celltype
+```
+
+```{r}
+umap_grid_w_violin_elane <- plot_grid(umap_cell_type, 
+                                      umap_elane_expr, 
+                                      violin_elane_expr_celltype,
+                                      ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_elane
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin_elane, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin_elane, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+## iGBM
+
+```{r}
+annotated_with_elane_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_elane_annot_nd.rds"))
+```
+
+```{r}
+umap_elane_expr_nd <- ggplot(annotated_with_elane_nd, 
+       aes(x = x, 
+           y = y, 
+           color = elane_expression)) +
+  rasterize(geom_point(alpha = 0.5)) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ELANE expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_elane_nd <- plot_grid(umap_cell_type_nd, 
+                                umap_elane_expr_nd, 
+                                ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_elane_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_scRNAseq_nd.pdf"), 
+       umap_grid_elane_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_expression_scRNAseq_nd.png"), 
+       umap_grid_elane_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+# Genearate scatter plot for lipocalin-2 (LCN2) expression
+
+The single-cell dataset does not seem to contain LCN2 expression data.
+
+# Overlay ELANE with ASAH1, MPO and CD14 expression
+
+## rGBM
+
+```{r}
+annotated_with_elane_pos <- annotated_with_elane %>%
+       mutate(elane_positive = if_else(elane_expression > 0, 
+                                       "ELANE +", 
+                                       NA)) %>% 
+       dplyr::select(-c(gene))
+
+annotated_with_asah1_elane <- left_join(annotated_with_asah1,
+                                        annotated_with_elane_pos)
+
+annotated_with_asah1_elane_only_elane <- annotated_with_asah1_elane %>%
+       filter(!is.na(elane_positive))
+
+annotated_with_mpo_only <- annotated_with_mpo %>%
+       filter(mpo_expression > 0)
+
+annotated_with_cd14_only <- annotated_with_cd14 %>%
+       filter(cd14_expression > 0)
+
+```
+
+```{r}
+```
+
+```{r}
+umap_asah1_elane_expr <- ggplot() +
+  rasterize(geom_point(data = annotated_with_asah1_elane, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = asah1_expression),
+       alpha = 0.5
+           )) +
+  scale_color_gradient(low = "grey", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ASAH1 expression\n[log2(CPM)]") +
+  new_scale_color() + 
+  rasterize(geom_point(data = annotated_with_asah1_elane_only_elane, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = elane_expression),
+       alpha = 0.5
+           )) +
+  scale_color_gradient(low = "grey", high = "green") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ELANE expression\n[log2(CPM)]") +
+  new_scale_color() + 
+  rasterize(geom_point(data = annotated_with_mpo_only, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = mpo_expression),
+       alpha = 0.5
+           )) +
+  scale_color_gradient(low = "grey", high = "blue") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "MPO expression\n[log2(CPM)]",
+       title = "Recurrent GBM") +
+  geom_text(data = cell_type_centroid, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") 
+  #theme(legend.position = "bottom")
+```
+
+```{r}
+umap_asah1_elane_expr
+```
+
+```{r}
+umap_grid_overlay_three <- plot_grid(umap_cell_type, 
+                                     umap_asah1_elane_expr, 
+                                     ncol = 2)
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_asah1_mpo_expression_scRNAseq.pdf"), 
+       umap_grid_overlay_three, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_asah1_mpo_expression_scRNAseq.png"), 
+       umap_grid_overlay_three, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_overlay_three
+```
+
+### Violin of overlay
+
+```{r}
+annotated_with_elane4violin2 <- dplyr::rename(annotated_with_elane4violin,
+                                              gene_expression = elane_expression)
+```
+
+```{r}
+annotated_with_asah14violin2 <- dplyr::rename(annotated_with_asah14violin,
+                                              gene_expression = asah1_expression)
+```
+
+```{r}
+annotated_with_mpo4violin2 <- dplyr::rename(annotated_with_mpo4violin,
+                                            gene_expression = mpo_expression)
+```
+
+```{r}
+annotated_with_asah1_elane_mpo <- bind_rows(
+       annotated_with_elane4violin2,
+       annotated_with_asah14violin2,
+       annotated_with_mpo4violin2
+       ) %>%
+       group_by(cluster, gene) %>%
+       filter(n() > 1) %>%
+       ungroup()
+```
+
+```{r}
+#View(annotated_with_asah1_elane_mpo)
+```
+
+```{r}
+violin_three_expr_celltype <- ggplot(annotated_with_asah1_elane_mpo, 
+       aes(x = cluster, 
+           y = gene_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "Gene expression [log2(CPM)]") +
+  facet_grid(.~gene,
+              scale = "free_x") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_three_expr_celltype
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_elane_asah1_mpo_expression_scRNAseq.pdf"), 
+       plot = violin_three_expr_celltype, 
+       device = "pdf",
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+dev.off()
+```
+
+## iGBM
+
+```{r}
+annotated_with_elane_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_elane_annot_nd.rds"))
+```
+
+```{r}
+annotated_with_mpo_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_mpo_annot_nd.rds"))
+```
+
+```{r}
+annotated_with_elane_pos_nd <- annotated_with_elane_nd %>%
+       mutate(elane_positive = if_else(elane_expression > 0, 
+                                       "ELANE +", 
+                                       NA)) %>% 
+       dplyr::select(-c(gene))
+
+annotated_with_asah1_elane_nd <- left_join(annotated_with_asah1_nd,
+                                           annotated_with_elane_pos_nd)
+
+annotated_with_asah1_elane_only_elane_nd <- annotated_with_asah1_elane_nd %>%
+       filter(!is.na(elane_positive))
+
+annotated_with_mpo_only_nd <- annotated_with_mpo_nd %>% 
+       filter(mpo_expression > 0)
+
+       annotated_with_mpo_only_nd
+```
+
+```{r}
+umap_asah1_elane_expr_nd <- ggplot() +
+  rasterize(geom_point(data = annotated_with_asah1_elane_nd, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = asah1_expression),
+       alpha = 0.5
+           )) +
+  scale_color_gradient(low = "grey", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ASAH1 expression\n[log2(CPM)]") +
+  new_scale_color() + 
+  rasterize(geom_point(data = annotated_with_asah1_elane_only_elane_nd, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = elane_expression),
+       alpha = 0.5
+           )) +
+  scale_color_gradient(low = "grey", high = "green") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ELANE expression\n[log2(CPM)]") +
+       new_scale_color() + 
+  rasterize(geom_point(data = annotated_with_mpo_only_nd, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = mpo_expression),
+       alpha = 0.5
+           )) +
+  geom_text(data = cell_type_centroid_nd, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") +
+  scale_color_gradient(low = "grey", high = "blue") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "MPO expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") 
+  #theme(legend.position = "bottom")
+```
+
+```{r}
+umap_asah1_elane_expr_nd
+```
+
+## UMAP overlay of rGBM and iGBM
+
+```{r}
+umap_grid_overlay_three_nd <- plot_grid(umap_asah1_elane_expr_nd,
+                                        umap_asah1_elane_expr, 
+                                        ncol = 2)
+```
+
+```{r}
+umap_grid_overlay_three_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_asah1_mpo_expression_scRNAseq_nd.pdf"), 
+       umap_grid_overlay_three_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_asah1_mpo_expression_scRNAseq_only_nd.pdf"), 
+       umap_asah1_elane_expr_nd, 
+       width = 34/2, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_elane_asah1_mpo_expression_scRNAseq_only_rgbm.pdf"), 
+       umap_asah1_elane_expr, 
+       width = 25, 
+       height = 25,
+       units = "cm")
+```
+
+# Generate scatter plot for CD68 expression
+
+## rGBM
+
+```{r}
+annotated_with_cd68 <- readRDS(here::here("rds/long_cpm_scrnaseq_cd68_annot.rds"))
+```
+
+```{r}
+umap_cd68_expr <- ggplot(annotated_with_cd68, 
+       aes(x = x, 
+           y = y, 
+           color = cd68_expression)) +
+  geom_point(alpha = 0.5) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD68 expression\n[log2(CPM)]") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_cd68 <- plot_grid(umap_cell_type, 
+                            umap_cd68_expr, 
+                            ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+#umap_grid_cd68
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_scRNAseq.pdf"), 
+       umap_grid_cd68, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_scRNAseq.png"), 
+       umap_grid_cd68, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+## Violin plot
+
+```{r}
+annotated_with_cd684violin <- annotated_with_cd68 %>%
+  filter(cd68_expression > 0) 
+```
+
+```{r}
+violin_cd68_expr_celltype <- ggplot(annotated_with_cd684violin, 
+       aes(x = cluster, 
+           y = cd68_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "CD68 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+#violin_cd68_expr_celltype
+```
+
+```{r}
+umap_grid_w_violin_cd68 <- plot_grid(umap_cell_type, 
+                                     umap_cd68_expr, 
+                                     violin_cd68_expr_celltype,
+                                     ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_cd68
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_violin_scRNAseq.pdf"), 
+       umap_grid_w_violin_cd68, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_violin_scRNAseq.png"), 
+       umap_grid_w_violin_cd68, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+## iGBM
+
+```{r}
+annotated_with_cd68_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_cd68_annot_nd.rds"))
+```
+
+```{r}
+umap_cd68_expr_nd <- ggplot(annotated_with_cd68_nd, 
+       aes(x = x, 
+           y = y, 
+           color = cd68_expression)) +
+  geom_point(alpha = 0.5) +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD68 expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_cd68_nd <- plot_grid(umap_cell_type_nd, 
+                               umap_cd68_expr_nd, 
+                               ncol = 2)
+```
+
+```{r fig.width=20, fig.height=10}
+umap_grid_cd68_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_scRNAseq_nd.pdf"), 
+       umap_grid_cd68_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_scRNAseq_nd.png"), 
+       umap_grid_cd68_nd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+### Violin plot
+
+```{r}
+annotated_with_cd684violin_nd <- annotated_with_cd68_nd %>%
+  filter(cd68_expression > 0) 
+```
+
+```{r}
+violin_cd68_expr_celltype_nd <- ggplot(annotated_with_cd684violin_nd, 
+       aes(x = cluster, 
+           y = cd68_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "CD68 expression [log2(CPM)]") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+#violin_cd68_expr_celltype_nd
+```
+
+```{r}
+umap_grid_w_violin_cd68_nd <- plot_grid(umap_cell_type_nd, 
+                                        umap_cd68_expr_nd, 
+                                        violin_cd68_expr_celltype_nd,
+                                        ncol = 2)
+```
+
+```{r fig.width=20, fig.height=20}
+umap_grid_w_violin_cd68_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_violin_scRNAseq_nd.pdf"), 
+       umap_grid_w_violin_cd68_nd, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cd68_expression_violin_scRNAseq_nd.png"), 
+       umap_grid_w_violin_cd68_nd, 
+       width = 34, 
+       height = 34,
+       units = "cm")
+```
+
+# Overlay IBA1 and CD68 expression
+
+## rGBM
+
+```{r}
+annotated_with_cd68_pos <- annotated_with_cd68 %>%
+       mutate(cd68_positive = if_else(cd68_expression > 0, 
+                                      "CD68 +", 
+                                      NA)) %>% 
+       dplyr::select(-c(gene))
+
+annotated_with_iba1_cd68 <- left_join(annotated_with_iba1,
+                                      annotated_with_cd68_pos)
+
+annotated_with_iba1_cd68_only_cd68 <- annotated_with_iba1_cd68 %>%
+       filter(!is.na(cd68_positive))
+
+annotated_with_iba1_only <- annotated_with_iba1 %>%
+       filter(iba1_expression > 0) 
+```
+
+```{r}
+umap_iba1_cd68_expr3 <- ggplot() +
+  rasterize(
+  geom_point(data = annotated_with_iba1_cd68, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = iba1_expression),
+       alpha = 0.2,
+       shape = 16
+           ),
+           dpi = 300
+           ) +
+  scale_color_gradient(low = "grey", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "IBA1 expression\n[log2(CPM)]") +
+  new_scale_color() + 
+  rasterize(
+  geom_point(data = annotated_with_iba1_cd68_only_cd68, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = cd68_expression),
+       alpha = 0.09,
+       shape = 16
+           ),
+           dpi = 300
+           ) +
+  scale_color_gradient(low = "grey", high = "green") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD68 expression\n[log2(CPM)]",
+       title = "Recurrent GBM")
+```
+
+```{r}
+umap_iba1_cd68_expr3
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_cd68_expression_scRNAseq.pdf"), 
+       umap_iba1_cd68_expr3, 
+       width = 34/2, 
+       height = 34/2,
+       units = "cm")
+```
+
+### Violin of overlay
+
+```{r}
+annotated_with_cd684violin2 <- dplyr::rename(annotated_with_cd684violin,
+                                              gene_expression = cd68_expression)
+```
+
+```{r}
+annotated_with_iba14violin2 <- dplyr::rename(annotated_with_iba14violin,
+                                              gene_expression = iba1_expression)
+```
+
+```{r}
+annotated_with_iba1_cd68 <- bind_rows(
+       annotated_with_cd684violin2,
+       annotated_with_iba14violin2
+       ) %>%
+       group_by(cluster, gene) %>%
+       filter(n() > 1) %>%
+       ungroup()
+```
+
+```{r}
+#View(annotated_with_iba1_cd68)
+```
+
+```{r}
+violin_two_expr_celltype <- ggplot(annotated_with_iba1_cd68, 
+       aes(x = cluster, 
+           y = gene_expression,
+           fill = cluster)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "Gene expression [log2(CPM)]") +
+  facet_grid(.~gene,
+              scale = "free_x") +
+  theme(legend.position = "none",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_iba1_cd68_expression_scRNAseq.pdf"), 
+       plot = violin_two_expr_celltype, 
+       device = "pdf",
+       width = 2*(34/3), 
+       height = 34/2,
+       units = "cm")
+```
+
+## iGBM
+
+```{r}
+annotated_with_cd68_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_cd68_annot_nd.rds"))
+```
+
+```{r}
+annotated_with_iba1_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_iba1_annot_nd.rds"))
+```
+
+```{r}
+annotated_with_cd68_pos_nd <- annotated_with_cd68_nd %>%
+       mutate(cd68_positive = if_else(cd68_expression > 0, 
+                                      "CD68 +", 
+                                      NA)) %>% 
+       dplyr::select(-c(gene))
+
+annotated_with_iba1_cd68_nd <- left_join(annotated_with_iba1_nd,
+                                         annotated_with_cd68_pos_nd)
+
+annotated_with_iba1_cd68_only_cd68_nd <- annotated_with_iba1_cd68_nd %>%
+       filter(!is.na(cd68_positive))
+
+annotated_with_iba1_only_nd <- annotated_with_iba1_nd %>%
+       filter(iba1_expression > 0) 
+```
+
+```{r}
+umap_iba1_cd68_expr3_nd <- ggplot() +
+  rasterize(
+  geom_point(data = annotated_with_iba1_cd68_nd, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = iba1_expression),
+       alpha = 0.2,
+       shape = 16
+           ),
+           dpi = 300
+           ) +
+  scale_color_gradient(low = "grey", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "IBA1 expression\n[log2(CPM)]") +
+  new_scale_color() + 
+  rasterize(
+  geom_point(data = annotated_with_iba1_cd68_only_cd68_nd, 
+       mapping = aes(x = x, 
+                     y = y,
+                     color = cd68_expression),
+       alpha = 0.09,
+       shape = 16
+           ),
+           dpi = 300
+           ) +
+  scale_color_gradient(low = "grey", high = "green") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "CD68 expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM")
+```
+
+```{r}
+umap_iba1_cd68_expr3_nd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_cd68_expression_scRNAseq_nd.pdf"), 
+       umap_iba1_cd68_expr3_nd, 
+       width = 34/2, 
+       height = 34/2,
+       units = "cm")
+```
+
+```{r}
+umap_iba1_cd68_both <- plot_grid(umap_iba1_cd68_expr3_nd,
+                                 umap_iba1_cd68_expr3)
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_iba1_cd68_expression_scRNAseq_both.pdf"), 
+       umap_iba1_cd68_both, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+## Violin plot of the overlay
+
+```{r}
+annot_w_iba1_CD68_igbm_n_rgbm <- bind_rows(
+       annotated_with_cd68 %>%
+       mutate(gbm_type = "rGBM",
+              expression = cd68_expression),
+       annotated_with_cd68_nd %>%
+       mutate(gbm_type = "iGBM",
+              expression = cd68_expression),
+       annotated_with_iba1 %>%
+       mutate(gbm_type = "rGBM",
+              expression = iba1_expression),
+       annotated_with_iba1_nd %>%
+       mutate(gbm_type = "iGBM",
+              expression = iba1_expression)
+              ) %>%
+                                        mutate(cluster = if_else(
+                                          cluster == "DC 1", 
+                                          "DC", 
+                                          cluster)) %>%
+                                        filter(
+                                          !cluster %in% c(
+                                                 "DC 2",
+                                                 "DC 3",
+                                                 "DC 4",
+                                                 "Mast cells",
+                                                 "Plasma B",
+                                                 "Regulatory T cells"
+                                          )
+                                        ) %>%
+              filter(expression > 5)
+
+unique(annot_w_iba1_CD68_igbm_n_rgbm$cluster)
+```
+
+```{r}
+violin_two_expr_celltype_igbm_n_rgbm <- ggplot(annot_w_iba1_CD68_igbm_n_rgbm, 
+       aes(x = cluster, 
+           y = expression,
+           fill = gbm_type)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "Gene expression [log2(CPM)]") +
+  facet_grid(.~gene,
+              scale = "free_x") +
+  theme(legend.position = "bottom",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_two_expr_celltype_igbm_n_rgbm
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_iba1_cd68_expression_scRNAseq_igbm_n_rgbm.pdf"), 
+       plot = violin_two_expr_celltype_igbm_n_rgbm, 
+       device = "pdf",
+       width = 2*(34/3), 
+       height = 34/3,
+       units = "cm")
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_iba1_cd68_expression_scRNAseq_igbm_n_rgbm.png"), 
+       plot = violin_two_expr_celltype_igbm_n_rgbm, 
+       device = "png",
+       width = 2*(34/3), 
+       height = 34/3,
+       units = "cm")
+```
+
+```{r}
+dev.off()
+```
+
+# rGBM vs iGBM comparisons
+
+## ASAH1 expression
+
+Load the data
+
+```{r}
+annotated_with_asah1_nd <- readRDS(here::here("rds/long_cpm_scrnaseq_asah1_annot_nd.rds"))
+```
+
+### UMAP of cell type based on iGBM (ND = newly diagnosed)
+
+```{r}
+cell_type_centroid_nd <- annotated_with_asah1_nd %>%
+  group_by(cluster) %>%
+  summarise(x = mean(x), 
+            y = mean(y)) %>%
+  ungroup()
+```
+
+```{r}
+umap_cell_type_nd <- ggplot(annotated_with_asah1_nd, 
+       aes(x = x, 
+           y = y, 
+           color = cluster)) +
+  ggrastr::rasterize(geom_point(alpha = 0.5)) +
+  geom_text(data = cell_type_centroid_nd, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "Cell type",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom") 
+```
+
+# UMAPs of cell type between iGBM and rGBM
+
+```{r}
+umap_grid_wnd <- plot_grid(umap_cell_type, 
+                           umap_cell_type_nd, 
+                           ncol = 2)
+```
+
+```{r}
+umap_grid_wnd
+```
+
+```{r}
+ggsave(here::here("figures/umap_grid_cell_type_iGBM_n_rGBM_scRNAseq.pdf"), 
+       umap_grid_wnd, 
+       width = 34, 
+       height = 34/2,
+       units = "cm")
+```
+
+## UMAPs with ASAH1 expression between iGBM and rGBM
+
+```{r}
+umap_asah1_expr_nd <- ggplot(annotated_with_asah1_nd, 
+       aes(x = x, 
+           y = y, 
+           color = asah1_expression)) +
+  ggrastr::rasterize(geom_point(alpha = 0.5)) +
+  geom_text(data = cell_type_centroid_nd, 
+            aes(x = x, 
+                y = y, 
+                label = cluster), 
+            size = 5,
+            color = "black") +
+  scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "UMAP_1", 
+       y = "UMAP_2", 
+       color = "ASAH1 expression\n[log2(CPM)]",
+       title = "Newly diagnosed GBM") +
+  theme(legend.position = "bottom")
+```
+
+```{r}
+umap_grid_asah1_nd <- plot_grid(umap_asah1_expr, 
+                                umap_asah1_expr_nd, 
+                                ncol = 2)
+```
+
+```{r}
+umap_grid_asah1_nd
+```
+
+## Violin plots of ASAH1 expression between iGBM and rGBM
+
+```{r}
+# merge tables from annotated abundances in iGBM and rGBM
+#View(annotated_with_asah1_nd)
+#View(annotated_with_asah1)
+
+str(annotated_with_asah1)
+str(annotated_with_asah1_nd)
+```
+
+```{r}
+annot_w_asah1_igbm_n_rgbm <- bind_rows(annotated_with_asah1 %>%
+                                        mutate(gbm_type = "rGBM"),
+                                       annotated_with_asah1_nd %>%
+                                        mutate(gbm_type = "iGBM")) %>%
+                                        mutate(cluster = if_else(
+                                          cluster == "DC 1", 
+                                          "DC", 
+                                          cluster)) %>%
+                                        filter(
+                                          !cluster %in% c(
+                                                 "DC 2",
+                                                 "DC 3",
+                                                 "DC 4",
+                                                 "Mast cells",
+                                                 "Plasma B",
+                                                 "Regulatory T cells"
+                                          )
+                                        )
+```
+
+```{r}
+annot_w_asah1_igbm_n_rgbm_1 <- annot_w_asah1_igbm_n_rgbm %>%
+  filter(asah1_expression > 2.5)
+```
+```{r}
+violin_asahl_expr_celltype <- ggplot(annot_w_asah1_igbm_n_rgbm_1, 
+       aes(x = cluster, 
+           y = asah1_expression,
+           fill = gbm_type)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  labs(x = "Cell type", 
+       y = "ASAH1 expression [log2(CPM)]") +
+  theme(legend.position = "bottom",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_asahl_expr_celltype
+```
+
+```{r}
+# fit a linear model to compare ASAH1 expression between iGBM and rGBM
+# within each cluster
+
+#View(annot_w_asah1_igbm_n_rgbm)
+
+```
+
+## Violin plots for ASAH1, ELANE and MPO expression between iGBM and rGBM
+
+```{r}
+#head(annot_w_asah1_elane_mpo_igbm_n_rgbm)
+
+annot_w_asah1_elane_mpo_igbm_n_rgbm <- bind_rows(
+       annotated_with_asah1 %>%
+       mutate(gbm_type = "rGBM",
+              expression = asah1_expression),
+       annotated_with_asah1_nd %>%
+       mutate(gbm_type = "iGBM",
+              expression = asah1_expression),
+       annotated_with_elane %>%
+       mutate(gbm_type = "rGBM",
+              expression = elane_expression),
+       annotated_with_elane_nd %>%
+       mutate(gbm_type = "iGBM",
+              expression = elane_expression),
+       annotated_with_mpo %>%
+       mutate(gbm_type = "rGBM",
+              expression = mpo_expression),
+       annotated_with_mpo_nd %>%
+       mutate(gbm_type = "iGBM",
+              expression = mpo_expression)
+              ) %>%
+                                        mutate(cluster = if_else(
+                                          cluster == "DC 1", 
+                                          "DC", 
+                                          cluster)) %>%
+                                        filter(
+                                          !cluster %in% c(
+                                                 "DC 2",
+                                                 "DC 3",
+                                                 "DC 4",
+                                                 "Mast cells",
+                                                 "Plasma B",
+                                                 "Regulatory T cells"
+                                          )
+                                        )
+
+unique(annot_w_asah1_elane_mpo_igbm_n_rgbm$cluster)
+```
+
+```{r}
+annot_w_asah1_elane_mpo_igbm_n_rgbm1 <- annot_w_asah1_elane_mpo_igbm_n_rgbm %>%
+  filter(expression > 2.5)
+```
+
+```{r}
+violin_three_expr_celltype <- ggplot(annot_w_asah1_elane_mpo_igbm_n_rgbm1, 
+       aes(x = cluster, 
+           y = expression,
+           fill = gbm_type)) +
+  geom_violin() +
+  #scale_color_gradient(low = "yellow", high = "red") +
+  facet_wrap(.~gene,
+              scale = "free_x",
+              ncol = 3) +
+  labs(x = "Cell type", 
+       y = "Gene expression [log2(CPM)]") +
+  theme(legend.position = "bottom",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6)
+        ) 
+```
+
+```{r}
+violin_three_expr_celltype
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_iba1_cd68_expression_scRNAseq_igbm_n_rgbm.pdf"), 
+       plot = violin_three_expr_celltype, 
+       device = "pdf",
+       width = 2*(34/3), 
+       height = 34/3,
+       units = "cm")
+```
+
+## ASAH1 expression in CD14+ cells
+
+### rGBM
+
+Filter CD14+ cells:
+
+```{r}
+max(annotated_with_cd14$cd14_expression)
+
+cd14_positive_rgbm <- annotated_with_cd14 %>%
+  filter(cluster %in% c(
+       "Monocytes",
+       "prol. TAM",
+       "TAM 1",
+       "TAM 2"
+  )) %>% 
+  pull(cell)
+
+annotated_with_asah1_cd14_posit_rgbm <- annotated_with_asah1 %>%
+       filter(cell %in% c(cd14_positive_rgbm))
+```
+
+### iGBM
+
+Filter CD14+ cells:
+
+```{r}
+max(annotated_with_cd14_nd$cd14_expression)
+
+cd14_positive_igbm <- annotated_with_cd14_nd %>%
+  filter(cluster %in% c(
+       "Monocytes",
+       "prol. TAM",
+       "TAM 1",
+       "TAM 2"
+  )) %>% 
+  pull(cell)
+
+annotated_with_asah1_cd14_posit_igbm <- annotated_with_asah1_nd %>%
+       filter(cell %in% c(cd14_positive_igbm)) 
+```
+
+### Violin plot of ASAH1 expression in CD14+ cells
+
+```{r}
+annotated_with_asah1_cd14_posit <- bind_rows(
+       annotated_with_asah1_cd14_posit_rgbm %>%
+       mutate(gbm_type = "rGBM"),
+       annotated_with_asah1_cd14_posit_igbm %>%
+       mutate(gbm_type = "iGBM")
+       ) %>%
+       group_by(cluster, gbm_type) %>%
+       filter(n() > 1) %>%
+       ungroup() %>% 
+       filter(asah1_expression > 2.5) %>%
+       mutate(cluster = if_else(
+         cluster == "DC 1", 
+         "DC", 
+         cluster)) %>%
+       filter(
+         !cluster %in% c(
+                "DC 2",
+                "DC 3",
+                "DC 4",
+                "Mast cells",
+                "Plasma B",
+                "Regulatory T cells",
+                "B cells",
+                "DC",
+                "NK cells",
+                "T cells"
+         )
+       )
+
+       annotated_with_asah1_cd14_posit
+```
+
+```{r}
+violin_asahl_expr_celltype_cd14 <- ggplot(annotated_with_asah1_cd14_posit, 
+       aes(x = cluster, 
+           y = asah1_expression,
+           fill = gbm_type)) +
+  geom_violin(width = 0.6) +
+  geom_boxplot(width = 0.4, ) +
+  stat_compare_means(aes(group = gbm_type), 
+                     label = "p.format") +
+  labs(x = "Cell type", 
+       y = "ASAH1 expression [log2(CPM)]") +
+  theme(legend.position = "bottom",
+        # tilt the x-axis labels 45 degrees
+        axis.text.x = element_text(angle = -45, 
+                                   hjust = 0,
+                                   vjust = 1.6))
+```
+
+```{r}
+dev.off()
+```
+
+```{r}
+violin_asahl_expr_celltype_cd14
+```
+
+```{r}
+ggsave(here::here("figures/violin_plots_asahl_expr_celltype_cd14_scRNAseq_igbm_n_rgbm.pdf"), 
+       plot = violin_asahl_expr_celltype_cd14, 
+       device = "pdf",
+       width = (2*(34/3))/2, 
+       height = 34/3,
+       units = "cm")
+```

### 2023-10-24 reports/html_gbm_recurrence_general_proteomics_manuscript_report.qmd
revision 2 commit

--- 2023-02-14 reports/gbm_recurrence_general_proteomics_manuscript_report.qmd
@@
 format:
-  gfm: 
+  html: 
     toc-depth: 4 # default is 3
+    toc-title: Contents
+    toc-location: left # default is right, option is 'body'
     number-sections: true
-    execute:
-      output: true
+    number-depth: 3
+    embed-resources: true
 always_allow_html: true
-editor: visual
+editor: source
 ---
@@
             position = position_nudge(x = 2)) +
-  labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
+  #labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
   xlab(paste("sPLS-DA Comp-1", 
@@
              adj.P.Val < 0.05,
-             Gene %in% c("ASAH1", "GPNMB", "SYNM", "CD14"))
-
-size <- 0.4
+             Gene %in% c("ASAH1"))
+                         #, "GPNMB", "SYNM", "CD14"))
+
+size <- 0.8
 
@@
                           Protein == "P05164" ~ "MPO"))
+
+slim_data4 <- slim_data3 %>% 
+  filter(Gene %in% c("ASAH1"))
+
 ```

### 2023-10-24 reports/html_gbm_recurrence_large_scale_proteolysis_manuscript_report.qmd
revision 2 commit

--- 2023-02-16 reports/gbm_recurrence_large_scale_proteolysis_manuscript_report.qmd
@@
 format:
-  gfm:
-    toc: true 
+  html: 
     toc-depth: 4 # default is 3
+    toc-title: Contents
+    toc-location: left # default is right, option is 'body'
     number-sections: true
-    execute:
-      output: true
+    number-depth: 3
+    embed-resources: true
 always_allow_html: true
@@
 #taking output from limma semi-specific
-log2_semi_tryptic <- dplyr::select(compar_tab_peptwo6_annn2,
+log2_semi_tryptic <- dplyr::select(compar_tab_peptwo6_annn,
                                    logFC, 
-                                   Protein, 
+                                   protein, 
                                    peptide, 
                                    specificity, 
-                                   is_terminal,
-                                   Gene) %>%
+                                   is_terminal) %>%
   filter(specificity == "semi_specific",
@@
 log2semipept2_log2protein_spec <- left_join(log2_fully_tryptic, 
-                                            log2_semi_tryptic) %>%
+                                            log2_semi_tryptic)# %>%
   na.omit() %>%
@@
 check_increased_pept <- cleavage_annoated_peptides %>%
-  filter(Peptide %in% increased_recwo6_pept_annn)
+  filter(Peptide %in% increased_recwo6_pept)
 ```
@@
 increased_Rec <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% increased_recwo6_pept_annn)
+                            Peptide %in% increased_recwo6_pept)
 
 decreased_Rec <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% decreased_recwo6_pept_annn)
-
-increased_Rec_4ice <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% increased_recwo6_pept_annn) %>%
-                    pull(cleave_area20)
-
-decreased_Rec_4ice <- filter(cleave_areas$cleave_area20,
-                            Peptide %in% decreased_recwo6_pept_annn) %>%
-                    pull(cleave_area20)
-```
-
-```{r}
+                            Peptide %in% decreased_recwo6_pept)
+
 write_tsv(x = increased_Rec,
@@
 
+increased_Rec_4ice <- filter(cleave_areas$cleave_area20,
+                            Peptide %in% increased_recwo6_pept) %>%
+                    pull(cleave_area20)
+
+decreased_Rec_4ice <- filter(cleave_areas$cleave_area20,
+                            Peptide %in% decreased_recwo6_pept) %>%
+                    pull(cleave_area20)
 ```
@@
   write_rds(form_peptidesincreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_increasedREC.rds"))
-
 } else {
@@
   write_rds(form_peptidesdecreased_4ice, file = here("results/semi_tryptic/formated_pept_daglogo_decreasedREC.rds"))
-
 } else {
@@
 write_tsv(x = diffabund_semis_fcs_annt,
-          file = here("suppl_tables/Supplementary_Table_3_differential_abundant_semispepecific_peptides.tsv"))
+          file = "suppl_tables/Supplementary_Table_3_differential_abundant_semispepecific_peptides.tsv")
 ```
@@
 write_tsv(x = gpnbm_supp_table,
-          file = here("suppl_tables/Supplementary_Table_4_gpnbm_semispepecific_peptides.tsv"))
-```
+          file = "suppl_tables/Supplementary_Table_4_gpnbm_semispepecific_peptides.tsv")
+```
